### 2019-03-25 article/article.tex
Add changes package

% Minimalistic template for a generic article, geared towards MDPI guidelines
% Use PDFLaTeX
\documentclass[a4paper,10pt]{article}
% Packages provided by the MDPI template already
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage{indentfirst}

% Custom packages
\usepackage[acronym,toc,shortcuts,nohypertypes=acronym]{glossaries}
\usepackage[colorlinks, allcolors=blue, unicode]{hyperref}
\usepackage{hyperxmp} % Import license metadata
\usepackage{titling} % Get the author for the metadata

% Track changes package, remove when final; using a drop-in rather than the old version in TeX Live
% Usage: \added \deleted \replaced{new}{old} \highlight \comment
% Each followed by [id=<your id>, comment={Comment, optional}]
\usepackage[xcolor={usenames,dvipsnames}]{changes}
\definechangesauthor[color=OliveGreen]{dm}
\definechangesauthor[color=NavyBlue]{jv}
\definechangesauthor[color=Purple]{nt}
\definechangesauthor[color=Mahogany]{mh}

% My own definition for the sections that start with a bold name but don't appear in TOC
\newcommand{\minisection}[1]{\medskip \textbf{#1:}}

% Metadata
\title{Fractional land cover classification method assessment using PROBA-V satellite data}
\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Jan Verbesselt, Martin Herold}

\hypersetup{
    pdflicenseurl={http://creativecommons.org/licenses/by-sa/4.0/},
    pdfcopyright={This work is licensed under the Creative Commons Attribution-ShareAlike 4.0 International License.},
    pdfauthor={\theauthor}, % These are supposed to be the default but don't seem to be
    pdftitle={\thetitle},
    pdflang={en-GB}
}

% Acronyms - cite with \ac{}
\newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
\makenoidxglossaries

% Use \begin{equation} for maths

\begin{document}

\maketitle

\linenumbers

\abstract{\comment[id=dm]{From the Master, will replace later}The current standard of land cover classification is to assign each pixel to one land cover class, which at coarse resolution causes loss of information about mixed land cover. Fuzzy land cover classification, which assigns fractions of each land cover class to each pixel, can deal with mixed pixels. However, so far its application has been limited to city-scale areas, four or fewer classes, and up to two algorithm comparisons. In this paper, the classification accuracy and processing speed were compared for three fuzzy classification machine learning algorithms: random forest regression, fuzzy \textit{c}-means and neural networks. The algorithms were used to classify the whole boreal-temperate forest gradient zone between Finland and Lithuania into nine land cover classes. Results showed that all of the tested algorithms are able to achieve similarly high classification accuracy. Random forest regression accuracy was the highest, but its processing speed was the lowest. The results are a milestone for creating a global fuzzy land cover classification product incorporating user-specific requirements.

\minisection{Keywords} Fuzzy land cover classification, machine learning, random forest, gradient boosting, neural networks, fuzzy c-means, PROBA-V}

\section{Introduction}

\begin{enumerate}
 \item Relevance of fractional LC mapping (link to user requirements and SDGs)
 \item Studies that have looked into fractional LC mapping so far
 \item Available fractional methods that were introduced in literature (wavelet transform etc.)
 \item Methods that are tested in this paper
\end{enumerate}

\added[id=dm]{\textbf{Objectives}}

\section{Materials and Methods}

\subsection{Reference data}

Description of the CGLOPS data, mentioning the balance and zero inflation problems, and the fact that the study area is Africa

\subsection{Preprocessing}

Not sure whether this should be its own section or just mentioned at the top:

\begin{enumerate}
 \item Temporal cloud filter for Proba-V
 \item Harmonic metrics extraction
 \item Terrain metrics calculation
\end{enumerate}

\subsection{Covariates/Features}

\begin{enumerate}
 \item GLSDEM
 \item Proba-V time series
 \item Perhaps also soil/climate data
\end{enumerate}

\subsection{Fractional land cover mapping methods}

\subsubsection{Logistic regression}

\subsubsection{Partial least squares regression}

\subsubsection{Fuzzy nearest prototype/centroid}

\subsubsection{Neural networks}

AKA multilayer perceptron

\subsubsection{Random forest regression}

Including variable importance

...and possibly more methods

\subsection{Validation/Accuracy Assessment}

Not sure if this should be a section, or just part of results:

\begin{enumerate}
 \item RMSE per class
 \item Sub-pixel confusion-uncertainty matrix
 \item Comparison with a control/intercept model (10\% of everything)
\end{enumerate}

\section{Results}

\begin{enumerate}
 \item RMSE per class comparison: no big differences
 \item Subpixel confusion matrix metrics: quite a big difference
 \item RF variable importance
 \item Truth:prediction scatterplots
\end{enumerate}

\section{Discussion}

\begin{enumerate}
 \item Fraction area accuracy goes up to 70\%, that is good considering that hard classification is not much better than that
 \item RF is the best, even though one model per class means that it does not take everything into account
 \item Accuracy assessment metric makes a big difference, useful to use subpixel confusion matrix for that
 \item Grasslands are hard to discern, water is easy, very few wetlands in Africa
 \item All covariates are important, problem may be getting more of them for the whole world
 \item Recommendations: more covariates, a fuzzy machine learning method that can deal with zeroes
\end{enumerate}

\section{Conclusions}

\begin{enumerate}
 \item Random forest is the best by some margin
 \item Subpixel confusion matrix is useful for differentiation
 \item All covariates are important
\end{enumerate}

\minisection{Author Contributions} Something

\minisection{Funding} Something

\minisection{Acknowledgements} Something

\minisection{Conflicts of Interest} The authors declare no conflict of interest.

%\section*{Abbreviations}

\printnoidxglossary[type=acronym]

\end{document}

### 2019-03-26 article/article.tex
Add abstract and objectives

@@
 \usepackage{titling} % Get the author for the metadata
+\usepackage{textcomp} % \texttimes
 
 % Track changes package, remove when final; using a drop-in rather than the old version in TeX Live
-% Usage: \added \deleted \replaced{new}{old} \highlight \comment
-% Each followed by [id=<your id>, comment={Comment, optional}]
+% Usage: \added, \deleted, \replaced{new}{old}, \highlight, \comment
+% All followed by an optional [id=<your id>, comment=<some comment>]
 \usepackage[xcolor={usenames,dvipsnames}]{changes}
-\definechangesauthor[color=OliveGreen]{dm}
+\definecolor{gem}{HTML}{003300}
+\definechangesauthor[color=gem]{dm}
 \definechangesauthor[color=NavyBlue]{jv}
@@
 \title{Fractional land cover classification method assessment using PROBA-V satellite data}
-\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Jan Verbesselt, Martin Herold}
+\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Jan Verbesselt}
+% * <Dainius Masiliūnas> 11:20:08 26 Mar 2019 UTC+0100:
+% TODO: fix newline issue
 
@@
 
-\abstract{\comment[id=dm]{From the Master, will replace later}The current standard of land cover classification is to assign each pixel to one land cover class, which at coarse resolution causes loss of information about mixed land cover.
-Fuzzy land cover classification, which assigns fractions of each land cover class to each pixel, can deal with mixed pixels.
-However, so far its application has been limited to city-scale areas, four or fewer classes, and up to two algorithm comparisons.
-In this paper, the classification accuracy and processing speed were compared for three fuzzy classification machine learning algorithms: random forest regression, fuzzy \textit{c}-means and neural networks.
-The algorithms were used to classify the whole boreal-temperate forest gradient zone between Finland and Lithuania into nine land cover classes.
-Results showed that all of the tested algorithms are able to achieve similarly high classification accuracy.
-Random forest regression accuracy was the highest, but its processing speed was the lowest.
-The results are a milestone for creating a global fuzzy land cover classification product incorporating user-specific requirements.
+\abstract{
+The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g.
+a pixel may be labelled as grassland, tree cover, or open water.
+However, in reality, the area covered by the pixel is hardly ever homogeneous, it may be e.g.
+trees surrounded by grass next to water.
+This mismatch between the reality and its representation in maps causes a loss in precision, as only the dominant land cover class can be represented in such ``hard'' classification maps.
+This problem is worse for coarse resolution land cover maps.
+A common workaround to this issue is to define mosaic classes, i.e.
+land cover classes that are combinations of other land cover classes, such as open forest (combination of grassland and tree cover).
+However, this approach leads to increasingly complex map legends and confusion about the definitions.
 
-\minisection{Keywords} Fuzzy land cover classification, machine learning, random forest, gradient boosting, neural networks, fuzzy c-means, PROBA-V}
+Our proposed solution to this issue is to employ fractional land cover mapping: for each pixel, define the proportion of the pixel that each considered land cover class occupies.
+This allows expressing an open forest as e.g.
+40\% grassland, 60\% tree cover, and 0\% of the rest of the land cover classes.
+This ``fuzzy'' classification not only allows to more precisely express the land cover that exists in reality, but also gives users the flexibility to derive their own maps with a customised legend.
+For instance, if a user defines open forest as one that has less than 50\% tree cover and at least 30\% of grassland, they can generate a map with this definition from land cover fraction data, rather than relying on the definition that the mapmakers provide.
+Exhaustive land cover fraction data is desired by a number of user communities, but to date there is no single global nor continental scale map that would provide fractions of each common land cover class.
+
+In this paper, we tested the application of fractional land cover mapping over the continent of Africa.
+We compared the performance of four very different machine learning algorithms: random forest regression, multilayer perceptron neural networks, partial least squares regression, fuzzy nearest prototype classification and logistic regression.
+In order to allow the algorithms to make good predictions about land cover fractions, six types of covariates were used: spectral bands and vegetation indices derived from the Proba-V 100 m TOC reflectance data; temporal features derived from the full time series of said imagery; terrestrial features from a digital elevation model; soil properties and climate data from global models, and location.
+We compared the contribution of each of the covariate types to improving the accuracy of the model predictions.
+
+One challenge when dealing with fractional data is the data imbalance: in most locations, only a single or a few classes dominate a pixel, with the rest taking up 0\% of the pixel.
+Values at the extremes (0\% and 100\%, i.e.
+pure pixels) are difficult for machine learning algorithms to estimate.
+We propose a new solution to this problem by combining three models: two classification models and one regression model.
+The first classification model determines whether a pixel is pure, and if so, it is processed using a classification model, otherwise it is processed with a regression model.
+
+% Validation
+% - 28k points
+% - Independent training and validation sets
+
+The models were trained on 26668 training points throughout Africa.
+The training points, collected as part of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project, describe the fraction that each land cover class (bare soil, crops, grassland, shrubs, trees, urban/built-up, water and wetlands) occupies within a 100\texttimes100 m area, which is aligned to the grid used for Proba-V 100 m products.
+After the models were trained, their land cover class fraction predictions were compared with an independent validation set of 3617 points, which are likewise aligned to the Proba-V grid.
+Overall as well as per-class RMSE, MAE, and ME statistics were used to compare the models.
+In addition, the subpixel confusion-uncertainty matrix was used to analyse which classes were the most difficult to separate for each model, as well as to derive overall, users' and producers' accuracies and the kappa statistic.
+The models were compared with each other, as well as against a control intercept-only model.
+
+Preliminary results show that the three-model approach increases the overall accuracy of the predictions by more accurately modelling the extreme values.
+From the machine learning models that were compared, random forest regression reaches the highest overall accuracy of 71\%.
+This is comparable to the reported accuracies of the current ``hard'' global land cover maps.
+Covariate importance analysis shows that spectral covariates are the most important for predicting land cover fractions, followed by the location, terrestrial and temporal covariates.
+
+The application of the method on data gathered throughout Africa shows the applicability of the method to a variety of land cover conditions, ranging from highly heterogeneous agricultural landscapes to homogeneous bodies of water.
+Given enough reference data from other parts of the world and enough computational power, the method could be applied globally to upscale it to a global fractional land cover map.
+Such a map would cater to users with different fields of focus.
+Among other benefits, more precise estimation of land cover allows for 1) improved climate models to help reach Sustainable Development Goals (for which land cover is a key variable), 2) facilitate the making of models for vegetation and urban area dynamics, and 3) help governments and land owners keep better track of relevant land cover and its change.
+In addition, this work contributes to the operationalisation of the CGLS-LC100 project, which aims to produce exhaustive global land cover fractions as an operational product.
+
+\minisection{Keywords} Fuzzy land cover classification, machine learning, random forest, gradient boosting, neural networks, fuzzy c-means, PROBA-V
+}
 
@@
 
-\added[id=dm]{\textbf{Objectives}}
+\added[id=dm]{
+\textbf{Objectives}:
+}
+
+\begin{enumerate}
+ \item Compare the performance of a variety of machine learning models for a continental-scale fractional land cover mapping task
+ \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
+ \item Investigate methods for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
+\end{enumerate}
 
@@
 
+\subsection{Covariates/Features}
+
+\begin{enumerate}
+ \item GLSDEM
+ \item Proba-V time series
+ \item SoilGrids and LandGIS
+ \item WorldClim
+\end{enumerate}
+
 \subsection{Preprocessing}
-
-Not sure whether this should be its own section or just mentioned at the top:
 
@@
  \item Terrain metrics calculation
-\end{enumerate}
-
-\subsection{Covariates/Features}
-
-\begin{enumerate}
- \item GLSDEM
- \item Proba-V time series
- \item Perhaps also soil/climate data
+ \item Additional climate parameter calculation
+ \item Manual model feature selection
 \end{enumerate}
@@
 \subsubsection{Logistic regression}
+
+\subsubsection{Lasso/ridge/elastic net regression}
 
@@
 
-...and possibly more methods
-
 \subsection{Validation/Accuracy Assessment}
@@
 \begin{enumerate}
- \item RMSE per class
- \item Sub-pixel confusion-uncertainty matrix
- \item Comparison with a control/intercept model (10\% of everything)
+ \item RMSE, MAE, ME per class
+ \item Sub-pixel confusion-uncertainty matrix: kappa, OA, PA, UA
+ \item Correlation matrix
+ \item Prediction accuracy at different parts of the predicted vs observed line
+ \item Comparison with a control/intercept model (10\% of everything)?
 \end{enumerate}
@@
  \item RF variable importance
- \item Truth:prediction scatterplots
+ \item Truth:prediction scatterplots (hex/box/bar)
+ \item Point map of Africa for the different classes
 \end{enumerate}
@@
 \begin{enumerate}
- \item Fraction area accuracy goes up to 70\%, that is good considering that hard classification is not much better than that
+ \item Fraction area accuracy goes up to 72\%, that is good considering that hard classification is not much better than that
  \item RF is the best, even though one model per class means that it does not take everything into account
  \item Accuracy assessment metric makes a big difference, useful to use subpixel confusion matrix for that
- \item Grasslands are hard to discern, water is easy, very few wetlands in Africa
- \item All covariates are important, problem may be getting more of them for the whole world
- \item Recommendations: more covariates, a fuzzy machine learning method that can deal with zeroes
+ \item Hardest to classify are grasslands and especially shrubs and urban
+ \item All covariates are important to a certain degree (show table of importances)
+ \item Histogram matching doesn't help any
 \end{enumerate}
@@
  \item Subpixel confusion matrix is useful for differentiation
- \item All covariates are important
+ \item All covariates are important but to a different degree
+ \item Using a multi-model approach improves MAE at the cost of RMSE, thus a trade-off
 \end{enumerate}
@@
 
-\minisection{Funding} Something
+\minisection{Funding} JRC CGLOPS
 
-\minisection{Acknowledgements} Something
+\minisection{Acknowledgements} VITO
 

### 2019-04-16 article/article.tex
Changes from papeeria
Add graphviz support
Include the graphics as static PDF (as Papeeria doesn't shell escape)
Addition of plots into the paper

@@
 \usepackage{textcomp} % \texttimes
+%\usepackage[pdf]{graphviz} % Importing dot files; can be replaced with generated PDFs
+\usepackage[norefs,nocites]{refcheck} % Develop: warn if figure is not referenced
 
@@
 
+\begin{figure}
+ %\inputdigraph[width=\textwidth]{article-figures/algorithms}{dot}
+ \includegraphics[width=\textwidth]{article-figures/algorithms}
+ \caption{Preprocessing chain.}
+ \label{fig-preprocessing}
+\end{figure}
+
 \subsection{Fractional land cover mapping methods}
@@
 Including variable importance
+
+\subsection{Model tuning (multi-step approach) to account for value imbalance}
+
+As the dataset is unbalanced (dominated by zeroes and hundreds) and thus the objective function tends to draw the model towards good prediction of 0/100, whereas in regression the middle is also important.
+Therefore we tried several approaches to increase the accuracy of the middle predictions.
+
+We compared a single model of the best algorithm, to a two-step approach: one model to predict zeroes, and one model to predict non-zeroes; and a three-step approach, with one model to predict pixel purity (i.e.
+if we face a classification or a regression problem), one to perform regression (on mixed pixels) and one to perform classification (on pure pixels).
+
+In addition, we investigated applying histogram matching, by matching the histogram of the predictions back to the histogram of the training set.
 
@@
 
+% 1:1 plots
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-points-a001s2}
+    \caption{Random Forest single model prediction correspondence to ground truth (raw points with overplotting).}
+    \label{raw-rf-1m-uncor}
+\end{figure}
+% Binplots
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n100s20}
+    \caption{Random Forest single model prediction correspondence to ground truth, 100x100 bins.}
+    \label{bin-rf-1m-uncor-n100}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n50s20sqrt}
+    \caption{Random Forest single model prediction correspondence to ground truth, 50x50 bins, square root transformation.}
+    \label{bin-rf-1m-uncor-n50sqrt}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n50s10sqrt}
+    \caption{Random Forest single model prediction correspondence to ground truth, 50x50 bins, square root transformation, smaller circle sizes.}
+    \label{bin-rf-1m-uncor-n50sqrts10}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-raster-n50sqrt}
+    \caption{Random Forest single model prediction correspondence to ground truth, square root transformation.}
+    \label{bin-rf-1m-uncor-rastersqrt}
+\end{figure}
+
+% Hexplots
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-1m-uncor-hex}
+    \caption{Random Forest single model prediction correspondence to ground truth.}
+    \label{hex-rf-1m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hex}
+    \caption{Random Forest three-step model prediction correspondence to ground truth.}
+    \label{hex-rf-3m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-1m-uncor-hm-hex}
+    \caption{Random Forest single model prediction (histogram matched) correspondence to ground truth.}
+    \label{hex-rf-1m-uncor-hm}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hm-hex}
+    \caption{Random Forest three-step model prediction (histogram matched) correspondence to ground truth.}
+    \label{hex-rf-3m-uncor-hm}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hma-hex}
+    \caption{Random Forest three-step model prediction (histogram matched middle, proportionally) correspondence to ground truth.}
+    \label{hex-rf-3m-uncor-hma}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hmb-hex}
+    \caption{Random Forest three-step model prediction (histogram matched non-extremes) correspondence to ground truth.}
+    \label{hex-rf-3m-uncor-hmb}
+\end{figure}
+
+% Boxplots
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-1m-uncor-box}
+    \caption{Random Forest single model prediction correspondence to ground truth.}
+    \label{box-rf-1m-uncor-hm}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-2m-uncor-box}
+    \caption{Random Forest two-step model prediction correspondence to ground truth.}
+    \label{box-rf-2m-uncor-hm}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-3m-uncor-box}
+    \caption{Random Forest three-step model prediction correspondence to ground truth.}
+    \label{box-rf-3m-uncor-hm}
+\end{figure}
+
+% Barplots
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-1m-uncor-bar}
+    \caption{Random Forest single model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-1m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-2m-uncor-bar}
+    \caption{Random Forest two-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-2m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-3m-uncor-bar}
+    \caption{Random Forest three-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-3m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-1m-uncor-hm-bar}
+    \caption{Random Forest single model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-1m-uncor-hm}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-3m-uncor-hm-bar}
+    \caption{Random Forest three-step model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-3m-uncor-hm}
+\end{figure}
+
+% Maps
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-1m-bubble}
+    \caption{Random Forest single model prediction residuals per class, spatially}
+    \label{resid-rf-1m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-3m-bubble}
+    \caption{Random Forest three-step model prediction residuals per class, spatially}
+    \label{resid-rf-3m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-validation}
+    \caption{Validation data spatially}
+    \label{raster-validation}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-rf-1m-uncor}
+    \caption{Random Forest single model predictions spatially}
+    \label{raster-rf-1m-uncor}
+\end{figure}
+
 \section{Discussion}
@@
 
-\minisection{Author Contributions} Something
+\minisection{Author Contributions} DM JV NT MH ML
 

### 2019-04-16 article/article.tex
Addition of accuracy statistics table

@@
 \usepackage{indentfirst}
+\usepackage{booktabs} % For \toprule, etc.
 
@@
 \newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
+\newacronym{OA}{OA}{Overall accuracy}
+\newacronym{UA}{UA}{User accuracy}
+\newacronym{PA}{PA}{Producer accuracy}
+\newacronym{RMSE}{RMSE}{Root mean squared error}
+\newacronym{MAE}{MAE}{Mean absolute error}
+\newacronym{ME}{ME}{Mean error}
+\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
 \makenoidxglossaries
@@
 \begin{enumerate}
- \item RMSE, MAE, ME per class
- \item Sub-pixel confusion-uncertainty matrix: kappa, OA, PA, UA
+ \item \ac{RMSE}, \ac{MAE}, \ac{ME} per class
+ \item Sub-pixel confusion-uncertainty matrix: kappa, \ac{OA}, \ac{PA}, \ac{UA}
  \item Correlation matrix
@@
 \end{enumerate}
+
+\subsection{Method accuracy comparison}
+
+The intercept model (always predicting the mean) achieves an \ac{RMSE} of 30\%, and \ac{OA}
+
+Logistic regression achieved RMSE of 18.7\%, MAE of 9.8\%,
+
+\begin{table}
+\centering
+\begin{tabular}{llllll}
+\toprule
+\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & \textbf{\acrshort{NSE}} & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
+\midrule
+Intercept
+& 30    &       &       & 25±4  & 0.1       \\
+Fuzzy nearest centroid
+& 22    &       &       & 49    & 0.41      \\
+Neural networks
+& 19.5  &       &       &       &           \\
+PLS regression
+& 19    &       &       & 47±5  & 0.36±0.07 \\
+Logistic regression
+& 18.7  & 9.8   &       & 66±4  & 0.57      \\
+Lasso regression
+& 18.5  &       &       &       &           \\
+Random forest
+& 16.6  & 9.2   & 0.67  & 68±4  & 0.60      \\
+\ensuremath{''} histogram matched
+& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
+\ensuremath{''} two-step
+& 18.3  & 8.0   & 0.59  & 72±2  & 0.63      \\
+\ensuremath{''} three-step
+& 18.6  & 8.2   & 0.59  & 71±4  & 0.64      \\
+\ensuremath{''} \ensuremath{''} histogram matched
+& 21.3  & 9.1   & 0.46  & 68±3  & 0.61      \\
+\bottomrule
+\end{tabular}
+\caption{Accuracy statistics of the models tested.}
+\label{tab-accuracy}
+\end{table}
 

### 2019-07-24 article/article.tex
Upgrade to TexLive 2019

@@
 
-\added[id=dm]{
+Global land cover maps
+
 \textbf{Objectives}:
-}
 
@@
 Intercept
-& 30    &       &       & 25±4  & 0.1       \\
+& 30.7  & 21.8  & 0     & 24±4  & 0.11±0.04 \\
 Fuzzy nearest centroid
@@
 Logistic regression
-& 18.7  & 9.8   &       & 66±4  & 0.57      \\
+& 20.0  & 10.7  & 0.53  & 66±4  & 0.57      \\
 Lasso regression
@@
 
-\minisection{Author Contributions} DM JV NT MH ML
+\minisection{Author Contributions} DM JV NT MH ML BK
 
@@
 
-\minisection{Acknowledgements} VITO
+\minisection{Acknowledgements} VITO, time series group (BB)
 

### 2019-07-29 article/article.tex
Changes from papeeria

@@
 \usepackage{booktabs} % For \toprule, etc.
+\usepackage[sort&compress,sectionbib]{natbib}
 
@@
 \abstract{
-The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g.
+\comment[id=dm]{The abstract is still a placeholder}The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g.
 a pixel may be labelled as grassland, tree cover, or open water.
@@
 \section{Introduction}
+% * <Dainius Masiliūnas> 14:50:56 29 Jul 2019 UTC+0200:
+% ~1250 words, 8 paragraphs + objectives
 
@@
  \item Methods that are tested in this paper
+ \item Global land cover maps
 \end{enumerate}
 
-Global land cover maps
+Global land cover mapping is very important from an ecological and earth systems point of view.
+Land cover maps are used for a variety of applications, from estimating area covered by forests \citep{bartalev2014probavboreal} to air quality modelling \citep{wiedinmyer2006airquality}.
+In addition, land cover maps, especially those of fine spatial resolution, have the potential to be used for land cover change monitoring \citep{defourny2012cci}.
+This is important for a number of land cover classes.
+For instance, forests are an important terrestrial carbon sink in the global carbon cycle, with different forests having different carbon sink capacities \citep{pan2011large}.
+Knowing their distribution and changes over time allows for more accurate climate change modelling.
+Another land cover class, wetlands, is key for maintaining biodiversity due to the uniqueness of wetland ecosystems that allows rare and protected species to live there.
+Wetlands are protected globally by the Ramsar convention \citep{davis1994ramsar}.
+
+The creation of a global land cover map from remote sensing data has been a goal of many different studies \citep{hansen2000hardtree}.
+Their results are nowadays freely available as satellite imagery products, such as Global Land Cover 2000 \citep{bartholome2005glc2000}, MODIS land cover \citep{friedl2010modis} and GlobCover \citep{arino2007globcover}.
+There are also maps based on the fusion of multiple other maps, like Geo-Wiki hybrid \citep{see2015hybrid}, and multiple sensors, like Land Cover CCI \citep{lccciguide} and GlobeLand30 \citep{chen2015globeland30}.
+However, these land cover maps all have certain drawbacks.
+
+The first common drawback of current global land cover classification maps is that their spatial resolution is low to medium.
+The majority of aforementioned land cover maps are derived from MODIS and MERIS sensor data.
+These sensors are only capable of capturing images with the finest pixels representing 300 by 300 metre areas (at the equator).
+Recent advances in satellite sensors and computing power would allow for land cover classification at a higher spatial and temporal resolution.
+For example, at the moment the PROBA-V mission by the European Space Agency produces imagery with as fine as 100 by 100 metre pixels (with additional products also providing 300 by 300 metre and 1 by 1 kilometre pixel size for comparison with previous sensors) \citep{probavguide}.
+It is well-suited for time series analysis, because it has an archive that goes back to 2013, as well as a fast revisit time of 2 days for full global coverage (1 day for locations above 35\textdegree{} latitude) \citep{dierckx2014probav}.
+
+The second drawback of current global land cover products is that their classification accuracy tends to be low, averaging at around 65\% \citep{tsendbazar2016integrating}.
+The third drawback is that all of these products use what is known as ``hard'' or ``crisp'' classification: each pixel is assigned to one particular land cover class only.
+The result of a hard classification is a thematic map.
+This classification type treats each pixel as homogeneous with regards to land cover, which is rarely the case in reality.
+In contrast, a ``fuzzy'' or ``soft'' classification, sometimes also called ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent} identifies the proportion of each land cover class within each pixel.
+The result of a fuzzy classification is one map per land cover class, showing the proportion of the pixel area covered by the land cover class.
+
+Hard classification is not well-suited for coarse resolution imagery, due to a large proportion of mixed pixels in it (representing areas of mixed land cover), compared to endmember pixels (or pure pixels, which are dominated by a single land cover class).
+When hard classification is attempted on such imagery, the accuracy of the result can be no higher than the area fraction that the dominant land cover type of the pixel occupies \citep{latifovic2004accuracy}.
+There have been attempts at increasing accuracy by defining mosaic classes, such as mixed forests (50\% needleleaf and 50\% broadleaf trees), but it leads to a proliferation of classes that increases map complexity and still does not deal with the core problem, thus not improving the classification accuracy by much \citep{tsendbazar2016comparative}.
+In contrast, ``fuzzy'' or ``soft'' classification results in each pixel containing information about the proportion of each class within that pixel, therefore operating on sub-pixel scales.
+This has the advantage of representing land cover more accurately, and gives the ability to represent mosaic classes as a combination of pure classes instead.
+Since the output of fuzzy classification is one raster per land cover class, with smooth transitions between classes, it is suitable for more in-depth analysis and user-specific visualisation criteria \citep{tsendbazar2016integrating}.
+Despite that, hard classification is still the most often used classification type due to the number of algorithms developed for it, ease of storing the result (it is thematic and thus takes little storage space), displaying it in a single map (albeit in a less accurate fashion) and performing accuracy assessment.
+
+There is a number of different classification algorithms, but only several of them are suitable to be used for fuzzy classification \citep{nath2014methods}.
+The two methods most commonly used in scientific literature are fuzzy \textit{c}-means and neural networks \citep{zhang2001fullyfuzzy}.
+Neural networks in particular are well-suited for fuzzy classification, since they allow multiple continuous output as well as input variables in a single model \citep{foody1997fuzzynnet}.
+Fuzzy \textit{c}-means, also known as fuzzy \textit{k}-means or soft \textit{k}-means, is a statistical method that relies on the proximity of pixels in feature space to class centroids, and thus is also suitable for fully fuzzy classification.
+However, since \textit{k}-means is an unsupervised classification algorithm, the ability to make use of training data in fuzzy \textit{c}-means is limited to determining class centroids with more precision \citep{hengl2004fuzzycmeans}.
+
+In addition, any other algorithms that provide a measure of uncertainty about class membership can be used, such as the ratio of individual tree votes of random forest \citep{breiman2001random} or class probabilities in gradient boosting \citep{friedman2001gradientboost}.
+While classification uncertainty is not a direct measure of class membership \citep{sytze2000fuzzyset}, they are nonetheless correlated.
+As such, it has been used by numerous authors as a proxy indicator of class membership, achieving satisfactory classification accuracy \citep{foody2002accuracy}.
+
+Furthermore, algorithms that can handle continuous variables, but only one response variable (such as random forest), can be used as well, by creating separate models for every class.
+This approach is called binary relevance \citep{karalas2016br}.
+Using this approach, the data needs to be post-processed after classification at a pixel level to conform to physical constraints (class membership must be between 0 and 100\% and sum up to 100\%).
+Random forest has been reported to give higher or equal accuracy results compared to other algorithms, like Support Vector Machines and individual decision trees, in hard classification scenarios using satellite imagery similar to that of PROBA-V \citep{duro2012algorithmcomparison}.
+Random forest regression was also shown to perform as well as other algorithms in fuzzy classification scenarios \citep{walton2008subpixelrf}, although it is used in much fewer studies on fuzzy classification than the other algorithms mentioned previously.
+Gradient boosting is an algorithm related to random forest, with the same advantages and disadvantages, but it is known to perform better in machine learning challenges \citep{chen2015higgs} and has not yet been tested on fuzzy land cover classification.
+
+Accuracy assessment is rather straight-forward for hard classification: typically a confusion matrix is employed for this purpose, showing how many pixels (or in some cases how much area \citep{stehman2009sampling}) in the image have been classified correctly, and how many incorrectly.
+Such a matrix makes it simple to tell which classes are hard to discern from one another, as well as allows for deriving statistics such as users' accuracy, producers' accuracy, and overall accuracy \citep{foody1996fuzzyevaluation}, as well as variance if probability sampling was used.
+However, a standard confusion matrix is not applicable in the context of fuzzy classification, since misclassification in this case is not absolute, but rather a matter of degree \citep{foody2002accuracy}.
+Several solutions, such as distance \citep{foody1996fuzzyevaluation}, cross-entropy and mutual information \citep{lu2007methods} have been suggested as alternatives for fuzzy classification accuracy assessment.
+
+Visualisation of the fuzzy classification results is also challenging, since each class effectively is a single-channel raster of its own.
+Three such classes can easily be combined into RGB channels for visualisation, but with more classes it is no longer possible.
+There have been attempts to develop a method based on the hue, saturation and intensity colour model to allow for a larger number of distinct classes to be visualised on a single raster \citep{hengl2004fuzzycmeans}.
+There is also a possibility of ``hardening'' the classification when high accuracy visualisation is not needed, and making use of multiple RGB rasters side-by-side when it is needed.
+
+All in all, even though global land cover mapping has been a focus of many other studies, existing global land cover maps still have room for improvement.
+This can be achieved by using data from the sensors of newer satellites such as PROBA-V, as well as performing fuzzy classification as opposed to the traditional hard classification.
+
+An important long-term goal of land cover mapping is to have a highly accurate fuzzy land cover classification method at a global scale.
+Since there are many different methods that could be used to achieve such a goal, and since the amount of data generated by satellites covering the entire world is enormous, this thesis focuses on comparing several different fuzzy classification methods on a subset of global imagery provided by the PROBA-V satellite.
+
+PROBA-V is a relatively new satellite, with few studies using it for land cover classification so far.
+A number of studies have used simulated PROBA-V data in preparation for its launch \citep{stathakis2014probavurban,roumenina2013probavcrops,bartalev2014probavboreal}.
+After launch, most studies have focused on its potential for crop classification \citep{roumenina2015probavcrops,durgun2016crop,lambert2016cropland}.
+All of these studies have only used the traditional hard classification.
+As such, the potential of using PROBA-V data for fuzzy classification in order to obtain a more accurate representation of land cover at global scale has never been tested before.
+
+Another problem of land cover classification is that there is high disagreement between different land cover products on mixed classes, such as mixed forests \citep{Herold2008lccomparison}.
+Mixed forests and wetlands are challenging to detect and classify using hard classification and optical remote sensing data, so it is important to check how well large-scale fuzzy classification performs on these classes.
+Therefore this thesis focused specifically on the region of ecotone from boreal forests to temperate broadleaf forests in Europe.
+This region also includes a large number of boreal wetlands.
+
+In addition, there is a wide variety of methods that can be used for fuzzy classification.
+Most studies on this topic focus on comparing only two methods at once.
+In this thesis, four of them have been tested: fuzzy \textit{c}-means, neural networks, random forest regression and multiclass gradient boosting.
+All of these methods are based on fundamentally different approaches to classification, so the results give insight into the possible performance of similar methods as well.
+In addition, the last of the mentioned methods has not yet been tested on fuzzy classification itself (although related methods have been).
+
+When it comes to global image processing, one problem is the amount of data and the time it takes to process it.
+Some algorithms are known to suffer from the curse of dimensionality, where a linear increase in input data results in a geometric increase in processing time \citep{walton2008subpixelrf}.
+Therefore it is important to also measure the processing time of different classification methods.
+
+In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
+In addition to spectral covariates (blue, red, near infrared, shortwave infrared bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}.
+Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}.
+Vegetation indices have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
+Therefore in this thesis such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.
+
+All in all the goal of this thesis was to improve large-scale land cover maps by performing fuzzy land cover classification in the boreal forest-temperate broadleaf forest gradient zone using PROBA-V satellite imagery.
+The research questions of the thesis are:
 
@@
 
+\bibliography{article-bib}
+
 \end{document}

### 2019-07-31 article/article.tex
Changes from papeeria

@@
 \usepackage[sort&compress,sectionbib]{natbib}
+\bibliographystyle{newapa}
 
@@
 \newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
+\newacronym{SDG}{SDG}{Sustainable Development Goal}
+\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
+\newacronym{CGLS}{CGLS-LC100}{Copernicus Global Land Services Land Cover}
+\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
+\newacronym{SMA}{SMA}{Spectral mixture analysis}
 \makenoidxglossaries
@@
 
-Global land cover mapping is very important from an ecological and earth systems point of view.
-Land cover maps are used for a variety of applications, from estimating area covered by forests \citep{bartalev2014probavboreal} to air quality modelling \citep{wiedinmyer2006airquality}.
-In addition, land cover maps, especially those of fine spatial resolution, have the potential to be used for land cover change monitoring \citep{defourny2012cci}.
-This is important for a number of land cover classes.
-For instance, forests are an important terrestrial carbon sink in the global carbon cycle, with different forests having different carbon sink capacities \citep{pan2011large}.
-Knowing their distribution and changes over time allows for more accurate climate change modelling.
-Another land cover class, wetlands, is key for maintaining biodiversity due to the uniqueness of wetland ecosystems that allows rare and protected species to live there.
-Wetlands are protected globally by the Ramsar convention \citep{davis1994ramsar}.
+Land cover, as one of the key variables for monitoring a number of \cglspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
+The most recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
 
-The creation of a global land cover map from remote sensing data has been a goal of many different studies \citep{hansen2000hardtree}.
-Their results are nowadays freely available as satellite imagery products, such as Global Land Cover 2000 \citep{bartholome2005glc2000}, MODIS land cover \citep{friedl2010modis} and GlobCover \citep{arino2007globcover}.
-There are also maps based on the fusion of multiple other maps, like Geo-Wiki hybrid \citep{see2015hybrid}, and multiple sensors, like Land Cover CCI \citep{lccciguide} and GlobeLand30 \citep{chen2015globeland30}.
-However, these land cover maps all have certain drawbacks.
+Most of these global land cover maps, as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
+However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
+This issue becomes worse at coarse resolutions and heterogeneous landscapes.
+It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus ultimately lead to an underestimation of tree cover at large scales.
 
-The first common drawback of current global land cover classification maps is that their spatial resolution is low to medium.
-The majority of aforementioned land cover maps are derived from MODIS and MERIS sensor data.
-These sensors are only capable of capturing images with the finest pixels representing 300 by 300 metre areas (at the equator).
-Recent advances in satellite sensors and computing power would allow for land cover classification at a higher spatial and temporal resolution.
-For example, at the moment the PROBA-V mission by the European Space Agency produces imagery with as fine as 100 by 100 metre pixels (with additional products also providing 300 by 300 metre and 1 by 1 kilometre pixel size for comparison with previous sensors) \citep{probavguide}.
-It is well-suited for time series analysis, because it has an archive that goes back to 2013, as well as a fast revisit time of 2 days for full global coverage (1 day for locations above 35\textdegree{} latitude) \citep{dierckx2014probav}.
+A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
+This is also called ``fuzzy'' or ``soft'' classification.
+Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g.
+\citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
+Nevertheless, to date there is \highlight[id=dm, comment=Depends on how we position CGLOPS]{no global map} that would provide fractions of every major land cover class in a single product.
+
+Land cover fraction mapping can be performed using different approaches and algorithms.
+In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
+Popular approaches include \ac{SMA} (e.g.
+\citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), random forest regression
 

### 2019-08-06 article/article.tex
Changes from papeeria

--- 2019-07-24 article/article.tex
@@
 \usepackage{booktabs} % For \toprule, etc.
+\usepackage[sort&compress,sectionbib]{natbib}
+\bibliographystyle{newapa}
 
@@
 \newacronym{GNSS}{GNSS}{Global Navigation Satellite System}
-\newacronym{OA}{OA}{Overall accuracy}
-\newacronym{UA}{UA}{User accuracy}
-\newacronym{PA}{PA}{Producer accuracy}
-\newacronym{RMSE}{RMSE}{Root mean squared error}
-\newacronym{MAE}{MAE}{Mean absolute error}
-\newacronym{ME}{ME}{Mean error}
+\newacronym{OA}{OA}{overall accuracy}
+\newacronym{UA}{UA}{user accuracy}
+\newacronym{PA}{PA}{producer accuracy}
+\newacronym{RMSE}{RMSE}{root mean squared error}
+\newacronym{MAE}{MAE}{mean absolute error}
+\newacronym{ME}{ME}{mean error}
 \newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
+\newacronym{SDG}{SDG}{Sustainable Development Goal}
+\newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
+\newacronym{CGLS}{CGLS-LC100}{Copernicus Global Land Services Land Cover}
+\newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
+\newacronym{SMA}{SMA}{spectral mixture analysis}
+\newacronym{SVM}{SVM}{support vector machine}
+\newacronym{MLP}{MLP}{multi-layer perceptron}
+\newacronym{NN}{NN}{neural network}
+\newacronym[text={random forest}]{RF}{RF}{random forest}
 \makenoidxglossaries
+% Disable abbreviating RF for now
+\glsunset{RF}
 
@@
  \item Methods that are tested in this paper
+ \item Global land cover maps
 \end{enumerate}
 
-Global land cover maps
+Land cover, as one of the key variables for monitoring a number of \cglspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
+The most recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
+
+Most of these global land cover maps, as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
+However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
+This issue becomes worse at coarse resolutions and heterogeneous landscapes.
+It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus ultimately lead to an underestimation of tree cover at large scales.
+
+A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
+This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}. A related term is ``super
+Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g.
+\citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
+Nevertheless, to date there is \highlight[id=dm, comment=Depends on how we position CGLOPS]{no global map} that would provide fractions of every major land cover class in a single product.
+
+Land cover fraction mapping can be performed using different approaches and algorithms.
+In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
+Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g.
+\citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
+
+%The second drawback of current global land cover products is that their classification accuracy tends to be low, averaging at around 65\% \citep{tsendbazar2016integrating}.
+The third drawback is that all of these products use what is known as ``hard'' or ``crisp'' classification: each pixel is assigned to one particular land cover class only.
+The result of a hard classification is a thematic map.
+This classification type treats each pixel as homogeneous with regards to land cover, which is rarely the case in reality.
+In contrast, a ``fuzzy'' or ``soft'' classification, sometimes also called ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent} identifies the proportion of each land cover class within each pixel.
+The result of a fuzzy classification is one map per land cover class, showing the proportion of the pixel area covered by the land cover class.
+
+%Hard classification is not well-suited for coarse resolution imagery, due to a large proportion of mixed pixels in it (representing areas of mixed land cover), compared to endmember pixels (or pure pixels, which are dominated by a single land cover class).
+When hard classification is attempted on such imagery, the accuracy of the result can be no higher than the area fraction that the dominant land cover type of the pixel occupies \citep{latifovic2004accuracy}.
+There have been attempts at increasing accuracy by defining mosaic classes, such as mixed forests (50\% needleleaf and 50\% broadleaf trees), but it leads to a proliferation of classes that increases map complexity and still does not deal with the core problem, thus not improving the classification accuracy by much \citep{tsendbazar2016comparative}.
+In contrast, ``fuzzy'' or ``soft'' classification results in each pixel containing information about the proportion of each class within that pixel, therefore operating on sub-pixel scales.
+This has the advantage of representing land cover more accurately, and gives the ability to represent mosaic classes as a combination of pure classes instead.
+Since the output of fuzzy classification is one raster per land cover class, with smooth transitions between classes, it is suitable for more in-depth analysis and user-specific visualisation criteria \citep{tsendbazar2016integrating}.
+Despite that, hard classification is still the most often used classification type due to the number of algorithms developed for it, ease of storing the result (it is thematic and thus takes little storage space), displaying it in a single map (albeit in a less accurate fashion) and performing accuracy assessment.
+
+%There is a number of different classification algorithms, but only several of them are suitable to be used for fuzzy classification \citep{nath2014methods}.
+The two methods most commonly used in scientific literature are fuzzy \textit{c}-means and neural networks \citep{zhang2001fullyfuzzy}.
+Neural networks in particular are well-suited for fuzzy classification, since they allow multiple continuous output as well as input variables in a single model \citep{foody1997fuzzynnet}.
+Fuzzy \textit{c}-means, also known as fuzzy \textit{k}-means or soft \textit{k}-means, is a statistical method that relies on the proximity of pixels in feature space to class centroids, and thus is also suitable for fully fuzzy classification.
+However, since \textit{k}-means is an unsupervised classification algorithm, the ability to make use of training data in fuzzy \textit{c}-means is limited to determining class centroids with more precision \citep{hengl2004fuzzycmeans}.
+
+%In addition, any other algorithms that provide a measure of uncertainty about class membership can be used, such as the ratio of individual tree votes of random forest \citep{breiman2001random} or class probabilities in gradient boosting \citep{friedman2001gradientboost}.
+While classification uncertainty is not a direct measure of class membership \citep{sytze2000fuzzyset}, they are nonetheless correlated.
+As such, it has been used by numerous authors as a proxy indicator of class membership, achieving satisfactory classification accuracy \citep{foody2002accuracy}.
+
+%Furthermore, algorithms that can handle continuous variables, but only one response variable (such as random forest), can be used as well, by creating separate models for every class.
+This approach is called binary relevance \citep{karalas2016br}.
+Using this approach, the data needs to be post-processed after classification at a pixel level to conform to physical constraints (class membership must be between 0 and 100\% and sum up to 100\%).
+Random forest has been reported to give higher or equal accuracy results compared to other algorithms, like Support Vector Machines and individual decision trees, in hard classification scenarios using satellite imagery similar to that of PROBA-V \citep{duro2012algorithmcomparison}.
+Random forest regression was also shown to perform as well as other algorithms in fuzzy classification scenarios \citep{walton2008subpixelrf}, although it is used in much fewer studies on fuzzy classification than the other algorithms mentioned previously.
+Gradient boosting is an algorithm related to random forest, with the same advantages and disadvantages, but it is known to perform better in machine learning challenges \citep{chen2015higgs} and has not yet been tested on fuzzy land cover classification.
+
+%Accuracy assessment is rather straight-forward for hard classification: typically a confusion matrix is employed for this purpose, showing how many pixels (or in some cases how much area \citep{stehman2009sampling}) in the image have been classified correctly, and how many incorrectly.
+Such a matrix makes it simple to tell which classes are hard to discern from one another, as well as allows for deriving statistics such as users' accuracy, producers' accuracy, and overall accuracy \citep{foody1996fuzzyevaluation}, as well as variance if probability sampling was used.
+However, a standard confusion matrix is not applicable in the context of fuzzy classification, since misclassification in this case is not absolute, but rather a matter of degree \citep{foody2002accuracy}.
+Several solutions, such as distance \citep{foody1996fuzzyevaluation}, cross-entropy and mutual information \citep{lu2007methods} have been suggested as alternatives for fuzzy classification accuracy assessment.
+
+%Visualisation of the fuzzy classification results is also challenging, since each class effectively is a single-channel raster of its own.
+Three such classes can easily be combined into RGB channels for visualisation, but with more classes it is no longer possible.
+There have been attempts to develop a method based on the hue, saturation and intensity colour model to allow for a larger number of distinct classes to be visualised on a single raster \citep{hengl2004fuzzycmeans}.
+There is also a possibility of ``hardening'' the classification when high accuracy visualisation is not needed, and making use of multiple RGB rasters side-by-side when it is needed.
+
+%All in all, even though global land cover mapping has been a focus of many other studies, existing global land cover maps still have room for improvement.
+This can be achieved by using data from the sensors of newer satellites such as PROBA-V, as well as performing fuzzy classification as opposed to the traditional hard classification.
+
+%An important long-term goal of land cover mapping is to have a highly accurate fuzzy land cover classification method at a global scale.
+Since there are many different methods that could be used to achieve such a goal, and since the amount of data generated by satellites covering the entire world is enormous, this thesis focuses on comparing several different fuzzy classification methods on a subset of global imagery provided by the PROBA-V satellite.
+
+%PROBA-V is a relatively new satellite, with few studies using it for land cover classification so far.
+A number of studies have used simulated PROBA-V data in preparation for its launch \citep{stathakis2014probavurban,roumenina2013probavcrops,bartalev2014probavboreal}.
+After launch, most studies have focused on its potential for crop classification \citep{roumenina2015probavcrops,durgun2016crop,lambert2016cropland}.
+All of these studies have only used the traditional hard classification.
+As such, the potential of using PROBA-V data for fuzzy classification in order to obtain a more accurate representation of land cover at global scale has never been tested before.
+
+%Another problem of land cover classification is that there is high disagreement between different land cover products on mixed classes, such as mixed forests \citep{Herold2008lccomparison}.
+Mixed forests and wetlands are challenging to detect and classify using hard classification and optical remote sensing data, so it is important to check how well large-scale fuzzy classification performs on these classes.
+Therefore this thesis focused specifically on the region of ecotone from boreal forests to temperate broadleaf forests in Europe.
+This region also includes a large number of boreal wetlands.
+
+%In addition, there is a wide variety of methods that can be used for fuzzy classification.
+Most studies on this topic focus on comparing only two methods at once.
+In this thesis, four of them have been tested: fuzzy \textit{c}-means, neural networks, random forest regression and multiclass gradient boosting.
+All of these methods are based on fundamentally different approaches to classification, so the results give insight into the possible performance of similar methods as well.
+In addition, the last of the mentioned methods has not yet been tested on fuzzy classification itself (although related methods have been).
+
+%When it comes to global image processing, one problem is the amount of data and the time it takes to process it.
+Some algorithms are known to suffer from the curse of dimensionality, where a linear increase in input data results in a geometric increase in processing time \citep{walton2008subpixelrf}.
+Therefore it is important to also measure the processing time of different classification methods.
+
+In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
+In addition to spectral covariates (blue, red, near infrared, shortwave infrared bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}.
+Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}.
+Vegetation indices have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
+Therefore in this thesis such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.
 
@@
 \begin{enumerate}
- \item Compare the performance of a variety of machine learning models for a continental-scale fractional land cover mapping task
+ \item Compare the performance of a variety of machine learning models for a global-scale land cover fraction mapping task
  \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
@@
 
-\section{Materials and Methods}
+\section{Data and Methods}
 
@@
 
-Description of the CGLOPS data, mentioning the balance and zero inflation problems, and the fact that the study area is Africa
+For training the global land cover fraction models, a training dataset collected by IIASA was used.
+This global point dataset provides information on nine class fractions of sampled pixels of the Proba-V 100 m product.
+The fraction estimate was acquired by high-resolution image interpretation by regional experts, for the year 2015.
+Each of the sampled Proba-V pixels was subdivided into a 10-by-10 subpixel grid and the area estimates were derived from the subpixel ratios.
 
-\subsection{Covariates/Features}
+For model validation, an independent global dataset collected by Wageningen University \& Research was used.
+The method of point collection is equivalent to the method used for training data collection, but performed by separate teams of regional experts.
+
+The data includes over 150\,000 training points and over 20\,000 validation points across the globe (see figure \ref{fig-reference-data}).
+Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel; the rest are set to zero.
+
+\begin{figure}
+ \includegraphics[width=\textwidth]{article-figures/maps/2019-08-06-training-and-validation}
+ \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): reference dataset.}
+ \label{fig-reference-data}
+\end{figure}
+
+\subsection{Model covariates}
+
+To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
+In this study, five groups of covariates were used: optical, terrain, soil, climate and location.
+All of these covariates had to be preprocessed before they could be input into the models.
+
+\subsubsection{Optical covariates}
+
+\begin{figure}
+ \includegraphics[width=\textwidth]{article-figures/flowcharts/preprocessing-optical}
+ \caption{Optical data preprocessing chain.}
+ \label{fig-preprocessing-optical}
+\end{figure}
 
@@
 
+%\begin{figure}
+ %\inputdigraph[width=\textwidth]{article-figures/algorithms}{dot}
+% \includegraphics[width=\textwidth]{article-figures/flowcharts/algorithms}
+% \caption{Preprocessing chain.}
+% \label{fig-preprocessing}
+%\end{figure}
+
 \begin{figure}
- %\inputdigraph[width=\textwidth]{article-figures/algorithms}{dot}
- \includegraphics[width=\textwidth]{article-figures/algorithms}
- \caption{Preprocessing chain.}
- \label{fig-preprocessing}
+ \includegraphics[width=\textwidth]{article-figures/flowcharts/processing}
+ \caption{Processing chain.}
+ \label{fig-processing}
 \end{figure}

### 2019-08-06 article/article.tex
Changes from papeeria

@@
 \newacronym{NN}{NN}{neural network}
+\newacronym{LOESS}{LOESS}{locally estimated scatterplot smoothing}
+\newacronym[firstplural={vegetation indices (VIs)}]{VI}{VI}{vegetation index}
+\newacronym{GLSDEM}{GLSDEM}{Global Land Survey Digital Elevation Model}
+\newacronym{TPI}{TPI}{Terrain Position Index}
 \newacronym[text={random forest}]{RF}{RF}{random forest}
@@
 Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}.
-Vegetation indices have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
+\Glspl{VI} have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
 Therefore in this thesis such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.
@@
 
+%\begin{enumerate}
+% \item GLSDEM
+% \item Proba-V time series
+% \item SoilGrids and LandGIS
+% \item WorldClim
+%\end{enumerate}
+
 \subsubsection{Optical covariates}
+% * <Dainius Masiliūnas> 16:04:43 06 Aug 2019 UTC+0200:
+% Maybe better to call it "Proba-V data" or such
+
+The entire archive of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product was used for this study.
+The time series of NDVI and each of the bands at the points of interest were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were too far from the fitted \ac{LOESS} curve over the blue band.
+
+The resulting cloud-free time series was then used to extract descriptive statistics (mean and interquartile range) of NDVI over the whole time series.
+Next, harmonic analysis was run in order to decompose the time series into sine and cosine components for two frequencies (annual and semiannual), as well as the trend and intercept of the model.
+From this, phase and amplitude for the two frequencies were calculated.
+
+Lastly, extra \glspl{VI} (NDMI, EVI, OSAVI and NIR\textsubscript{v}) were calculated.
+Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season of 2015.
+
+The optical preprocessing chain is displayed graphically in figure \ref{fig-preprocessing-optical}.
+All of the outputs were then used in the processing chain.
 
@@
 
-\begin{enumerate}
- \item GLSDEM
- \item Proba-V time series
- \item SoilGrids and LandGIS
- \item WorldClim
-\end{enumerate}
+\subsubsection{Terrain covariates}
 
-\subsection{Preprocessing}
+The NASA \ac{GLSDEM} 90 m product was used as the elevation covariate.
+In addition, terrain parameters were calculated out of elevation: slope, aspect, roughness and \ac{TPI}.
+All terrain covariates were then resampled to the Proba-V grid.
+
+\subsubsection{Climate covariates}
+
+The climate data from WorldClim 2.0 30 s product was used.
+Biophysical parameters were calculated from the data as well.
+Some additional biophysical parameters were also calculated, namely the means, minimums and maximums over the whole time series of solar radiation, wind speed and water vapour pressure.
+
+\subsubsection{Soil covariates}
+
+\subsubsection{Location covariates}
+
+The latitude, longitude and absolute latitude of the points were also included as covariates.
+
+\subsection{Covariate selection}
 

### 2020-04-28 article/article.tex
Changes from papeeria

@@
 % Metadata
-\title{Fractional land cover classification method assessment using PROBA-V satellite data}
-\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Jan Verbesselt}
+\title{Global land cover fraction mapping: methods, variables, challenges}
+\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Marcel Buchhorn, Jan Verbesselt}
 % * <Dainius Masiliūnas> 11:20:08 26 Mar 2019 UTC+0100:
@@
 \newacronym{LC-CCI}{LC-CCI}{ESA Climate Change Initiative Land Cover}
-\newacronym{CGLS}{CGLS-LC100}{Copernicus Global Land Services Land Cover}
+\newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
+\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
 \newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
+\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
 \newacronym{SMA}{SMA}{spectral mixture analysis}
@@
 \newacronym{TPI}{TPI}{Terrain Position Index}
-\newacronym[text={random forest}]{RF}{RF}{random forest}
+\newacronym{OSAVI}{OSAVI}{Optimised Soil-Adjusted Vegetation Index}
+\newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
+\newacronym{NIR}{NIR}{near infra-red}
+\newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
+\newacronym{FNC}{FNC}{fuzzy nearest centroid}
+\newacronym{PLS}{PLS}{partial least squares}
+\newacronym{MLR}{MLR}{multinomial logistic regression}
+\newacronym{RF}{RF}{random forest}
+\newacronym{TOC}{TOC}{Top-of-Canopy}
+\newacronym{CART}{CART}{Classification and regression tree}
 \makenoidxglossaries
-% Disable abbreviating RF for now
-\glsunset{RF}
+% Disable abbreviating RF for now [text={random forest}]
+%\glsunset{RF}
 
@@
 
-\abstract{
-\comment[id=dm]{The abstract is still a placeholder}The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g.
+\abstract{The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g.
 a pixel may be labelled as grassland, tree cover, or open water.
@@
 For instance, if a user defines open forest as one that has less than 50\% tree cover and at least 30\% of grassland, they can generate a map with this definition from land cover fraction data, rather than relying on the definition that the mapmakers provide.
-Exhaustive land cover fraction data is desired by a number of user communities, but to date there is no single global nor continental scale map that would provide fractions of each common land cover class.
+Thus exhaustive land cover fraction data is desired by a number of user communities.
 
-In this paper, we tested the application of fractional land cover mapping over the continent of Africa.
-We compared the performance of four very different machine learning algorithms: random forest regression, multilayer perceptron neural networks, partial least squares regression, fuzzy nearest prototype classification and logistic regression.
+In this paper, we compared the performance of eight very machine learning algorithms for land cover fraction mapping: random forest regression, multilayer perceptron neural networks, partial least squares regression, fuzzy nearest prototype classification, multinomial logistic regression, lasso regression, cubist regression and support vector regression.
 In order to allow the algorithms to make good predictions about land cover fractions, six types of covariates were used: spectral bands and vegetation indices derived from the Proba-V 100 m TOC reflectance data; temporal features derived from the full time series of said imagery; terrestrial features from a digital elevation model; soil properties and climate data from global models, and location.
@@
 
-The models were trained on 26668 training points throughout Africa.
+The models were trained on 138164 training points throughout the globe.
 The training points, collected as part of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project, describe the fraction that each land cover class (bare soil, crops, grassland, shrubs, trees, urban/built-up, water and wetlands) occupies within a 100\texttimes100 m area, which is aligned to the grid used for Proba-V 100 m products.
-After the models were trained, their land cover class fraction predictions were compared with an independent validation set of 3617 points, which are likewise aligned to the Proba-V grid.
+After the models were trained, their land cover class fraction predictions were compared with an independent validation set of 20705 points, which are likewise aligned to the Proba-V grid.
 Overall as well as per-class RMSE, MAE, and ME statistics were used to compare the models.
@@
 
-Preliminary results show that the three-model approach increases the overall accuracy of the predictions by more accurately modelling the extreme values.
-From the machine learning models that were compared, random forest regression reaches the highest overall accuracy of 71\%.
+Results show that the three-model approach increases the overall accuracy of the predictions by more accurately modelling the extreme values.
+From the machine learning models that were compared, random forest and cubist regressions reaches the highest overall accuracy of 72\%.
 This is comparable to the reported accuracies of the current ``hard'' global land cover maps.
-Covariate importance analysis shows that spectral covariates are the most important for predicting land cover fractions, followed by the location, terrestrial and temporal covariates.
+Random Forest covariate importance analysis shows that spectral covariates are the most important for predicting land cover fractions, followed by the location, terrestrial and temporal covariates.
 
-The application of the method on data gathered throughout Africa shows the applicability of the method to a variety of land cover conditions, ranging from highly heterogeneous agricultural landscapes to homogeneous bodies of water.
-Given enough reference data from other parts of the world and enough computational power, the method could be applied globally to upscale it to a global fractional land cover map.
-Such a map would cater to users with different fields of focus.
+The application of the method on data gathered throughout the globe shows the applicability of the method to a variety of land cover conditions, ranging from highly heterogeneous agricultural landscapes to homogeneous bodies of water.
+This research proves that machine learning algorithms can be applied globally to create a global fractional land cover product.
+Such a product would cater to users with different fields of focus.
 Among other benefits, more precise estimation of land cover allows for 1) improved climate models to help reach Sustainable Development Goals (for which land cover is a key variable), 2) facilitate the making of models for vegetation and urban area dynamics, and 3) help governments and land owners keep better track of relevant land cover and its change.
@@
 
+\pagebreak
+\tableofcontents
+% * <Dainius Masiliūnas> 11:40:16 21 Apr 2020 UTC+0200:
+% Temporary, to make it easier to understand the structure
+\pagebreak
+
 \section{Introduction}
-% * <Dainius Masiliūnas> 14:50:56 29 Jul 2019 UTC+0200:
-% ~1250 words, 8 paragraphs + objectives
 
@@
 
-Land cover, as one of the key variables for monitoring a number of \cglspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
-The most recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
+Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
+Some of the recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS-LC100}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
 
@@
 \citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
-Nevertheless, to date there is \highlight[id=dm, comment=Depends on how we position CGLOPS]{no global map} that would provide fractions of every major land cover class in a single product.
+Nevertheless, to date only \ac{CGLS-LC100} provides a global map with fractions of every major land cover class.
 
-Land cover fraction mapping can be performed using different approaches and algorithms.
+Land cover fraction mapping can be performed using a variety of different approaches and algorithms.
 In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
@@
 \citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
+So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor a global scale.
 
@@
 \Glspl{VI} have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
-Therefore in this thesis such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.
+Therefore in this paper such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.
 
-\textbf{Objectives}:
+All in all, with this study we aim to investigate and compare methods for performing land cover fraction mapping, at a global scale, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
+We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hamper land cover fraction mapping.
+Therefore the objectives of our study are:
 
@@
  \item Compare the performance of a variety of machine learning models for a global-scale land cover fraction mapping task
+ \item Investigate approaches for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
  \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
- \item Investigate methods for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
 \end{enumerate}
@@
 
-For training the global land cover fraction models, a training dataset collected by IIASA was used.
+The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS} project.
+The training set was generated by experts in \ac{IIASA}.
 This global point dataset provides information on nine class fractions of sampled pixels of the Proba-V 100 m product.
-The fraction estimate was acquired by high-resolution image interpretation by regional experts, for the year 2015.
+The fraction estimate was acquired by high-resolution image interpretation by experts, for the year 2015.
 Each of the sampled Proba-V pixels was subdivided into a 10-by-10 subpixel grid and the area estimates were derived from the subpixel ratios.
 
-For model validation, an independent global dataset collected by Wageningen University \& Research was used.
-The method of point collection is equivalent to the method used for training data collection, but performed by separate teams of regional experts.
+For model validation, an independent global dataset of \ac{CGLS}, generated by Wageningen University \& Research \citep{tsendbazar_developing_2018}, was used.
+The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.
 
-The data includes over 150\,000 training points and over 20\,000 validation points across the globe (see figure \ref{fig-reference-data}).
-Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel; the rest are set to zero.
+The data includes over 150\,000 training points and over 20\,000 validation points across the globe, describing the fractions of 12 classes (see figure \ref{fig-reference-data}) as they were in the year 2015.
+Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel.
+This means that for a given pixel, only a few of the 12 classes would normally have a fraction that is above 0\%.
+This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
+Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
+In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
+But for user communities, correct prediction of non-zero values is more important.
 
@@
  \includegraphics[width=\textwidth]{article-figures/maps/2019-08-06-training-and-validation}
- \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): reference dataset.}
+ \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): validation dataset.}
  \label{fig-reference-data}
@@
 To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
-In this study, five groups of covariates were used: optical, terrain, soil, climate and location.
+In this study, six groups of covariates were used: spectral, harmonic, terrain, soil, climate and location (see table \ref{tab-inputdata}).
 All of these covariates had to be preprocessed before they could be input into the models.
 
-%\begin{enumerate}
-% \item GLSDEM
-% \item Proba-V time series
-% \item SoilGrids and LandGIS
-% \item WorldClim
-%\end{enumerate}
+\begin{table}
+\centering
+    \begin{tabular}{llp{6cm}}
+         \toprule
+         \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
+         \midrule
+         Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
+         Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
+         Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
+         Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
+         Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
+         Location & 3 & None (intrinsic) \\
+         \bottomrule
+    \end{tabular}
+    \label{tab-inputdata}
+    \caption{Data sources for the covariates input into the model.}
+\end{table}
 
@@
 
-The entire archive of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product was used for this study.
+The entire archive (2014-03-11 to 2019-07-16) of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{probavguide2} was used for this study.
 The time series of NDVI and each of the bands at the points of interest were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were too far from the fitted \ac{LOESS} curve over the blue band.
@@
 Lastly, extra \glspl{VI} (NDMI, EVI, OSAVI and NIR\textsubscript{v}) were calculated.
-Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season of 2015.
+Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season (three months) of 2015.
 
@@
 
-The NASA \ac{GLSDEM} 90 m product was used as the elevation covariate.
-In addition, terrain parameters were calculated out of elevation: slope, aspect, roughness and \ac{TPI}.
+The ASTER GDEM v003 \citep{ASTGTM003} product was used as the elevation covariate.
+In addition, terrain parameters were calculated out of elevation: slope, aspect and \ac{TPI}.
 All terrain covariates were then resampled to the Proba-V grid.
@@
 
-The climate data from WorldClim 2.0 30 s product was used.
+The climate data from WorldClim 2.0 30 s product \citep{worldclim2} was used.
 Biophysical parameters were calculated from the data as well.
@@
 
-\subsubsection{Location covariates}
+Two products were used for data on soils: SoilGrids and LandGIS.
+SoilGrids is based on a machine learning model that predicts soil properties at various soil depths globally at a 250 m resolution \citep{hengl_soilgrids250m_2017}
+LandGIS is a successor to SoilGrids using a refined machine learning methodology \citep{hengl_predictive_2019}, however, it does not yet provide as many covariates as SoilGrids.
 
-The latitude, longitude and absolute latitude of the points were also included as covariates.
+In the creation of SoilGrids, a land cover map (based on MODIS) was used.
+In order to avoid circular dependencies, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
+In addition, the taxonomy covariates were also excluded, since they are categorical data derived from the numerical soil property data.
+
+\subsubsection{Other covariates}
+
+The latitude, longitude and absolute latitude of the points were also included as location covariates.
+Lastly, as part of the SoilGrids dataset, the MODIS land surface temperature product (MOD11A2, \citealp{wan_mod11a2_2015}) data monthly composites for 2011 were also used as additional covariates.
 
@@
 
-\begin{enumerate}
- \item Temporal cloud filter for Proba-V
- \item Harmonic metrics extraction
- \item Terrain metrics calculation
- \item Additional climate parameter calculation
- \item Manual model feature selection
-\end{enumerate}
+In total, 313 covariates were used for predicting the land cover fractions.
+However, many of them are colinear with one another, and thus variable selection was used to remove colinear covariates.
+Covariates that had a Pearson's correlation coefficient $r$ above 0.9 were excluded in an iterative process.
+After that, covariates with a Spearman's $\rho$ above 0.9 were likewise excluded.
+
+Most of the colinear covariates were soil covariates at different depths.
+Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
+Similarly, climate data was colinear between subsequent months.
+Thus January and August data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
+
+While \ac{OSAVI} was calculated as a covariate, ultimately it was dropped due to too high correlation with \ac{NDVI}.
+Similarly, the spectral bands of Proba-V were also highly correlated, so only the red and \ac{NIR} bands were kept.
+
+After the covariate selection process, 67 covariates remained.
+These covariates include data from each of the covariate categories.
+See table \ref{tab-inputdata} for an overview.
 
@@
  \includegraphics[width=\textwidth]{article-figures/flowcharts/processing}
+% * <Dainius Masiliūnas> 13:45:26 13 Aug 2019 UTC+0200:
+% TODO: optimise style for readability.
  \caption{Processing chain.}
@@
 
-\subsection{Fractional land cover mapping methods}
+\subsection{Land cover fraction mapping methods}
 
-\subsubsection{Logistic regression}
+For deriving land cover fractions, we compared a wide array of machine learning regression methods: \ac{FNC}, \ac{MLP} \glspl{NN}, \ac{PLS} regression, Lasso regression, \ac{MLR}, \ac{RF} regression.
+% * <Dainius Masiliūnas> 16:05:16 07 Aug 2019 UTC+0200:
+% Perhaps also SVM and MESMA
+In addition, as a baseline we also compare the results with an intercept model (all fractions always predicted to be equal).
+
+All of the models were made using the free and open-source statistical software R \citep{r_2019}.
 
@@
 
-\subsubsection{Partial least squares regression}
+Ridge regression is a regularisation method which reduces large coefficients of model covariates in order to prevent overfitting.
+Lasso regression is based on the same principle, except it reduces the coefficients of each model covariate to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
+Elastic net regression is a method combining the two, where model coefficients may be reduced or set to zero.
+This has the effect of the coefficients of correlated covariates being set to a similar value.
 
-\subsubsection{Fuzzy nearest prototype/centroid}
+Since we performed a manual covariate selection prior to running regularisation to remove highly correlated covariates, and having a further reduced set of covariates is useful for other algorithms, for this method we chose Lasso regression.
+Lasso regression builds upon least squares, and thus the classes are predicted separately to each other, using the binary relevance approach as presented by \citet{karalas2016br}.
+Therefore, the output fractions do not necessarily sum to 100\%.
+As such, they were linearly scaled.
+The lasso regression was performed using the \texttt{glmnet} package \citep{glmnet}.
 
-\subsubsection{Neural networks}
+\subsubsection{\Glsentryfull{PLS} regression}
 
-AKA multilayer perceptron
+\ac{PLS} regression is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space \citep{martens_multivariate_1992}.
+As such, it is well suited for multicollinear data.
+Due to this property, the \highlight[id=nt,comment={would you not agree for better comparability just use the same set of covariates?}]{full set of covariates} were used as input to this model, rather than the reduced (uncorrelated) set.
+\ac{PLS} regression was performed using the package \texttt{pls} \citep{pls}.
 
-\subsubsection{Random forest regression}
+\subsubsection{\Glsentryfull{FNC} regression}
 
-Including variable importance
+\Gls{FNC} regression, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from the class centroid in feature space.
+This method always produces output that sums to 100\%.
+In this study, we used an implementation adapted from the package GSIF \citep{hengl2004fuzzycmeans}.
+
+\subsubsection{\Glsentryfull{MLR}}
+
+While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
+This probability can be used as a proxy for the fraction of the pixel that the class covers.
+These probabilities always add up to 100\%.
+However, an \ac{MLR} can only be trained on labelled data, rather than fractions.
+Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
+This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
+The logistic regression implementation was provided by the \texttt{nnet} package \citep{nnet}.
+
+\subsubsection{\Glsentryfullpl{NN}}
+
+\Glspl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\%.
+In this study, we used a multilayer perceptron with two layers of 256 neurons and softmax activation.
+The \texttt{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics card to accelerate the neural network training process.
+
+\subsubsection{\Glsentryfull{RF} regression}
+
+\ac{RF} regression is a popular method for land cover classification that works by building a number of decision trees based on random subsets of the input training data, and averaging the ``votes'' of these individual trees.
+\ac{RF} can only take one variable as the response variable, therefore there are two ways to obtain land cover fraction predictions from an \ac{RF} model.
+One way is analogous to how we performed \ac{MLR}: giving labels of the dominant land cover class as the response variable, and then considering the proportion of votes from the trees as the proportions of land cover classes in the pixel.
+The second way is binary relevance, where a separate model is trained to predict the presence of each class.
+Then the predictions are scaled to sum up to 100\%.
+In this study, we chose the second approach, so that the model for each class would be able to learn from the fractions of all the input data, rather than just the labels of the dominant land cover class.
+The processing was done using the \texttt{ranger} package \citep{ranger}.
+
+\ac{RF} was also used for assessing the covariate importance.
+Since the number of covariates is imbalanced by group (i.e.
+there are merely 3 location covariates and 168 soil covariates, which are often collinear), covariate importance was assessed by permuting all covariates in a group and reporting the decrease in \ac{RMSE} resulting from that.
 
@@
 
-As the dataset is unbalanced (dominated by zeroes and hundreds) and thus the objective function tends to draw the model towards good prediction of 0/100, whereas in regression the middle is also important.
+After determining the method that performs the best, we attempted to tune it further in an attempt to solve the dataset balance (zero inflation) problem.
+As the dataset is dominated by zeroes and hundreds, minimising the objective function tends to draw the model towards predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
+This is not desirable for users of land cover fraction data, as the added value of fractions is the information about the middle of the range; otherwise, discrete classification would be just as good.
 Therefore we tried several approaches to increase the accuracy of the middle predictions.
@@
 
-Not sure if this should be a section, or just part of results:
+To assess the performance of the models, we used a number of statistical measures:
 
 \begin{enumerate}
- \item \ac{RMSE}, \ac{MAE}, \ac{ME} per class
- \item Sub-pixel confusion-uncertainty matrix: kappa, \ac{OA}, \ac{PA}, \ac{UA}
- \item Correlation matrix
- \item Prediction accuracy at different parts of the predicted vs observed line
- \item Comparison with a control/intercept model (10\% of everything)?
+ \item \Gls{RMSE}, \ac{MAE}, \ac{ME}, overall and per class.
+The overall measures are calculated by taking the mean of all class points pooled together, rather than taking a mean of the per-class means.
+ \item \Gls{NSE} was used to determine the goodness of fit.
+For regression output, \ac{NSE} is a method of calculating the coefficient of determination $R^2$, and is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
+ \item \Gls{SCM} \citep{silvan-cardenas_sub-pixel_2008}, and the metrics derived from it: \ac{OA}, as well as \ac{PA} and \ac{UA} per class.
+The \gls{SCM} is an adaptation of the confusion matrix concept to fractional data, where the diagonal of the matrix expresses the overlap (agreement) of the target and predicted classes, and the off-diagonal gives an estimation of which classes are confused with which.
+When using the MID-PROD operators as suggested by \citet{silvan-cardenas_sub-pixel_2008}, the \ac{SCM} provides an uncertainty measure that comes from the fact that there is no unique solution for which part of the modelled fractions is confused with which part of the validation fractions.
+ \item Correlation matrix of predicted vs observed class fractions.
+ \item Prediction accuracy at different parts of the predicted vs observed line.
+We calculated the accuracy of prediction of zeroes, hundreds, and whether the point falls in the expected half of the 1:1 line (above 50\% or below 50\%).
+The numbers obtained are averages and standard deviations of the aforementioned binomial values over individual class fractions, as well as overall.
+\item Lastly, spatial residuals, i.e. the overestimation and underestimation of each class fraction on a map.
 \end{enumerate}
@@
 
-The intercept model (always predicting the mean) achieves an \ac{RMSE} of 30\%, and \ac{OA}
+Overall model accuracy statistics are reported in table \ref{tab-accuracy}.
+\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
+\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.
 
-Logistic regression achieved RMSE of 18.7\%, MAE of 9.8\%,
+The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
+Predictions per class were made using the binary relevance method, as the models predict a single outcome unlike the \gls{NN} and \gls{MLR} models.
+This shows that despite the model training procedure involving less data for each class trained and the rescaling to 100\% requirement, the advantages of these models outweigh the disadvantages of the binary relevance method.
+% * <Dainius Masiliūnas> 18:32:47 15 Apr 2020 UTC+0200:
+% Discussion point: is a multiple Y RF possible?
+
+All tested models had difficulty discerning the vegetation classes (crops, grassland, shrub, tree) between each other, and less so with the non-vegetated land cover (water, bare soil, urban).
 
@@
 \textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & \textbf{\acrshort{NSE}} & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
+% * <Dainius Masiliūnas> 14:04:01 13 Aug 2019 UTC+0200:
+% Death to kappa
 \midrule
 Intercept
-& 30.7  & 21.8  & 0     & 24±4  & 0.11±0.04 \\
-Fuzzy nearest centroid
-& 22    &       &       & 49    & 0.41      \\
-Neural networks
-& 19.5  &       &       &       &           \\
-PLS regression
-& 19    &       &       & 47±5  & 0.36±0.07 \\
-Logistic regression
-& 20.0  & 10.7  & 0.53  & 66±4  & 0.57      \\
+& 29.9  & 21.4  & 0     & 26±5  & 0.13±0.07 \\
+\Gls{PLS} regression
+& 27.2  & 17.8  & 0.17  & 38±4  & 0.21±0.06 \\
+\Gls{FNC}
+& 24.1  & 13.7  & 0.35  & 52±4  & 0.41±0.06 \\
 Lasso regression
-& 18.5  &       &       &       &           \\
-Random forest
-& 16.6  & 9.2   & 0.67  & 68±4  & 0.60      \\
-\ensuremath{''} histogram matched
-& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
+& 22.1  & 13.3  & 0.45  & 54±4  & 0.40±0.05 \\
+\Gls{MLR}
+& 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
+\Glspl{NN}
+% * <Dainius Masiliūnas> 18:06:02 15 Apr 2020 UTC+0200:
+% Double-check NN results - were they updated?
+& 23.8  & 10.0  & 0.37  & 65±2  & 0.54±0.02 \\
+\Gls{SVM} regression
+& 20.7  & 8.9   & 0.52  & 69±2  & 0.58±0.03 \\
+Cubist regression
+& 18.1  & 8.1   & 0.63  & 72±2  & 0.63±0.03 \\
+\Gls{RF} regression
+& 17.3  & 9.4   & 0.66  & 67±4  & 0.57±0.05 \\
+%\ensuremath{''} histogram matched
+%& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
 \ensuremath{''} two-step
-& 18.3  & 8.0   & 0.59  & 72±2  & 0.63      \\
+& 19.9  & 8.2   & 0.56  & 71±2  & 0.62±0.02 \\
 \ensuremath{''} three-step
-& 18.6  & 8.2   & 0.59  & 71±4  & 0.64      \\
-\ensuremath{''} \ensuremath{''} histogram matched
-& 21.3  & 9.1   & 0.46  & 68±3  & 0.61      \\
+& 19.4  & 8.4   & 0.58  & 71±3  & 0.62±0.04 \\
+\ensuremath{''} median vote
+& 20.7  & 7.9   & 0.52  & 71±1  & 0.63±0.02 \\
+\ensuremath{''} \ensuremath{''} two-step
+& 20.0  & 8.1   & 0.54  & 72±1  & 0.63±0.02 \\
+\ensuremath{''} \ensuremath{''} three-step
+& 20.2  & 7.9   & 0.54  & 72±2  & 0.64±0.02 \\
+%\ensuremath{''} \ensuremath{''} histogram matched
+%& 21.3  & 9.1   & 0.46  & 68±3  & 0.61      \\
 \bottomrule
@@
 \end{table}
+
+% Barplots
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-1m-uncor-bar}
+    \caption{Random Forest single model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-1m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-2m-uncor-bar}
+    \caption{Random Forest two-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-2m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-3m-uncor-bar}
+    \caption{Random Forest three-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-3m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-1m-uncor-hm-bar}
+    \caption{Random Forest single model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-1m-uncor-hm}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-3m-uncor-hm-bar}
+    \caption{Random Forest three-step model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
+    \label{bar-rf-3m-uncor-hm}
+\end{figure}
+
+% Maps
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-1m-bubble}
+    \caption{Random Forest single model prediction residuals per class, spatially}
+    \label{resid-rf-1m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-3m-bubble}
+    \caption{Random Forest three-step model prediction residuals per class, spatially}
+    \label{resid-rf-3m-uncor}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-validation}
+    \caption{Validation data spatially}
+    \label{raster-validation}
+\end{figure}
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-rf-1m-uncor}
+    \caption{Random Forest single model predictions spatially}
+    \label{raster-rf-1m-uncor}
+\end{figure}
+
+\subsection{Adjustment for zero inflation}
 
@@
 
-% Barplots
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-1m-uncor-bar}
-    \caption{Random Forest single model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-1m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-2m-uncor-bar}
-    \caption{Random Forest two-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-2m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-3m-uncor-bar}
-    \caption{Random Forest three-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-3m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-1m-uncor-hm-bar}
-    \caption{Random Forest single model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-1m-uncor-hm}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-3m-uncor-hm-bar}
-    \caption{Random Forest three-step model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-3m-uncor-hm}
-\end{figure}
-
-% Maps
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-1m-bubble}
-    \caption{Random Forest single model prediction residuals per class, spatially}
-    \label{resid-rf-1m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-3m-bubble}
-    \caption{Random Forest three-step model prediction residuals per class, spatially}
-    \label{resid-rf-3m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-validation}
-    \caption{Validation data spatially}
-    \label{raster-validation}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-rf-1m-uncor}
-    \caption{Random Forest single model predictions spatially}
-    \label{raster-rf-1m-uncor}
-\end{figure}
+\subsection{Covariate importance}
 
@@
 \begin{enumerate}
+ \item Fraction mapping methods:
+ \begin{enumerate}
+    \item RF is the best, even though one model per class means that it does not take everything into account
     \item Fraction area accuracy goes up to 72\%, that is good considering that hard classification is not much better than that
- \item RF is the best, even though one model per class means that it does not take everything into account
+    \item Hardest to classify are grasslands and especially shrubs and urban
     \item Accuracy assessment metric makes a big difference, useful to use subpixel confusion matrix for that
- \item Hardest to classify are grasslands and especially shrubs and urban
+ \end{enumerate}
+ \item Optimising for zero inflation:
+ \begin{enumerate}
+    \item Multi-step models improve MAE at the cost of RMSE and NSE
+    \item Histogram matching doesn't help any
+ \end{enumerate}
+ \item Covariate importance:
+ \begin{enumerate}
     \item All covariates are important to a certain degree (show table of importances)
- \item Histogram matching doesn't help any
+ \end{enumerate}
 \end{enumerate}
@@
 
-\minisection{Author Contributions} DM JV NT MH ML BK
+\minisection{Author Contributions} DM JV NT MH ML MB BS RK
 
@@
 
-\minisection{Acknowledgements} VITO, time series group (BB)
+\minisection{Acknowledgements} VITO, Terrascope, time series group (BB, BK)
 

### 2020-04-29 article/article.tex
Changes from papeeria

@@
 
-\abstract{The current common practice for producing global land cover maps is to represent land cover classes as discrete units, e.g.
-a pixel may be labelled as grassland, tree cover, or open water.
-However, in reality, the area covered by the pixel is hardly ever homogeneous, it may be e.g.
-trees surrounded by grass next to water.
-This mismatch between the reality and its representation in maps causes a loss in precision, as only the dominant land cover class can be represented in such ``hard'' classification maps.
-This problem is worse for coarse resolution land cover maps.
-A common workaround to this issue is to define mosaic classes, i.e.
-land cover classes that are combinations of other land cover classes, such as open forest (combination of grassland and tree cover).
+\abstract{Currently most global land cover maps represent land cover classes as discrete units.
+However, in reality pixels often contain mixed land cover.
+This causes loss of precision as only the dominant land cover class may be represented.
+A common workaround to this issue is to define mosaic classes that are combinations of pure land cover classes mixed with certain ratios, e.g.
+open forest as a combination of grassland and trees.
 However, this approach leads to increasingly complex map legends and confusion about the definitions.
 
-Our proposed solution to this issue is to employ fractional land cover mapping: for each pixel, define the proportion of the pixel that each considered land cover class occupies.
-This allows expressing an open forest as e.g.
-40\% grassland, 60\% tree cover, and 0\% of the rest of the land cover classes.
-This ``fuzzy'' classification not only allows to more precisely express the land cover that exists in reality, but also gives users the flexibility to derive their own maps with a customised legend.
-For instance, if a user defines open forest as one that has less than 50\% tree cover and at least 30\% of grassland, they can generate a map with this definition from land cover fraction data, rather than relying on the definition that the mapmakers provide.
-Thus exhaustive land cover fraction data is desired by a number of user communities.
+Our proposed solution is to map land cover fractions: define the proportion that each land cover class occupies within the pixel.
+Such fuzzy classification both increases precision of the map, and also gives users the flexibility to derive their own maps with a customised legend, rather than relying on the definition that the mapmakers provide.
+Thus exhaustive land cover fraction maps are desired by a number of user communities.
 
-In this paper, we compared the performance of eight very machine learning algorithms for land cover fraction mapping: random forest regression, multilayer perceptron neural networks, partial least squares regression, fuzzy nearest prototype classification, multinomial logistic regression, lasso regression, cubist regression and support vector regression.
-In order to allow the algorithms to make good predictions about land cover fractions, six types of covariates were used: spectral bands and vegetation indices derived from the Proba-V 100 m TOC reflectance data; temporal features derived from the full time series of said imagery; terrestrial features from a digital elevation model; soil properties and climate data from global models, and location.
-We compared the contribution of each of the covariate types to improving the accuracy of the model predictions.
+We compared the performance of eight machine learning regression algorithms for land cover fraction mapping.
+We used and assessed six types of covariates for training the models: spectral bands and vegetation indices derived from Proba-V 100 m TOC reflectance; temporal features from its time series; terrestrial features from ASTER GDEM; soil properties from SoilGrids; climate data from WorldClim; and location.
+The models were trained on 138164 training points throughout the globe, describing land cover fractions of pixels aligned to the Proba-V 100 m grid.
+After training, their land cover class fraction predictions were compared with an independent validation set of 20705 points, likewise aligned to the Proba-V grid.
 
-One challenge when dealing with fractional data is the data imbalance: in most locations, only a single or a few classes dominate a pixel, with the rest taking up 0\% of the pixel.
-Values at the extremes (0\% and 100\%, i.e.
-pure pixels) are difficult for machine learning algorithms to estimate.
-We propose a new solution to this problem by combining three models: two classification models and one regression model.
-The first classification model determines whether a pixel is pure, and if so, it is processed using a classification model, otherwise it is processed with a regression model.
+One challenge for fraction mapping models is data sparsity: most fractions are 0\%, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
+We propose a new solution by combining three models.
+A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.
 
-% Validation
-% - 28k points
-% - Independent training and validation sets
+Results show that the three-model approach results in better estimation of extreme values (lower MAE of 8.4 vs 9.4) at the cost of misclassification (higher RMSE of 19.4 vs 17.3).
+From the compared models, random forest and cubist regressions reach the highest overall accuracy of 72\%, comparable to the reported accuracies of the current discrete global land cover maps.
+The models were driven primarily by spectral remote sensing data, followed by soil and climate data.
 
-The models were trained on 138164 training points throughout the globe.
-The training points, collected as part of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project, describe the fraction that each land cover class (bare soil, crops, grassland, shrubs, trees, urban/built-up, water and wetlands) occupies within a 100\texttimes100 m area, which is aligned to the grid used for Proba-V 100 m products.
-After the models were trained, their land cover class fraction predictions were compared with an independent validation set of 20705 points, which are likewise aligned to the Proba-V grid.
-Overall as well as per-class RMSE, MAE, and ME statistics were used to compare the models.
-In addition, the subpixel confusion-uncertainty matrix was used to analyse which classes were the most difficult to separate for each model, as well as to derive overall, users' and producers' accuracies and the kappa statistic.
-The models were compared with each other, as well as against a control intercept-only model.
-
-Results show that the three-model approach increases the overall accuracy of the predictions by more accurately modelling the extreme values.
-From the machine learning models that were compared, random forest and cubist regressions reaches the highest overall accuracy of 72\%.
-This is comparable to the reported accuracies of the current ``hard'' global land cover maps.
-Random Forest covariate importance analysis shows that spectral covariates are the most important for predicting land cover fractions, followed by the location, terrestrial and temporal covariates.
-
-The application of the method on data gathered throughout the globe shows the applicability of the method to a variety of land cover conditions, ranging from highly heterogeneous agricultural landscapes to homogeneous bodies of water.
-This research proves that machine learning algorithms can be applied globally to create a global fractional land cover product.
-Such a product would cater to users with different fields of focus.
-Among other benefits, more precise estimation of land cover allows for 1) improved climate models to help reach Sustainable Development Goals (for which land cover is a key variable), 2) facilitate the making of models for vegetation and urban area dynamics, and 3) help governments and land owners keep better track of relevant land cover and its change.
-In addition, this work contributes to the operationalisation of the CGLS-LC100 project, which aims to produce exhaustive global land cover fractions as an operational product.
+This research proves that machine learning algorithms can be applied globally to a wide variety of land cover to create an exhaustive global land cover fraction map.
+This expresses land cover more precisely, and empowers users to create their own discrete maps, which helps land cover monitoring and climate modelling communities.
+In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product.
 

### 2020-04-29 article/article.tex
Make abstract more concise

@@
 
-\abstract{Currently most global land cover maps represent land cover classes as discrete units.
-However, in reality pixels often contain mixed land cover.
-This causes loss of precision as only the dominant land cover class may be represented.
-A common workaround to this issue is to define mosaic classes that are combinations of pure land cover classes mixed with certain ratios, e.g.
-open forest as a combination of grassland and trees.
-However, this approach leads to increasingly complex map legends and confusion about the definitions.
+\abstract{Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio.
+In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision, reduces legend complexity, and allows users the flexibility of defining their own mosaic discrete classes using thresholds they define themselves.
+To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.
 
-Our proposed solution is to map land cover fractions: define the proportion that each land cover class occupies within the pixel.
-Such fuzzy classification both increases precision of the map, and also gives users the flexibility to derive their own maps with a customised legend, rather than relying on the definition that the mapmakers provide.
-Thus exhaustive land cover fraction maps are desired by a number of user communities.
+We compared the performance of eight machine learning regression algorithms for land cover fraction mapping, for the best model assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
 
-We compared the performance of eight machine learning regression algorithms for land cover fraction mapping.
-We used and assessed six types of covariates for training the models: spectral bands and vegetation indices derived from Proba-V 100 m TOC reflectance; temporal features from its time series; terrestrial features from ASTER GDEM; soil properties from SoilGrids; climate data from WorldClim; and location.
-The models were trained on 138164 training points throughout the globe, describing land cover fractions of pixels aligned to the Proba-V 100 m grid.
-After training, their land cover class fraction predictions were compared with an independent validation set of 20705 points, likewise aligned to the Proba-V grid.
+The models were trained on over 130\,000 points and validated on a separate dataset of over 20\,000 points of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project.
+Both datasets are global and aligned with the Proba-V 100 m grid.
 
 One challenge for fraction mapping models is data sparsity: most fractions are 0\%, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
-We propose a new solution by combining three models.
+We proposed a new solution by combining three models.
 A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.
 
-Results show that the three-model approach results in better estimation of extreme values (lower MAE of 8.4 vs 9.4) at the cost of misclassification (higher RMSE of 19.4 vs 17.3).
+Results showed that the three-model approach results in better estimation of extreme values (lower MAE of 8.4 vs 9.4) at the cost of misclassification (higher RMSE of 19.4 vs 17.3).
 From the compared models, random forest and cubist regressions reach the highest overall accuracy of 72\%, comparable to the reported accuracies of the current discrete global land cover maps.
@@
 This expresses land cover more precisely, and empowers users to create their own discrete maps, which helps land cover monitoring and climate modelling communities.
-In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product.
+In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product, and to assess the accuracy of land cover fraction maps both thematically and spatially.
 

### 2020-05-13 article/article.tex
Writing results section

@@
 \newacronym{CART}{CART}{Classification and regression tree}
+\newacronym{GLM}{GLM}{General linear regression model}
 \makenoidxglossaries
@@
 \subsection{Covariate selection}
+\label{sec-covariate-selection}
 
@@
 \begin{enumerate}
- \item RMSE per class comparison: no big differences
- \item Subpixel confusion matrix metrics: quite a big difference
- \item RF variable importance
- \item Truth:prediction scatterplots (hex/box/bar)
- \item Point map of Africa for the different classes
+ \item Differences between all models, showing the RF has lowest RMSE
+ \item Variable importance for RF; stressing importance of RS data by a model with only RS variables
+ \item Spatial wall-to-wall demo map and spatial errors
 \end{enumerate}
@@
 
-Overall model accuracy statistics are reported in table \ref{tab-accuracy}.
+Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
 \Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
@@
 
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/boxplots/2020-05-08-model-comparison-bar}
+    \caption{Comparison of the distribution of absolute errors between different models tested, ordered by decreasing overall \gls{MAE}.
+Dotted lines indicate \gls{MAE}, box indicates median and 75th/25th percentiles.}
+    \label{fig-models}
+\end{figure}
+
 The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
-Predictions per class were made using the binary relevance method, as the models predict a single outcome unlike the \gls{NN} and \gls{MLR} models.
-This shows that despite the model training procedure involving less data for each class trained and the rescaling to 100\% requirement, the advantages of these models outweigh the disadvantages of the binary relevance method.
-% * <Dainius Masiliūnas> 18:32:47 15 Apr 2020 UTC+0200:
-% Discussion point: is a multiple Y RF possible?
+Predictions per class were made using the binary relevance method (one model per class), as these models are univariate.
+This shows that despite the binary relevance limitations of each model only being aware of its relevant class and not the others, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.
 
-All tested models had difficulty discerning the vegetation classes (crops, grassland, shrub, tree) between each other, and less so with the non-vegetated land cover (water, bare soil, urban).
+Multivariate linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
+The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
+This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
+It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.
+
+\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
+This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
+This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
+
+\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods, but better than all of the multivariate linear models.
+
+There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large misclassifications or not.
+Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
+Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk misclassification, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely wrong.
+
+All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil).
+This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
+In contrast, the tree fraction has a lot of samples with 100\% fraction and it deviates a lot per each sample, making it a highly unpredictable class from the data alone.
+Most models showed good improvement in predicting tree cover compared to the intercept model, indicating that the covariates used were useful in telling trees apart from other classes.
+In contrast, there were only moderate improvements for predicting the grass fraction, as it is difficult for the classifiers to distinguish this class with the given input covariates.
 
@@
 & 29.9  & 21.4  & 0     & 26±5  & 0.13±0.07 \\
-\Gls{PLS} regression
-& 27.2  & 17.8  & 0.17  & 38±4  & 0.21±0.06 \\
 \Gls{FNC}
-& 24.1  & 13.7  & 0.35  & 52±4  & 0.41±0.06 \\
+& 24.4  & 13.5  & 0.33  & 53±4  & 0.42±0.06 \\
+General linear model and \Gls{PLS}
+& 21.6  & 12.7  & 0.48  & 56±4  & 0.43±0.05 \\
 Lasso regression
-& 22.1  & 13.3  & 0.45  & 54±4  & 0.40±0.05 \\
+& 21.4  & 12.6  & 0.49  & 56±4  & 0.43±0.05 \\
 \Gls{MLR}
 & 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
+Multivariate RF
+& 18.7  & 10.0  & 0.61  & 65±3  & 0.55±0.05 \\
 \Glspl{NN}
@@
 % Double-check NN results - were they updated?
-& 23.8  & 10.0  & 0.37  & 65±2  & 0.54±0.02 \\
+& 22.7  & 9.2   & 0.43  & 68±1  & 0.57±0.02 \\
 \Gls{SVM} regression
@@
 
-% Barplots
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-1m-uncor-bar}
-    \caption{Random Forest single model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-1m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-2m-uncor-bar}
-    \caption{Random Forest two-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-2m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-19-rf-3m-uncor-bar}
-    \caption{Random Forest three-step model prediction accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-3m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-1m-uncor-hm-bar}
-    \caption{Random Forest single model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-1m-uncor-hm}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/barplots/2019-03-22-rf-3m-uncor-hm-bar}
-    \caption{Random Forest three-step model prediction (histogram matched) accuracies per class, per predicted magnitude (extremes vs middle)}
-    \label{bar-rf-3m-uncor-hm}
-\end{figure}
-
 % Maps
@@
 \subsection{Adjustment for zero inflation}
-
-% 1:1 plots
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-points-a001s2}
-    \caption{Random Forest single model prediction correspondence to ground truth (raw points with overplotting).}
-    \label{raw-rf-1m-uncor}
-\end{figure}
-% Binplots
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n100s20}
-    \caption{Random Forest single model prediction correspondence to ground truth, 100x100 bins.}
-    \label{bin-rf-1m-uncor-n100}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n50s20sqrt}
-    \caption{Random Forest single model prediction correspondence to ground truth, 50x50 bins, square root transformation.}
-    \label{bin-rf-1m-uncor-n50sqrt}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-point-n50s10sqrt}
-    \caption{Random Forest single model prediction correspondence to ground truth, 50x50 bins, square root transformation, smaller circle sizes.}
-    \label{bin-rf-1m-uncor-n50sqrts10}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/binplots/2019-04-12-rf-1m-uncor-raster-n50sqrt}
-    \caption{Random Forest single model prediction correspondence to ground truth, square root transformation.}
-    \label{bin-rf-1m-uncor-rastersqrt}
-\end{figure}
-
-% Hexplots
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-1m-uncor-hex}
-    \caption{Random Forest single model prediction correspondence to ground truth.}
-    \label{hex-rf-1m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hex}
-    \caption{Random Forest three-step model prediction correspondence to ground truth.}
-    \label{hex-rf-3m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-1m-uncor-hm-hex}
-    \caption{Random Forest single model prediction (histogram matched) correspondence to ground truth.}
-    \label{hex-rf-1m-uncor-hm}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hm-hex}
-    \caption{Random Forest three-step model prediction (histogram matched) correspondence to ground truth.}
-    \label{hex-rf-3m-uncor-hm}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hma-hex}
-    \caption{Random Forest three-step model prediction (histogram matched middle, proportionally) correspondence to ground truth.}
-    \label{hex-rf-3m-uncor-hma}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/hexplots/2019-03-22-rf-3m-uncor-hmb-hex}
-    \caption{Random Forest three-step model prediction (histogram matched non-extremes) correspondence to ground truth.}
-    \label{hex-rf-3m-uncor-hmb}
-\end{figure}
-
-% Boxplots
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-1m-uncor-box}
-    \caption{Random Forest single model prediction correspondence to ground truth.}
-    \label{box-rf-1m-uncor-hm}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-2m-uncor-box}
-    \caption{Random Forest two-step model prediction correspondence to ground truth.}
-    \label{box-rf-2m-uncor-hm}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=0.8\textwidth]{article-figures/boxplots/2019-03-19-rf-3m-uncor-box}
-    \caption{Random Forest three-step model prediction correspondence to ground truth.}
-    \label{box-rf-3m-uncor-hm}
-\end{figure}
 

### 2020-05-14 article/article.tex
Changes from papeeria

@@
 \subsection{Model tuning (multi-step approach) to account for value imbalance}
+\label{sec-multistep}
 
@@
 As the dataset is dominated by zeroes and hundreds, minimising the objective function tends to draw the model towards predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
-This is not desirable for users of land cover fraction data, as the added value of fractions is the information about the middle of the range; otherwise, discrete classification would be just as good.
+This is not desirable for users of land cover fraction data, as the added value of fraction information is the information about the middle of the range; otherwise, discrete classification would be just as good.
 Therefore we tried several approaches to increase the accuracy of the middle predictions.
@@
 Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk misclassification, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely wrong.
+
+Overall accuracy and kappa statistics from the \gls{SCM} coincided with the \gls{MAE} scores, whereas \gls{NSE} coincided with the \gls{RMSE} scores.
+This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
+The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
 
@@
 
+Since \gls{RF} regression resulted in the lowest RMSE, it was used for testing the effect of the multi-step approach detailed in section \ref{sec-multistep}.
+The overall results can be seen at the bottom of table \ref{tab-accuracy}.
+
+For an \gls{RF} regression when the mean of all tree votes was used as the indicator for the land cover fraction, using a two-step approach improved the \gls{MAE} of the model, while increasing \gls{RMSE}.
+This is because the false positives in the first step of the two-step model (the binary zero/non-zero classification) result in some high fractions nevertheless being set to zero.
+But on average it does improve the accuracy, since without this adjustment, \gls{RF} regression would almost never assign the precise value of 0\% to a fraction, and thus result in some error for all pixels.
+With this adjustment, it is possible to predict the result perfectly in the cases where the fraction is 0\%, which is the most common case.
+
+However, the two-step approach combined with binary relevance also introduced a drawback in that there were cases where all land cover class fractions were predicted to be 0\% in the first step.
+In such cases it was not possible to determine the land cover fractions, and to make them sup up to 100\%, they were all set to equal parts ($100\% / 7$), which introduced further error.
+
+The three-step approach solved this issue, since the first step would not predict zeroes but rather purity.
+In the case of pure pixels, they would be passed to \gls{RF} classification, which would always return the most likely pure class.
+For mean vote \gls{RF}, the result was not significantly different from the two-step approach.
+While this approach improved also the case of predicting the 100\% fractions precisely, the addition of another model caused extra error that would propagate through the process, negating the gain.
+
+Another option explored for dealing with the issue of zero inflation was to use the median of all tree votes in \gls{RF} instead of the mean.
+This had a comparable effect to using the multi-step approach, as it resulted in a lot more precise predictions for fractions with 0\%.
+However, in the cases with high uncertainty, the median vote was much more likely to predict a 100\% fraction, and thus was also more likely to misclassify a pixel completely.
+Thus it resulted in much higher \gls{RMSE} and much lower \gls{MAE}.
+Using a multi-step approach together with median vote resulted in little difference, as the two approaches have a similar effect.
+The three-step approach was an overall improvement to the single-step approach, albeit a small one.
+
 \subsection{Covariate importance}

### 2020-05-27 article/article.tex
Writing first draft of discussion, conclusions

@@
 \usepackage[sort&compress,sectionbib]{natbib}
-\bibliographystyle{newapa}
+\usepackage{graphicx}
+\bibliographystyle{apalike}
 
@@
 \usepackage[norefs,nocites]{refcheck} % Develop: warn if figure is not referenced
-
-% Track changes package, remove when final; using a drop-in rather than the old version in TeX Live
-% Usage: \added, \deleted, \replaced{new}{old}, \highlight, \comment
-% All followed by an optional [id=<your id>, comment=<some comment>]
-\usepackage[xcolor={usenames,dvipsnames}]{changes}
-\definecolor{gem}{HTML}{003300}
-\definechangesauthor[color=gem]{dm}
-\definechangesauthor[color=NavyBlue]{jv}
-\definechangesauthor[color=Purple]{nt}
-\definechangesauthor[color=Mahogany]{mh}
 
@@
 \newacronym{NDVI}{NDVI}{Normalised Difference Vegetation Index}
+\newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
+\newacronym{NIRv}{NIRv}{\gls{NIR} of vegetation}
 \newacronym{NIR}{NIR}{near infra-red}
@@
 \linenumbers
+
+\minisection{Highlights}
+
+\begin{itemize}
+    \item 11 statistical and machine learning models tested for land cover fraction mapping
+    \item Random forest regression performed the best, with RMSE of 17.3 globally
+    \item A three-step regression method proposed, which results in the lowest MAE of 7.9
+    \item Remote sensing data was the most important for model predictions, but all auxiliary data also contributed significantly
+    \item Global wall-to-wall land cover fraction maps produced and their errors assessed
+\end{itemize}
 
@@
 
-\begin{enumerate}
- \item Relevance of fractional LC mapping (link to user requirements and SDGs)
- \item Studies that have looked into fractional LC mapping so far
- \item Available fractional methods that were introduced in literature (wavelet transform etc.)
- \item Methods that are tested in this paper
- \item Global land cover maps
-\end{enumerate}
+% \begin{enumerate}
+%  \item Relevance of fractional LC mapping (link to user requirements and SDGs)
+%  \item Studies that have looked into fractional LC mapping so far
+%  \item Available fractional methods that were introduced in literature (wavelet transform etc.)
+%  \item Methods that are tested in this paper
+%  \item Global land cover maps
+% \end{enumerate}
 
 Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
-Some of the recent achievements have been the \ac{LC-CCI} map \highlight[id=dm, comment=Is there a better citation?]{\citep{defourny2012cci}} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \highlight[id=dm, comment=ditto]{\ac{CGLS-LC100}} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
+Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \ac{CGLS-LC100} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
 
@@
 A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
-This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}. A related term is ``super
+This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
+Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but this concept relates also to the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.
+
 Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g.
@@
 So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor a global scale.
-
-%The second drawback of current global land cover products is that their classification accuracy tends to be low, averaging at around 65\% \citep{tsendbazar2016integrating}.
-The third drawback is that all of these products use what is known as ``hard'' or ``crisp'' classification: each pixel is assigned to one particular land cover class only.
-The result of a hard classification is a thematic map.
-This classification type treats each pixel as homogeneous with regards to land cover, which is rarely the case in reality.
-In contrast, a ``fuzzy'' or ``soft'' classification, sometimes also called ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent} identifies the proportion of each land cover class within each pixel.
-The result of a fuzzy classification is one map per land cover class, showing the proportion of the pixel area covered by the land cover class.
-
-%Hard classification is not well-suited for coarse resolution imagery, due to a large proportion of mixed pixels in it (representing areas of mixed land cover), compared to endmember pixels (or pure pixels, which are dominated by a single land cover class).
-When hard classification is attempted on such imagery, the accuracy of the result can be no higher than the area fraction that the dominant land cover type of the pixel occupies \citep{latifovic2004accuracy}.
-There have been attempts at increasing accuracy by defining mosaic classes, such as mixed forests (50\% needleleaf and 50\% broadleaf trees), but it leads to a proliferation of classes that increases map complexity and still does not deal with the core problem, thus not improving the classification accuracy by much \citep{tsendbazar2016comparative}.
-In contrast, ``fuzzy'' or ``soft'' classification results in each pixel containing information about the proportion of each class within that pixel, therefore operating on sub-pixel scales.
-This has the advantage of representing land cover more accurately, and gives the ability to represent mosaic classes as a combination of pure classes instead.
-Since the output of fuzzy classification is one raster per land cover class, with smooth transitions between classes, it is suitable for more in-depth analysis and user-specific visualisation criteria \citep{tsendbazar2016integrating}.
-Despite that, hard classification is still the most often used classification type due to the number of algorithms developed for it, ease of storing the result (it is thematic and thus takes little storage space), displaying it in a single map (albeit in a less accurate fashion) and performing accuracy assessment.
-
-%There is a number of different classification algorithms, but only several of them are suitable to be used for fuzzy classification \citep{nath2014methods}.
-The two methods most commonly used in scientific literature are fuzzy \textit{c}-means and neural networks \citep{zhang2001fullyfuzzy}.
-Neural networks in particular are well-suited for fuzzy classification, since they allow multiple continuous output as well as input variables in a single model \citep{foody1997fuzzynnet}.
-Fuzzy \textit{c}-means, also known as fuzzy \textit{k}-means or soft \textit{k}-means, is a statistical method that relies on the proximity of pixels in feature space to class centroids, and thus is also suitable for fully fuzzy classification.
-However, since \textit{k}-means is an unsupervised classification algorithm, the ability to make use of training data in fuzzy \textit{c}-means is limited to determining class centroids with more precision \citep{hengl2004fuzzycmeans}.
-
-%In addition, any other algorithms that provide a measure of uncertainty about class membership can be used, such as the ratio of individual tree votes of random forest \citep{breiman2001random} or class probabilities in gradient boosting \citep{friedman2001gradientboost}.
-While classification uncertainty is not a direct measure of class membership \citep{sytze2000fuzzyset}, they are nonetheless correlated.
-As such, it has been used by numerous authors as a proxy indicator of class membership, achieving satisfactory classification accuracy \citep{foody2002accuracy}.
-
-%Furthermore, algorithms that can handle continuous variables, but only one response variable (such as random forest), can be used as well, by creating separate models for every class.
-This approach is called binary relevance \citep{karalas2016br}.
-Using this approach, the data needs to be post-processed after classification at a pixel level to conform to physical constraints (class membership must be between 0 and 100\% and sum up to 100\%).
-Random forest has been reported to give higher or equal accuracy results compared to other algorithms, like Support Vector Machines and individual decision trees, in hard classification scenarios using satellite imagery similar to that of PROBA-V \citep{duro2012algorithmcomparison}.
-Random forest regression was also shown to perform as well as other algorithms in fuzzy classification scenarios \citep{walton2008subpixelrf}, although it is used in much fewer studies on fuzzy classification than the other algorithms mentioned previously.
-Gradient boosting is an algorithm related to random forest, with the same advantages and disadvantages, but it is known to perform better in machine learning challenges \citep{chen2015higgs} and has not yet been tested on fuzzy land cover classification.
-
-%Accuracy assessment is rather straight-forward for hard classification: typically a confusion matrix is employed for this purpose, showing how many pixels (or in some cases how much area \citep{stehman2009sampling}) in the image have been classified correctly, and how many incorrectly.
-Such a matrix makes it simple to tell which classes are hard to discern from one another, as well as allows for deriving statistics such as users' accuracy, producers' accuracy, and overall accuracy \citep{foody1996fuzzyevaluation}, as well as variance if probability sampling was used.
-However, a standard confusion matrix is not applicable in the context of fuzzy classification, since misclassification in this case is not absolute, but rather a matter of degree \citep{foody2002accuracy}.
-Several solutions, such as distance \citep{foody1996fuzzyevaluation}, cross-entropy and mutual information \citep{lu2007methods} have been suggested as alternatives for fuzzy classification accuracy assessment.
-
-%Visualisation of the fuzzy classification results is also challenging, since each class effectively is a single-channel raster of its own.
-Three such classes can easily be combined into RGB channels for visualisation, but with more classes it is no longer possible.
-There have been attempts to develop a method based on the hue, saturation and intensity colour model to allow for a larger number of distinct classes to be visualised on a single raster \citep{hengl2004fuzzycmeans}.
-There is also a possibility of ``hardening'' the classification when high accuracy visualisation is not needed, and making use of multiple RGB rasters side-by-side when it is needed.
-
-%All in all, even though global land cover mapping has been a focus of many other studies, existing global land cover maps still have room for improvement.
-This can be achieved by using data from the sensors of newer satellites such as PROBA-V, as well as performing fuzzy classification as opposed to the traditional hard classification.
-
-%An important long-term goal of land cover mapping is to have a highly accurate fuzzy land cover classification method at a global scale.
-Since there are many different methods that could be used to achieve such a goal, and since the amount of data generated by satellites covering the entire world is enormous, this thesis focuses on comparing several different fuzzy classification methods on a subset of global imagery provided by the PROBA-V satellite.
-
-%PROBA-V is a relatively new satellite, with few studies using it for land cover classification so far.
-A number of studies have used simulated PROBA-V data in preparation for its launch \citep{stathakis2014probavurban,roumenina2013probavcrops,bartalev2014probavboreal}.
-After launch, most studies have focused on its potential for crop classification \citep{roumenina2015probavcrops,durgun2016crop,lambert2016cropland}.
-All of these studies have only used the traditional hard classification.
-As such, the potential of using PROBA-V data for fuzzy classification in order to obtain a more accurate representation of land cover at global scale has never been tested before.
-
-%Another problem of land cover classification is that there is high disagreement between different land cover products on mixed classes, such as mixed forests \citep{Herold2008lccomparison}.
-Mixed forests and wetlands are challenging to detect and classify using hard classification and optical remote sensing data, so it is important to check how well large-scale fuzzy classification performs on these classes.
-Therefore this thesis focused specifically on the region of ecotone from boreal forests to temperate broadleaf forests in Europe.
-This region also includes a large number of boreal wetlands.
-
-%In addition, there is a wide variety of methods that can be used for fuzzy classification.
-Most studies on this topic focus on comparing only two methods at once.
-In this thesis, four of them have been tested: fuzzy \textit{c}-means, neural networks, random forest regression and multiclass gradient boosting.
-All of these methods are based on fundamentally different approaches to classification, so the results give insight into the possible performance of similar methods as well.
-In addition, the last of the mentioned methods has not yet been tested on fuzzy classification itself (although related methods have been).
-
-%When it comes to global image processing, one problem is the amount of data and the time it takes to process it.
-Some algorithms are known to suffer from the curse of dimensionality, where a linear increase in input data results in a geometric increase in processing time \citep{walton2008subpixelrf}.
-Therefore it is important to also measure the processing time of different classification methods.
 
@@
 
-For deriving land cover fractions, we compared a wide array of machine learning regression methods: \ac{FNC}, \ac{MLP} \glspl{NN}, \ac{PLS} regression, Lasso regression, \ac{MLR}, \ac{RF} regression.
-% * <Dainius Masiliūnas> 16:05:16 07 Aug 2019 UTC+0200:
-% Perhaps also SVM and MESMA
+For deriving land cover fractions, we compared a wide array of machine learning regression methods: \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression, \ac{MLR}, \ac{MLP} \glspl{NN}, \glspl{SVM}, Cubist and \ac{RF} regression.
 In addition, as a baseline we also compare the results with an intercept model (all fractions always predicted to be equal).
@@
 All of the models were made using the free and open-source statistical software R \citep{r_2019}.
+
+\subsubsection{\Glsentryfull{GLM}}
+
+\Gls{GLM}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
+In this study, we used \gls{GLM} as a baseline model.
+\Gls{GLM} does not normalise the output to add up to 100\%, and also allows negative predictions.
+Therefore all predictions < 0 were capped to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
+
+We also tested whether dropping any covariates would result in better predictions, using a sphericity test with Greenhouse-Geisser correction to compare models with different covariates.
+However, only three covariates could be dropped and it did not improve the predictions, therefore we decided to keep all the covariates in the model.
+\Gls{GLM} is implemented in base R \citep{r_2019}.
 
@@
 
-Since we performed a manual covariate selection prior to running regularisation to remove highly correlated covariates, and having a further reduced set of covariates is useful for other algorithms, for this method we chose Lasso regression.
-Lasso regression builds upon least squares, and thus the classes are predicted separately to each other, using the binary relevance approach as presented by \citet{karalas2016br}.
-Therefore, the output fractions do not necessarily sum to 100\%.
-As such, they were linearly scaled.
+We chose lasso regression, so that it would be possible to identify if there are any covariates that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
+The $\lambda$ parameter was chosen using cross-validation.
+
 The lasso regression was performed using the \texttt{glmnet} package \citep{glmnet}.
+This method is multivariate in the same way that \gls{GLM} is, and the output was clamped to positive numbers and rescaled to add up to 100\% the same way.
 
@@
 As such, it is well suited for multicollinear data.
-Due to this property, the \highlight[id=nt,comment={would you not agree for better comparability just use the same set of covariates?}]{full set of covariates} were used as input to this model, rather than the reduced (uncorrelated) set.
+This method is also multivariate in the same way that \gls{GLM} is, and the same method was used to clamp and rescale the predictions.
 \ac{PLS} regression was performed using the package \texttt{pls} \citep{pls}.
@@
 This method always produces output that sums to 100\%.
+
 In this study, we used an implementation adapted from the package GSIF \citep{hengl2004fuzzycmeans}.
+We tested both using a logistic regression to estimate the class centroids, as well as using a weighted average of the training data to do so, and ended up using the latter method as the result was more accurate.
 
@@
 
-\Glspl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\%.
-In this study, we used a multilayer perceptron with two layers of 256 neurons and softmax activation.
+\Glspl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for any postprocessing.
+In this study, we used a multilayer perceptron with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+Softmax activation was used to get the output.
+The Nadam optimiser \citep{dozat_incorporating_2016} was used, with \gls{MAE} as the loss function.
 The \texttt{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics card to accelerate the neural network training process.
@@
 
-\ac{RF} regression is a popular method for land cover classification that works by building a number of decision trees based on random subsets of the input training data, and averaging the ``votes'' of these individual trees.
+\ac{RF} regression is a popular method for land cover classification that works by building a number of decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual trees \citep{breiman2001random}.
 \ac{RF} can only take one variable as the response variable, therefore there are two ways to obtain land cover fraction predictions from an \ac{RF} model.
 One way is analogous to how we performed \ac{MLR}: giving labels of the dominant land cover class as the response variable, and then considering the proportion of votes from the trees as the proportions of land cover classes in the pixel.
-The second way is binary relevance, where a separate model is trained to predict the presence of each class.
-Then the predictions are scaled to sum up to 100\%.
-In this study, we chose the second approach, so that the model for each class would be able to learn from the fractions of all the input data, rather than just the labels of the dominant land cover class.
+The second way is binary relevance, where a separate model is trained to predict the presence of each class \citet{karalas2016br}.
+Then the predictions are scaled linearly to sum up to 100\%.
+In the case of \gls{RF}, output values do not need to be clipped as their range is always within the constraints of the input data range.
+In this study, we chose the binary relevance approach, so that the model for each class would be able to learn from the fractions of all the input data, rather than just the labels of the dominant land cover class.
 The processing was done using the \texttt{ranger} package \citep{ranger}.
@@
 \ac{RF} was also used for assessing the covariate importance.
+Permutation importance was used: each covariate was shuffled in turn, and the resulting decrease in \gls{RMSE} was calculated, comparing against the validation data set.
 Since the number of covariates is imbalanced by group (i.e.
-there are merely 3 location covariates and 168 soil covariates, which are often collinear), covariate importance was assessed by permuting all covariates in a group and reporting the decrease in \ac{RMSE} resulting from that.
+there are merely 3 location covariates and 22 spectral covariates), and there may still be colinear covariates, we also tested calculating covariate importance by permuting all covariates in a group together.
+However, the result was similar to the sum of permutation importance of each individual covariates in the group, which showed that the covariate selection process had eliminated all colinear covariates.
+Thus we reported the more detailed covariate importance permuted per individual covariate.
+
+\subsubsection{Cubist regression}
+
+Cubist regression \citep{cubist} is based on \gls{RF} regression, with the major difference being that instead of using a threshold of a covariate for a split in the decision tree, Cubist instead uses a linear regression based on a subset of the data relevant for the split in question.
+In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previous ones.
+For our dataset, 10 committees gave the best result.
+Like with \gls{RF}, Cubist was used with the binary relevance method.
+The \texttt{Cubist} package \citep{cubistpackage} was used for this.
+
+\subsubsection{\Glsentryfull{SVM} regression}
+
+\Glspl{SVM} are machine learning models that attempt to find the optimal boundary between the class clusters in feature space by constructing an optimal dividing hyperplane.
+For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
+We used the binary relevance method, clamped the output values to the 0-100\% range and linearly scaled them to add up to 100\%.
+The package \texttt{liquidSVM} \citep{liquidSVM} was suitable for this purpose, as it is optimised for processing large datasets such as ours.
+
+\subsubsection{Other methods}
+
+A few more methods were considered, but in the end turned out to be unworkable with our dataset.
+Multivariate \gls{RF} in available in R provided by the \texttt{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
+However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
+
+Another model that was considered was SuperLearner \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
+Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
+Whereas setting the number of folds to a lower number resulted in worse results than when using a single classifier.
+
+Lastly, multiple-endmember spectral unmixing using non-negative least squares was also attempted, but the results were significantly worse than even the intercept model, possibly because the class fractions are not a linear combination of the covariates.
 
@@
 In addition, we investigated applying histogram matching, by matching the histogram of the predictions back to the histogram of the training set.
+However, that turned out to decrease accuracy in all cases, as the cases where a model predicts a value close to the true value would be set to a more extreme value to compensate for other predictions that were not as accurate to begin with.
+Therefore in the end histogram matching was not used.
 
@@
 \section{Results}
-
-\begin{enumerate}
- \item Differences between all models, showing the RF has lowest RMSE
- \item Variable importance for RF; stressing importance of RS data by a model with only RS variables
- \item Spatial wall-to-wall demo map and spatial errors
-\end{enumerate}
 
@@
 \textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & \textbf{\acrshort{NSE}} & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
-% * <Dainius Masiliūnas> 14:04:01 13 Aug 2019 UTC+0200:
-% Death to kappa
 \midrule
@@
 & 24.4  & 13.5  & 0.33  & 53±4  & 0.42±0.06 \\
-General linear model and \Gls{PLS}
+\Gls{GLM}, \Gls{PLS}, Lasso
 & 21.6  & 12.7  & 0.48  & 56±4  & 0.43±0.05 \\
-Lasso regression
-& 21.4  & 12.6  & 0.49  & 56±4  & 0.43±0.05 \\
+%Lasso regression
+%& 21.4  & 12.6  & 0.49  & 56±4  & 0.43±0.05 \\
 \Gls{MLR}
 & 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
-Multivariate RF
-& 18.7  & 10.0  & 0.61  & 65±3  & 0.55±0.05 \\
+%Multivariate RF
+%& 18.7  & 10.0  & 0.61  & 65±3  & 0.55±0.05 \\
 \Glspl{NN}
-% * <Dainius Masiliūnas> 18:06:02 15 Apr 2020 UTC+0200:
-% Double-check NN results - were they updated?
 & 22.7  & 9.2   & 0.43  & 68±1  & 0.57±0.02 \\
@@
 \Gls{RF} regression
-& 17.3  & 9.4   & 0.66  & 67±4  & 0.57±0.05 \\
+& \textbf{17.3}  & 9.4   & \textbf{0.66}  & 67±4  & 0.57±0.05 \\
 %\ensuremath{''} histogram matched
@@
 \ensuremath{''} \ensuremath{''} three-step
-& 20.2  & 7.9   & 0.54  & 72±2  & 0.64±0.02 \\
+& 20.2  & \textbf{7.9}   & 0.54  & \textbf{72±2}  & \textbf{0.64±0.02} \\
 %\ensuremath{''} \ensuremath{''} histogram matched
@@
 \end{table}
-
-% Maps
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-1m-bubble}
-    \caption{Random Forest single model prediction residuals per class, spatially}
-    \label{resid-rf-1m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-03-26-rf-3m-bubble}
-    \caption{Random Forest three-step model prediction residuals per class, spatially}
-    \label{resid-rf-3m-uncor}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-validation}
-    \caption{Validation data spatially}
-    \label{raster-validation}
-\end{figure}
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/maps/2019-04-10-rasterised-rf-1m-uncor}
-    \caption{Random Forest single model predictions spatially}
-    \label{raster-rf-1m-uncor}
-\end{figure}
 
@@
 
+\Gls{RF} permutation importance was performed to test which variables were important for \gls{RF} regression predictions.
+The single-step \gls{RF} model was selected for this as it achieved the lowest \gls{RMSE}.
+The results are shown in figure \ref{fig-varimp}.
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/heatmaps/2020-05-12-varimp-heatmap-top10}
+    \caption{Random Forest single model variable importance, top 10 covariates per category.}
+    \label{fig-varimp}
+\end{figure}
+
+The overall most important variables were the maximum and minimum NDVI over the whole time series of Proba-V imagery.
+They were followed by the median NDMI and NIRv over the whole time series.
+These vegetation indices were of lower importance only for the shrub and crop classes.
+These two classes are complicated to distinguish from other natural vegetation, and thus benefited from a wide variety of additional covariates, including location and climate information.
+
+To distinguish crops in particular, soil information such as pH, bulk density and sand fraction was useful for the model.
+Crops are usually grown in fertile soils that can sustain them, and due to the fact that cropland is further managed, the soil is also altered to be more fertile, which affects these soil properties.
+In addition, the crop prediction model benefited from harmonic metrics derived from time series the most.
+The second order amplitude and sine of the harmonic model of the time series allows the model to detect areas with a double harvest throughout the year, which is indicative of crops.
+
+Climate data was also most useful for predicting crops, especially the mean temperature and the closely (exponentially) related mean of water vapour pressure.
+All of the natural precipitation class predictions benefited from (log-transformed) annual precipitation, solar irradiance in January and isothermality.
+
+Location data was most useful for crop prediction, and to a lesser extent for grassland, shrubland and tree prediction.
+These classes tend to be spatially clustered into large areas.
+In contract, location information was not useful for inland water prediction as it is a rare class that takes little area.
+Both absolute and regular latitude was useful for the prediction, and latitude was more useful than longitude for the predictions.
+
+Terrain information was the least useful covariate category.
+It was most useful for crops and water, as crops are inconvenient to grow on slopes or in high altitudes.
+Aspect and \gls{TPI} were only useful for the water class, as standing water has no aspect and a \gls{TPI} of zero, as it has a slope of zero too.
+
+\subsection{Spatial predictions and accuracy}
+
+We used the single-step \gls{RF} model to predict the land cover at a global scale to obtain wall-to-wall global land cover fraction maps.
+See figure \ref{fig-walltowall} for a visualisation (sampled at 0.2 degrees) of all the fraction layers combined.
+The wall-to-wall fraction maps reveal how land cover fraction mapping is capable of expressing gradients and mixed land cover.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{article-figures/maps/2020-05-25-walltowall}
+    \caption{Wall-to-wall predictions of all seven land cover class fractions using the single-step \gls{RF} regression model.}
+    \label{fig-walltowall}
+\end{figure}
+
+In addition, to see whether there were any patterns in the accuracy of the predictions globally, we produced spatial accuracy maps based on the predictions in locations where we had independent validation data.
+This is presented in raster format in figure \ref{fig-resid-rf-1m-uncor}.
+
+\begin{figure}
+    \centering
+    \includegraphics[height=0.8\textheight]{article-figures/maps/2020-05-22-spatialerrors}
+    \caption{Random Forest single model prediction absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.}
+    \label{fig-resid-rf-1m-uncor}
+\end{figure}
+
 \section{Discussion}
 
-\begin{enumerate}
- \item Fraction mapping methods:
- \begin{enumerate}
-    \item RF is the best, even though one model per class means that it does not take everything into account
-    \item Fraction area accuracy goes up to 72\%, that is good considering that hard classification is not much better than that
-    \item Hardest to classify are grasslands and especially shrubs and urban
-    \item Accuracy assessment metric makes a big difference, useful to use subpixel confusion matrix for that
- \end{enumerate}
- \item Optimising for zero inflation:
- \begin{enumerate}
-    \item Multi-step models improve MAE at the cost of RMSE and NSE
-    \item Histogram matching doesn't help any
- \end{enumerate}
- \item Covariate importance:
- \begin{enumerate}
-    \item All covariates are important to a certain degree (show table of importances)
- \end{enumerate}
-\end{enumerate}
+% \begin{enumerate}
+%  \item Fraction mapping methods (Methods):
+%  \begin{enumerate}
+%     \item RF is the best, even though one model per class means that it does not take everything into account
+%     \item Fraction area accuracy goes up to 72\%, that is good considering that hard classification is not much better than that
+%     \item Hardest to classify are grasslands and especially shrubs and urban
+%     \item Accuracy assessment metric makes a big difference, useful to use subpixel confusion matrix for that
+%  \end{enumerate}
+%  \item Optimising for zero inflation (Challenges):
+%  \begin{enumerate}
+%     \item Multi-step models improve MAE at the cost of RMSE and NSE
+%     \item Histogram matching doesn't help any
+%  \end{enumerate}
+%  \item Covariate importance (Variables):
+%  \begin{enumerate}
+%     \item All covariates are important to a certain degree (show table of importances)
+%  \end{enumerate}
+% \end{enumerate}
+
+\subsection{Methods}
+
+\Gls{RF} regression using binary relevance performed the best of all the tested algorithms.
+It was also adaptive to two different goals: minimising \gls{RMSE} (default) and minimising \gls{MAE} (with median voting).
+The three-step Median \gls{RF} model achieved an overall accuracy of 72±2, which is high considering that discrete global land cover maps typically achieve around 70\% accuracy \citep{herold_challenges_2008}.
+In contrast, land cover fraction mapping is expected to be more challenging than discrete classification.
+This is because land cover fractions contain the information for all of the land cover classes considered in each pixel, and this information needs to be retrieved from discrete pixels, with no prior information about class allocation within the pixel.
+
+Compared to other studies that compared different machine learning algorithms for land cover fraction classification, we have achieved a much lower \gls{MAE}.
+\citet{li_monitoring_2018} compared Cubist and \gls{RF} regression for water fraction classification and achieved \gls{MAE} as low as 7.52\% and \gls{RMSE} of 10.39\% with Cubist.
+Our equivalent results were 2.57\% \gls{MAE} and 11.51\% RMSE, respectively.
+We have confirmed that Cubist works slightly better than \gls{RF} regression for this class (\gls{RF} regression result is 11.65\% \gls{RMSE}, 3.48\% \gls{MAE}), but overall for all classes \gls{RF} regression nevertheless edges out (Table \ref{tab-accuracy}).
+Similarly, \citet{walton2008subpixelrf} assessed the accuracy of urban land cover mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which was also observed as performing better than \gls{RF} regression and \glspl{SVM}.
+Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the urban class, and we also observed that Cubist performed slightly better than the other two algorithms for this class.
+\citet{okujeni_comparison_2014} also performed land cover fraction mapping over an urban area, but their results indicated that \gls{SVM} regression performed better than \gls{RF}.
+This could be due to much finer resolution (3.6–9 m) used in their study, and the associated difference in class definitions.
+The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, and thus different balance of the datasets between the studies.
+The other studies focused on a single particular class, whereas both urban and water were relatively rare classes in our global study, and therefore the dataset had much more zero fractions in comparison, which on average are easy to predict.
+
+\subsection{Variables}
+
+Remote sensing variables were the most important variables in our study, especially the maximum and minimum \gls{NDVI} over the entire (temporally-cleaned) time series.
+Multiple vegetation indices were useful, with \gls{NDMI} and \gls{NIRv} median over the time series being still much more important than any other variable from other groups.
+Nevertheless, even though a lot of the other covariates were of much lower importance, they nevertheless contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of covariates that were correlated over 0.9 as explained in section \ref{sec-covariate-selection}).
+This is also due to the large variety of land cover classes in the study, since a covariate is useful if it helps predict any of the land cover classes better.
+The result that including more covariates is beneficial, but remote sensing data is the most important, is in line with the conclusions of \citet{li_monitoring_2018} and \citet{hengl_soilgrids250m_2017}.
+The latter study also shows high importance of climate data, however, it is focused on soils, whereas climate has more effect on soils than on land cover.
+Methods based on decision trees help with effectively using covariates that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any covariates as non-informative.
+
+\subsection{Challenges}
+
+One challenge in land cover fraction prediction is the tendency of the models to favour the mean over the extremes, as that minimises \gls{RMSE} that may arise from incorrect predictions.
+However, that leads to increased fuzziness of the result, where pixels with high uncertainty are marked as a mix of many classes, and zero fractions are rarely predicted.
+Our proposed multi-step model adjusts the balance the other way round.
+As it is a combination of one or two classification models and one regression model, the prediction has a lot more pure pixels compared to a single regression model.
+This works very well in reducing \gls{MAE} (and improving the related metrics from the \gls{SCM}), as the especially common case of zero is captured better.
+On the other hand, it comes at a cost of higher \gls{RMSE} and lower \gls{NSE}.
+This is because in highly uncertain cases, the model does a best guess at a pure class, and often it is incorrect, which leads to 100\% error in those cases, which is highly penalised by \gls{RMSE}.
+A similar effect is seen when using techniques such as median vote for \gls{RF} regression.
+This allows for predicting zero fractions correctly more often, but also misclassify completely more often.
+Our results showed that median vote has a stronger effect for reducing \gls{MAE} than the multi-step approach.
+Combining both methods is beneficial to further reduce \gls{RMSE} for median vote predictions.
+
+Therefore the use of a multi-step approach depends on what is more important for the user.
+If an occasional complete misclassification is acceptable, then a multi-step model provides an overall more accurate result, especially for zero fractions.
+On the other hand, a single-step approach emphasises the strength of land cover fraction mapping in being able to express gradual changes over space better, and avoids large misclassifications.
+We expect a multi-step approach to be more suitable for fine resolution mapping, where more pure pixels can be expected, and single-step approach to be more useful for coarse resolution mapping where mixed pixels are the norm.
+
+There is an inherent challenge in land cover mapping for discerning classes that are related, e.g. grass and shrubs.
+Grass was particularly difficult to map, in part due to confusion between grass, crops and shrubs.
+These classes are similar not just to regression algorithms, but also to expert interpreters, which may lead to higher uncertainties also in the training and validation data for these classes.
+Another challenge is class imbalance.
+For example, the urban class is rare, it hardly ever forms a 100\% fraction and its extent small.
+That makes it easy to classify, as a fraction of 0\% is in most cases a good guess, but it that makes it not useful for user needs.
+This challenge is further exacerbated by the training dataset containing relatively few points in urban areas to begin with.
+As such, having a more balanced dataset may lead to a better representation of the accuracy of each class and help with class predictions.
+
+In this study, model optimisation and dataset sizes were also a challenge.
+Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming covariates or replacing missing values).
+Since the large dataset size increases the training time for the models, parameters to tune have to be chosen carefully.
+As such, it is possible to miss a particular parameter combination that may further improve the results.
+On the other hand, our results showed that the differences between the methods are not very high.
+The differences in model parametrisations are expected to be even lower.
+
+The scale of the study poses performance challenges, as covariates that the models use need to be available at a global scale.
+Our results showed that all of the covariate groups used in the study were useful, therefore leaving a covariate out means sacrificing some predictive power.
+The gain would be less time needed for processing, as that covariate would no longer need to be downloaded, preprocessed and processed.
+In addition, model training performance (time for training and memory usage) may be an important consideration for global land cover mapping, as it may limit the scope of what models may be used for this task.
 
@@
 
-\begin{enumerate}
- \item Random forest is the best by some margin
- \item Subpixel confusion matrix is useful for differentiation
- \item All covariates are important but to a different degree
- \item Using a multi-model approach improves MAE at the cost of RMSE, thus a trade-off
-\end{enumerate}
+% \begin{enumerate}
+%  \item Random forest is the best
+%  \item All covariates are important but to a different degree, remote sensing parameters are most important
+%  \item Using a multi-model approach improves MAE at the cost of RMSE, thus a trade-off
+% \end{enumerate}
 
-\minisection{Author Contributions} DM JV NT MH ML MB BS RK
+In this study, we compared 10 algorithms for land cover fraction mapping and created a demonstation map showing the result of the best performing model by \gls{RMSE}.
+We also introduced a multi-step approach to try and tackle the issue of correctly predicting the extreme fraction values of 0\% and 100\%.
+The best performing models were based on \gls{CART} tree ensembles, with \gls{RF} regression performing the best overall.
+A single-step \gls{RF} regression model using majority voting resulted in the best results for RMSE (17.3\%) and \gls{NSE} (0.66).
+Using median voting combined with the three-step approach obtained the best results for \gls{MAE} (7.9\%) and \gls{SCM} \gls{OA} (72±2\%).
+Remote sensing covariates were the most important in the models, although all other types of covariates (climate, soil, terrain) also contributed to the model significantly.
+This demonstrates the possibility of exhaustive global land cover fraction mapping using currently available methods, achieving a comparable accuracy with discrete land cover maps, but offering better precision.
+It also allows the users to define their own thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
 
-\minisection{Funding} JRC CGLOPS
+This work paves the way for operationalisation of global land cover fraction mapping.
+Future work on this field should investigate the implementation of this approach on different sensors, so as to investigate how useful is land cover fraction mapping also on finer resolution imagery, such as Sentinel-2.
+There is also a need to investigate how land cover change can be captured and represented using land cover fractions, so that gradual change over time could be expressed.
+
+\minisection{Author Contributions}
+
+\begin{itemize}
+    \item Conceptualisation: MH NT JV
+    \item Data curation: DM
+    \item Formal Analysis: DM
+    \item Funding acquisition: JV MH BS
+    \item Investigation: DM ML NT
+    \item Methodology: DM MH NT BK LL
+    \item Project administration: MH NT JV
+    \item Resources: MH BS
+    \item Software: DM
+    \item Supervision: MH NT JV
+    \item Validation: NT DM
+    \item Visualisation: DM MH
+    \item Writing – original draft: DM
+    \item Writing – review \& editing: NT JV MH MB BS
+\end{itemize}
+
+DM JV NT MH ML MB BS RK
 

### 2020-07-16 article/article.tex
Draft 2

@@
-% Minimalistic template for a generic article, geared towards MDPI guidelines
 % Use PDFLaTeX
-\documentclass[a4paper,10pt]{article}
-% Packages provided by the MDPI template already
+\documentclass[review,authoryear,3p]{elsarticle}
+\journal{Remote Sensing of Environment}
+\bibliographystyle{model5-names}\biboptions{authoryear}
+
 \usepackage[T1]{fontenc}
@@
 \usepackage{booktabs} % For \toprule, etc.
-\usepackage[sort&compress,sectionbib]{natbib}
+%\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
 \usepackage{graphicx}
-\bibliographystyle{apalike}
+\usepackage{adjustbox} % For resizing tables
+\usepackage{array} % For mid-alignment (m{xcm}) in tables
+\usepackage{rotating} % Sideways figures
+%\bibliographystyle{apalike} % elsarticle defined at the top
 
@@
 \usepackage[colorlinks, allcolors=blue, unicode]{hyperref}
-\usepackage{hyperxmp} % Import license metadata
-\usepackage{titling} % Get the author for the metadata
+%\usepackage{hyperxmp} % Import license metadata % Conflicts with elsarticle
+%\usepackage{titling} % Get the author for the metadata % Conflicts with elsarticle
 \usepackage{textcomp} % \texttimes
 %\usepackage[pdf]{graphviz} % Importing dot files; can be replaced with generated PDFs
-\usepackage[norefs,nocites]{refcheck} % Develop: warn if figure is not referenced
+%\usepackage[norefs,nocites]{refcheck} % Develop: warn if figure is not referenced
 
 % My own definition for the sections that start with a bold name but don't appear in TOC
-\newcommand{\minisection}[1]{\medskip \textbf{#1:}}
+\newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}
 
-% Metadata
-\title{Global land cover fraction mapping: methods, variables, challenges}
-\author{Dainius Masiliūnas, Nandin-Erdene Tsendbazar, Martin Herold, \\ Myroslava Lesiv, Marcel Buchhorn, Jan Verbesselt}
-% * <Dainius Masiliūnas> 11:20:08 26 Mar 2019 UTC+0100:
-% TODO: fix newline issue
+% Metadata moved to frontmatter for elsarticle
 
-\hypersetup{
-    pdflicenseurl={http://creativecommons.org/licenses/by-sa/4.0/},
-    pdfcopyright={This work is licensed under the Creative Commons Attribution-ShareAlike 4.0 International License.},
-    pdfauthor={\theauthor}, % These are supposed to be the default but don't seem to be
-    pdftitle={\thetitle},
-    pdflang={en-GB}
-}
+% Requires hyperxmp and titling
+% \hypersetup{
+%     pdflicenseurl={http://creativecommons.org/licenses/by-sa/4.0/},
+%     pdfcopyright={This work is licensed under the Creative Commons Attribution-ShareAlike 4.0 International License.},
+%     pdfauthor={\theauthor}, % These are supposed to be the default but don't seem to be
+%     pdftitle={\thetitle},
+%     pdflang={en-GB}
+% }
 
@@
 \newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
+\newacronym{WUR}{WUR}{Wageningen University \& Research}
 \newacronym{SMA}{SMA}{spectral mixture analysis}
@@
 \newacronym{GLM}{GLM}{General linear regression model}
-\makenoidxglossaries
+\newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
+%\makenoidxglossaries
 % Disable abbreviating RF for now [text={random forest}]
@@
 
-\maketitle
+\begin{frontmatter} % elsarticle only
 
-\linenumbers
+\title{Global land characterisation using cover fractions at 100m resolution}
 
-\minisection{Highlights}
+\author[WURGRS]{Dainius Masiliūnas\corref{ca}}
+\ead{dainius.masiliunas@wur.nl}
+\cortext[ca]{Corresponding author}
 
-\begin{itemize}
-    \item 11 statistical and machine learning models tested for land cover fraction mapping
-    \item Random forest regression performed the best, with RMSE of 17.3 globally
-    \item A three-step regression method proposed, which results in the lowest MAE of 7.9
-    \item Remote sensing data was the most important for model predictions, but all auxiliary data also contributed significantly
-    \item Global wall-to-wall land cover fraction maps produced and their errors assessed
-\end{itemize}
+\author[WURGRS]{Nandin-Erdene Tsendbazar}
+\ead{nandin.tsendbazar@wur.nl}
 
-\abstract{Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio.
-In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision, reduces legend complexity, and allows users the flexibility of defining their own mosaic discrete classes using thresholds they define themselves.
+\author[WURGRS]{Martin Herold}
+\ead{martin.herold@wur.nl}
+
+\author[IIASA]{Myroslava Lesiv}
+\ead{lesiv@iiasa.ac.at}
+
+\author[VITO]{Marcel Buchhorn}
+\ead{marcel.buchhorn@vito.be}
+
+\author[WURGRS]{Jan Verbesselt}
+\ead{jan.verbesselt@wur.nl}
+
+\address[WURGRS]{Wageningen University \& Research, Laboratory of Geo-Information Science and Remote Sensing, Droevendaalsesteeg 3, 6708 PB Wageningen, the Netherlands}
+\address[IIASA]{International Institute for Applied Systems Analysis (IIASA), Schlossplatz 1, A-2361 Laxenburg, Austria}
+\address[VITO]{Flemish Institute for Technological Research (VITO), Boeretang 200, BE-2400 Mol, Belgium}
+
+\begin{abstract} % elsarticle: environment instead of a command \abstract{
+Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio.
+In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity.
 To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.
 
-We compared the performance of eight machine learning regression algorithms for land cover fraction mapping, for the best model assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
+We compared the performance of three classes of machine learning regression algorithms for land cover fraction mapping, for the best model assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe at a sampling interval of 0.2 degrees.
 
@@
 
-Results showed that the three-model approach results in better estimation of extreme values (lower MAE of 8.4 vs 9.4) at the cost of misclassification (higher RMSE of 19.4 vs 17.3).
-From the compared models, random forest and cubist regressions reach the highest overall accuracy of 72\%, comparable to the reported accuracies of the current discrete global land cover maps.
-The models were driven primarily by spectral remote sensing data, followed by soil and climate data.
+Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%.
+Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved by a tuned random forest model with our proposed three-model approach, which results in better estimation of extreme values at the cost of more misclassification.
+The models were driven primarily by optical remote sensing data, followed by soil and climate data.
+Herbaceous cover was the most challenging to discern for all models, whereas built-up and water were the easiest to estimate.
 
-This research proves that machine learning algorithms can be applied globally to a wide variety of land cover to create an exhaustive global land cover fraction map.
-This expresses land cover more precisely, and empowers users to create their own discrete maps, which helps land cover monitoring and climate modelling communities.
+This research proves that machine learning algorithms can be applied globally to a wide variety of land cover classes to create an exhaustive global land cover fraction map.
+This expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which helps to customise the result for a diverse range of uses.
 In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product, and to assess the accuracy of land cover fraction maps both thematically and spatially.
+\end{abstract}
 
-\minisection{Keywords} Fuzzy land cover classification, machine learning, random forest, gradient boosting, neural networks, fuzzy c-means, PROBA-V
-}
+\begin{keyword}
+Land cover fraction mapping\sep machine learning\sep random forest\sep Cubist\sep neural networks\sep fuzzy nearest centroid\sep lasso regression\sep PROBA-V
+\end{keyword}
 
-\pagebreak
-\tableofcontents
-% * <Dainius Masiliūnas> 11:40:16 21 Apr 2020 UTC+0200:
-% Temporary, to make it easier to understand the structure
-\pagebreak
+\end{frontmatter} % end elsarticle only
+
+%\maketitle % elsarticle doesn't use it
+
+\linenumbers
+
+\minisection{Highlights}
+
+\begin{itemize}
+    \item Three groups of statistical and machine learning models tested for deriving seven land cover fractions at a global scale
+    \item Random forest regression performed the best (17.3 RMSE or 7.9 MAE)
+    \item Remote sensing data was the most important for model predictions with varying features relevant for different fraction predictions
+    \item Independent validation results show highest accuracies for built-up and water fractions, lower accuracies are found for grasses and shrubs
+    \item Comprehensive multi-fraction global land cover characterisation shown as an alternative to discrete classification
+\end{itemize}
+
+%\pagebreak
+%\tableofcontents
+%\pagebreak
 
@@
 
-% \begin{enumerate}
-%  \item Relevance of fractional LC mapping (link to user requirements and SDGs)
-%  \item Studies that have looked into fractional LC mapping so far
-%  \item Available fractional methods that were introduced in literature (wavelet transform etc.)
-%  \item Methods that are tested in this paper
-%  \item Global land cover maps
-% \end{enumerate}
+Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received increasing amounts of attention, also as new satellites are launched and the capacity for land monitoring increases.
+In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
+Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \ac{CGLS-LC100} map \citep{buchhorn_moderate_2019} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
 
-Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received attention in recent years, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
-Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at maintaining a long medium-resolution time series of global land cover for the climate community, \ac{CGLS-LC100} map that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
-
-Most of these global land cover maps, as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
+All of these global land cover maps (with the exception of \ac{CGLS-LC100} fraction layers), as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
 However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
@@
 \citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
-So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor a global scale.
+So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor at global scale.
 
@@
 
-All in all, with this study we aim to investigate and compare methods for performing land cover fraction mapping, at a global scale, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
-We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hamper land cover fraction mapping.
+Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
+It requires a comprehensible dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
+For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
+In this study, we made use of an independent validation set of over 20\,000 points collected across the globe using the same legend, grid and spatial resolution.
+The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
+We used the validation data to assess model accuracy both thematically and spatially.
+
+All in all, with this study we aim to investigate and compare methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
+We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
 Therefore the objectives of our study are:
@@
 \begin{enumerate}
- \item Compare the performance of a variety of machine learning models for a global-scale land cover fraction mapping task
- \item Investigate approaches for reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
+ \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping
+ \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
  \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
+ \item Assess the accuracy of the models from a thematic and spatial point of view, taking various user needs into account
 \end{enumerate}
@@
 
-For model validation, an independent global dataset of \ac{CGLS}, generated by Wageningen University \& Research \citep{tsendbazar_developing_2018}, was used.
+For model validation, an independent global dataset of \ac{CGLS}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
 The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.
 
-The data includes over 150\,000 training points and over 20\,000 validation points across the globe, describing the fractions of 12 classes (see figure \ref{fig-reference-data}) as they were in the year 2015.
-Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy a single pixel.
-This means that for a given pixel, only a few of the 12 classes would normally have a fraction that is above 0\%.
-This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
-Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
-In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
-But for user communities, correct prediction of non-zero values is more important.
+The data includes over 150\,000 training points and over 21\,000 validation points across the globe, describing the fractions of 12 classes in the year 2015.
+The classification scheme follows the \gls{LCCS} \citep{buchhorn_moderate_2019}.
+However, some of the classes were very detailed and had too few observations to classify well, thus we merged the classes into a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
+The unknown class was discarded: points with the dominant land cover class marked as unknown were not used.
+For points with a minority fraction of unknown, the remaining classes were linearly rescaled to add up to 100\%.
+Thus in the end the training dataset size became 138\,164 and the validation dataset size became 20\,705. See figure \ref{fig-reference-data} for the spatial distribution of the points.
+
+%Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy space within an area of a single pixel.
+%This means that for a given pixel, only a few of the classes would normally have a fraction that is above 0\%.
+%This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
+%Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
+%In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
+%But for user communities, correct prediction of non-zero values is more important.
 
 \begin{figure}
- \includegraphics[width=\textwidth]{article-figures/maps/2019-08-06-training-and-validation}
- \caption{Reference data points with the dominant land cover class. (a): training dataset. (b): validation dataset.}
+ \includegraphics[width=\textwidth]{article/article-figures/maps/2020-07-06-training-and-validation}
+ \caption{Points representing 100x100 m areas at which land cover reference data was used in the study.
+The colours represent the dominant land cover class at each point.
+(a): training dataset, collected by \gls{IIASA}, 138\,164 points used in this study.
+(b): validation dataset, collected by \gls{WUR}, 20\,705 points used in this study.}
  \label{fig-reference-data}
+\end{figure}
+
+\begin{figure}
+ \centering
+ \includegraphics[width=0.75\textwidth]{article-figures/flowcharts/2020-07-10-flowchart}
+ \caption{Processing workflow, from the raw input data to model accuracy assessment and wall-to-wall map output.}
+ \label{fig-processing}
 \end{figure}
@@
 To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
-In this study, six groups of covariates were used: spectral, harmonic, terrain, soil, climate and location (see table \ref{tab-inputdata}).
+See figure \ref{fig-processing} for an overview of the whole processing chain.
+
+In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see table \ref{tab-inputdata}).
 All of these covariates had to be preprocessed before they could be input into the models.
 
-\begin{table}
-\centering
-    \begin{tabular}{llp{6cm}}
-         \toprule
-         \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
-         \midrule
-         Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
-         Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
-         Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
-         Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
-         Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
-         Location & 3 & None (intrinsic) \\
-         \bottomrule
-    \end{tabular}
-    \label{tab-inputdata}
-    \caption{Data sources for the covariates input into the model.}
-\end{table}
+% \begin{table}
+% \centering
+%     \begin{tabular}{llp{6cm}}
+%          \toprule
+%          \textbf{Category} & \textbf{Number} & \textbf{Data source} \\
+%          \midrule
+%          Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
+%          Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
+%          Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
+%          Climate & 21 & WorldClim 2.0 \citep{worldclim2} \\
+%          Soil & 8 & SoilGrids \citep{hengl_soilgrids250m_2017} \\
+%          Location & 3 & None (intrinsic) \\
+%          \bottomrule
+%     \end{tabular}
+%     \label{tab-inputdata}
+%     \caption{Data sources for the covariates input into the model.}
+% \end{table}
 
-\subsubsection{Optical covariates}
-% * <Dainius Masiliūnas> 16:04:43 06 Aug 2019 UTC+0200:
-% Maybe better to call it "Proba-V data" or such
+%\subsubsection{Vegetation indices and temporal metrics}
+\label{sec-temporal-filter}
 
 The entire archive (2014-03-11 to 2019-07-16) of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{probavguide2} was used for this study.
-The time series of NDVI and each of the bands at the points of interest were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were too far from the fitted \ac{LOESS} curve over the blue band.
+While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of Proba-V was chosen to obtain more robust statistics for the temporal metrics.
+Land cover change is a relatively rare phenomenon and thus is not likely to affect more than 2\% of the training points within the time period.
+Nevertheless, this may result in poorer model performance compared to if reference data for each year would be available.
 
-The resulting cloud-free time series was then used to extract descriptive statistics (mean and interquartile range) of NDVI over the whole time series.
-Next, harmonic analysis was run in order to decompose the time series into sine and cosine components for two frequencies (annual and semiannual), as well as the trend and intercept of the model.
-From this, phase and amplitude for the two frequencies were calculated.
+The time series of each of the Proba-V bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue band.
+The resulting cloud-free time series was then used to extract descriptive statistics (median and interquartile range) of NDVI, NDMI, EVI, OSAVI and NIR\textsubscript{v} over the whole time series, as well as for each season separately.
+Next, harmonic analysis was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
+From this, phase and amplitude for the two orders were calculated.
+Lastly, the minimum and maximum of the NDVI time series was also retained as covariates.
 
-Lastly, extra \glspl{VI} (NDMI, EVI, OSAVI and NIR\textsubscript{v}) were calculated.
-Each of these \glspl{VI} was then composited into a single image that represents the leaf-on season (three months) of 2015.
+% \begin{figure}
+%  \includegraphics[width=\textwidth]{article-figures/flowcharts/preprocessing-optical}
+%  \caption{Optical data preprocessing chain.}
+%  \label{fig-preprocessing-optical}
+% \end{figure}
 
-The optical preprocessing chain is displayed graphically in figure \ref{fig-preprocessing-optical}.
-All of the outputs were then used in the processing chain.
+%\subsubsection{Terrain covariates}
 
-\begin{figure}
- \includegraphics[width=\textwidth]{article-figures/flowcharts/preprocessing-optical}
- \caption{Optical data preprocessing chain.}
- \label{fig-preprocessing-optical}
-\end{figure}
+For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to Proba-V 100 m grid and used as the elevation covariate.
+In addition, GDAL was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
+%All terrain covariates were then resampled to the Proba-V grid.
 
-\subsubsection{Terrain covariates}
+%\subsubsection{Climate covariates}
 
-The ASTER GDEM v003 \citep{ASTGTM003} product was used as the elevation covariate.
-In addition, terrain parameters were calculated out of elevation: slope, aspect and \ac{TPI}.
-All terrain covariates were then resampled to the Proba-V grid.
+For climate data, WorldClim 2.0 30 s product \citep{worldclim2} was used.
+It includes monthly temperature, precipitation, solar radiation, wind speed and water vapour pressure data.
+In additional, 19 bioclimatic parameters were calculated from the data, using the dismo package \citep{dismo}.
+Some additional biophysical parameters were also calculated, namely all of the climate data during the coldest, warmest, driest and wettest months of the year at each location, as well as yearly averages of each of the original parameters.
 
-\subsubsection{Climate covariates}
+%\subsubsection{Soil covariates}
 
-The climate data from WorldClim 2.0 30 s product \citep{worldclim2} was used.
-Biophysical parameters were calculated from the data as well.
-Some additional biophysical parameters were also calculated, namely the means, minimums and maximums over the whole time series of solar radiation, wind speed and water vapour pressure.
+SoilGrids \citep{hengl_soilgrids250m_2017} was used for getting data about soils.
+It is based on a random forest model that predicts soil properties at various soil depths globally at a 250 m resolution.
+In the creation of SoilGrids, a land cover map (based on MODIS) had been used, and so, in order to avoid circular inference, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
+The soil taxonomy covariates were also excluded, since they are categorical derivatives from the numerical soil property data and thus do not contribute to land cover fraction prediction.
 
-\subsubsection{Soil covariates}
+%\subsubsection{Location covariates}
 
-Two products were used for data on soils: SoilGrids and LandGIS.
-SoilGrids is based on a machine learning model that predicts soil properties at various soil depths globally at a 250 m resolution \citep{hengl_soilgrids250m_2017}
-LandGIS is a successor to SoilGrids using a refined machine learning methodology \citep{hengl_predictive_2019}, however, it does not yet provide as many covariates as SoilGrids.
-
-In the creation of SoilGrids, a land cover map (based on MODIS) was used.
-In order to avoid circular dependencies, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
-In addition, the taxonomy covariates were also excluded, since they are categorical data derived from the numerical soil property data.
-
-\subsubsection{Other covariates}
-
-The latitude, longitude and absolute latitude of the points were also included as location covariates.
-Lastly, as part of the SoilGrids dataset, the MODIS land surface temperature product (MOD11A2, \citealp{wan_mod11a2_2015}) data monthly composites for 2011 were also used as additional covariates.
+The latitude, longitude and absolute latitude of the reference points were also included when training the models, so that the models could learn spatial patterns.
 
@@
 Similarly, climate data was colinear between subsequent months.
-Thus January and August data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
-
-While \ac{OSAVI} was calculated as a covariate, ultimately it was dropped due to too high correlation with \ac{NDVI}.
-Similarly, the spectral bands of Proba-V were also highly correlated, so only the red and \ac{NIR} bands were kept.
+Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
+In addition, all of the spectral bands of Proba-V were also highly correlated with each other as well as to vegetation indices, so only vegetation indices were kept.
 
@@
 
-\begin{figure}
- \includegraphics[width=\textwidth]{article-figures/flowcharts/processing}
-% * <Dainius Masiliūnas> 13:45:26 13 Aug 2019 UTC+0200:
-% TODO: optimise style for readability.
- \caption{Processing chain.}
- \label{fig-processing}
-\end{figure}
-
 \subsection{Land cover fraction mapping methods}
 
-For deriving land cover fractions, we compared a wide array of machine learning regression methods: \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression, \ac{MLR}, \ac{MLP} \glspl{NN}, \glspl{SVM}, Cubist and \ac{RF} regression.
-In addition, as a baseline we also compare the results with an intercept model (all fractions always predicted to be equal).
+For deriving land cover fractions, we compared a wide array of machine learning regression methods.
+They can be broadly divided into three categories.
+Tested linear models were \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression and \ac{MLR}.
+Non-tree machine learning models tested were \ac{MLP} \glspl{NN} and \glspl{SVM}.
+Lastly, machine learning models based on \glspl{CART} tested were Cubist and \ac{RF} regressions.
+In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
 
@@
 
-\subsubsection{\Glsentryfull{GLM}}
+\subsubsection{Linear models}
+%\subsubsection{\Glsentryfull{GLM}}
 
-\Gls{GLM}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
+\Glsentryfull{GLM}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
 In this study, we used \gls{GLM} as a baseline model.
-\Gls{GLM} does not normalise the output to add up to 100\%, and also allows negative predictions.
-Therefore all predictions < 0 were capped to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
-
-We also tested whether dropping any covariates would result in better predictions, using a sphericity test with Greenhouse-Geisser correction to compare models with different covariates.
+We also tested whether dropping any additional covariates from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction to compare models with different covariates.
 However, only three covariates could be dropped and it did not improve the predictions, therefore we decided to keep all the covariates in the model.
@@
 
-\subsubsection{Lasso/ridge/elastic net regression}
+%\subsubsection{Lasso/ridge/elastic net regression}
 
-Ridge regression is a regularisation method which reduces large coefficients of model covariates in order to prevent overfitting.
+Ridge regression is a linear model regularisation method which reduces large coefficients of model covariates in order to prevent overfitting.
 Lasso regression is based on the same principle, except it reduces the coefficients of each model covariate to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
@@
 This has the effect of the coefficients of correlated covariates being set to a similar value.
-
 We chose lasso regression, so that it would be possible to identify if there are any covariates that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
 The $\lambda$ parameter was chosen using cross-validation.
+The lasso regression was performed using the \texttt{glmnet} package \citep{glmnet}.
 
-The lasso regression was performed using the \texttt{glmnet} package \citep{glmnet}.
-This method is multivariate in the same way that \gls{GLM} is, and the output was clamped to positive numbers and rescaled to add up to 100\% the same way.
+%\subsubsection{\Glsentryfull{PLS} regression}
 
-\subsubsection{\Glsentryfull{PLS} regression}
-
-\ac{PLS} regression is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space \citep{martens_multivariate_1992}.
+\Glsentryfull{PLS} regression is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space \citep{martens_multivariate_1992}.
 As such, it is well suited for multicollinear data.
-This method is also multivariate in the same way that \gls{GLM} is, and the same method was used to clamp and rescale the predictions.
 \ac{PLS} regression was performed using the package \texttt{pls} \citep{pls}.
 
-\subsubsection{\Glsentryfull{FNC} regression}
+All of the aforementioned linear models do not normalise the output to add up to 100\%, and also allow negative predictions.
+Therefore for these models, all predictions < 0 were set to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
 
-\Gls{FNC} regression, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from the class centroid in feature space.
+%\subsubsection{\Glsentryfull{FNC} regression}
+
+\Glsentryfull{FNC} regression, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
 This method always produces output that sums to 100\%.
-
 In this study, we used an implementation adapted from the package GSIF \citep{hengl2004fuzzycmeans}.
@@
 
-\subsubsection{\Glsentryfull{MLR}}
+%\subsubsection{\Glsentryfull{MLR}}
 
+Lastly, we tested \glsentryfull{MLR}.
 While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
@@
 
-\subsubsection{\Glsentryfullpl{NN}}
+\subsubsection{Non-tree machine learning methods}
+%\subsubsection{\Glsentryfullpl{NN}}
 
-\Glspl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for any postprocessing.
-In this study, we used a multilayer perceptron with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+\Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
+In this study, we used a \glsentryfull{MLP} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
 Softmax activation was used to get the output.
 The Nadam optimiser \citep{dozat_incorporating_2016} was used, with \gls{MAE} as the loss function.
-The \texttt{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics card to accelerate the neural network training process.
+The \texttt{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics processing unit to accelerate the neural network training process.
 
-\subsubsection{\Glsentryfull{RF} regression}
+%\subsubsection{\Glsentryfull{SVM} regression}
 
-\ac{RF} regression is a popular method for land cover classification that works by building a number of decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual trees \citep{breiman2001random}.
+\Glsentryfullpl{SVM} are machine learning models that attempt to find the optimal boundary between the class clusters in feature space by constructing a dividing hyperplane.
+For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
+We used the binary relevance method, clamped the output values to the 0-100\% range and linearly scaled them to add up to 100\%.
+The package \texttt{liquidSVM} \citep{liquidSVM} was suitable for this purpose, as it is optimised for processing large datasets such as ours.
+
+\subsubsection{Tree-based machine learning methods}
+%\subsubsection{\Glsentryfull{RF} regression}
+
+\Glsentryfull{RF} regression is a popular method for land cover classification that works by building a number of \gls{CART} decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual trees \citep{breiman2001random}.
 \ac{RF} can only take one variable as the response variable, therefore there are two ways to obtain land cover fraction predictions from an \ac{RF} model.
@@
 
-\ac{RF} was also used for assessing the covariate importance.
+As the results for \gls{RF} regression proved the best for \gls{RMSE}, it was also used for assessing the covariate importance.
 Permutation importance was used: each covariate was shuffled in turn, and the resulting decrease in \gls{RMSE} was calculated, comparing against the validation data set.
 Since the number of covariates is imbalanced by group (i.e.
-there are merely 3 location covariates and 22 spectral covariates), and there may still be colinear covariates, we also tested calculating covariate importance by permuting all covariates in a group together.
-However, the result was similar to the sum of permutation importance of each individual covariates in the group, which showed that the covariate selection process had eliminated all colinear covariates.
+there are merely 3 location covariates and 22 vegetation index metrics), and there may still be colinear covariates, we also tested calculating covariate importance by permuting all covariates in a group together.
+However, the result was similar to the sum of permutation importance of each individual covariates in the group, which showed that the covariate selection process had successfully eliminated all colinear covariates.
 Thus we reported the more detailed covariate importance permuted per individual covariate.
 
-\subsubsection{Cubist regression}
+%\subsubsection{Cubist regression}
 
-Cubist regression \citep{cubist} is based on \gls{RF} regression, with the major difference being that instead of using a threshold of a covariate for a split in the decision tree, Cubist instead uses a linear regression based on a subset of the data relevant for the split in question.
+Lastly, we tested Cubist regression \citep{cubist}.
+It is based on \gls{RF} regression, with the major difference being that instead of using a threshold of a covariate for a split in the decision tree, Cubist instead uses a linear regression based on a subset of the data relevant for the split in question.
 In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previous ones.
@@
 
-\subsubsection{\Glsentryfull{SVM} regression}
-
-\Glspl{SVM} are machine learning models that attempt to find the optimal boundary between the class clusters in feature space by constructing an optimal dividing hyperplane.
-For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
-We used the binary relevance method, clamped the output values to the 0-100\% range and linearly scaled them to add up to 100\%.
-The package \texttt{liquidSVM} \citep{liquidSVM} was suitable for this purpose, as it is optimised for processing large datasets such as ours.
-
 \subsubsection{Other methods}
@@
 A few more methods were considered, but in the end turned out to be unworkable with our dataset.
-Multivariate \gls{RF} in available in R provided by the \texttt{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
+Multivariate \gls{RF} is available in R provided by the \texttt{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
 However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
@@
 Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
-Whereas setting the number of folds to a lower number resulted in worse results than when using a single classifier.
+Attempts to reduce training time by lowering the number of folds resulted in worse results than when using a single classifier.
 
-Lastly, multiple-endmember spectral unmixing using non-negative least squares was also attempted, but the results were significantly worse than even the intercept model, possibly because the class fractions are not a linear combination of the covariates.
+Lastly, multiple-endmember spectral unmixing using non-negative least squares \citep{franc2005sequential} was also attempted, but the results were significantly worse than even the intercept model, possibly because the class fractions are not a linear combination of the covariates.
 
@@
 
-After determining the method that performs the best, we attempted to tune it further in an attempt to solve the dataset balance (zero inflation) problem.
-As the dataset is dominated by zeroes and hundreds, minimising the objective function tends to draw the model towards predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
+After determining the method that performs the best, we attempted to tune it further in an attempt to solve the balance (zero inflation) problem in the training dataset.
+As the dataset describes fractions of each land cover class at each point, most of the locations consist of a mix of only a few classes, and the rest of the classes are zero.
+As the dataset is dominated by zeroes, minimising the objective function tends to draw the model towards predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
 This is not desirable for users of land cover fraction data, as the added value of fraction information is the information about the middle of the range; otherwise, discrete classification would be just as good.
@@
 
-We compared a single model of the best algorithm, to a two-step approach: one model to predict zeroes, and one model to predict non-zeroes; and a three-step approach, with one model to predict pixel purity (i.e.
-if we face a classification or a regression problem), one to perform regression (on mixed pixels) and one to perform classification (on pure pixels).
+For the best performing algorithm, we compared: (1) a single model, (2) a two-model approach: (a) one model to predict zeroes, and (b) one model to predict non-zeroes; and (3) a three-model approach, with (a) one model to predict pixel purity (i.e.
+whether we face a classification or a regression problem), (b) one to perform regression (on mixed pixels as determined by (a)), and (c) one model to perform classification (on pure pixels as determined by (a)).
 
-In addition, we investigated applying histogram matching, by matching the histogram of the predictions back to the histogram of the training set.
-However, that turned out to decrease accuracy in all cases, as the cases where a model predicts a value close to the true value would be set to a more extreme value to compensate for other predictions that were not as accurate to begin with.
-Therefore in the end histogram matching was not used.
+%In addition, we investigated applying histogram matching, by matching the histogram of the predictions back to the histogram of the training set.
+%However, that turned out to decrease accuracy in all cases, as the cases where a model predicts a value close to the true value would be set to a more extreme value to compensate for other predictions that were not as accurate to begin with.
+%Therefore in the end histogram matching was not used.
 
@@
 The overall measures are calculated by taking the mean of all class points pooled together, rather than taking a mean of the per-class means.
- \item \Gls{NSE} was used to determine the goodness of fit.
+ \item \Gls{NSE} \citep{nash1970river} was used to determine the goodness of fit.
 For regression output, \ac{NSE} is a method of calculating the coefficient of determination $R^2$, and is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
  \item \Gls{SCM} \citep{silvan-cardenas_sub-pixel_2008}, and the metrics derived from it: \ac{OA}, as well as \ac{PA} and \ac{UA} per class.
-The \gls{SCM} is an adaptation of the confusion matrix concept to fractional data, where the diagonal of the matrix expresses the overlap (agreement) of the target and predicted classes, and the off-diagonal gives an estimation of which classes are confused with which.
-When using the MID-PROD operators as suggested by \citet{silvan-cardenas_sub-pixel_2008}, the \ac{SCM} provides an uncertainty measure that comes from the fact that there is no unique solution for which part of the modelled fractions is confused with which part of the validation fractions.
+The \gls{SCM} is an adaptation of the confusion matrix concept to fractional data.
+When using the MIN-PROD composite operator as suggested by \citet{silvan-cardenas_sub-pixel_2008}, the diagonal of the matrix expresses the maximum overlap (agreement) of the target and predicted class fractions, and the off-diagonal is an expression of which classes the non-overlapping fractions should belong to by calculating the expected value of overlap (product of reference and predicted class fraction).
+For instance, a pixel of 60\% grass and 40\% shrub, when predicted as 40\% grass and 60\% shrub, would have $min(60\%, 40\%)=40\%$ in the diagonals and $\frac{20\%\times20\%}{20\%}=20\%$ in the off-diagonals.
+For cases where the allocation of the overestimated and underestimated parts of the fraction do not have a unique solution, the \ac{SCM} indicates the expected value and an associated uncertainty measure.
  \item Correlation matrix of predicted vs observed class fractions.
- \item Prediction accuracy at different parts of the predicted vs observed line.
-We calculated the accuracy of prediction of zeroes, hundreds, and whether the point falls in the expected half of the 1:1 line (above 50\% or below 50\%).
-The numbers obtained are averages and standard deviations of the aforementioned binomial values over individual class fractions, as well as overall.
-\item Lastly, spatial residuals, i.e. the overestimation and underestimation of each class fraction on a map.
+ \item Boxplots showing the distribution of predictions, in bins for each 10\% of the predictions, against the validation data distribution, and a comparison with the 1:1 line.
+\item Lastly, spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point, rasterised into a map.
 \end{enumerate}
@@
 \subsection{Method accuracy comparison}
-
-Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
-\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
-\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/boxplots/2020-05-08-model-comparison-bar}
-    \caption{Comparison of the distribution of absolute errors between different models tested, ordered by decreasing overall \gls{MAE}.
-Dotted lines indicate \gls{MAE}, box indicates median and 75th/25th percentiles.}
-    \label{fig-models}
-\end{figure}
-
-The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
-Predictions per class were made using the binary relevance method (one model per class), as these models are univariate.
-This shows that despite the binary relevance limitations of each model only being aware of its relevant class and not the others, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.
-
-Multivariate linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
-The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
-This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
-It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.
-
-\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
-This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
-This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
-
-\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods, but better than all of the multivariate linear models.
-
-There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large misclassifications or not.
-Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
-Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk misclassification, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely wrong.
-
-Overall accuracy and kappa statistics from the \gls{SCM} coincided with the \gls{MAE} scores, whereas \gls{NSE} coincided with the \gls{RMSE} scores.
-This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
-The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
-
-All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil).
-This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
-In contrast, the tree fraction has a lot of samples with 100\% fraction and it deviates a lot per each sample, making it a highly unpredictable class from the data alone.
-Most models showed good improvement in predicting tree cover compared to the intercept model, indicating that the covariates used were useful in telling trees apart from other classes.
-In contrast, there were only moderate improvements for predicting the grass fraction, as it is difficult for the classifiers to distinguish this class with the given input covariates.
 
@@
 Cubist regression
-& 18.1  & 8.1   & 0.63  & 72±2  & 0.63±0.03 \\
+& 18.1  & 8.1   & 0.63  & \textbf{72±2}  & 0.63±0.03 \\
 \Gls{RF} regression
@@
 %& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
+{''} RS-only
+& 18.4  & 10.3  & 0.62  & 64±4  & 0.54±0.05 \\
 \ensuremath{''} two-step
@@
 \ensuremath{''} median vote
-& 20.7  & 7.9   & 0.52  & 71±1  & 0.63±0.02 \\
+& 20.7  & \textbf{7.9}   & 0.52  & 71±1  & 0.63±0.02 \\
 \ensuremath{''} \ensuremath{''} two-step
-& 20.0  & 8.1   & 0.54  & 72±1  & 0.63±0.02 \\
+& 20.0  & 8.1   & 0.54  & \textbf{72±1}  & 0.63±0.02 \\
 \ensuremath{''} \ensuremath{''} three-step
@@
 \end{tabular}
-\caption{Accuracy statistics of the models tested.}
+\caption{Accuracy statistics of the models tested. Best performing statistics are highlighted.}
 \label{tab-accuracy}
 \end{table}
+
+Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
+\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
+\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.
+This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
+The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{article/article-figures/barplots/2020-06-03-model-comparison-bar}
+    \caption{Comparison of \gls{RMSE} (top) and \gls{MAE} (bottom) per class of the best performing models in their category, and the intercept-only model as a reference.}
+    \label{fig-models}
+\end{figure}
+
+The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
+Predictions per class were made using the binary relevance method (one model per class), as these models are univariate.
+This shows that despite the binary relevance limitations of each model only being aware of its relevant class and not the others, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.
+
+Multivariate linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
+The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
+This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
+It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.
+
+\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
+This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
+This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
+
+\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods, but better than all of the multivariate linear models.
+
+There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large misclassifications or not.
+Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
+Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk misclassification, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely wrong.
+
+All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil).
+This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
+In contrast, the tree fraction has a lot of samples with 100\% fraction and it deviates a lot per each sample, making it a highly unpredictable class from the data alone.
+Most models showed good improvement in predicting tree cover compared to the intercept model, indicating that the covariates used were useful in distinguishing tree cover from other classes and quantifying its proportion in the pixel.
+In contrast, there were only moderate improvements for predicting the grass fraction, as it is difficult for the classifiers to distinguish this class with the given input covariates.
 
@@
 
-For an \gls{RF} regression when the mean of all tree votes was used as the indicator for the land cover fraction, using a two-step approach improved the \gls{MAE} of the model, while increasing \gls{RMSE}.
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{article/article-figures/barplots/2020-06-04-rf-comparison-bar}
+    \caption{Comparison of \gls{RF} regression models (intercept model shown as a reference).
+1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
+Mean and median are the tree vote summary statistics.}
+    \label{fig-randomforest}
+\end{figure}
+
+For a \gls{RF} regression when the mean of all tree votes was used as the indicator for the land cover fraction, using a two-step approach improved the \gls{MAE} of the model, while increasing \gls{RMSE} (see figure \ref{fig-randomforest}).
 This is because the false positives in the first step of the two-step model (the binary zero/non-zero classification) result in some high fractions nevertheless being set to zero.
@@
 However, the two-step approach combined with binary relevance also introduced a drawback in that there were cases where all land cover class fractions were predicted to be 0\% in the first step.
-In such cases it was not possible to determine the land cover fractions, and to make them sup up to 100\%, they were all set to equal parts ($100\% / 7$), which introduced further error.
+In such cases it was not possible to determine the land cover fractions, and to make them sum up to 100\%, they were all set to equal parts ($100\% / 7$), which introduced further error.
 
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/heatmaps/2020-05-12-varimp-heatmap-top10}
-    \caption{Random Forest single model variable importance, top 10 covariates per category.}
+    \centering
+    \includegraphics[width=12cm]{article-figures/heatmaps/2020-06-19-varimp-heatmap-top5}
+    \caption{Random Forest single model variable importance, top 5 covariates per category.
+Categories are ordered by cumulative importance.
+The values represent decrease in RMSE when the covariate is permuted.}
     \label{fig-varimp}
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{article-figures/maps/2020-05-25-walltowall}
-    \caption{Wall-to-wall predictions of all seven land cover class fractions using the single-step \gls{RF} regression model.}
-    \label{fig-walltowall}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\textwidth]{article-figures/maps/2020-05-25-walltowall}
+%     \caption{Wall-to-wall predictions of all seven land cover class fractions using the single-step \gls{RF} regression model.}
+%     \label{fig-walltowall}
+% \end{figure}
 
 In addition, to see whether there were any patterns in the accuracy of the predictions globally, we produced spatial accuracy maps based on the predictions in locations where we had independent validation data.
-This is presented in raster format in figure \ref{fig-resid-rf-1m-uncor}.
+This is also presented in raster format in figure \ref{fig-walltowall}.
 
-\begin{figure}
+\begin{sidewaysfigure}
     \centering
-    \includegraphics[height=0.8\textheight]{article-figures/maps/2020-05-22-spatialerrors}
-    \caption{Random Forest single model prediction absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.}
-    \label{fig-resid-rf-1m-uncor}
-\end{figure}
+    \includegraphics[width=\textwidth]{article/article-figures/maps/2020-06-19-walltowall.png}
+    \caption{Random Forest single model predictions per class.
+Left column: predicted fractions.
+Middle column: absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.
+Right column: distribution of predicted versus true values, shown as box plots with bins each 10\%.}
+    \label{fig-walltowall}
+\end{sidewaysfigure}
 
@@
 
-\Gls{RF} regression using binary relevance performed the best of all the tested algorithms.
+In this study, we have tested a number of machine learning methods for comprehensive land cover class fraction prediction at the global scale.
+Since global land cover fraction mapping is a novel field, there are few other studies or products that our results can be directly compared with.
+Most products in this field are made available as layers for a particular land cover class fraction, rather than being comprehensive and guaranteeing a sum to 100\%.
+In addition, the documentation for these products rarely report accuracy metrics, and none use an independent validation set for this purpose.
+Lastly, these products are generated for different time scales than our study, both in terms of input time series as well as the time for which the model training data has been collected.
+
+%The three-step Median \gls{RF} model achieved an overall accuracy of 72±2\%, which is high considering that discrete global land cover maps typically achieve around 70\% accuracy \citep{herold_challenges_2008}.
+%In contrast, land cover fraction mapping is expected to be more challenging than discrete classification.
+%This is because land cover fractions contain the information for all of the land cover classes considered in each pixel, and this information needs to be retrieved from discrete pixels, with no prior information about class allocation within the pixel.
+
+Compared to a global land cover product that specialises in a specific land cover class, namely tree cover, the \gls{RF} one-step mean model had a higher RMSE (19.2\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (17.4\%), however, the methodology for the validation differs significantly from the one used in our study.
+\citet{feng_earth_2016} assessed a discrete (binary) GLCF forest cover product for three different years using a more similar methodology, and yielded similar results to ours: a user's accuracy of 98\% and producer's accuracy of 80\%, compared to the \gls{RF} three-step median tree cover user's accuracy of 81\% and producer's accuracy of 86\%.
+
+Our results showed that \gls{RF} regression using binary relevance performed the best of all the tested algorithms.
 It was also adaptive to two different goals: minimising \gls{RMSE} (default) and minimising \gls{MAE} (with median voting).
-The three-step Median \gls{RF} model achieved an overall accuracy of 72±2, which is high considering that discrete global land cover maps typically achieve around 70\% accuracy \citep{herold_challenges_2008}.
-In contrast, land cover fraction mapping is expected to be more challenging than discrete classification.
-This is because land cover fractions contain the information for all of the land cover classes considered in each pixel, and this information needs to be retrieved from discrete pixels, with no prior information about class allocation within the pixel.
+Both of these goals may be valuable to different user communities.%, depending on whether areas with high uncertainty in the predicted land cover should be treated as a mix of covers, or a guess should be made at a single most likely cover.
 
@@
 \citet{li_monitoring_2018} compared Cubist and \gls{RF} regression for water fraction classification and achieved \gls{MAE} as low as 7.52\% and \gls{RMSE} of 10.39\% with Cubist.
-Our equivalent results were 2.57\% \gls{MAE} and 11.51\% RMSE, respectively.
+Our respective results were 2.57\% \gls{MAE} and 11.51\% RMSE.
 We have confirmed that Cubist works slightly better than \gls{RF} regression for this class (\gls{RF} regression result is 11.65\% \gls{RMSE}, 3.48\% \gls{MAE}), but overall for all classes \gls{RF} regression nevertheless edges out (Table \ref{tab-accuracy}).
+One important difference between the studies is that we focused on comprehensive land cover, rather than a particular class, which has an effect on the model training, and our focus was on a global rather than regional scale, has an effect on the resulting statistics due to higher sample site diversity.
+
 Similarly, \citet{walton2008subpixelrf} assessed the accuracy of urban land cover mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which was also observed as performing better than \gls{RF} regression and \glspl{SVM}.
@@
 This could be due to much finer resolution (3.6–9 m) used in their study, and the associated difference in class definitions.
-The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, and thus different balance of the datasets between the studies.
-The other studies focused on a single particular class, whereas both urban and water were relatively rare classes in our global study, and therefore the dataset had much more zero fractions in comparison, which on average are easy to predict.
+The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, namely their focus on fine scale predictions over a local area for the urban class in particular, in contrast to our global focus with comprehensive classes, leading to a different balance of the datasets between the studies.
+The other studies focused on a single particular class, whereas both urban and water were relatively rare classes in our global study, and therefore the dataset had much more zero fractions in comparison, which on average is easier to predict.
 
@@
 
-Remote sensing variables were the most important variables in our study, especially the maximum and minimum \gls{NDVI} over the entire (temporally-cleaned) time series.
+Remote sensing variables were the most important variables in our study, especially the maximum and minimum \gls{NDVI} over the entire time series.
+Note that these values are taken from a time series that has undergone temporal outlier removal, as detailed in section \ref{sec-temporal-filter}, and thus roughly correspond to the 5th and 95th percentiles of the data without additional temporal filtering.
 Multiple vegetation indices were useful, with \gls{NDMI} and \gls{NIRv} median over the time series being still much more important than any other variable from other groups.
-Nevertheless, even though a lot of the other covariates were of much lower importance, they nevertheless contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of covariates that were correlated over 0.9 as explained in section \ref{sec-covariate-selection}).
+Nevertheless, even though a lot of the other covariates were of much lower importance, they  contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of covariates that were correlated over 0.9 as explained in section \ref{sec-covariate-selection}).
 This is also due to the large variety of land cover classes in the study, since a covariate is useful if it helps predict any of the land cover classes better.
@@
 This is because in highly uncertain cases, the model does a best guess at a pure class, and often it is incorrect, which leads to 100\% error in those cases, which is highly penalised by \gls{RMSE}.
+Due to this effect, the resulting map is closer to a discrete classification map, with less expressive transition gradients between land cover classes.
+Thus we chose to present the single-model \gls{RF} regression result in figure \ref{fig-walltowall}.
+
 A similar effect is seen when using techniques such as median vote for \gls{RF} regression.
-This allows for predicting zero fractions correctly more often, but also misclassify completely more often.
+This allows for predicting zero fractions correctly more often, but also misclassifies completely more often.
 Our results showed that median vote has a stronger effect for reducing \gls{MAE} than the multi-step approach.
@@
 
-Therefore the use of a multi-step approach depends on what is more important for the user.
+These findings show that the use of a multi-step approach depends on what is more important for the user.
 If an occasional complete misclassification is acceptable, then a multi-step model provides an overall more accurate result, especially for zero fractions.
@@
 We expect a multi-step approach to be more suitable for fine resolution mapping, where more pure pixels can be expected, and single-step approach to be more useful for coarse resolution mapping where mixed pixels are the norm.
+
+Machine learning algorithms pose several challenges that are inherent to how the models are constructed.
+The trade-off between minimising \gls{RMSE} and minimising \gls{MAE} comes from the chosen loss function.
+In addition, the result always tends towards the mean in cases of high uncertainty.
+The models predict in areas with a high cover of a fraction, such as shrublands, a lower proportion than expected, whereas in cases with a low cover, such as for fraction of built-up in dark-soil deserts, higher fractions are predicted.
+This is due to a guess of the mean being less penalising than predicting the extremes incorrectly; e.g.
+for a case of 50\% shrub cover, predicting 100\% shrubs would be a larger mistake (and thus lead to higher RMSE) than predicting 15\% shrubs.
+Likewise, predicting 0\% built-up in dark bare soils risks a case where it truly would be 100\% built-up, so on average predicting 15\% built-up in this case lowers the possible error.
 
@@
 This challenge is further exacerbated by the training dataset containing relatively few points in urban areas to begin with.
-As such, having a more balanced dataset may lead to a better representation of the accuracy of each class and help with class predictions.
+As such, having a more balanced training dataset may lead to a better representation of the accuracy of each class and help with class predictions.
 
@@
 
+Several challenges are yet to be tackled in this field.
+Finer spatial resolution mapping, such as 10 m mapping using Sentinel-2 data, is a future research direction, where pixels will be even more likely to be covered by a single class.
+Such future developments would be more likely to benefit from a multi-model approach or optimisation for \gls{MAE}.
+Alternatively, finer spatial resolution sensors would allow mapping fractions at coarser resolutions in a more precise way, as it would be possible for the model to directly make use of data from the sensor about subpixels of the mapped pixels.
+Another future work direction is land cover change mapping over time.
+Land cover fraction information would allow tracking gradual changes, such as regrowth, over time better.
+The challenge here is that the higher uncertainty about fraction estimates may cause the time series of land cover fractions to fluctuate, making it difficult to determine trends.
+
 \section{Conclusions}
@@
 
-In this study, we compared 10 algorithms for land cover fraction mapping and created a demonstation map showing the result of the best performing model by \gls{RMSE}.
+In this study, we compared three classes of algorithms for land cover fraction mapping and created a demonstation map showing the result of the best performing model by \gls{RMSE}.
 We also introduced a multi-step approach to try and tackle the issue of correctly predicting the extreme fraction values of 0\% and 100\%.
@@
 Remote sensing covariates were the most important in the models, although all other types of covariates (climate, soil, terrain) also contributed to the model significantly.
-This demonstrates the possibility of exhaustive global land cover fraction mapping using currently available methods, achieving a comparable accuracy with discrete land cover maps, but offering better precision.
+This demonstrates the possibility of exhaustive global land cover fraction mapping using currently available methods, offering better precision than discrete land cover maps.
 It also allows the users to define their own thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
+Thus this work paves the way for operationalisation of global land cover fraction mapping.
 
-This work paves the way for operationalisation of global land cover fraction mapping.
-Future work on this field should investigate the implementation of this approach on different sensors, so as to investigate how useful is land cover fraction mapping also on finer resolution imagery, such as Sentinel-2.
-There is also a need to investigate how land cover change can be captured and represented using land cover fractions, so that gradual change over time could be expressed.
+% Converted from role-centric to person-centric as per Elsevier guidelines: https://www.elsevier.com/authors/journal-authors/policies-and-ethics/credit-author-statement
+\minisection{Author Contributions}
+\textbf{Dainius Masiliūnas}: Methodology, Formal Analysis, Software, Writing – Original Draft, Data Curation.
+\textbf{Nandin-Erdene Tsendbazar}: Investigation, Data Curation, Supervision, Conceptualisation, Writing – review \& editing.
+\textbf{Martin Herold}: Conceptualisation, Project Administration, Supervision, Methodology.
+\textbf{Myroslava Lesiv}: Investigation, Data Curation.
+\textbf{Marcel Buchhorn}: Writing – review \& editing.
+\textbf{Jan Verbesselt}: Conceptualisation, Methodology, Supervision, Writing – review \& editing.
 
-\minisection{Author Contributions}
+% \begin{itemize}
+%     \item Conceptualisation: MH NT JV
+%     \item Data curation: DM
+%     \item Formal Analysis: DM
+%     \item Funding acquisition: JV MH BS
+%     \item Investigation: DM ML NT
+%     \item Methodology: DM MH JV NT BK LL
+%     \item Project administration: MH NT JV
+%     \item Resources: MH BS
+%     \item Software: DM
+%     \item Supervision: MH NT JV
+%     \item Validation: NT DM
+%     \item Visualisation: DM MH
+%     \item Writing – original draft: DM
+%     \item Writing – review \& editing: NT JV MH MB BS
+% \end{itemize}
 
-\begin{itemize}
-    \item Conceptualisation: MH NT JV
-    \item Data curation: DM
-    \item Formal Analysis: DM
-    \item Funding acquisition: JV MH BS
-    \item Investigation: DM ML NT
-    \item Methodology: DM MH NT BK LL
-    \item Project administration: MH NT JV
-    \item Resources: MH BS
-    \item Software: DM
-    \item Supervision: MH NT JV
-    \item Validation: NT DM
-    \item Visualisation: DM MH
-    \item Writing – original draft: DM
-    \item Writing – review \& editing: NT JV MH MB BS
-\end{itemize}
+\minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {Proba-V} data, and IIASA for providing the model training dataset.
+In addition, we thank Benjamin Kellenberger for help on tuning neural network parameterisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
 
-DM JV NT MH ML MB BS RK
+%\minisection{Conflicts of Interest} The authors declare no conflict of interest.
 
-\minisection{Acknowledgements} VITO, Terrascope, time series group (BB, BK)
-
-\minisection{Conflicts of Interest} The authors declare no conflict of interest.
+%\minisection{Appendix 1} % Not sure how to typeset this, as it does not fit under the section header here.
+\begin{table}
+\begin{center}
+\begin{adjustbox}{max width=8cm}
+\begin{tabular}{m{12cm}l}
+\toprule
+\multicolumn{1}{l}{Category \& data source}&\multicolumn{1}{c}{Covariate}\tabularnewline
+\midrule
+Location (3, intrinsic)&
+\begin{tabular}{@{}l@{}}longitude\tabularnewline
+latitude\tabularnewline
+absolute latitude
+\end{tabular}\tabularnewline
+\midrule
+Vegetation indices (22, derived from Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
+\begin{tabular}{@{}l@{}}
+minimum NDVI\tabularnewline
+maximum NDVI\tabularnewline
+median NDMI\tabularnewline
+NDMI yearly IQR\tabularnewline
+NDMI March-May IQR\tabularnewline
+NDMI June-August IQR\tabularnewline
+NDMI September-November IQR\tabularnewline
+NDMI December-February IQR\tabularnewline
+OSAVI March-May IQR\tabularnewline
+OSAVI June-August IQR\tabularnewline
+OSAVI September-November IQR\tabularnewline
+OSAVI December-February IQR\tabularnewline
+EVI March-May IQR\tabularnewline
+EVI June-August IQR\tabularnewline
+EVI September-November IQR\tabularnewline
+EVI December-February IQR\tabularnewline
+median NIRv\tabularnewline
+NIRv yearly IQR\tabularnewline
+NIRv March-May IQR\tabularnewline
+NIRv June-August IQR\tabularnewline
+NIRv September-November IQR\tabularnewline
+NIRv December-February IQR
+\end{tabular}\tabularnewline
+\midrule
+Temporal metrics (9, derived from a harmonic model over time series of Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
+\begin{tabular}{@{}l@{}}
+NDVI order 1 cosine\tabularnewline
+NDVI order 1 sine\tabularnewline
+NDVI order 2 cosine\tabularnewline
+NDVI order 2 sine\tabularnewline
+NDVI trend coefficient\tabularnewline
+NDVI order 1 phase\tabularnewline
+NDVI order 1 amplitude\tabularnewline
+NDVI order 2 phase\tabularnewline
+NDVI order 2 amplitude
+\end{tabular}\tabularnewline
+\midrule
+Terrain (4, ASTER GDEM V003 \citep{ASTGTM003})&
+\begin{tabular}{@{}l@{}}
+elevation\tabularnewline
+slope (log-transformed)\tabularnewline
+aspect\tabularnewline
+terrain position index
+\end{tabular}\tabularnewline
+\midrule
+Climate (21, WorldClim 2.0 \citep{worldclim2})&
+\begin{tabular}{@{}l@{}}
+January precipitation (log)\tabularnewline
+April precipitation (log)\tabularnewline
+July precipitation (log)\tabularnewline
+October precipitation (log)\tabularnewline
+January solar irradiance\tabularnewline
+July solar irradiance\tabularnewline
+mean temperature\tabularnewline
+temperature monthly range\tabularnewline
+isothermality\tabularnewline
+temperature annual range\tabularnewline
+annual precipitation (log)\tabularnewline
+temperature seasonality\tabularnewline
+minimum solar irradiance\tabularnewline
+maximum solar irradiance\tabularnewline
+mean solar irradiance\tabularnewline
+mean windspeed\tabularnewline
+mean water vapour pressure\tabularnewline
+coldest month precipitation (log)\tabularnewline
+warmest month precipitation (log)\tabularnewline
+wettest month solar irradiance\tabularnewline
+driest month solar irradiance
+\end{tabular}\tabularnewline
+\midrule
+Soil (8, SoilGrids \citep{hengl_soilgrids250m_2017})&
+\begin{tabular}{@{}l@{}}
+soil available water\tabularnewline
+soil bulk density\tabularnewline
+soil cation exchange capacity (log)\tabularnewline
+soil clay fraction\tabularnewline
+soil coarse fragments (log)\tabularnewline
+soil pH\tabularnewline
+soil sand fraction\tabularnewline
+soil water wilting point
+\end{tabular}\tabularnewline
+\bottomrule
+\end{tabular}
+\end{adjustbox}
+\label{tab-inputdata}
+\caption{List of covariates input into the models, and their data sources.}
+\end{center}
+\end{table}
 
@@
 
-\printnoidxglossary[type=acronym]
+%\printnoidxglossary[type=acronym]
 
@@
 \end{document}
+\ead{marcel.buchhorn@vito.be}

### 2020-07-22 article/article.tex
Resolving most comments of draft 2

@@
 \newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
-\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
+%\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
 \newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
@@
 \newacronym{NDMI}{NDMI}{Normalised Difference Moisture Index}
-\newacronym{NIRv}{NIRv}{\gls{NIR} of vegetation}
+\newacronym{NIRv}{NIR\textsubscript{v}}{\gls{NIR} of vegetation}
+\newacronym{EVI}{EVI}{Enhanced Vegetation Index}
 \newacronym{NIR}{NIR}{near infra-red}
@@
 \newacronym{TOC}{TOC}{Top-of-Canopy}
-\newacronym{CART}{CART}{Classification and regression tree}
+\newacronym{CART}{CART}{classification and regression tree}
 \newacronym{GLM}{GLM}{General linear regression model}
 \newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
+\newacronym{GDAL}{GDAL}{Geospatial Data Abstraction Library}
 %\makenoidxglossaries
@@
 
-We compared the performance of three classes of machine learning regression algorithms for land cover fraction mapping, for the best model assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe at a sampling interval of 0.2 degrees.
-
-The models were trained on over 130\,000 points and validated on a separate dataset of over 20\,000 points of the Copernicus Global Land System: Dynamic Land Cover (CGLS-LC100) project.
-Both datasets are global and aligned with the Proba-V 100 m grid.
-
-One challenge for fraction mapping models is data sparsity: most fractions are 0\%, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
+We compared the performance of three types of machine learning regression algorithms for land cover fraction mapping.
+For the best model, we assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe at a sampling interval of 0.2 degrees.
+The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points of the \gls{CGLS-LC100} project.
+Both datasets are global and aligned with the PROBA-V 100 m grid.
+
+One challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
 We proposed a new solution by combining three models.
@@
 Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%.
-Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved by a tuned random forest model with our proposed three-model approach, which results in better estimation of extreme values at the cost of more misclassification.
+Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved by a tuned random forest model with our proposed three-model approach.
 The models were driven primarily by optical remote sensing data, followed by soil and climate data.
@@
 This research proves that machine learning algorithms can be applied globally to a wide variety of land cover classes to create an exhaustive global land cover fraction map.
-This expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which helps to customise the result for a diverse range of uses.
-In addition, this work contributes to efforts to produce exhaustive global land cover fractions as an operational product, and to assess the accuracy of land cover fraction maps both thematically and spatially.
+Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which helps to customise the result for a diverse range of uses.
+In addition, this work directly contributes to the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project, as well as to the accuracy assessment of land cover fraction maps both thematically and spatially.
 \end{abstract}
@@
 \begin{itemize}
-    \item Three groups of statistical and machine learning models tested for deriving seven land cover fractions at a global scale
+    \item 3 machine learning model groups tested for deriving 7 land cover fractions globally
     \item Random forest regression performed the best (17.3 RMSE or 7.9 MAE)
-    \item Remote sensing data was the most important for model predictions with varying features relevant for different fraction predictions
-    \item Independent validation results show highest accuracies for built-up and water fractions, lower accuracies are found for grasses and shrubs
-    \item Comprehensive multi-fraction global land cover characterisation shown as an alternative to discrete classification
+    \item Optical remote sensing data was the most important input for model predictions
+    \item Independent validation showed highest accuracies for built-up and water fractions
+    \item Multi-fraction land cover mapping shown as an alternative to discrete classification
 \end{itemize}
@@
 In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
-Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \ac{CGLS-LC100} map \citep{buchhorn_moderate_2019} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
-
-All of these global land cover maps (with the exception of \ac{CGLS-LC100} fraction layers), as well as previous global land cover maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided as discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
+Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
+
+Except for the \ac{CGLS-LC100} product with its cover fraction layers, all other global land cover maps, including previous maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
 However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
@@
 \citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
-Nevertheless, to date only \ac{CGLS-LC100} provides a global map with fractions of every major land cover class.
+Nevertheless, to date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.
 
@@
 Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g.
-\citealp{yang_landsat_2012, hobbs2003linear, shimabukuro1991least}), \ac{RF} regression \citep{schwieder_estimating_2014}, \gls{SVM} regression \citep{schwieder_estimating_2014}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
+\citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression \citep{walton2008subpixelrf}, \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
 So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor at global scale.
@@
 In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
-In addition to spectral covariates (blue, red, near infrared, shortwave infrared bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}.
+In addition to optical covariates (blue, red, near infrared, shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}.
 Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}.
-\Glspl{VI} have been used to improve the detection of wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
-Therefore in this paper such covariates have also been used in order to increase land cover classification accuracy, and their importance for prediction was analysed.
+\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are often used to improve land cover class separability, e.g.
+for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
+Therefore in this paper such covariates have also been used in order to increase the accuracy of land cover fraction regression model predictions, and their importance for the predictions was analysed.
 
 Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
-It requires a comprehensible dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
+It requires a comprehensive dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
 For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
-In this study, we made use of an independent validation set of over 20\,000 points collected across the globe using the same legend, grid and spatial resolution.
+Because of these challenges, the information about the accuracy of the existing global land cover products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
+Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
+In this study, we made use of an independent validation set collected as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019} that includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as in our study.
 The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
@@
  \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
- \item Assess the accuracy of the models from a thematic and spatial point of view, taking various user needs into account
+ \item Assess the accuracy of the models from a thematic and spatial point of view
 \end{enumerate}
@@
 
-The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS} project.
+The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020, nandin-erdene_tsendbazar_copernicus_2019}.
 The training set was generated by experts in \ac{IIASA}.
-This global point dataset provides information on nine class fractions of sampled pixels of the Proba-V 100 m product.
+This global point dataset provides information on nine class fractions of sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
 The fraction estimate was acquired by high-resolution image interpretation by experts, for the year 2015.
-Each of the sampled Proba-V pixels was subdivided into a 10-by-10 subpixel grid and the area estimates were derived from the subpixel ratios.
-
-For model validation, an independent global dataset of \ac{CGLS}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
+Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.
+
+For model validation, an independent global dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
 The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.
@@
 The data includes over 150\,000 training points and over 21\,000 validation points across the globe, describing the fractions of 12 classes in the year 2015.
-The classification scheme follows the \gls{LCCS} \citep{buchhorn_moderate_2019}.
-However, some of the classes were very detailed and had too few observations to classify well, thus we merged the classes into a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
+The classification scheme follows the \gls{LCCS} \citep{buchhorn_copernicus_2020}.
+However, due to limited number of observations for some rare classes, we merged them to get a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous vegetation (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
 The unknown class was discarded: points with the dominant land cover class marked as unknown were not used.
@@
  \includegraphics[width=\textwidth]{article/article-figures/maps/2020-07-06-training-and-validation}
- \caption{Points representing 100x100 m areas at which land cover reference data was used in the study.
+ \caption{Sample points representing 100x100 m areas at which land cover reference data was used in the study.
 The colours represent the dominant land cover class at each point.
 (a): training dataset, collected by \gls{IIASA}, 138\,164 points used in this study.
-(b): validation dataset, collected by \gls{WUR}, 20\,705 points used in this study.}
+(b): validation dataset, collected by \gls{WUR}, 20\,705 points used in this study.
+Both datasets were collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020}.}
  \label{fig-reference-data}
 \end{figure}
+
+\subsection{Model covariates}
+
+To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
+See figure \ref{fig-processing} for an overview of the whole processing chain.
+
+In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see table \ref{tab-inputdata}).
+All of these covariates had to be preprocessed before they could be input into the models.
 
@@
 
-\subsection{Model covariates}
-
-To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
-See figure \ref{fig-processing} for an overview of the whole processing chain.
-
-In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see table \ref{tab-inputdata}).
-All of these covariates had to be preprocessed before they could be input into the models.
-
 % \begin{table}
@@
 %          \midrule
-%          Spectral & 22 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
-%          Harmonic & 9 & Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
+%          Spectral & 22 & PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
+%          Harmonic & 9 & PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2} \\
 %          Terrain & 4 & ASTER GDEM V003 \citep{ASTGTM003} \\
@@
 
-The entire archive (2014-03-11 to 2019-07-16) of the Proba-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{probavguide2} was used for this study.
-While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of Proba-V was chosen to obtain more robust statistics for the temporal metrics.
-Land cover change is a relatively rare phenomenon and thus is not likely to affect more than 2\% of the training points within the time period.
+The entire archive (2014-03-11 to 2019-07-16) of the PROBA-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{dierckx2014probav,probavguide2} was used for this study.
+While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of PROBA-V was chosen to obtain more robust statistics for the temporal metrics.
+Land cover change is a relatively rare phenomenon and thus, according to tests performed in the making of \gls{CGLS-LC100} collection 3, is not likely to affect more than 2\% of the training points.
 Nevertheless, this may result in poorer model performance compared to if reference data for each year would be available.
 
-The time series of each of the Proba-V bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue band.
-The resulting cloud-free time series was then used to extract descriptive statistics (median and interquartile range) of NDVI, NDMI, EVI, OSAVI and NIR\textsubscript{v} over the whole time series, as well as for each season separately.
-Next, harmonic analysis was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
+The time series of each of the PROBA-V spectral bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue reflectance band.
+The resulting cloud-free time series was then used to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}.
+Next, the \glspl{VI} were used to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
+Next, harmonic analysis \citep{jakubauskas2001harmonic} was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
 From this, phase and amplitude for the two orders were calculated.
@@
 
-% \begin{figure}
-%  \includegraphics[width=\textwidth]{article-figures/flowcharts/preprocessing-optical}
-%  \caption{Optical data preprocessing chain.}
-%  \label{fig-preprocessing-optical}
-% \end{figure}
-
 %\subsubsection{Terrain covariates}
 
-For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to Proba-V 100 m grid and used as the elevation covariate.
-In addition, GDAL was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
-%All terrain covariates were then resampled to the Proba-V grid.
+For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to PROBA-V 100 m grid and used as the elevation covariate.
+In addition, \gls{GDAL} \citep{gdal} was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
+%All terrain covariates were then resampled to the PROBA-V grid.
 
@@
 
-For climate data, WorldClim 2.0 30 s product \citep{worldclim2} was used.
+For climate data, WorldClim 2.0 30 second product \citep{worldclim2} was used.
 It includes monthly temperature, precipitation, solar radiation, wind speed and water vapour pressure data.
-In additional, 19 bioclimatic parameters were calculated from the data, using the dismo package \citep{dismo}.
+In addition, 19 bioclimatic parameters were calculated from the data, using the dismo package \citep{dismo}.
 Some additional biophysical parameters were also calculated, namely all of the climate data during the coldest, warmest, driest and wettest months of the year at each location, as well as yearly averages of each of the original parameters.
@@
 
-SoilGrids \citep{hengl_soilgrids250m_2017} was used for getting data about soils.
-It is based on a random forest model that predicts soil properties at various soil depths globally at a 250 m resolution.
+SoilGrids \citep{hengl_soilgrids250m_2017} was used for variables related to soils.
+SoilGrids is based on a random forest model that predicts soil properties at various soil depths globally at a 250 m resolution.
 In the creation of SoilGrids, a land cover map (based on MODIS) had been used, and so, in order to avoid circular inference, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
@@
 
-In total, 313 covariates were used for predicting the land cover fractions.
-However, many of them are colinear with one another, and thus variable selection was used to remove colinear covariates.
-Covariates that had a Pearson's correlation coefficient $r$ above 0.9 were excluded in an iterative process.
-After that, covariates with a Spearman's $\rho$ above 0.9 were likewise excluded.
-
-Most of the colinear covariates were soil covariates at different depths.
+In total, 313 covariates were generated in pre-processing.
+However, many of them are collinear with one another, which prolongs training time for machine learning models and leads to unreliable coefficient estimation and increased error variance for linear models.
+Thus, variable selection was used to remove collinear covariates before predicting the land cover fractions.
+Covariates that had a Pearson's correlation coefficient $r$ \citep{pearson_notes_1895} above 0.9 were excluded in an iterative process.
+After that, covariates with a Spearman's rank correlation $\rho$ \citep{spearman1904rank} above 0.9 were likewise excluded.
+
+Most of the collinear covariates were soil covariates at different depths.
 Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
-Similarly, climate data was colinear between subsequent months.
+Similarly, climate data was collinear between subsequent months.
 Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
-In addition, all of the spectral bands of Proba-V were also highly correlated with each other as well as to vegetation indices, so only vegetation indices were kept.
+In addition, all of the spectral bands of PROBA-V were also highly correlated with each other as well as to \glspl{VI}, so only \glspl{VI} were kept.
 
@@
 For deriving land cover fractions, we compared a wide array of machine learning regression methods.
-They can be broadly divided into three categories.
-Tested linear models were \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression and \ac{MLR}.
-Non-tree machine learning models tested were \ac{MLP} \glspl{NN} and \glspl{SVM}.
-Lastly, machine learning models based on \glspl{CART} tested were Cubist and \ac{RF} regressions.
+We broadly divided the methods into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
+See table \ref{tab-methods} for the full list of methods that were compared in this study.
+%They can be broadly divided into three categories.
+%Tested linear models were \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression and \ac{MLR}.
+%Non-tree machine learning models tested were \ac{MLP} \glspl{NN} and \glspl{SVM}.
+%Lastly, machine learning models based on \glspl{CART} tested were Cubist and \ac{RF} regressions.
 In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
@@
 All of the models were made using the free and open-source statistical software R \citep{r_2019}.
+
+\begin{table}
+\centering
+    \begin{tabular}{m{4cm}ll}
+         \toprule
+         \textbf{Category} & \textbf{Name} & \textbf{Reference} \tabularnewline
+         \midrule
+         Linear models &
+         \begin{tabular}{@{}l@{}}
+         \Glsentryfull{FNC}\tabularnewline
+         \Glsentryfull{GLM}\tabularnewline
+         \Glsentryfull{PLS} regression \tabularnewline
+         Lasso regression \tabularnewline
+         \Glsentryfull{MLR}
+         \end{tabular}
+         & 
+         \begin{tabular}{@{}l@{}}
+         \citealt{keller_fuzzy_1985} \tabularnewline
+         \citealt{neter_applied_1996} \tabularnewline
+         \citealt{wold_pls-regression_2001} \tabularnewline
+         \citealt{tibshirani_regression_1996} \tabularnewline
+         \citealt{theil_multinomial_1969}
+         \end{tabular}
+         \tabularnewline \midrule
+         Machine learning models based on decision trees &
+         \begin{tabular}{@{}l@{}}
+         \Glsentryfull{RF} regression\tabularnewline
+         Cubist regression
+         \end{tabular}
+         &
+         \begin{tabular}{@{}l@{}}
+         \citealt{breiman2001random} \tabularnewline
+         \citealt{cubist}
+         \end{tabular}
+         \tabularnewline \midrule
+         Other machine learning models &
+         \begin{tabular}{@{}l@{}}
+         \Glsentryfullpl{MLP}\tabularnewline
+         \Glsentryfull{SVM} regression
+         \end{tabular}
+         & 
+         \begin{tabular}{@{}l@{}}
+         \citealt{dreyfus_artificial_1990} \tabularnewline
+         \citealt{suykens_least_1999}
+         \end{tabular}
+         \tabularnewline \bottomrule
+    \end{tabular}
+    \caption{List of regression methods for land cover fraction estimation tested in this study.}
+    \label{tab-methods}
+\end{table}
 
@@
 
-\Glsentryfull{GLM}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
+\Glsentryfull{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
 In this study, we used \gls{GLM} as a baseline model.
@@
 This has the effect of the coefficients of correlated covariates being set to a similar value.
-We chose lasso regression, so that it would be possible to identify if there are any covariates that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
+We chose lasso regression \citep{tibshirani_regression_1996}, so that it would be possible to identify if there are any covariates that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
 The $\lambda$ parameter was chosen using cross-validation.
@@
 
-\Glsentryfull{PLS} regression is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space \citep{martens_multivariate_1992}.
+\Glsentryfull{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space.
 As such, it is well suited for multicollinear data.
@@
 
-\Glsentryfull{FNC} regression, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
+\Glsentryfull{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
 This method always produces output that sums to 100\%.
@@
 
-Lastly, we tested \glsentryfull{MLR}.
+Lastly, we tested \glsentryfull{MLR} \citep{theil_multinomial_1969}.
 While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
@@
 \Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
-In this study, we used a \glsentryfull{MLP} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+In this study, we used a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
 Softmax activation was used to get the output.
@@
 
-A few more methods were considered, but in the end turned out to be unworkable with our dataset.
+Additional methods were considered, but in the end turned out to be unworkable with our dataset.
 Multivariate \gls{RF} is available in R provided by the \texttt{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
@@
 
-\subsection{Validation/Accuracy Assessment}
+\subsection{Validation/}
 
@@
 \subsection{Method accuracy comparison}
+
+Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
+\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
+\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.
+This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
+The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
 
@@
 
-Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
-\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
-\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.
-This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
-The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
-
 \begin{figure}
@@
 
-All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil).
+All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil), as shown in figure \ref{fig-models}.
 This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
@@
     \centering
-    \includegraphics[width=12cm]{article-figures/heatmaps/2020-06-19-varimp-heatmap-top5}
+    \includegraphics[width=14cm]{article-figures/heatmaps/2020-06-19-varimp-heatmap-top5}
     \caption{Random Forest single model variable importance, top 5 covariates per category.
@@
 
-The overall most important variables were the maximum and minimum NDVI over the whole time series of Proba-V imagery.
+The overall most important variables were the maximum and minimum NDVI over the whole time series of PROBA-V imagery.
 They were followed by the median NDMI and NIRv over the whole time series.
@@
 This is also presented in raster format in figure \ref{fig-walltowall}.
+
+The global patterns of the land cover fractions are shown to express biotic gradients.
+For instance, gradients are visible between communities dominated by shrubs and ones dominated by herbaceous vegetation, such as in south and east Africa.
+Likewise, the gradient of tree cover from 100\% in the African tropics to 0\% in the sub-Sahara region is evident.
+Herbaceous cover in sub-Sahara shows an asymmetric gradient: the cover is highest at the 14th-15th latitude and decreases quickly to the north, up to the 18th latitude, but slowly to the south, until the 5th latitude.
+Inland water shows up as more discrete, as it naturally forms discrete patches and mixed pixels with water are more rare.
+Built-up area is relatively rare and is rarely forms a 100\% fraction, as urban areas are in most places interspersed with greenery within the footprint of the pixel.
+
+The spatial pattern of map accuracy shows that the land cover of areas with pure land cover was the easiest to predict, such as tropical forests (100\% tree cover) and deserts (100\% bare land).
+Conversely, it was more difficult to predict fractions in areas with mixed land cover, as isolating individual fractions becomes more challenging.
+In addition, land cover fractions in the extreme latitudes were more difficult to predict as well.
+There less training data was available, owing to the lack of high-resolution imagery for image interpretation in these areas.
+
+The distribution of the predictions was relatively even across the whole range for herbaceous vegetation and trees, but uneven for bare land and especially built-up area.
+Built-up fraction was never predicted over 65\% and thus was more often underestimating in cases where the predicted fraction was above 15\%.
+The predictions of each class showed underestimation of high fractions and overestimation of low fractions, i.e. tending towards the mean.
 
@@
 
-Compared to a global land cover product that specialises in a specific land cover class, namely tree cover, the \gls{RF} one-step mean model had a higher RMSE (19.2\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (17.4\%), however, the methodology for the validation differs significantly from the one used in our study.
+Compared to a global land cover product that specialises in a specific land cover class, namely tree cover, the \gls{RF} one-step mean model had a higher RMSE (19.2\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (16.8\%).
+However, the methodology for the validation differed significantly from the one used in our study, as it compared the predicted fractions to fractions as computed from lidar datasets within several local study areas, rather than using global expert image interpretation as we did.
 \citet{feng_earth_2016} assessed a discrete (binary) GLCF forest cover product for three different years using a more similar methodology, and yielded similar results to ours: a user's accuracy of 98\% and producer's accuracy of 80\%, compared to the \gls{RF} three-step median tree cover user's accuracy of 81\% and producer's accuracy of 86\%.
@@
 
-In this study, we compared three classes of algorithms for land cover fraction mapping and created a demonstation map showing the result of the best performing model by \gls{RMSE}.
-We also introduced a multi-step approach to try and tackle the issue of correctly predicting the extreme fraction values of 0\% and 100\%.
+In this study, we compared three groups of algorithms for land cover fraction mapping and created a demonstation map showing the result of the best performing model by \gls{RMSE}.
+We also introduced a multi-step approach to tackle the issue of correctly predicting the extreme fraction values of 0\% and 100\%.
 The best performing models were based on \gls{CART} tree ensembles, with \gls{RF} regression performing the best overall.
@@
 It also allows the users to define their own thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
-Thus this work paves the way for operationalisation of global land cover fraction mapping.
+This work paves the way for operationalisation of global land cover fraction mapping.
 
@@
 
-\minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {Proba-V} data, and IIASA for providing the model training dataset.
+\minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {PROBA-V} data, and IIASA for providing the model training dataset.
 In addition, we thank Benjamin Kellenberger for help on tuning neural network parameterisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
@@
 \midrule
-Vegetation indices (22, derived from Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
+Vegetation indices (22, derived from PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
 \begin{tabular}{@{}l@{}}
@@
 \midrule
-Temporal metrics (9, derived from a harmonic model over time series of Proba-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
+Temporal metrics (9, derived from a harmonic model over time series of PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
 \begin{tabular}{@{}l@{}}
@@
 \end{adjustbox}
+\caption{List of covariates input into the models, and their data sources.}
 \label{tab-inputdata}
-\caption{List of covariates input into the models, and their data sources.}
 \end{center}
@@
 \end{document}
-\ead{marcel.buchhorn@vito.be}

### 2020-08-05 article/article.tex
Draft 3

@@
 \usepackage{amsmath}
-\usepackage{indentfirst}
+%\usepackage{indentfirst}
 \usepackage{booktabs} % For \toprule, etc.
+\usepackage{textcomp} % for degree sign
 %\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
 \usepackage{graphicx}
-\usepackage{adjustbox} % For resizing tables
+%\usepackage{adjustbox} % For resizing tables
 \usepackage{array} % For mid-alignment (m{xcm}) in tables
@@
 
+% Put table captions on top
+\usepackage{floatrow}
+\floatsetup[table]{capposition=top}
+
 % My own definition for the sections that start with a bold name but don't appear in TOC
 \newcommand{\minisection}[1]{\paragraph{#1}}%{\medskip \textbf{#1:}}
+% Package names
+\newcommand{\cran}[1]{CRAN: #1}
 
@@
 \newacronym{NIR}{NIR}{near infra-red}
+\newacronym{SIF}{SIF}{sun-induced chlorophyll fluorescence}
+\newacronym{GPP}{GPP}{gross primary productivity}
+\newacronym{MSI}{MSI}{multi-spectral instrument}
+\newacronym{SAR}{SAR}{synthetic aperture radar}
 \newacronym{SCM}{SCM}{subpixel confusion-uncertainty matrix}
@@
 \newacronym{CART}{CART}{classification and regression tree}
-\newacronym{GLM}{GLM}{General linear regression model}
+\newacronym{GLM}{GLM}{general linear regression model}
 \newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
@@
 \begin{abstract} % elsarticle: environment instead of a command \abstract{
-Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a particular ratio.
+Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a predetermined ratio.
 In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity.
@@
 
-One challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
+One challenge for models for land cover fraction mapping is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
 We proposed a new solution by combining three models.
@@
 
-\minisection{Highlights}
-
-\begin{itemize}
-    \item 3 machine learning model groups tested for deriving 7 land cover fractions globally
-    \item Random forest regression performed the best (17.3 RMSE or 7.9 MAE)
-    \item Optical remote sensing data was the most important input for model predictions
-    \item Independent validation showed highest accuracies for built-up and water fractions
-    \item Multi-fraction land cover mapping shown as an alternative to discrete classification
-\end{itemize}
-
-%\pagebreak
-%\tableofcontents
-%\pagebreak
-
 \section{Introduction}
@@
 In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
-Some of the recent achievements have been the \ac{LC-CCI} map \citep{defourny2012cci} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
+Some of the recent achievements have been the \ac{LC-CCI} map \citep{esa_land_2017} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
 
@@
 This issue becomes worse at coarse resolutions and heterogeneous landscapes.
-It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus ultimately lead to an underestimation of tree cover at large scales.
+It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus underestimate tree cover in the pixel.
+These systematic errors then add up when scaling the result to the entire region.
 
-A potential solution to this issue is land cover fraction mapping, where instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
+A potential solution to this issue is to characterise land cover using cover fractions.
+In this approach, instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
+That way, the land cover models work not on pixel labels, such as ``forest'', but on land cover characteristics, such as tree cover, defined as the area of the pixel covered by tree canopies, as well as herbaceous cover, defined as the area not covered by woody vegetation.
 This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
-Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but this concept relates also to the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.
+Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but these concepts include the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.
 
-Land cover fraction mapping has been attempted in the past, both in research \citep{adams_classification_1995, foody1996fuzzyevaluation, colditz_land_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018} and production (e.g.
-\citealp{hansen2000hardtree, hansen_continuous_2011, pengra_global_2015, hansen_global_2003, sexton_global_2013}).
-Nevertheless, to date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.
+Land cover fraction mapping has been attempted in the past.
+Most of the previous research has focused on deriving land cover fractions of 3-6 classes at a local scale  \citep{adams_classification_1995, foody1996fuzzyevaluation, walton2008subpixelrf, hansen_continuous_2011, sharma_assessing_2011, uma_shankar_wavelet-fuzzy_2011, dwivedi_optimisation_2012, lizarazo_quantitative_2012, gessner_estimating_2013, okujeni_generalizing_2018}, less often at a regional scale and with more detailed classes \citep{colditz_land_2011}.
+The methods for assessing the accuracy of the results vary greatly between the different studies.
+In addition, global land cover fraction products have emerged, but focused on a particular class, such as tree cover \citep{hansen_global_2003, Hansen2013forestchange}, water \citep{schroeder_development_2015} and urban area \citep{gao_mapping_2020}.
+To date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.
 
@@
 \citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression \citep{walton2008subpixelrf}, \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
-So far other studies have only compared a few of these methods at once, and never with an exhaustive set of land cover classes nor at global scale.
+So far other studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.
 
 In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
-In addition to optical covariates (blue, red, near infrared, shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of crops \citep{jakubauskas2001harmonic}.
-Elevation covariates (elevation, slope, aspect) are known to improve the separability between tree classes \citep{burrough2001fuzzy}.
+In addition to optical covariates (such as blue, red, \gls{NIR} and shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
+Elevation covariates (elevation, slope, aspect) are known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
 \Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are often used to improve land cover class separability, e.g.
@@
 
+A common issue with the use of land cover fraction data as input into regression models is data imbalance.
+The more classes are mapped, the more often there will be 0\% of a given class in any particular pixel, leading to zero inflation.
+In addition, in homogeneous areas, 100\% cover of a single class is common.
+This leaves little data in the middle for the regression models to learn from.
+Conversely, regression models tend to favour predictions closer to the mean and rarely predict extreme values.
+We investigated this issue and suggested ways to deal with it.
+
 Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
@@
 For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
-Because of these challenges, the information about the accuracy of the existing global land cover products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
+Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
 Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
@@
 
-All in all, with this study we aim to investigate and compare methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
-We also aim to suggest ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
-Therefore the objectives of our study are:
+All in all, in this study we investigated and compared methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
+We also investigated ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
+Therefore the objectives of our study were to:
 
 \begin{enumerate}
- \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping
- \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean
- \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions
- \item Assess the accuracy of the models from a thematic and spatial point of view
+ \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping.
+ \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean.
+ \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions.
+ \item Assess the accuracy of the models from a thematic and spatial point of view.
 \end{enumerate}
@@
 The training set was generated by experts in \ac{IIASA}.
-This global point dataset provides information on nine class fractions of sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
-The fraction estimate was acquired by high-resolution image interpretation by experts, for the year 2015.
+This global point dataset provides information on fractions of each class within sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
+The fraction estimates were acquired by high-resolution image interpretation by experts, for the year 2015.
 Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.
 
-For model validation, an independent global dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
+For model accuracy assessment, an independent global validation dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
 The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.
@@
 Thus in the end the training dataset size became 138\,164 and the validation dataset size became 20\,705. See figure \ref{fig-reference-data} for the spatial distribution of the points.
-
-%Since the data describes fractions, it is zero-inflated, since it is relatively rare for more than just a few land cover classes to occupy space within an area of a single pixel.
-%This means that for a given pixel, only a few of the classes would normally have a fraction that is above 0\%.
-%This leads to a dataset that is zero-inflated (also known as sparse), with over half of all values for a pixel being zero.
-%Such an imbalance is a problem for statistical and machine learning techniques that are tuned by minimising the error of the output, since they are heavily biased towards a good prediction of zeroes.
-%In other words, when validating against a zero-inflated dataset, a prediction of zero is a safe bet.
-%But for user communities, correct prediction of non-zero values is more important.
 
@@
 See figure \ref{fig-processing} for an overview of the whole processing chain.
+The processing was carried out in R \citep{r_2019} and the resulting code has been made openly available in \citet{dainius_masiliunas_2020_3973123}.
 
-In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see table \ref{tab-inputdata}).
-All of these covariates had to be preprocessed before they could be input into the models.
+In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
+These covariates had to be preprocessed before they could be input into the models.
 
@@
  \centering
- \includegraphics[width=0.75\textwidth]{article-figures/flowcharts/2020-07-10-flowchart}
+ \includegraphics[width=0.6\textwidth]{article-figures/flowcharts/2020-07-10-flowchart}
  \caption{Processing workflow, from the raw input data to model accuracy assessment and wall-to-wall map output.}
@@
 
-%\subsubsection{Vegetation indices and temporal metrics}
 \label{sec-temporal-filter}
@@
 While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of PROBA-V was chosen to obtain more robust statistics for the temporal metrics.
-Land cover change is a relatively rare phenomenon and thus, according to tests performed in the making of \gls{CGLS-LC100} collection 3, is not likely to affect more than 2\% of the training points.
-Nevertheless, this may result in poorer model performance compared to if reference data for each year would be available.
+Land cover change is a relatively rare phenomenon, according to tests performed in the making of \gls{CGLS-LC100} collection 3, and so we expected that the difference caused by land cover change to the input data would be smaller than the margin of error of the model output.
+Nevertheless, this may result in poorer model performance compared to if up-to-date reference data for each year would be available.
 
@@
 Next, harmonic analysis \citep{jakubauskas2001harmonic} was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
-From this, phase and amplitude for the two orders were calculated.
-Lastly, the minimum and maximum of the NDVI time series was also retained as covariates.
-
-%\subsubsection{Terrain covariates}
+From this harmonic analysis we obtained minimum and maximum values of the \gls{NDVI} time series, and calculated the phase and amplitude for the two harmonic orders from the respective sine and cosine components.
 
@@
 In addition, \gls{GDAL} \citep{gdal} was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
-%All terrain covariates were then resampled to the PROBA-V grid.
-
-%\subsubsection{Climate covariates}
 
@@
 
-%\subsubsection{Soil covariates}
-
 SoilGrids \citep{hengl_soilgrids250m_2017} was used for variables related to soils.
@@
 The soil taxonomy covariates were also excluded, since they are categorical derivatives from the numerical soil property data and thus do not contribute to land cover fraction prediction.
-
-%\subsubsection{Location covariates}
 
@@
 In total, 313 covariates were generated in pre-processing.
-However, many of them are collinear with one another, which prolongs training time for machine learning models and leads to unreliable coefficient estimation and increased error variance for linear models.
+However, many of them were collinear with one another.
+Multicollinearity prolongs training time for machine learning models and leads to unreliable coefficient estimation and increased error variance for linear models.
 Thus, variable selection was used to remove collinear covariates before predicting the land cover fractions.
@@
 
-Most of the collinear covariates were soil covariates at different depths.
+The covariates to exclude were selected manually, to avoid interpretation difficulties that arise from automatic selection procedures.
+The majority of the collinear covariates were soil metrics predicted at different depths.
 Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
@@
 Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
-In addition, all of the spectral bands of PROBA-V were also highly correlated with each other as well as to \glspl{VI}, so only \glspl{VI} were kept.
+Our initial correlation analysis also showed that the spectral bands of PROBA-V were highly correlated with each other as well as to \glspl{VI}, so only \glspl{VI} were used as covariates.
 
@@
 These covariates include data from each of the covariate categories.
-See table \ref{tab-inputdata} for an overview.
-
-%\begin{figure}
- %\inputdigraph[width=\textwidth]{article-figures/algorithms}{dot}
-% \includegraphics[width=\textwidth]{article-figures/flowcharts/algorithms}
-% \caption{Preprocessing chain.}
-% \label{fig-preprocessing}
-%\end{figure}
+See supplementary material, table S1 for an overview.
 
@@
 We broadly divided the methods into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
+We tuned the methods to select optimal parameters for each.
 See table \ref{tab-methods} for the full list of methods that were compared in this study.
-%They can be broadly divided into three categories.
-%Tested linear models were \ac{FNC}, \ac{GLM}, \ac{PLS} regression, Lasso regression and \ac{MLR}.
-%Non-tree machine learning models tested were \ac{MLP} \glspl{NN} and \glspl{SVM}.
-%Lastly, machine learning models based on \glspl{CART} tested were Cubist and \ac{RF} regressions.
 In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
-
 All of the models were made using the free and open-source statistical software R \citep{r_2019}.
@@
 \subsubsection{Linear models}
-%\subsubsection{\Glsentryfull{GLM}}
 
-\Glsentryfull{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to a regular linear regression that allows for multiple outcomes.
-In this study, we used \gls{GLM} as a baseline model.
-We also tested whether dropping any additional covariates from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction to compare models with different covariates.
+The \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to the standard linear regression that allows for multiple outcomes.
+In this study, we used a \gls{GLM} as a baseline model.
+We also tested whether dropping any additional covariates from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction \citep{greenhouse_methods_1959} to compare models with different covariates.
 However, only three covariates could be dropped and it did not improve the predictions, therefore we decided to keep all the covariates in the model.
 \Gls{GLM} is implemented in base R \citep{r_2019}.
-
-%\subsubsection{Lasso/ridge/elastic net regression}
 
@@
 The $\lambda$ parameter was chosen using cross-validation.
-The lasso regression was performed using the \texttt{glmnet} package \citep{glmnet}.
+The lasso regression was performed using the \cran{glmnet} package \citep{glmnet}.
 
-%\subsubsection{\Glsentryfull{PLS} regression}
-
-\Glsentryfull{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space.
+\Gls{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space.
 As such, it is well suited for multicollinear data.
-\ac{PLS} regression was performed using the package \texttt{pls} \citep{pls}.
+\ac{PLS} regression was performed using the package \cran{pls} \citep{pls}.
 
@@
 
-%\subsubsection{\Glsentryfull{FNC} regression}
-
-\Glsentryfull{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
+\Gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
 This method always produces output that sums to 100\%.
@@
 
-%\subsubsection{\Glsentryfull{MLR}}
-
-Lastly, we tested \glsentryfull{MLR} \citep{theil_multinomial_1969}.
+Lastly, we tested \gls{MLR} \citep{theil_multinomial_1969}.
 While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
@@
 These probabilities always add up to 100\%.
-However, an \ac{MLR} can only be trained on labelled data, rather than fractions.
+However, an \gls{MLR} can only be trained on labelled data, rather than fractions.
 Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
 This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
-The logistic regression implementation was provided by the \texttt{nnet} package \citep{nnet}.
+The logistic regression implementation was provided by the \cran{nnet} package \citep{nnet}.
 
 \subsubsection{Non-tree machine learning methods}
-%\subsubsection{\Glsentryfullpl{NN}}
 
 \Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
-In this study, we used a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+In this study, after performing initial parameter tests, we used a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+We chose the Nadam optimiser \citep{dozat_incorporating_2016}, with \gls{MAE} as the loss function.
 Softmax activation was used to get the output.
-The Nadam optimiser \citep{dozat_incorporating_2016} was used, with \gls{MAE} as the loss function.
-The \texttt{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics processing unit to accelerate the neural network training process.
-
-%\subsubsection{\Glsentryfull{SVM} regression}
+The \cran{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics processing unit to accelerate the neural network training process.
 
@@
 For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
-We used the binary relevance method, clamped the output values to the 0-100\% range and linearly scaled them to add up to 100\%.
-The package \texttt{liquidSVM} \citep{liquidSVM} was suitable for this purpose, as it is optimised for processing large datasets such as ours.
+We used the binary relevance method, where separate models are trained per class that predict the proportion of the class in the pixel \citep{karalas2016br}.
+We clamped the output values to the 0-100\% range and linearly scaled them to add up to 100\%.
+The package \cran{liquidSVM} \citep{liquidSVM} was suitable for this purpose, due to its optimisations for processing large datasets such as ours.
 
 \subsubsection{Tree-based machine learning methods}
-%\subsubsection{\Glsentryfull{RF} regression}
 
@@
 One way is analogous to how we performed \ac{MLR}: giving labels of the dominant land cover class as the response variable, and then considering the proportion of votes from the trees as the proportions of land cover classes in the pixel.
-The second way is binary relevance, where a separate model is trained to predict the presence of each class \citet{karalas2016br}.
-Then the predictions are scaled linearly to sum up to 100\%.
-In the case of \gls{RF}, output values do not need to be clipped as their range is always within the constraints of the input data range.
+The second way is the binary relevance method as with \glspl{SVM}.
+In the case of \gls{RF}, output values do not need to be clamped as their range is always within the constraints of the input data range, but still need to be scaled to add up to 100\%.
 In this study, we chose the binary relevance approach, so that the model for each class would be able to learn from the fractions of all the input data, rather than just the labels of the dominant land cover class.
-The processing was done using the \texttt{ranger} package \citep{ranger}.
+The processing was done using the \cran{ranger} package \citep{ranger}.
 
@@
 Since the number of covariates is imbalanced by group (i.e.
-there are merely 3 location covariates and 22 vegetation index metrics), and there may still be colinear covariates, we also tested calculating covariate importance by permuting all covariates in a group together.
-However, the result was similar to the sum of permutation importance of each individual covariates in the group, which showed that the covariate selection process had successfully eliminated all colinear covariates.
+there are merely 3 location covariates and 22 vegetation index metrics), and there may still be collinear covariates, we also tested calculating covariate importance by permuting all covariates in a group together.
+However, the result was similar to the sum of permutation importance of each individual covariates in the group, which showed that the covariate selection process had successfully eliminated all collinear covariates.
 Thus we reported the more detailed covariate importance permuted per individual covariate.
-
-%\subsubsection{Cubist regression}
 
@@
 Like with \gls{RF}, Cubist was used with the binary relevance method.
-The \texttt{Cubist} package \citep{cubistpackage} was used for this.
+The \cran{Cubist} package \citep{cubistpackage} was used for this.
 
@@
 Additional methods were considered, but in the end turned out to be unworkable with our dataset.
-Multivariate \gls{RF} is available in R provided by the \texttt{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
+Multivariate \gls{RF} is available in R provided by the \cran{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
 However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
 
-Another model that was considered was SuperLearner \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
+Another model that was considered was \cran{SuperLearner} \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
 Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
@@
 
-%In addition, we investigated applying histogram matching, by matching the histogram of the predictions back to the histogram of the training set.
-%However, that turned out to decrease accuracy in all cases, as the cases where a model predicts a value close to the true value would be set to a more extreme value to compensate for other predictions that were not as accurate to begin with.
-%Therefore in the end histogram matching was not used.
+\subsection{Accuracy assessment}
 
-\subsection{Validation/}
+To assess the performance of the models, we used a number of statistical measures.
+First, we calculated \gls{RMSE}, \ac{MAE} and \ac{ME}, both separately per class, and also pooled overall.
+These overall measures were calculated by taking the mean of all class points pooled together, rather than taking a mean of the per-class means.
 
-To assess the performance of the models, we used a number of statistical measures:
+Next, the \gls{NSE} \citep{nash1970river} was used to determine the goodness of fit.
+For regression output, \ac{NSE} is a method of calculating the coefficient of determination $R^2$, and is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
 
-\begin{enumerate}
- \item \Gls{RMSE}, \ac{MAE}, \ac{ME}, overall and per class.
-The overall measures are calculated by taking the mean of all class points pooled together, rather than taking a mean of the per-class means.
- \item \Gls{NSE} \citep{nash1970river} was used to determine the goodness of fit.
-For regression output, \ac{NSE} is a method of calculating the coefficient of determination $R^2$, and is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
- \item \Gls{SCM} \citep{silvan-cardenas_sub-pixel_2008}, and the metrics derived from it: \ac{OA}, as well as \ac{PA} and \ac{UA} per class.
+Furthermore, we calculated a \gls{SCM} \citep{silvan-cardenas_sub-pixel_2008}, and the metrics derived from it: \ac{OA}, as well as \ac{PA} and \ac{UA} per class.
 The \gls{SCM} is an adaptation of the confusion matrix concept to fractional data.
-When using the MIN-PROD composite operator as suggested by \citet{silvan-cardenas_sub-pixel_2008}, the diagonal of the matrix expresses the maximum overlap (agreement) of the target and predicted class fractions, and the off-diagonal is an expression of which classes the non-overlapping fractions should belong to by calculating the expected value of overlap (product of reference and predicted class fraction).
+We used the MIN-PROD composite operator as recommended by \citet{silvan-cardenas_sub-pixel_2008}.
+When using this operator, the diagonal of the matrix expresses the maximum overlap (agreement) of the target and predicted class fractions, and the off-diagonal is an expression of which classes the non-overlapping fractions should belong to by calculating the expected value of overlap (product of reference and predicted class fraction).
 For instance, a pixel of 60\% grass and 40\% shrub, when predicted as 40\% grass and 60\% shrub, would have $min(60\%, 40\%)=40\%$ in the diagonals and $\frac{20\%\times20\%}{20\%}=20\%$ in the off-diagonals.
 For cases where the allocation of the overestimated and underestimated parts of the fraction do not have a unique solution, the \ac{SCM} indicates the expected value and an associated uncertainty measure.
- \item Correlation matrix of predicted vs observed class fractions.
- \item Boxplots showing the distribution of predictions, in bins for each 10\% of the predictions, against the validation data distribution, and a comparison with the 1:1 line.
-\item Lastly, spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point, rasterised into a map.
-\end{enumerate}
+
+Moreover, we produced boxplots showing the distribution of predictions, in bins for each 10\% of the predictions, against the validation data distribution, and a comparison with the 1:1 line.
+Lastly, we produced spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point.
+We rasterised the result into a map for ease of view, taking the mean value where several points fell into the same raster cell.
 
@@
 Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
-\Gls{RF} regression with no adjustments (single model) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
-\gls{NSE} results follow the \gls{RMSE} results, and the \gls{OA} from the subpixel confusion-uncertainty matrix follow the \gls{MAE} results.
+\Gls{RF} regression with no adjustments (single-step) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
+\gls{NSE} results follow the \gls{RMSE} results, and the \glsentryfull{OA} from the \glsentryfull{SCM} follow the \gls{MAE} results.
 This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
@@
 
+The models that performed the best were the two tree-based machine learning models: \gls{RF} and Cubist.
+In the case of these models, predictions per class were made using the binary relevance method (one model per class), as they are univariate.
+This shows that despite the binary relevance limitations, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.
+
+The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
+The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
+This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
+It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.
+
+\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
+This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
+This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
+
+\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods.
+Nevertheless, it performed better than all of the multivariate linear models.
+
 \begin{table}
 \centering
+%\adjustbox{max width=0.5\textwidth}{
 \begin{tabular}{llllll}
@@
 & 21.6  & 12.7  & 0.48  & 56±4  & 0.43±0.05 \\
-%Lasso regression
-%& 21.4  & 12.6  & 0.49  & 56±4  & 0.43±0.05 \\
 \Gls{MLR}
 & 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
-%Multivariate RF
-%& 18.7  & 10.0  & 0.61  & 65±3  & 0.55±0.05 \\
 \Glspl{NN}
@@
 & \textbf{17.3}  & 9.4   & \textbf{0.66}  & 67±4  & 0.57±0.05 \\
-%\ensuremath{''} histogram matched
-%& 22.0  & 9.7   & 0.43  & 66±2  & 0.58±0.03 \\
-{''} RS-only
+\ensuremath{''}
+only RS covariates
 & 18.4  & 10.3  & 0.62  & 64±4  & 0.54±0.05 \\
@@
 & 20.2  & \textbf{7.9}   & 0.54  & \textbf{72±2}  & \textbf{0.64±0.02} \\
-%\ensuremath{''} \ensuremath{''} histogram matched
-%& 21.3  & 9.1   & 0.46  & 68±3  & 0.61      \\
 \bottomrule
 \end{tabular}
-\caption{Accuracy statistics of the models tested. Best performing statistics are highlighted.}
+%}
+\caption{Accuracy statistics of the models tested. Best performing statistics are highlighted. ``Only RS covariates'' stands for a model trained only with \glspl{VI} and temporal metrics.}
 \label{tab-accuracy}
@@
 
-The models that performed the best (\gls{RF} and Cubist) are related models based on \gls{CART} ensembles.
-Predictions per class were made using the binary relevance method (one model per class), as these models are univariate.
-This shows that despite the binary relevance limitations of each model only being aware of its relevant class and not the others, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.
+There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
+Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
+Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely incorrect.
+Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
 
-Multivariate linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
-The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
-This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
-It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.
-
-\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
-This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
-This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
-
-\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods, but better than all of the multivariate linear models.
-
-There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large misclassifications or not.
-Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
-Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk misclassification, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely wrong.
-
-All tested models had difficulty discerning the vegetation classes (grassland, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (water, urban, bare soil), as shown in figure \ref{fig-models}.
+All tested models had difficulty discerning the vegetation classes (herbaceous, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
 This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
-In contrast, the tree fraction has a lot of samples with 100\% fraction and it deviates a lot per each sample, making it a highly unpredictable class from the data alone.
+In contrast, the tree class had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
 Most models showed good improvement in predicting tree cover compared to the intercept model, indicating that the covariates used were useful in distinguishing tree cover from other classes and quantifying its proportion in the pixel.
-In contrast, there were only moderate improvements for predicting the grass fraction, as it is difficult for the classifiers to distinguish this class with the given input covariates.
+In contrast, there were only moderate improvements for predicting the herbaceous vegetation fraction, as it is more difficult for the classifiers to distinguish this class with the given input covariates.
 
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{article/article-figures/barplots/2020-06-04-rf-comparison-bar}
-    \caption{Comparison of \gls{RF} regression models (intercept model shown as a reference).
-1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
-Mean and median are the tree vote summary statistics.}
-    \label{fig-randomforest}
-\end{figure}
-
-For a \gls{RF} regression when the mean of all tree votes was used as the indicator for the land cover fraction, using a two-step approach improved the \gls{MAE} of the model, while increasing \gls{RMSE} (see figure \ref{fig-randomforest}).
+For an \gls{RF} regression that uses the mean statistic for ensembling the tree votes, using a two-step approach improved the \gls{MAE} of the model, while increasing \gls{RMSE} (see figure \ref{fig-randomforest}).
 This is because the false positives in the first step of the two-step model (the binary zero/non-zero classification) result in some high fractions nevertheless being set to zero.
@@
 
-Another option explored for dealing with the issue of zero inflation was to use the median of all tree votes in \gls{RF} instead of the mean.
+Another option explored for dealing with the issue of zero inflation was to use the median statistic for ensembling tree votes in \gls{RF}, instead of the mean.
 This had a comparable effect to using the multi-step approach, as it resulted in a lot more precise predictions for fractions with 0\%.
-However, in the cases with high uncertainty, the median vote was much more likely to predict a 100\% fraction, and thus was also more likely to misclassify a pixel completely.
+However, in the cases with high uncertainty, the median vote was much more likely to predict a 100\% fraction, and thus was also more likely to predict 100\% of the wrong class.
 Thus it resulted in much higher \gls{RMSE} and much lower \gls{MAE}.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{article/article-figures/barplots/2020-06-04-rf-comparison-bar}
+    \caption{Comparison of \gls{RF} regression models (intercept model shown as a reference).
+1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
+Mean and median are the tree vote summary statistics.}
+    \label{fig-randomforest}
+\end{figure}
+
 \subsection{Covariate importance}
@@
 \Gls{RF} permutation importance was performed to test which variables were important for \gls{RF} regression predictions.
-The single-step \gls{RF} model was selected for this as it achieved the lowest \gls{RMSE}.
+The single-step \gls{RF} model was selected for this, as it achieved the lowest \gls{RMSE}.
 The results are shown in figure \ref{fig-varimp}.
@@
     \includegraphics[width=14cm]{article-figures/heatmaps/2020-06-19-varimp-heatmap-top5}
-    \caption{Random Forest single model variable importance, top 5 covariates per category.
+    \caption{Random Forest single-step approach variable importance, top 5 covariates per category.
 Categories are ordered by cumulative importance.
-The values represent decrease in RMSE when the covariate is permuted.}
+The values represent increase in RMSE for the given class when the given covariate is permuted.}
     \label{fig-varimp}
@@
 
-The overall most important variables were the maximum and minimum NDVI over the whole time series of PROBA-V imagery.
-They were followed by the median NDMI and NIRv over the whole time series.
+The overall most important variables were the maximum and minimum \gls{NDVI} over the whole time series of PROBA-V imagery.
+They were followed by the median \gls{NDMI} and \gls{NIRv} over the whole time series.
 These vegetation indices were of lower importance only for the shrub and crop classes.
@@
 Climate data was also most useful for predicting crops, especially the mean temperature and the closely (exponentially) related mean of water vapour pressure.
-All of the natural precipitation class predictions benefited from (log-transformed) annual precipitation, solar irradiance in January and isothermality.
+All of the natural vegetation class predictions benefited from (log-transformed) annual precipitation, solar irradiance in January and isothermality information.
 
@@
 These classes tend to be spatially clustered into large areas.
-In contract, location information was not useful for inland water prediction as it is a rare class that takes little area.
 Both absolute and regular latitude was useful for the prediction, and latitude was more useful than longitude for the predictions.
+In contrast, location information was not useful for inland water prediction, as it is a rare and spatially dispersed class that takes little area.
 
@@
 We used the single-step \gls{RF} model to predict the land cover at a global scale to obtain wall-to-wall global land cover fraction maps.
-See figure \ref{fig-walltowall} for a visualisation (sampled at 0.2 degrees) of all the fraction layers combined.
+See figure \ref{fig-walltowall} for a visualisation (sampled at 0.2 degrees) of all of the fraction layers separately, and the supplementary materials for the full resolution raster itself.
 The wall-to-wall fraction maps reveal how land cover fraction mapping is capable of expressing gradients and mixed land cover.
-
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=\textwidth]{article-figures/maps/2020-05-25-walltowall}
-%     \caption{Wall-to-wall predictions of all seven land cover class fractions using the single-step \gls{RF} regression model.}
-%     \label{fig-walltowall}
-% \end{figure}
 
@@
 Likewise, the gradient of tree cover from 100\% in the African tropics to 0\% in the sub-Sahara region is evident.
-Herbaceous cover in sub-Sahara shows an asymmetric gradient: the cover is highest at the 14th-15th latitude and decreases quickly to the north, up to the 18th latitude, but slowly to the south, until the 5th latitude.
+Herbaceous cover in sub-Sahara shows an asymmetric gradient: the cover is highest at around 14-15\textdegree{} N and decreases quickly towards the north, becoming zero around 18\textdegree{} N; but decreases slowly towards the south, reaching all the way to 5\textdegree{} N.
 Inland water shows up as more discrete, as it naturally forms discrete patches and mixed pixels with water are more rare.
-Built-up area is relatively rare and is rarely forms a 100\% fraction, as urban areas are in most places interspersed with greenery within the footprint of the pixel.
+Built-up area is relatively rare and rarely forms a 100\% fraction, as urban areas tend to include both built-up area and greenery within the footprint of the pixel.
 
 The spatial pattern of map accuracy shows that the land cover of areas with pure land cover was the easiest to predict, such as tropical forests (100\% tree cover) and deserts (100\% bare land).
-Conversely, it was more difficult to predict fractions in areas with mixed land cover, as isolating individual fractions becomes more challenging.
+Conversely, it was more difficult to predict fractions in areas with mixed land cover, as isolating individual fractions from pixel-level information is more challenging.
 In addition, land cover fractions in the extreme latitudes were more difficult to predict as well.
-There less training data was available, owing to the lack of high-resolution imagery for image interpretation in these areas.
+In these areas less training data was available, owing to the lack of high-resolution imagery there.
 
@@
 The predictions of each class showed underestimation of high fractions and overestimation of low fractions, i.e. tending towards the mean.
+As we can see from the box widths of the boxplots in figure \ref{fig-walltowall}, the number of predictions were much more even across the entire range for herbaceous vegetation compared to shrubs.
+Shrub fractions were never predicted above 85\%, and the low fraction predictions dominated, showing that the model expected that shrubs would be mixed with other classes and form small fractions only.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{article/article-figures/maps/2020-06-19-walltowall.png}
+    \includegraphics[width=\textwidth]{article/article-figures/maps/2020-07-29-walltowall.png}
     \caption{Random Forest single model predictions per class.
-Left column: predicted fractions.
-Middle column: absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.
-Right column: distribution of predicted versus true values, shown as box plots with bins each 10\%.}
+Top row: predicted fractions, each given its own colour (``Overall'' shows a hardened map of dominant land cover using the same colour scheme).
+Middle row: absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.
+Bottom row: distribution of predicted versus true values, shown as box plots with bins each 10\%, widths representing sample size.}
     \label{fig-walltowall}
@@
 
-% \begin{enumerate}
-%  \item Fraction mapping methods (Methods):
-%  \begin{enumerate}
-%     \item RF is the best, even though one model per class means that it does not take everything into account
-%     \item Fraction area accuracy goes up to 72\%, that is good considering that hard classification is not much better than that
-%     \item Hardest to classify are grasslands and especially shrubs and urban
-%     \item Accuracy assessment metric makes a big difference, useful to use subpixel confusion matrix for that
-%  \end{enumerate}
-%  \item Optimising for zero inflation (Challenges):
-%  \begin{enumerate}
-%     \item Multi-step models improve MAE at the cost of RMSE and NSE
-%     \item Histogram matching doesn't help any
-%  \end{enumerate}
-%  \item Covariate importance (Variables):
-%  \begin{enumerate}
-%     \item All covariates are important to a certain degree (show table of importances)
-%  \end{enumerate}
-% \end{enumerate}
+\subsection{Methods and their accuracies}
 
-\subsection{Methods}
-
-In this study, we have tested a number of machine learning methods for comprehensive land cover class fraction prediction at the global scale.
+In this study, we have tested a number of machine learning methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale.
 Since global land cover fraction mapping is a novel field, there are few other studies or products that our results can be directly compared with.
-Most products in this field are made available as layers for a particular land cover class fraction, rather than being comprehensive and guaranteeing a sum to 100\%.
+Most products in this field are made available as layers for a particular land cover class fraction, rather than being thematically comprehensive and guaranteeing a sum of all fractions to 100\%.
 In addition, the documentation for these products rarely report accuracy metrics, and none use an independent validation set for this purpose.
@@
 
-%The three-step Median \gls{RF} model achieved an overall accuracy of 72±2\%, which is high considering that discrete global land cover maps typically achieve around 70\% accuracy \citep{herold_challenges_2008}.
-%In contrast, land cover fraction mapping is expected to be more challenging than discrete classification.
-%This is because land cover fractions contain the information for all of the land cover classes considered in each pixel, and this information needs to be retrieved from discrete pixels, with no prior information about class allocation within the pixel.
-
-Compared to a global land cover product that specialises in a specific land cover class, namely tree cover, the \gls{RF} one-step mean model had a higher RMSE (19.2\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (16.8\%).
+Compared to a global land cover product that specialises in a specific land cover class, namely tree cover, the \gls{RF} one-step mean model had a higher \gls{RMSE} (19.2\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (16.8\%).
 However, the methodology for the validation differed significantly from the one used in our study, as it compared the predicted fractions to fractions as computed from lidar datasets within several local study areas, rather than using global expert image interpretation as we did.
-\citet{feng_earth_2016} assessed a discrete (binary) GLCF forest cover product for three different years using a more similar methodology, and yielded similar results to ours: a user's accuracy of 98\% and producer's accuracy of 80\%, compared to the \gls{RF} three-step median tree cover user's accuracy of 81\% and producer's accuracy of 86\%.
+A previous study on the MODIS vegetation continuous fields product by \citet{montesano_modis_2009} used a more similar methodology as we did, i.e.
+expert interpreters estimating cover fractions from a 100 by 100 point subdivisions of pixels using high-resolution imagery, and reported an R² of 0.57, \gls{RMSE} of 13.4, RMSD of 21.3, slope from a linear regression of 0.5 and intercept of 18.4.
+In comparison, our results of a single-step \gls{RF} regression for the tree cover class was 0.77 for both \gls{NSE} and linear regression R², \gls{RMSE} of 19.3, slope of 0.7 and intercept of 8.8.
+Thus while our model appears to perform better, it is important to note that the scopes of the studies were different, as \citet{montesano_modis_2009} only evaluated boreal regions rather than the entire globe, and used MODIS 500 m data rather than PROBA-V 100 m data.
 
 Our results showed that \gls{RF} regression using binary relevance performed the best of all the tested algorithms.
-It was also adaptive to two different goals: minimising \gls{RMSE} (default) and minimising \gls{MAE} (with median voting).
-Both of these goals may be valuable to different user communities.%, depending on whether areas with high uncertainty in the predicted land cover should be treated as a mix of covers, or a guess should be made at a single most likely cover.
+It was also adaptive to two different goals: minimising \gls{RMSE} (with the default, mean voting) and minimising \gls{MAE} (with median voting).
+Both of these goals may be valuable to different user communities, depending on whether areas with high uncertainty in the predicted land cover should be treated as a mix of covers, or a guess should be made at a single most likely cover.
+The former is more likely to be useful for the modelling community that deals with uncertainty with probabilistic frameworks, and the latter may be more useful for policymakers and land owners who are more concerned with what land cover is most likely to be on the ground.
+
+Our results also showed that the three-step approach introduced in this study, when combined with median voting, resulted in a decrease in \gls{RMSE} without affecting \gls{MAE}.
+Thus, it achieved the best combination that is optimised towards reducing \gls{MAE}.
+The three-step approach can also be used with any algorithm that provides both classification and regression modes, or with two separate unrelated classification and regression algorithms.
 
@@
 Our respective results were 2.57\% \gls{MAE} and 11.51\% RMSE.
-We have confirmed that Cubist works slightly better than \gls{RF} regression for this class (\gls{RF} regression result is 11.65\% \gls{RMSE}, 3.48\% \gls{MAE}), but overall for all classes \gls{RF} regression nevertheless edges out (Table \ref{tab-accuracy}).
-One important difference between the studies is that we focused on comprehensive land cover, rather than a particular class, which has an effect on the model training, and our focus was on a global rather than regional scale, has an effect on the resulting statistics due to higher sample site diversity.
+Our findings confirm their conclusion that Cubist works slightly better than \gls{RF} regression for this class (\gls{RF} regression result is 11.65\% \gls{RMSE}, 3.48\% \gls{MAE}), but overall when considering all of the classes, \gls{RF} regression nevertheless edges out (see table \ref{tab-accuracy}).
+One important difference between the studies is that we focused on thematically comprehensive land cover, rather than a particular class, which has an effect on the training data balance.
+Also, our focus was on a global rather than regional scale, which has an effect on the resulting statistics due to higher sample site diversity.
 
-Similarly, \citet{walton2008subpixelrf} assessed the accuracy of urban land cover mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which was also observed as performing better than \gls{RF} regression and \glspl{SVM}.
+At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which was also observed as performing better than \gls{RF} regression and \glspl{SVM}.
 Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the urban class, and we also observed that Cubist performed slightly better than the other two algorithms for this class.
-\citet{okujeni_comparison_2014} also performed land cover fraction mapping over an urban area, but their results indicated that \gls{SVM} regression performed better than \gls{RF}.
+Both \citet{okujeni_comparison_2014} and \citet{foody1997fuzzynnet} also performed land cover fraction mapping over an urban area.
+The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF}.
 This could be due to much finer resolution (3.6–9 m) used in their study, and the associated difference in class definitions.
+\citet{foody1997fuzzynnet} reported an \gls{RMSE} of 16.6\% using \gls{MLP} \glspl{NN}, lower than our results (22.7\%).
 The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, namely their focus on fine scale predictions over a local area for the urban class in particular, in contrast to our global focus with comprehensive classes, leading to a different balance of the datasets between the studies.
-The other studies focused on a single particular class, whereas both urban and water were relatively rare classes in our global study, and therefore the dataset had much more zero fractions in comparison, which on average is easier to predict.
 
-\subsection{Variables}
+\subsection{Variable importance}
 
@@
 Nevertheless, even though a lot of the other covariates were of much lower importance, they  contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of covariates that were correlated over 0.9 as explained in section \ref{sec-covariate-selection}).
+Methods based on decision trees help with effectively using covariates that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any covariates as non-informative.
 This is also due to the large variety of land cover classes in the study, since a covariate is useful if it helps predict any of the land cover classes better.
-The result that including more covariates is beneficial, but remote sensing data is the most important, is in line with the conclusions of \citet{li_monitoring_2018} and \citet{hengl_soilgrids250m_2017}.
-The latter study also shows high importance of climate data, however, it is focused on soils, whereas climate has more effect on soils than on land cover.
-Methods based on decision trees help with effectively using covariates that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any covariates as non-informative.
 
-\subsection{Challenges}
+The result that including more covariates is beneficial, but remote sensing data is the most important, is in line with the conclusions of e.g.
+\citet{li_monitoring_2018} and \citet{hengl_soilgrids250m_2017}.
+Remote sensing data is also unique in that it forms a time series, which both allows us to extract more information from it, such as the temporal metrics used in this study, but also to monitor land cover change over time.
+In addition to remote sensing data, \citet{hengl_soilgrids250m_2017} also noted high importance of climate data, however, it is focused on soils, whereas climate has more effect on long-term processes such as soil formation than on land cover.
+
+\subsection{Challenges and future outlook}
 
@@
 On the other hand, it comes at a cost of higher \gls{RMSE} and lower \gls{NSE}.
-This is because in highly uncertain cases, the model does a best guess at a pure class, and often it is incorrect, which leads to 100\% error in those cases, which is highly penalised by \gls{RMSE}.
+This is because in highly uncertain cases, the model does a best guess at a pure class, and due to the high uncertainty often the guess is incorrect.
+This leads to 100\% error in those cases, which is highly penalised by \gls{RMSE}.
 Due to this effect, the resulting map is closer to a discrete classification map, with less expressive transition gradients between land cover classes.
@@
 A similar effect is seen when using techniques such as median vote for \gls{RF} regression.
-This allows for predicting zero fractions correctly more often, but also misclassifies completely more often.
+This allows for predicting zero fractions correctly more often, but likewise more often causes a prediction of 100\% of the wrong class.
 Our results showed that median vote has a stronger effect for reducing \gls{MAE} than the multi-step approach.
@@
 These findings show that the use of a multi-step approach depends on what is more important for the user.
-If an occasional complete misclassification is acceptable, then a multi-step model provides an overall more accurate result, especially for zero fractions.
-On the other hand, a single-step approach emphasises the strength of land cover fraction mapping in being able to express gradual changes over space better, and avoids large misclassifications.
+If an occasional prediction of 100\% of the wrong class is acceptable, then a multi-step model provides an overall more accurate result, especially for zero fractions.
+On the other hand, a single-step approach emphasises the strength of land cover fraction mapping in being able to express gradual changes over space better, and avoids large errors.
 We expect a multi-step approach to be more suitable for fine resolution mapping, where more pure pixels can be expected, and single-step approach to be more useful for coarse resolution mapping where mixed pixels are the norm.
@@
 This is due to a guess of the mean being less penalising than predicting the extremes incorrectly; e.g.
-for a case of 50\% shrub cover, predicting 100\% shrubs would be a larger mistake (and thus lead to higher RMSE) than predicting 15\% shrubs.
+for a case of 50\% shrub cover, predicting 100\% shrubs would be a larger mistake (and thus lead to higher \gls{RMSE}) than predicting 15\% shrubs.
 Likewise, predicting 0\% built-up in dark bare soils risks a case where it truly would be 100\% built-up, so on average predicting 15\% built-up in this case lowers the possible error.
+The three-step approach tilts the balance in the other direction, as the first step determines that the pixel is pure, and the classification step determines that it is more likely to be bare soil than built-up area, but it may also misclassify in other cases.
+If not pure, the regression step still tends to predict towards the mean due to the loss function, so the challenge of underestimating large fractions and overestimating small ones remains.
 
-There is an inherent challenge in land cover mapping for discerning classes that are related, e.g. grass and shrubs.
-Grass was particularly difficult to map, in part due to confusion between grass, crops and shrubs.
+There is an inherent challenge in land cover mapping for discerning classes that are related, e.g. herbaceous vegetation and shrubs.
+Herbaceous vegetation was particularly difficult to map, in part due to confusion between herbaceous vegetation, cropland and shrubs.
 These classes are similar not just to regression algorithms, but also to expert interpreters, which may lead to higher uncertainties also in the training and validation data for these classes.
+The difference between the definition shrubs and herbaceous vegetation largely comes down to height, therefore dynamic information about vegetation height would allow mapping these classes more accurately.
+However, this would require either photogrammetry techniques (that typically rely on much finer spatial resolution and more overlap between the scenes) to reconstruct vegetation height, or the use of non-optical sensors such as \gls{SAR} interferometry.
+Another way to differentiate between the vegetation classes could be to make use of hyperspectral data, which allows differentiation between different kinds of vegetation based on their photosynthesis processes, which affect light absorption.
+Emerging new high level hyperspectral products, such as \gls{SIF} or \gls{GPP}, such as ones based on the upcoming FLEX satellite, could allow for an easy way to incorporate this extra information into land cover models.
+
 Another challenge is class imbalance.
-For example, the urban class is rare, it hardly ever forms a 100\% fraction and its extent small.
-That makes it easy to classify, as a fraction of 0\% is in most cases a good guess, but it that makes it not useful for user needs.
-This challenge is further exacerbated by the training dataset containing relatively few points in urban areas to begin with.
+For example, the built-up class rarely forms a 100\% fraction.
+That makes it easy to classify, as a fraction of 0\% is in most cases a good guess, but that makes it not useful for user needs.
+This challenge is further exacerbated by the training dataset containing relatively few points in built-up areas to begin with.
 As such, having a more balanced training dataset may lead to a better representation of the accuracy of each class and help with class predictions.
@@
 As such, it is possible to miss a particular parameter combination that may further improve the results.
-On the other hand, our results showed that the differences between the methods are not very high.
+On the other hand, our results showed that the differences between the best performing methods are not very high.
 The differences in model parametrisations are expected to be even lower.
@@
 
-Several challenges are yet to be tackled in this field.
-Finer spatial resolution mapping, such as 10 m mapping using Sentinel-2 data, is a future research direction, where pixels will be even more likely to be covered by a single class.
+Several more challenges are yet to be tackled in this field, but doing so is becoming more and more feasible over time.
+Finer spatial resolution mapping, such as 10 m mapping using Sentinel-2 data, is a future research direction, where the pixel footprint will be even more likely to be covered by 100\% of a single class.
 Such future developments would be more likely to benefit from a multi-model approach or optimisation for \gls{MAE}.
 Alternatively, finer spatial resolution sensors would allow mapping fractions at coarser resolutions in a more precise way, as it would be possible for the model to directly make use of data from the sensor about subpixels of the mapped pixels.
-Another future work direction is land cover change mapping over time.
-Land cover fraction information would allow tracking gradual changes, such as regrowth, over time better.
-The challenge here is that the higher uncertainty about fraction estimates may cause the time series of land cover fractions to fluctuate, making it difficult to determine trends.
+
+Another benefit of using different optical sensors is higher spectral resolution.
+As PROBA-V only measures four spectral bands, the amount of information that can be retrieved from them is limited.
+Sensors such as Sentinel-2 \gls{MSI} would allow for computing a more diverse range of \glspl{VI}, such as ones based on the slope of the red edge, potentially allowing for a better distinction between different vegetation types.
+
+With more availability of such additional remote sensing data, it becomes increasingly more feasible to perform land cover change mapping, to make it possible to monitor land cover changes over time.
+Land cover fraction information would allow better tracking of gradual changes, such as regrowth, over time.
+The challenge here is that the higher uncertainty about fraction estimates may cause the time series of land cover fractions to fluctuate, making it difficult to determine robust trends.
 
@@
 
-% \begin{enumerate}
-%  \item Random forest is the best
-%  \item All covariates are important but to a different degree, remote sensing parameters are most important
-%  \item Using a multi-model approach improves MAE at the cost of RMSE, thus a trade-off
-% \end{enumerate}
-
-In this study, we compared three groups of algorithms for land cover fraction mapping and created a demonstation map showing the result of the best performing model by \gls{RMSE}.
-We also introduced a multi-step approach to tackle the issue of correctly predicting the extreme fraction values of 0\% and 100\%.
+In this study, we compared three groups of algorithms for land cover fraction mapping and created a demonstation map showing the global distribution of land cover in fraction layers, using the best performing model according to \gls{RMSE}.
+We also introduced a multi-step approach to tackle the issue of accurately predicting the extreme fraction values of 0\% and 100\%.
 The best performing models were based on \gls{CART} tree ensembles, with \gls{RF} regression performing the best overall.
@@
 Remote sensing covariates were the most important in the models, although all other types of covariates (climate, soil, terrain) also contributed to the model significantly.
-This demonstrates the possibility of exhaustive global land cover fraction mapping using currently available methods, offering better precision than discrete land cover maps.
-It also allows the users to define their own thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
-This work paves the way for operationalisation of global land cover fraction mapping.
+
+These findings directly contribute towards the operationalisation of global land cover fraction mapping by analysing thematically exhaustive global land cover fraction mapping using currently available methods.
+Information on land cover fractions offers better precision than discrete land cover maps, and allows the users to define their own thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
+Furthermore, given the recent advances in optical sensor spatial resolution and the resulting increase in pixel purity, the multi-step model approach may become more important in the future.
+Lastly, due to the advances in spatial and spectral resolution, longer imagery time series of existing sensors, and the availability of non-optical sensor data, this work paves the way towards operational land cover fraction change mapping, which would allow monitoring gradual land cover change.
 
@@
 
-% \begin{itemize}
-%     \item Conceptualisation: MH NT JV
-%     \item Data curation: DM
-%     \item Formal Analysis: DM
-%     \item Funding acquisition: JV MH BS
-%     \item Investigation: DM ML NT
-%     \item Methodology: DM MH JV NT BK LL
-%     \item Project administration: MH NT JV
-%     \item Resources: MH BS
-%     \item Software: DM
-%     \item Supervision: MH NT JV
-%     \item Validation: NT DM
-%     \item Visualisation: DM MH
-%     \item Writing – original draft: DM
-%     \item Writing – review \& editing: NT JV MH MB BS
-% \end{itemize}
-
 \minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {PROBA-V} data, and IIASA for providing the model training dataset.
-In addition, we thank Benjamin Kellenberger for help on tuning neural network parameterisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
+In addition, we thank Bruno Smets for ideas and support from the \gls{CGLS-LC100} project, Benjamin Kellenberger for help on tuning \gls{NN} parameterisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
 
@@
 
-%\minisection{Appendix 1} % Not sure how to typeset this, as it does not fit under the section header here.
-\begin{table}
-\begin{center}
-\begin{adjustbox}{max width=8cm}
-\begin{tabular}{m{12cm}l}
-\toprule
-\multicolumn{1}{l}{Category \& data source}&\multicolumn{1}{c}{Covariate}\tabularnewline
-\midrule
-Location (3, intrinsic)&
-\begin{tabular}{@{}l@{}}longitude\tabularnewline
-latitude\tabularnewline
-absolute latitude
-\end{tabular}\tabularnewline
-\midrule
-Vegetation indices (22, derived from PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
-\begin{tabular}{@{}l@{}}
-minimum NDVI\tabularnewline
-maximum NDVI\tabularnewline
-median NDMI\tabularnewline
-NDMI yearly IQR\tabularnewline
-NDMI March-May IQR\tabularnewline
-NDMI June-August IQR\tabularnewline
-NDMI September-November IQR\tabularnewline
-NDMI December-February IQR\tabularnewline
-OSAVI March-May IQR\tabularnewline
-OSAVI June-August IQR\tabularnewline
-OSAVI September-November IQR\tabularnewline
-OSAVI December-February IQR\tabularnewline
-EVI March-May IQR\tabularnewline
-EVI June-August IQR\tabularnewline
-EVI September-November IQR\tabularnewline
-EVI December-February IQR\tabularnewline
-median NIRv\tabularnewline
-NIRv yearly IQR\tabularnewline
-NIRv March-May IQR\tabularnewline
-NIRv June-August IQR\tabularnewline
-NIRv September-November IQR\tabularnewline
-NIRv December-February IQR
-\end{tabular}\tabularnewline
-\midrule
-Temporal metrics (9, derived from a harmonic model over time series of PROBA-V 100m \ac{TOC} reflectance v1.02 \citep{probavguide2})&
-\begin{tabular}{@{}l@{}}
-NDVI order 1 cosine\tabularnewline
-NDVI order 1 sine\tabularnewline
-NDVI order 2 cosine\tabularnewline
-NDVI order 2 sine\tabularnewline
-NDVI trend coefficient\tabularnewline
-NDVI order 1 phase\tabularnewline
-NDVI order 1 amplitude\tabularnewline
-NDVI order 2 phase\tabularnewline
-NDVI order 2 amplitude
-\end{tabular}\tabularnewline
-\midrule
-Terrain (4, ASTER GDEM V003 \citep{ASTGTM003})&
-\begin{tabular}{@{}l@{}}
-elevation\tabularnewline
-slope (log-transformed)\tabularnewline
-aspect\tabularnewline
-terrain position index
-\end{tabular}\tabularnewline
-\midrule
-Climate (21, WorldClim 2.0 \citep{worldclim2})&
-\begin{tabular}{@{}l@{}}
-January precipitation (log)\tabularnewline
-April precipitation (log)\tabularnewline
-July precipitation (log)\tabularnewline
-October precipitation (log)\tabularnewline
-January solar irradiance\tabularnewline
-July solar irradiance\tabularnewline
-mean temperature\tabularnewline
-temperature monthly range\tabularnewline
-isothermality\tabularnewline
-temperature annual range\tabularnewline
-annual precipitation (log)\tabularnewline
-temperature seasonality\tabularnewline
-minimum solar irradiance\tabularnewline
-maximum solar irradiance\tabularnewline
-mean solar irradiance\tabularnewline
-mean windspeed\tabularnewline
-mean water vapour pressure\tabularnewline
-coldest month precipitation (log)\tabularnewline
-warmest month precipitation (log)\tabularnewline
-wettest month solar irradiance\tabularnewline
-driest month solar irradiance
-\end{tabular}\tabularnewline
-\midrule
-Soil (8, SoilGrids \citep{hengl_soilgrids250m_2017})&
-\begin{tabular}{@{}l@{}}
-soil available water\tabularnewline
-soil bulk density\tabularnewline
-soil cation exchange capacity (log)\tabularnewline
-soil clay fraction\tabularnewline
-soil coarse fragments (log)\tabularnewline
-soil pH\tabularnewline
-soil sand fraction\tabularnewline
-soil water wilting point
-\end{tabular}\tabularnewline
-\bottomrule
-\end{tabular}
-\end{adjustbox}
-\caption{List of covariates input into the models, and their data sources.}
-\label{tab-inputdata}
-\end{center}
-\end{table}
+% NOTE: Appendix is in supplement.tex
 

### 2020-08-05 article/highlights.tex
Draft 3

\documentclass[review,authoryear,3p]{elsarticle}
\journal{Remote Sensing of Environment}
\begin{document}
\section*{Highlights}
\begin{itemize}
    \item 3 machine learning model groups tested for deriving 7 land cover fractions globally
    \item Random forest regression performed the best
    \item Optical remote sensing data was the most important input for model predictions
    \item Independent validation showed highest accuracies for built-up and water fractions
    \item Multi-fraction land cover mapping shown as an alternative to discrete classification
\end{itemize}

\end{document}

### 2020-08-05 article/article.tex
Final changes

@@
 \newacronym{CGLS-LC100}{CGLS-LC100}{Copernicus Global Land Services 100 m Land Cover}
-%\newacronym{CGLS}{CGLS}{Copernicus Global Land Services}
 \newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
@@
 %\makenoidxglossaries
-% Disable abbreviating RF for now [text={random forest}]
-%\glsunset{RF}
-
-% Use \begin{equation} for maths
 
@@
 \begin{keyword}
-Land cover fraction mapping\sep machine learning\sep random forest\sep Cubist\sep neural networks\sep fuzzy nearest centroid\sep lasso regression\sep PROBA-V
+PROBA-V\sep global land cover mapping\sep land cover fraction mapping\sep time series analysis\sep machine learning\sep random forest\sep Cubist\sep support vector regression\sep neural network\sep spatial accuracy
 \end{keyword}
@@
 It includes monthly temperature, precipitation, solar radiation, wind speed and water vapour pressure data.
-In addition, 19 bioclimatic parameters were calculated from the data, using the dismo package \citep{dismo}.
+In addition, 19 bioclimatic parameters were calculated from the data, using the \cran{dismo} package \citep{dismo}.
 Some additional biophysical parameters were also calculated, namely all of the climate data during the coldest, warmest, driest and wettest months of the year at each location, as well as yearly averages of each of the original parameters.
@@
 
-In total, 313 covariates were generated in pre-processing.
+In total, 313 covariates were generated in preprocessing.
 However, many of them were collinear with one another.
@@
 This method always produces output that sums to 100\%.
-In this study, we used an implementation adapted from the package GSIF \citep{hengl2004fuzzycmeans}.
+In this study, we used an implementation adapted from the package \cran{GSIF} \citep{hengl2004fuzzycmeans}.
 We tested both using a logistic regression to estimate the class centroids, as well as using a weighted average of the training data to do so, and ended up using the latter method as the result was more accurate.
@@
 
-In this study, we compared three groups of algorithms for land cover fraction mapping and created a demonstation map showing the global distribution of land cover in fraction layers, using the best performing model according to \gls{RMSE}.
+In this study, we compared three groups of algorithms for land cover fraction mapping and created a demonstration map showing the global distribution of land cover in fraction layers, using the best performing model according to \gls{RMSE}.
 We also introduced a multi-step approach to tackle the issue of accurately predicting the extreme fraction values of 0\% and 100\%.
@@
 % Converted from role-centric to person-centric as per Elsevier guidelines: https://www.elsevier.com/authors/journal-authors/policies-and-ethics/credit-author-statement
-\minisection{Author Contributions}
-\textbf{Dainius Masiliūnas}: Methodology, Formal Analysis, Software, Writing – Original Draft, Data Curation.
+%\minisection{Author Contributions}
+%\textbf{Dainius Masiliūnas}: Methodology, Formal Analysis, Software, Writing – Original Draft, Data Curation.
 \textbf{Nandin-Erdene Tsendbazar}: Investigation, Data Curation, Supervision, Conceptualisation, Writing – review \& editing.
@@
 \minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {PROBA-V} data, and IIASA for providing the model training dataset.
-In addition, we thank Bruno Smets for ideas and support from the \gls{CGLS-LC100} project, Benjamin Kellenberger for help on tuning \gls{NN} parameterisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
-
-%\minisection{Conflicts of Interest} The authors declare no conflict of interest.
+In addition, we thank Bruno Smets for ideas and support from the \gls{CGLS-LC100} project, Benjamin Kellenberger for help on tuning \gls{NN} parametrisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
 
@@
 
+%\listoffigures
+
 \end{document}

### 2020-08-05 article/credit.tex
Final changes

\documentclass[review,authoryear,3p]{elsarticle}
\journal{Remote Sensing of Environment}
\begin{document}

\section*{Author Contributions}
\textbf{Dainius Masiliūnas}: Methodology, Formal Analysis, Software, Writing – Original Draft, Data Curation. \textbf{Nandin-Erdene Tsendbazar}: Investigation, Data Curation, Supervision, Conceptualisation, Writing – review \& editing. \textbf{Martin Herold}: Conceptualisation, Project Administration, Supervision, Methodology. \textbf{Myroslava Lesiv}: Investigation, Data Curation. \textbf{Marcel Buchhorn}: Writing – review \& editing. \textbf{Jan Verbesselt}: Conceptualisation, Methodology, Supervision, Writing – review \& editing.

\end{document}

### 2020-08-05 article/article.tex
Fix file paths

@@
 \begin{figure}
- \includegraphics[width=\textwidth]{article/article-figures/maps/2020-07-06-training-and-validation}
+ \includegraphics[width=\textwidth]{article-figures/maps/2020-07-06-training-and-validation}
  \caption{Sample points representing 100x100 m areas at which land cover reference data was used in the study.
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{article/article-figures/barplots/2020-06-03-model-comparison-bar}
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
     \caption{Comparison of \gls{RMSE} (top) and \gls{MAE} (bottom) per class of the best performing models in their category, and the intercept-only model as a reference.}
@@
     \centering
-    \includegraphics[width=\textwidth]{article/article-figures/barplots/2020-06-04-rf-comparison-bar}
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-04-rf-comparison-bar}
     \caption{Comparison of \gls{RF} regression models (intercept model shown as a reference).
@@
     \centering
-    \includegraphics[width=\textwidth]{article/article-figures/maps/2020-07-29-walltowall.png}
+    \includegraphics[width=\textwidth]{article-figures/maps/2020-07-29-walltowall.png}
     \caption{Random Forest single model predictions per class.

### 2020-11-18 article/article.tex
Draft 2 of first review round

@@
 \usepackage{graphicx}
-%\usepackage{adjustbox} % For resizing tables
+\usepackage{adjustbox} % For resizing tables
 \usepackage{array} % For mid-alignment (m{xcm}) in tables
@@
 % Package names
-\newcommand{\cran}[1]{CRAN: #1}
+\newcommand{\cran}[1]{CRAN:~#1}
 
@@
 \newacronym{ME}{ME}{mean error}
-\newacronym{NSE}{NSE}{Nash–Sutcliffe model efficiency coefficient}
+\newacronym{RRMSE}{RRMSE}{relative root mean squared error}
+\newacronym{RMAE}{RMAE}{relative mean absolute error}
+\newacronym{RME}{RME}{relative mean error}
+\newacronym{NSE}{NSE}{the Nash–Sutcliffe model efficiency coefficient}
 \newacronym{SDG}{SDG}{Sustainable Development Goal}
@@
 \newacronym{FROM-GLC}{FROM-GLC10}{Finer Resolution Observation and Monitoring Global Land Cover}
-\newacronym{IIASA}{IIASA}{International Institute for Applied Systems Analysis}
+\newacronym{IIASA}{IIASA}{the International Institute for Applied Systems Analysis}
 \newacronym{WUR}{WUR}{Wageningen University \& Research}
@@
 \newacronym{LCCS}{LCCS}{UN Land Cover Classification Scheme}
-\newacronym{GDAL}{GDAL}{Geospatial Data Abstraction Library}
+\newacronym{GDAL}{GDAL}{the Geospatial Data Abstraction Library}
+\newacronym{GFCC}{GFCC}{Global Forest Cover Change}
+\newacronym{GHSL}{GHSL}{Global Human Settlement Layer}
+\newacronym{GSW}{GSW}{Global Surface Water}
 %\makenoidxglossaries
@@
 
-\title{Global land characterisation using cover fractions at 100m resolution}
+\title{Global land characterisation using land cover fractions at 100 m resolution}
 
@@
 Currently most global land cover maps are produced with discrete classes, which express the dominant land cover class in each pixel, or a combination of several classes at a predetermined ratio.
-In contrast, land cover fraction mapping allows for expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity.
+In contrast, land cover fraction mapping enables expressing the proportion of each pure class in each pixel, which increases precision and reduces legend complexity.
 To map land cover fractions, regression rather than classification algorithms are needed, and multiple approaches are available for this task.
 
-We compared the performance of three types of machine learning regression algorithms for land cover fraction mapping.
-For the best model, we assessed the importance of six types of training variables, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe at a sampling interval of 0.2 degrees.
-The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points of the \gls{CGLS-LC100} project.
-Both datasets are global and aligned with the PROBA-V 100 m grid.
+We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping.
+For the model with the lowest MAE, we assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
+The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points, provided by the \gls{CGLS-LC100} project.
+Both datasets are global and aligned with the PROBA-V 100 m UTM grid.
 
-One challenge for models for land cover fraction mapping is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% are difficult to predict well.
+A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% fractions are difficult to predict accurately.
 We proposed a new solution by combining three models.
@@
 Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%.
-Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved by a tuned random forest model with our proposed three-model approach.
-The models were driven primarily by optical remote sensing data, followed by soil and climate data.
-Herbaceous cover was the most challenging to discern for all models, whereas built-up and water were the easiest to estimate.
+Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest with our proposed three-model approach.
+This model was driven primarily by optical remote sensing data, followed by climate data.
 
-This research proves that machine learning algorithms can be applied globally to a wide variety of land cover classes to create an exhaustive global land cover fraction map.
-Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which helps to customise the result for a diverse range of uses.
-In addition, this work directly contributes to the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project, as well as to the accuracy assessment of land cover fraction maps both thematically and spatially.
+This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions.
+Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which enables customising the result for a diverse range of uses.
+The three-step approach enables mapping 0\% and 100\% fractions more accurately, and thus has already been taken up in the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project.
+This study also includes an accuracy assessment of land cover fraction maps both thematically and spatially, which could be taken up by future land cover fraction mapping efforts.
 \end{abstract}
@@
 
-Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has received increasing amounts of attention, also as new satellites are launched and the capacity for land monitoring increases.
-In this context, new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
-Some of the recent achievements have been the \ac{LC-CCI} map \citep{esa_land_2017} that aims at producing a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} map \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that aims at a finer spatial resolution and high quality with yearly updates since 2015, and the \ac{FROM-GLC} map \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
+Land cover, as one of the key variables for monitoring a number of \glspl{SDG}, has lately received more attention due to increased availability of higher spatial and temporal resolution satellite data.
+In this context, the capacity for land monitoring has increased, and new global land cover maps have emerged to better map the current land cover of the world, as well as to track land cover change.
+Some of the recent achievements have been the \ac{LC-CCI} product \citep{esa_land_2017} that provides a long-term set of consistent global annual medium-resolution land cover maps aimed at the climate community, \glsentryfull{CGLS-LC100} product \citep{buchhorn_moderate_2019, buchhorn_copernicus_2020} that provides a finer spatial resolution and higher quality with yearly updates since 2015, and the \ac{FROM-GLC} product \citep{fromglc2019} that showcases the potential of land cover mapping at 10 m resolution.
 
-Except for the \ac{CGLS-LC100} product with its cover fraction layers, all other global land cover maps, including previous maps such as by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single class.
-However, this results in less precise estimation of land cover due to mixed pixels, where in reality there are multiple land cover classes in the area covered by a single pixel.
-This issue becomes worse at coarse resolutions and heterogeneous landscapes.
+Except for the cover fraction layers of the \ac{CGLS-LC100} product, all other global land cover products that include all major land cover classes, such as the ones described by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single land cover class.
+Such discrete classification oversimplifies reality, as mixed pixels that are covered by multiple land cover classes are a common occurrence. 
+This issue is exacerbated at coarse resolutions and over heterogeneous landscapes.
 It may result in biases, for instance, a sparse forest may be classified as grassland, ignoring the relatively few trees in the area, and thus underestimate tree cover in the pixel.
@@
 A potential solution to this issue is to characterise land cover using cover fractions.
-In this approach, instead of a discrete class, the proportion of every class in the legend is reported for every pixel of the map.
-That way, the land cover models work not on pixel labels, such as ``forest'', but on land cover characteristics, such as tree cover, defined as the area of the pixel covered by tree canopies, as well as herbaceous cover, defined as the area not covered by woody vegetation.
-This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' classification or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
-Related terms are ``super-resolution mapping'' and ``subpixel mapping'', but these concepts include the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.
+In this approach, instead of a single discrete class, the proportion of every class in the legend is reported for every pixel of the map.
+That way, the land cover models work not on pixel labels, such as ``forest'', but on land cover characteristics, such as tree cover, defined as the area of the pixel covered by tree canopies, or herbaceous cover, defined as the area not covered by woody vegetation.
+This is also called ``fuzzy'' or ``soft'' classification, and sometimes ``subpixel'' mapping or ``linear mixture modelling'' \citep{Okeke2006fuzzyexponent}.
+%A related term is ``super-resolution mapping'', but this concept includes the allocation of land cover fractions inside a pixel spatially, rather than just determining the fraction ratio per pixel.
 
@@
 The methods for assessing the accuracy of the results vary greatly between the different studies.
-In addition, global land cover fraction products have emerged, but focused on a particular class, such as tree cover \citep{hansen_global_2003, Hansen2013forestchange}, water \citep{schroeder_development_2015} and urban area \citep{gao_mapping_2020}.
-To date only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.
+In addition, global land cover fraction products have emerged, but focused on a particular class, such as tree cover \citep{hansen_global_2003, Hansen2013forestchange, townshend_global_2017}, water \citep{schroeder_development_2015, pekel_high-resolution_2016} and urban area \citep{corbane_automated_2019, gao_mapping_2020, gong_annual_2020}.
+To date, only \ac{CGLS-LC100} \citep{marcel_buchhorn_copernicus_2019} provides global maps with fractions of every major land cover class \citep{nandin-erdene_tsendbazar_copernicus_2019}.
 
@@
 In its core, it is a regression rather than a classification problem, as the output is a fraction of a label rather than a label itself.
-Popular approaches include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g.
-\citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression \citep{walton2008subpixelrf}, \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
-So far other studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.
-
-In addition to classification methods, one aspect that is important for classification accuracy is the data (covariates) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
-In addition to optical covariates (such as blue, red, \gls{NIR} and shortwave infrared spectral bands), temporal covariates (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
-Elevation covariates (elevation, slope, aspect) are known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
-\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are often used to improve land cover class separability, e.g.
-for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
-Therefore in this paper such covariates have also been used in order to increase the accuracy of land cover fraction regression model predictions, and their importance for the predictions was analysed.
+Methods that have been tested in previous studies include fuzzy nearest centroid regression \citep{zhang2001fullyfuzzy}, \ac{SMA} (e.g.
+\citealp{shimabukuro1991least, adams_classification_1995, hobbs2003linear, yang_landsat_2012}), \ac{RF} regression (e.g.
+\citealp{walton2008subpixelrf}), \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
+The previous studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.
 
 A common issue with the use of land cover fraction data as input into regression models is data imbalance.
-The more classes are mapped, the more often there will be 0\% of a given class in any particular pixel, leading to zero inflation.
-In addition, in homogeneous areas, 100\% cover of a single class is common.
+The more classes are mapped, the more more likely it is that one or more classes are not present in a given pixel (have a 0\% fraction), leading to zero inflation.
+This is especially the case in homogeneous areas, where we can find not only an inflation of 0\% fractions, but also an inflation in 100\% fractions.
 This leaves little data in the middle for the regression models to learn from.
 Conversely, regression models tend to favour predictions closer to the mean and rarely predict extreme values.
-We investigated this issue and suggested ways to deal with it.
+In this study, we propose the use of a hierarchical multi-step modelling approach to better predict these extreme values.
 
-Lastly, accuracy assessment by itself is often times challenging, especially at a global scale.
-It requires a comprehensive dataset across the globe that would be comparable with the modelled land cover, and yet independent of the training data used to create it.
-For land cover fraction mapping it is even more challenging, as fraction information needs to be available, or alternatively fine spatial resolution data from which the fractions can be calculated.
-Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited, which in turn is problematic for the users to decide on what product suits their needs the best.
+In addition to classification methods, one aspect that is important for classification accuracy is the data (features) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
+In addition to optical features (such as spectral bands), temporal features (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
+\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are commonly used to improve land cover class separability, e.g.
+for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
+Elevation features (elevation, slope, aspect) are also known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
+%In this paper, we also used such features for increasing the accuracy of land cover fraction regression model predictions.
+%, and their importance for the predictions was analysed.
+
+Lastly, model accuracy assessment by itself is often challenging, especially at a global scale.
+It requires a comprehensive dataset across the globe that would be comparable with the training data, and yet independent of it.
+Accuracy assessment of land cover fractions is even more challenging, as it requires a dataset that either provides fraction information, or fine spatial resolution data from which it can be calculated.
+Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited.
+This makes it difficult for the users to decide whether a given product suits their needs.
 Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
-In this study, we made use of an independent validation set collected as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019} that includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as in our study.
-The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
-We used the validation data to assess model accuracy both thematically and spatially.
 
-All in all, in this study we investigated and compared methods for performing global land cover fraction mapping, and with a comprehensive set of methods as well as a wide variety of input covariates, so as to improve land cover fraction map production.
-We also investigated ways of dealing with the balance issues inherent to land cover fraction reference data that hampers land cover fraction mapping, and to assess the accuracy of the result.
+%In this study, we made use of a validation dataset that was collected independently from the training dataset, as part of the \gls{CGLS-LC100} project \citep{nandin-erdene_tsendbazar_copernicus_2019}.
+%The dataset includes a total of over 20\,000 points collected across the globe, using the same legend, grid and spatial resolution as the training dataset.
+%The validation data was collected at \gls{WUR}, whereas the training data was collected at \gls{IIASA}, ensuring independence of the datasets.
+%We used the validation data to assess model accuracy both thematically and spatially.
+
+In this study we investigated and compared a variety of methods for performing global land cover fraction mapping.
+Furthermore, we proposed a method to enhance land cover fraction mapping further by dealing with data imbalance, which is inherent in land cover fraction training data, and tested it on the method that achieved the highest accuracy in the method comparison.
+Lastly, we investigated the effect of a wide variety of input features on each mapped class fraction, to gain insight into what data was key for the model with the highest accuracy to achieve the best result.
 Therefore the objectives of our study were to:
@@
 \begin{enumerate}
- \item Compare the performance of a variety of machine learning regression models for global-scale land cover fraction mapping.
- \item Investigate approaches for tuning the models by reducing bias in the predictions with regards to zero inflation and predictions tending towards the mean.
- \item Compare the importance of covariates and covariate groups for the prediction of land cover fractions.
- \item Assess the accuracy of the models from a thematic and spatial point of view.
+ \item Compare the performance of a variety of machine learning regression models for mapping fractions of land cover on a global scale.
+ \item Investigate approaches for reducing bias in the model predictions with regards to zero inflation and predictions tending towards the mean.
+ \item Analyse the importance of features input into the highest accuracy model for the prediction of each land cover fraction.
+ \item Assess the accuracy of the models from a thematic and spatial point of view, comparing it to existing global land cover products.
 \end{enumerate}
@@
 The reference data (for model training and validation) used in this study was collected as part of the \ac{CGLS-LC100} project \citep{buchhorn_copernicus_2020, nandin-erdene_tsendbazar_copernicus_2019}.
-The training set was generated by experts in \ac{IIASA}.
-This global point dataset provides information on fractions of each class within sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
-The fraction estimates were acquired by high-resolution image interpretation by experts, for the year 2015.
-Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.
-
-For model accuracy assessment, an independent global validation dataset of \ac{CGLS-LC100}, generated by \gls{WUR} \citep{tsendbazar_developing_2018}, was used.
-The method of point collection is equivalent to the method used for training data collection, but performed independently by a different group of regional experts.
-
 The data includes over 150\,000 training points and over 21\,000 validation points across the globe, describing the fractions of 12 classes in the year 2015.
 The classification scheme follows the \gls{LCCS} \citep{buchhorn_copernicus_2020}.
-However, due to limited number of observations for some rare classes, we merged them to get a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous vegetation (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
-The unknown class was discarded: points with the dominant land cover class marked as unknown were not used.
+However, due to the limited number of observations for some rare classes, we merged them to get a total of seven: bare land (including snow and ice), cropland (including shifting cultivation), herbaceous vegetation (including wetland, lichen and moss), shrubs, trees, built-up and inland water.
+The ``unknown'' class was discarded: points with the dominant land cover class marked as unknown were not used.
 For points with a minority fraction of unknown, the remaining classes were linearly rescaled to add up to 100\%.
@@
 
+These global datasets were generated by performing high-resolution satellite imagery interpretation by a team of experts, using the GeoWiki platform.
+Each sampled point corresponded to a single 100~m by 100~m pixel of the PROBA-V 100~m UTM grid \citep{buchhorn_copernicus_2020}.
+The area of each of these sampled pixels was subdivided into 100 subpixels at 10~m by 10~m spatial resolution.
+The subpixels were labelled by the experts, and then converted into land cover fraction estimates by calculating the proportion of subpixels that each land cover class covers in the pixel.
+The satellite imagery that the experts interpreted corresponded to the year 2015.
+
+%These global point datasets provide information on fractions of each class within sampled pixels of the PROBA-V 100 m product \citep{dierckx2014probav}.
+%The fraction estimates were acquired by high-resolution image interpretation by experts, for the year 2015.
+%Each of the sampled PROBA-V pixels was subdivided into a 10-by-10 m subpixel grid and the class fraction estimates were derived from the subpixel ratios.
+
+The training set was generated by a team at \ac{IIASA}, whereas the validation dataset was generated by a team at \gls{WUR}.
+The validation dataset was first developed over Africa \citep{tsendbazar_developing_2018}, and was later expanded to cover the whole world \citep{buchhorn_copernicus_2020}.
+The class definitions and tools used to collect the data was equivalent for both datasets, but performed independently by a separate group of regional experts to ensure independence of the data, and using a different sampling method.
+The validation dataset was generated separately using stratified random sampling and follows the CEOS Land Product Validation guidelines, which focus on independent and statistically rigorous accuracy assessment.
+In contrast, the training dataset uses a mix of systematic sampling and additional random sampling in areas that exhibit higher heterogeneity, so as to provide the algorithms with enough data to train in all areas of the world.
+
 \begin{figure}
- \includegraphics[width=\textwidth]{article-figures/maps/2020-07-06-training-and-validation}
- \caption{Sample points representing 100x100 m areas at which land cover reference data was used in the study.
+ \includegraphics[width=\textwidth]{article-figures/maps/2020-10-12-training-and-validation}
+ \caption{Sample points representing 100x100~m areas at which land cover reference data was used in the study.
 The colours represent the dominant land cover class at each point.
@@
 
-\subsection{Model covariates}
+\subsection{Model training features}
 
-To predict land cover fractions in unsampled areas, the models learn from covariates present at the locations of the training data, and then the covariates are used for prediction in unsampled areas.
-See figure \ref{fig-processing} for an overview of the whole processing chain.
+See figure \ref{fig-processing} for an overview of the whole processing chain used in this study.
 The processing was carried out in R \citep{r_2019} and the resulting code has been made openly available in \citet{dainius_masiliunas_2020_3973123}.
 
-In this study, six groups of covariates were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
-These covariates had to be preprocessed before they could be input into the models.
+To train the models and predict land cover fractions in unsampled locations, six groups of features were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
+These features had to be preprocessed before they could be input into the models.
 
@@
 %     \label{tab-inputdata}
-%     \caption{Data sources for the covariates input into the model.}
+%     \caption{Data sources for the features input into the model.}
 % \end{table}
@@
 
-The entire archive (2014-03-11 to 2019-07-16) of the PROBA-V 100 m Level 3 Top-of-Canopy 5-day composite product \citep{dierckx2014probav,probavguide2} was used for this study.
-While the reference data corresponded to the land cover at the year 2015 specifically, the whole time series of PROBA-V was chosen to obtain more robust statistics for the temporal metrics.
+We chose to use the entire archive (2014-03-11 to 2019-07-16) of the PROBA-V 100~m Level~3 Top-of-Canopy 5-day composite product \citep{dierckx2014probav,probavguide2} for this study.
+The PROBA-V archive provides a relatively long history of frequent (daily or 2-day) observations, which is beneficial for time series analysis, as there are more observations of the land surface in cloudy areas, and a dense time series of observations can be acquired to generate robust temporal metrics.
+While the reference data corresponded to the land cover at the year 2015 specifically, we chose to use the whole time series of PROBA-V data to obtain more robust statistics for the temporal metrics.
+The long time series makes the temporal outlier filtering step more reliable, increases the robustness of the fitted harmonic model, and removes the effects of interannual variability and seasonality in the calculated descriptive statistics, as described in the following sections.
 Land cover change is a relatively rare phenomenon, according to tests performed in the making of \gls{CGLS-LC100} collection 3, and so we expected that the difference caused by land cover change to the input data would be smaller than the margin of error of the model output.
@@
 
-The time series of each of the PROBA-V spectral bands were cloud masked, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the fitted \ac{LOESS} curve over the blue reflectance band.
-The resulting cloud-free time series was then used to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}.
-Next, the \glspl{VI} were used to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
-Next, harmonic analysis \citep{jakubauskas2001harmonic} was run on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
-From this harmonic analysis we obtained minimum and maximum values of the \gls{NDVI} time series, and calculated the phase and amplitude for the two harmonic orders from the respective sine and cosine components.
+We masked out the clouds from the time series of each of the four PROBA-V spectral bands, first by applying the status mask provided with the product itself, and then by running a temporal cloud filter to remove the remaining outliers that were further than 2 standard deviations away from the \ac{LOESS} curve fitted over the blue reflectance band.
+We then used the resulting cloud-free time series to generate the following \glspl{VI}: \gls{NDVI}, \gls{NDMI}, \gls{EVI}, \gls{OSAVI} and \gls{NIRv}.
+These \glspl{VI} are commonly used in land cover mapping to aid in discerning the vegetation classes among each other.
+Next, we used the \gls{VI} time series to calculate the descriptive statistics (median and interquartile range) over the whole time series, as well as for each phenological season separately.
+We included the resulting metrics as vegetation index features.
+Next, we ran harmonic analysis \citep{jakubauskas2001harmonic} on NDVI in order to decompose the time series into sine and cosine components for two frequency orders (annual and semiannual), as well as the trend and intercept of the model.
+The temporal metrics from the harmonic model quantify the seasonality of the area and allow differentiating between vegetation with different seasonality, such as crops.
+From this harmonic analysis we obtained the minimum and maximum values of the \gls{NDVI} time series, which we included as additional vegetation index features.
+We also obtained the the trend and intercept components from the harmonic model.
+Lastly, we calculated the phase and amplitude for the two harmonic orders from the respective sine and cosine components of the model.
+We used these sine, cosine, trend and intercept components, as well as the phase and amplitude of the two harmonic orders, as temporal features in our study.
+For a list of features that were ultimately used as input to the models, see supplementary material, table S1.
 
-For elevation data, the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) was resampled to PROBA-V 100 m grid and used as the elevation covariate.
-In addition, \gls{GDAL} \citep{gdal} was used to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
+To generate elevation features, we obtained the ASTER GDEM v003 \citep{ASTGTM003} product (30 m) and resampled it to the PROBA-V 100~m grid.
+We used the result directly as the terrain elevation feature.
+In addition, we used \gls{GDAL} \citep{gdal} to calculate terrain parameters out of elevation: slope, aspect and \ac{TPI}.
 
-For climate data, WorldClim 2.0 30 second product \citep{worldclim2} was used.
+We chose the WorldClim~2.0 30~second product \citep{worldclim2} as a source of climate features.
 It includes monthly temperature, precipitation, solar radiation, wind speed and water vapour pressure data.
-In addition, 19 bioclimatic parameters were calculated from the data, using the \cran{dismo} package \citep{dismo}.
-Some additional biophysical parameters were also calculated, namely all of the climate data during the coldest, warmest, driest and wettest months of the year at each location, as well as yearly averages of each of the original parameters.
+In addition to these features, we calculated 19 bioclimatic parameters from the data, using the \cran{dismo} package \citep{dismo}.
+We also calculated some additional biophysical parameters, namely all of the climate variables during the coldest, warmest, driest and wettest months of the year at each location, as well as the yearly averages of the climate variables.
 
-SoilGrids \citep{hengl_soilgrids250m_2017} was used for variables related to soils.
-SoilGrids is based on a random forest model that predicts soil properties at various soil depths globally at a 250 m resolution.
-In the creation of SoilGrids, a land cover map (based on MODIS) had been used, and so, in order to avoid circular inference, the covariates that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
-The soil taxonomy covariates were also excluded, since they are categorical derivatives from the numerical soil property data and thus do not contribute to land cover fraction prediction.
+We used SoilGrids \citep{hengl_soilgrids250m_2017} to obtain features related to soils.
+SoilGrids is based on a random forest model that predicts soil properties at various soil depths globally at a 250~m resolution.
+In the creation of SoilGrids, a land cover map (based on MODIS) had been used, and so, in order to avoid circular inference, the features that are significantly influenced by the land cover map as detailed in \citet{hengl_soilgrids250m_2017} were excluded.
+The soil taxonomy features were also excluded, since they are categorical derivatives from the numerical soil property data and thus do not contribute to land cover fraction prediction.
 
-The latitude, longitude and absolute latitude of the reference points were also included when training the models, so that the models could learn spatial patterns.
+Lastly, we also included the latitude, longitude and absolute latitude of the reference points as location features when training the models, so that the models could learn spatial patterns.
 
-\subsection{Covariate selection}
+\subsection{Feature selection}
 \label{sec-covariate-selection}
 
-In total, 313 covariates were generated in preprocessing.
+In total, we generated 313 features in preprocessing.
 However, many of them were collinear with one another.
 Multicollinearity prolongs training time for machine learning models and leads to unreliable coefficient estimation and increased error variance for linear models.
-Thus, variable selection was used to remove collinear covariates before predicting the land cover fractions.
-Covariates that had a Pearson's correlation coefficient $r$ \citep{pearson_notes_1895} above 0.9 were excluded in an iterative process.
-After that, covariates with a Spearman's rank correlation $\rho$ \citep{spearman1904rank} above 0.9 were likewise excluded.
+Thus, we employed variable selection to remove collinear features before predicting the land cover fractions.
+Features that had a Pearson's correlation coefficient $r$ \citep{pearson_notes_1895} above 0.9 were excluded in an iterative process.
+After that, features with a Spearman's rank correlation $\rho$ \citep{spearman1904rank} above 0.9 were likewise excluded.
 
-The covariates to exclude were selected manually, to avoid interpretation difficulties that arise from automatic selection procedures.
-The majority of the collinear covariates were soil metrics predicted at different depths.
-Therefore the 10 cm depth covariates were left in, and the other depths excluded, as long as $r$ was above 0.9.
+We manually selected the features to exclude, to avoid interpretation difficulties that arise from automatic selection procedures.
+The majority of the collinear features were soil metrics predicted at different depths.
+Therefore, we left in the 10 cm depth features, and excluded the other depths, as long as $r$ was above 0.9.
 Similarly, climate data was collinear between subsequent months.
-Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with other months' data.
-Our initial correlation analysis also showed that the spectral bands of PROBA-V were highly correlated with each other as well as to \glspl{VI}, so only \glspl{VI} were used as covariates.
+Thus January and July data was preferred, as these months represent different extremes of the year and are less correlated with data of the other months.
+Our initial correlation analysis also showed that the spectral bands of PROBA-V were highly correlated with each other as well as to \glspl{VI}, so we only used \glspl{VI} as features.
 
-After the covariate selection process, 67 covariates remained.
-These covariates include data from each of the covariate categories.
-See supplementary material, table S1 for an overview.
+After the feature selection process, 67 features remained.
+These features include data from each of the feature categories.
+See supplementary material, table S1 for an overview of all of the features that remained and thus used in model training and prediction.
 
@@
 
-For deriving land cover fractions, we compared a wide array of machine learning regression methods.
-We broadly divided the methods into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
-We tuned the methods to select optimal parameters for each.
-See table \ref{tab-methods} for the full list of methods that were compared in this study.
-In addition, as a baseline we also compared the results with an intercept model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
-All of the models were made using the free and open-source statistical software R \citep{r_2019}.
+We compared a wide array of machine learning regression methods for deriving land cover fractions.
+The tested methods can be broadly divided into three types: linear models, machine learning models based on \glspl{CART}, and machine learning models not based on \glspl{CART}.
+See table \ref{tab-methods} for the full list of methods that we compared in this study.
+In addition, as a baseline we also compared the results with a trivial equal proportion model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
+We tuned each algorithm to select optimal parameters, and postprocessed the output of each algorithm as necessary to ensure that all land cover fractions in each pixel add up to 100\%.
+Namely, if the model output for any class was outside of the 0-100\% range, the values were clamped to that range, and if the values did not add up to 100\%, they were linearly rescaled so that they would.
+All of the model building and data analysis was performed using the free and open-source statistical software R \citep{r_2019}.
 
@@
          \begin{tabular}{@{}l@{}}
-         \Glsentryfullpl{MLP}\tabularnewline
+         \Gls{MLP} \glsentryfullpl{NN}\tabularnewline
          \Glsentryfull{SVM} regression
@@
          \end{tabular}
+         \tabularnewline \midrule
+         Ensemble learning &
+         Super Learner
+         & 
+         \citealt{laan_super_2007}
          \tabularnewline \bottomrule
@@
 
-The \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to the standard linear regression that allows for multiple outcomes.
-In this study, we used a \gls{GLM} as a baseline model.
-We also tested whether dropping any additional covariates from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction \citep{greenhouse_methods_1959} to compare models with different covariates.
-However, only three covariates could be dropped and it did not improve the predictions, therefore we decided to keep all the covariates in the model.
-\Gls{GLM} is implemented in base R \citep{r_2019}.
+We chose to compare five types of linear models, to have a baseline for a comparison with the nonlinear machine learning models.
+The most simple model we selected was the \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression.
+It is an extension to the standard linear regression that allows for multiple outcomes, and is implemented in base R \citep{r_2019}.
+Next, we tested two linear models that include input data regularisation in the model itself: lasso regression \citep{tibshirani_regression_1996}, implemented in the \cran{glmnet} package \citep{glmnet}, and \gls{PLS} regression \citep{wold_pls-regression_2001}, implemented in the \cran{pls} package \citep{pls}.
+We also tested \gls{MLR} \citep{theil_multinomial_1969}, provided by the \cran{nnet} package \citep{nnet}.
+\gls{MLR} is usually used for classification and is fit using land cover class labels rather than fractions, but the output includes probabilities for each class that add up to 100\%.
+We fit \gls{MLR} using the dominant land cover class as a label for the pixel and used the class probabilities as a proxy for land cover fractions.
+Lastly, we tested \gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, which is provided in the \cran{GSIF} package \citep{hengl2004fuzzycmeans}.
+It is a simple regression method, where the land cover fractions in a pixel are determined by the distance of the pixel from the centroids of each class in feature space.
 
-Ridge regression is a linear model regularisation method which reduces large coefficients of model covariates in order to prevent overfitting.
-Lasso regression is based on the same principle, except it reduces the coefficients of each model covariate to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
-Elastic net regression is a method combining the two, where model coefficients may be reduced or set to zero.
-This has the effect of the coefficients of correlated covariates being set to a similar value.
-We chose lasso regression \citep{tibshirani_regression_1996}, so that it would be possible to identify if there are any covariates that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
-The $\lambda$ parameter was chosen using cross-validation.
-The lasso regression was performed using the \cran{glmnet} package \citep{glmnet}.
+%The \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to the standard linear regression that allows for multiple outcomes.
+%In this study, we used a \gls{GLM} as a baseline model.
+%We also tested whether dropping any additional features from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction \citep{greenhouse_methods_1959} to compare models with different features.
+%However, only three features could be dropped and it did not improve the predictions, therefore we decided to keep all the features in the model.
+%\Gls{GLM} is implemented in base R \citep{r_2019}.
 
-\Gls{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input covariates into latent variables that explain the most variance of the response variable in multidimensional space.
-As such, it is well suited for multicollinear data.
-\ac{PLS} regression was performed using the package \cran{pls} \citep{pls}.
+%Ridge regression is a linear model regularisation method which reduces large coefficients of model features in order to prevent overfitting.
+%Lasso regression is based on the same principle, except it reduces the coefficients of each model feature to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
+%Elastic net regression is a method combining the two, where model coefficients may be reduced or set to zero.
+%This has the effect of the coefficients of correlated features being set to a similar value.
+%We chose lasso regression \citep{tibshirani_regression_1996}, so that it would be possible to identify if there are any features that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
+%The $\lambda$ parameter was chosen using cross-validation.
+%The lasso regression was performed using the \cran{glmnet} package \citep{glmnet}.
 
-All of the aforementioned linear models do not normalise the output to add up to 100\%, and also allow negative predictions.
-Therefore for these models, all predictions < 0 were set to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
+%\Gls{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input features into latent variables that explain the most variance of the response variable in multidimensional space.
+%As such, it is well suited for multicollinear data.
+%\ac{PLS} regression was performed using the package \cran{pls} \citep{pls}.
 
-\Gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
-This method always produces output that sums to 100\%.
-In this study, we used an implementation adapted from the package \cran{GSIF} \citep{hengl2004fuzzycmeans}.
-We tested both using a logistic regression to estimate the class centroids, as well as using a weighted average of the training data to do so, and ended up using the latter method as the result was more accurate.
+%All of the aforementioned linear models do not normalise the output to add up to 100\%, and also allow negative predictions.
+%Therefore for these models, all predictions < 0 were set to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
 
-Lastly, we tested \gls{MLR} \citep{theil_multinomial_1969}.
-While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
-This probability can be used as a proxy for the fraction of the pixel that the class covers.
-These probabilities always add up to 100\%.
-However, an \gls{MLR} can only be trained on labelled data, rather than fractions.
-Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
-This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
-The logistic regression implementation was provided by the \cran{nnet} package \citep{nnet}.
+%\Gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
+%This method always produces output that sums to 100\%.
+%In this study, we used an implementation adapted from the package \cran{GSIF} \citep{hengl2004fuzzycmeans}.
+%We tested both using a logistic regression to estimate the class centroids, as well as using a weighted average of the training data to do so, and ended up using the latter method as the result was more accurate.
+
+%Lastly, we tested \gls{MLR} \citep{theil_multinomial_1969}.
+%While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
+%This probability can be used as a proxy for the fraction of the pixel that the class covers.
+%These probabilities always add up to 100\%.
+%However, an \gls{MLR} can only be trained on labelled data, rather than fractions.
+%Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
+%This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
+%The logistic regression implementation was provided by the \cran{nnet} package \citep{nnet}.
 
@@
 \Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
-In this study, after performing initial parameter tests, we used a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
-We chose the Nadam optimiser \citep{dozat_incorporating_2016}, with \gls{MAE} as the loss function.
-Softmax activation was used to get the output.
-The \cran{keras} package \citep{keras}, built upon TensorFlow, was used as the \ac{NN} implementation, and for the ability to make use of a graphics processing unit to accelerate the neural network training process.
+In this study, after performing tuning, we ended up using a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+We used the Nadam optimiser \citep{dozat_incorporating_2016} with \gls{MAE} as the loss function to optimise the \gls{NN}, and softmax activation for the output.
+The models were trained using the \cran{keras} package \citep{keras}, built upon TensorFlow that enables the use of a graphics processing unit to accelerate the \gls{NN} training process.
 
@@
 For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
-We used the binary relevance method, where separate models are trained per class that predict the proportion of the class in the pixel \citep{karalas2016br}.
-We clamped the output values to the 0-100\% range and linearly scaled them to add up to 100\%.
-The package \cran{liquidSVM} \citep{liquidSVM} was suitable for this purpose, due to its optimisations for processing large datasets such as ours.
+As \gls{SVM} models are univariate, we used the binary relevance method \citep{karalas2016br}: training separate models per class that predict a single class, and then combining the results.
+We found the package \cran{liquidSVM} \citep{liquidSVM} suitable for this purpose, as it is optimised for processing large datasets such as ours.
 
@@
 
-\Glsentryfull{RF} regression is a popular method for land cover classification that works by building a number of \gls{CART} decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual trees \citep{breiman2001random}.
-\ac{RF} can only take one variable as the response variable, therefore there are two ways to obtain land cover fraction predictions from an \ac{RF} model.
-One way is analogous to how we performed \ac{MLR}: giving labels of the dominant land cover class as the response variable, and then considering the proportion of votes from the trees as the proportions of land cover classes in the pixel.
-The second way is the binary relevance method as with \glspl{SVM}.
-In the case of \gls{RF}, output values do not need to be clamped as their range is always within the constraints of the input data range, but still need to be scaled to add up to 100\%.
-In this study, we chose the binary relevance approach, so that the model for each class would be able to learn from the fractions of all the input data, rather than just the labels of the dominant land cover class.
+\Glsentryfull{RF} regression is a popular method for land cover classification that works by building a number of \gls{CART} decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual decision trees \citep{breiman2001random}.
+\ac{RF} is univariate, therefore we again used the binary relevance approach.
 The processing was done using the \cran{ranger} package \citep{ranger}.
 
-As the results for \gls{RF} regression proved the best for \gls{RMSE}, it was also used for assessing the covariate importance.
-Permutation importance was used: each covariate was shuffled in turn, and the resulting decrease in \gls{RMSE} was calculated, comparing against the validation data set.
-Since the number of covariates is imbalanced by group (i.e.
-there are merely 3 location covariates and 22 vegetation index metrics), and there may still be collinear covariates, we also tested calculating covariate importance by permuting all covariates in a group together.
-However, the result was similar to the sum of permutation importance of each individual covariates in the group, which showed that the covariate selection process had successfully eliminated all collinear covariates.
-Thus we reported the more detailed covariate importance permuted per individual covariate.
+Lastly, we tested Cubist regression \citep{cubist}.
+It is based on \gls{RF} regression, but instead of using a threshold of a feature to split the decision tree, Cubist uses a linear regression based on a subset of the data relevant for the split in question.
+In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previously generated ones.
+Model tuning led to us using 10 committees.
+Cubist predictions were also made using the binary relevance method.
+The algorithm is provided by the \cran{Cubist} package \citep{cubistpackage}.
 
-Lastly, we tested Cubist regression \citep{cubist}.
-It is based on \gls{RF} regression, with the major difference being that instead of using a threshold of a covariate for a split in the decision tree, Cubist instead uses a linear regression based on a subset of the data relevant for the split in question.
-In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previous ones.
-For our dataset, 10 committees gave the best result.
-Like with \gls{RF}, Cubist was used with the binary relevance method.
-The \cran{Cubist} package \citep{cubistpackage} was used for this.
+\subsubsection{Ensemble learning}
 
-\subsubsection{Other methods}
+Lastly, we made an ensemble from the two machine learning models that produced the lowest \gls{RMSE} and \gls{MAE} respectively.
+We used the \cran{sl3} package \citep{sl3} to create a hierarchical ensemble, where the two models were cross-validated using 10-fold cross-validation to obtain relative weights of each model for each land cover class, following the super learner algorithm \citep{laan_super_2007}.
+The predictions of the two models, along with the weights of the models, were then used as input features for another \gls{RF} regression metalearner.
+The output from this metalearner is the final prediction of the whole ensemble.
+As the ensembled methods are univariate, we used the binary relevance method in this case as well.
 
-Additional methods were considered, but in the end turned out to be unworkable with our dataset.
-Multivariate \gls{RF} is available in R provided by the \cran{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
-However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
+%\subsubsection{Other methods}
 
-Another model that was considered was \cran{SuperLearner} \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
-Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
-Attempts to reduce training time by lowering the number of folds resulted in worse results than when using a single classifier.
+%Additional methods were considered, but in the end turned out to be unworkable with our dataset.
+%Multivariate \gls{RF} is available in R provided by the \cran{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
+%However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
 
-Lastly, multiple-endmember spectral unmixing using non-negative least squares \citep{franc2005sequential} was also attempted, but the results were significantly worse than even the intercept model, possibly because the class fractions are not a linear combination of the covariates.
+%Another model that was considered was \cran{SuperLearner} \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
+%Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
+%Attempts to reduce training time by lowering the number of folds resulted in worse results than when using a single classifier.
 
-\subsection{Model tuning (multi-step approach) to account for value imbalance}
+%Lastly, multiple-endmember spectral unmixing using non-negative least squares \citep{franc2005sequential} was also attempted, but the results were significantly worse than even the equal proportions model, possibly because the class fractions are not a linear combination of the features.
+
+\subsection{Multi-step approach to account for value imbalance}
 \label{sec-multistep}
 
-After determining the method that performs the best, we attempted to tune it further in an attempt to solve the balance (zero inflation) problem in the training dataset.
-As the dataset describes fractions of each land cover class at each point, most of the locations consist of a mix of only a few classes, and the rest of the classes are zero.
-As the dataset is dominated by zeroes, minimising the objective function tends to draw the model towards predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
+After determining the method with the lowest \gls{RMSE}, we attempted to improve it further in an attempt to solve the dataset balance issue, namely the high frequency of 0\% and 100\% fractions.
+As the dataset describes fractions of each land cover class at each point, most of the locations consist of a mix of only a few classes, and the fraction of the rest of the classes is zero at that location.
+If the pixel is pure, then one land cover class will be 100\% and the rest 0\%, which is also a common case.
+This leads to the dataset getting dominated by zeroes.
+In that case, minimising the objective function of a machine learning model leads to prioritising predicting 0\% fractions well, and ignoring the prediction errors in the middle of the 0-100\% range.
 This is not desirable for users of land cover fraction data, as the added value of fraction information is the information about the middle of the range; otherwise, discrete classification would be just as good.
-Therefore we tried several approaches to increase the accuracy of the middle predictions.
+Conversely, 0\% by itself is rarely predicted precisely, because when the value is uncertain, predictions tend towards the mean.
+Therefore we tried several approaches to deal with data imbalance by employing a hierarchical combination of machine learning models.
 
-For the best performing algorithm, we compared: (1) a single model, (2) a two-model approach: (a) one model to predict zeroes, and (b) one model to predict non-zeroes; and (3) a three-model approach, with (a) one model to predict pixel purity (i.e.
-whether we face a classification or a regression problem), (b) one to perform regression (on mixed pixels as determined by (a)), and (c) one model to perform classification (on pure pixels as determined by (a)).
+Taking the algorithm with the lowest \gls{RMSE}, we compared three approaches.
+(a): a single regression model trained on all data.
+(b): a two-model approach in two steps.
+Step 1: a binary classification model for each class to predict zeroes, trained on a generated dataset that, based on the land cover fraction values, had labels ``zero'' and ``non-zero''.
+Step 2: a model to predict non-zeroes, trained on all of the non-zero fraction values.
+For the combined prediction, all points that were predicted as ``zero'' in step 1 were set to 0\%, otherwise the value from the model in step 2 was used.
+(c): a three-model approach using three steps.
+Step 1: a binary classification model to predict pixel purity (i.e.
+whether we face a classification or a regression problem), trained on a generated dataset that had labels ``pure'' for points that had a single land cover fraction above a purity threshold, e.g.
+95\%, and ``non-pure'' for points that do not.
+Step 2: a model to perform regression on mixed pixels (as determined in step 1).
+Step 3: a model to perform classification on pure pixels (as determined in step 1), resulting in a prediction of 100\% fraction of the predicted discrete class and 0\% fractions for all other classes.
+The combined prediction is the combination of the results of steps 2 and 3, as determined by step 1.
+
+For the three-step approach, we also tested the effect of the fraction threshold for when we consider a pixel ``pure''.
+The lower the threshold, the more pixels are considered pure and the more often the classification model will be selected, as opposed to the regression model.
+We also evaluated the accuracy metrics of the separate steps of the multi-step models.
+
+Lastly, we compared the results of our proposed multi-step approach with an approach that uses the median for ensembling tree votes, instead of the mean.
+The median vote leads to predicting the extreme fractions of 0\% and 100\% more often, since if the majority of the decision trees vote for one of the two extremes, it gets selected as the output value.
+Finally, we also investigated the combination of both approaches.
 
@@
 To assess the performance of the models, we used a number of statistical measures.
-First, we calculated \gls{RMSE}, \ac{MAE} and \ac{ME}, both separately per class, and also pooled overall.
+We started with the statistics of assessing land cover fraction model accuracy that are the most commonly used in this field: \glsentryfull{RMSE}, \glsentryfull{MAE} and \glsentryfull{ME}.
+\gls{MAE} represents the average difference between the predicted and the reference land cover fractions.
+In our case, its unit is percentage points.
+\gls{RMSE} squares the errors, therefore giving a larger penalty for large errors, and thus is always higher than \gls{MAE}.
+These statistics are relatively straightforward to calculate and interpret.
+In the case of land cover fraction mapping, \gls{RMSE} is very sensitive to errors where a pixel is entirely mapped as a different class (i.e. 100\% instead of 0\%).
+\gls{MAE} is more lenient and not as influenced by a small number of such large misclassifications.
+Thus it is more indicative of the overall model accuracy, whereas \gls{RMSE} is more indicative of the presence of large errors.
+
+We calculated \gls{RMSE}, \ac{MAE} and \ac{ME}, both separately per class, and also pooled overall.
 These overall measures were calculated by taking the mean of all class points pooled together, rather than taking a mean of the per-class means.
+In addition, we calculated the \gls{RRMSE}, \gls{RMAE} and \gls{RME} for each class by dividing the absolute measures by the mean fraction of each class.
+The relative statistics give an extra penalty for poor predictions of rare land cover fractions (i.e.
+those that are absent from most pixels), to account for the issue that a prediction of constant 0\% would lead to low \gls{RMSE} and \gls{MAE} for rare class fractions.
 
-Next, the \gls{NSE} \citep{nash1970river} was used to determine the goodness of fit.
-For regression output, \ac{NSE} is a method of calculating the coefficient of determination $R^2$, and is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
+Next, we estimated the goodness of fit of the models by calculating the coefficient of determination $R^2$ of the models in two ways.
+The first way is \gls{NSE} \citep{nash1970river}, which is equivalent to an $R^2$ of a linear regression model whose intercept and slope terms are predetermined and are equal to 0 and 1, respectively.
+This metric shows how far away the predicted values are from the 1:1 line with the reference values.
+The value range of \gls{NSE} is $(-\infty, 1]$, where a value of 0 means that the predicted values are no better than predicting the mean value.
+The second way is to calculate the $R^2$ of an Ordinary Least Squares regression that estimates the intercept and slope, rather than presetting it.
+This method always gives higher $R^2$ values than \gls{NSE}, as it allows the regression line to have more flexibility.
+The two ways of estimating the coefficient of determination are herewith represented as $R^2_{NSE}$ and $R^2_{OLS}$, respectively.
 
@@
 
-Moreover, we produced boxplots showing the distribution of predictions, in bins for each 10\% of the predictions, against the validation data distribution, and a comparison with the 1:1 line.
+Moreover, to show the variation in predicted land cover fraction values depending on the magnitude of the fractions, we produced boxplots showing the distribution of the predictions.
+The boxplots are binned for each 10\% of the predictions.
+The 1:1 line indicates how well do the distributions of the predictions and the reference data match.
+
 Lastly, we produced spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point.
-We rasterised the result into a map for ease of view, taking the mean value where several points fell into the same raster cell.
+We rasterised the result into a map for ease of view.
+If multiple points fell into the same raster cell, we reported the mean value.
+
+To put our results in a wider context, we compared them with the currently available global products for specific land cover classes.
+For this comparison, we used the same validation dataset as for our models.
+Since the validation dataset reflects land cover of the year 2015, we chose to compare our results with four products produced for or around the year 2015.
+Therefore we chose the \gls{GFCC} forest cover product for 2015 \citep{townshend_global_2017}, the \gls{GSW} water occurrence history product for 2015 \citep{pekel_high-resolution_2016}, and two products for the comparison with the built-up class: \gls{GHSL} built-up for 2014 \citep{corbane_automated_2019} and \gls{FROM-GLC} impervious surface change layer for 2015 \citep{gong_annual_2020}.
+As the latter two products show the change in built-up cover over time, we reclassified them to have a value of 100 if the area was built-up in 2014 or 2015, respectively.
+For \gls{FROM-GLC}, we assumed that areas that had been covered by impervious surface at any time prior to 2015 continued to be covered by impervious surface in 2015.
+\gls{GSW} data was also reclassified to 100 if the water was permanently present and 50 if it was seasonally present.
+Next, for validation purposes we resampled each product to the PROBA-V 100 m grid using the bilinear resampling method, so that the cover fractions get aggregated over the same areas as our own data.
+Then we compared the product values with our validation data point values.
+
+\subsection{Feature importance}
+
+To calculate feature importance, we performed permutation importance on the model that had the lowest \gls{MAE}.
+We shuffled the values of each feature in turn, and the models made predictions based on all of the features, including the shuffled one.
+This results in a decrease in the accuracy of predictions, as the feature in question no longer contains meaningful information.
+We then recorded the resulting increase in \gls{MAE} compared to the validation set as a measure of feature importance, both per class and overall for all classes combined.
 
@@
 
-Overall model accuracy statistics are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
-\Gls{RF} regression with no adjustments (single-step) obtained the lowest \gls{RMSE}, but lowest \gls{MAE} was obtained with \gls{RF} regression using the three-step model and median voting.
-\gls{NSE} results follow the \gls{RMSE} results, and the \glsentryfull{OA} from the \glsentryfull{SCM} follow the \gls{MAE} results.
-This is due to the \gls{NSE} being a representation of how close the data is to the 1:1 line with the validation data, where big outliers have a large effect on the value.
+The overall accuracy statistics of the compared models are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
+Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote, and by \gls{MAE} when using a median vote.
+The combination of the three-step model and the median vote resulted in a decrease in \gls{RMSE} without impacting \gls{MAE}.
+
+The results show that the $R^2$ statistics, especially $R^2_{NSE}$, are in agreement with the \gls{RMSE} statistics, and the \glsentryfull{OA} from the \glsentryfull{SCM} is in agreement with the \gls{MAE} statistics.
+$R^2$ is a representation of how closely the predictions correlate with the validation data, therefore big outliers have a large effect on the value.
 The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
 
-The models that performed the best were the two tree-based machine learning models: \gls{RF} and Cubist.
-In the case of these models, predictions per class were made using the binary relevance method (one model per class), as they are univariate.
-This shows that despite the binary relevance limitations, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and requiring a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models outweigh the disadvantages of the binary relevance method.
+The baseline models that performed the best were the two tree-based machine learning models: \gls{RF} regression and Cubist.
+Both of these models are univariate, therefore predictions per class had to be made using the binary relevance method (one model per class).
+This shows that despite the limitations of the binary relevance method, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and the need for a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models and the flexibility of fitting the models for each class outweigh the disadvantages of the binary relevance method.
 
-The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed significantly worse, as they are limited to a linear approach.
-The results of all of the linear models were very similar, which shows that there was no added value in \gls{PLS} and Lasso extensions over the basic \gls{GLM}.
-This may be due to the additional preprocessing step of covariate selection as detailed in section \ref{sec-covariate-selection}.
-It resulted in all remaining covariates being important enough to the prediction of land cover fractions, so that removing any more of them hurts rather than improves model performance.
+The super learner method that combines both \gls{RF} regression and Cubist regression resulted in a model that is in between the two ensembled models in terms of accuracy.
+Its \gls{RMSE} was below that of Cubist regression, but above that of \gls{RF} regression.
+Conversely, the \gls{MAE} was below that of \gls{RF} regression (single model), but above that of Cubist regression.
+This shows that the metalearner (another \gls{RF} model) could not differentiate between the two ensembled models well enough to select the model that is the most accurate for a given class.
+Rather, it weighted in both of the models' predictions, losing the advantages of the models when taken separately.
+
+The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed the worst, as they are limited to a linear approach.
+The results of all of the linear models were very similar.
+There was no added value to \gls{PLS} and Lasso regressions over the basic \gls{GLM}.
+This is likely due to the additional preprocessing step of feature selection, as detailed in section \ref{sec-covariate-selection}.
+Lasso and \gls{PLS} regressions are extensions to \gls{GLM} that include a regularisation step, which proved to be unnecessary if all of the highly correlated features are already manually removed.
 
 \Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
-This indicates that the shape of the covariate data in feature space is too complex to capture merely with a centroid approach.
+This indicates that the shape of the feature data in feature space is too complex to capture merely with a centroid approach.
 This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
@@
 %\adjustbox{max width=0.5\textwidth}{
-\begin{tabular}{llllll}
+\begin{tabular}{lllllll}
 \toprule
-\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & \textbf{\acrshort{NSE}} & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
+\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & $\mathbf{R^2_{NSE}}$ & $\mathbf{R^2_{OLS}}$ & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
 \midrule
-Intercept
-& 29.9  & 21.4  & 0     & 26±5  & 0.13±0.07 \\
+Equal proportions
+& 29.9  & 21.4  & 0     & — & 26±5  & 0.13±0.07 \\
 \Gls{FNC}
-& 24.4  & 13.5  & 0.33  & 53±4  & 0.42±0.06 \\
+& 24.4  & 13.5  & 0.33  & 0.35 & 53±4  & 0.42±0.06 \\
 \Gls{GLM}, \Gls{PLS}, Lasso
-& 21.6  & 12.7  & 0.48  & 56±4  & 0.43±0.05 \\
+& 21.6  & 12.7  & 0.48  & 0.49 & 56±4  & 0.43±0.05 \\
 \Gls{MLR}
-& 21.6  & 12.1  & 0.48  & 58±4  & 0.46±0.06 \\
-\Glspl{NN}
-& 22.7  & 9.2   & 0.43  & 68±1  & 0.57±0.02 \\
+& 21.6  & 12.1  & 0.48  & 0.48 & 58±4  & 0.46±0.06 \\
+\Gls{MLP} \glspl{NN}
+& 22.7  & 9.2   & 0.43  & 0.52 & 68±1  & 0.57±0.02 \\
 \Gls{SVM} regression
-& 20.7  & 8.9   & 0.52  & 69±2  & 0.58±0.03 \\
+& 20.7  & 8.9   & 0.52  & 0.56 & 69±2  & 0.58±0.03 \\
 Cubist regression
-& 18.1  & 8.1   & 0.63  & \textbf{72±2}  & 0.63±0.03 \\
+& 18.1  & 8.1   & 0.63  & 0.65 & \textbf{72±2}  & 0.63±0.03 \\
 \Gls{RF} regression
-& \textbf{17.3}  & 9.4   & \textbf{0.66}  & 67±4  & 0.57±0.05 \\
+& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} & 67±4  & 0.57±0.05 \\
 \ensuremath{''}
-only RS covariates
-& 18.4  & 10.3  & 0.62  & 64±4  & 0.54±0.05 \\
+only RS features
+& 18.4  & 10.3  & 0.62  & 0.64 & 64±4  & 0.54±0.05 \\
 \ensuremath{''} two-step
-& 19.9  & 8.2   & 0.56  & 71±2  & 0.62±0.02 \\
+& 19.9  & 8.2   & 0.56  & 0.60 & 71±2  & 0.62±0.02 \\
 \ensuremath{''} three-step
-& 19.4  & 8.4   & 0.58  & 71±3  & 0.62±0.04 \\
+& 19.4  & 8.4   & 0.58  & 0.60 & 71±3  & 0.62±0.04 \\
 \ensuremath{''} median vote
-& 20.7  & \textbf{7.9}   & 0.52  & 71±1  & 0.63±0.02 \\
+& 20.7  & \textbf{7.9}   & 0.52  & 0.60 & 71±1  & 0.63±0.02 \\
 \ensuremath{''} \ensuremath{''} two-step
-& 20.0  & 8.1   & 0.54  & \textbf{72±1}  & 0.63±0.02 \\
+& 20.0  & 8.1   & 0.54  & 0.60 & \textbf{72±1}  & 0.63±0.02 \\
 \ensuremath{''} \ensuremath{''} three-step
-& 20.2  & \textbf{7.9}   & 0.54  & \textbf{72±2}  & \textbf{0.64±0.02} \\
+& 20.2  & \textbf{7.9}   & 0.54  & 0.60 & \textbf{72±2}  & \textbf{0.64±0.02} \\
+Cubist + RF ensemble
+& 17.7  & 8.6   & 0.65  & 0.65 & 70±3 & 0.61±0.04 \\
 \bottomrule
@@
 %}
-\caption{Accuracy statistics of the models tested. Best performing statistics are highlighted. ``Only RS covariates'' stands for a model trained only with \glspl{VI} and temporal metrics.}
+\caption{Accuracy statistics of the tested models. Best performing statistics are highlighted. ``Only RS features'' stands for a model trained only with \glspl{VI} and temporal metrics.}
 \label{tab-accuracy}
@@
     \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
-    \caption{Comparison of \gls{RMSE} (top) and \gls{MAE} (bottom) per class of the best performing models in their category, and the intercept-only model as a reference.}
+    \caption{Comparison of absolute \gls{RMSE} (top) and absolute \gls{MAE} (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
     \label{fig-models}
@@
 
-There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
-Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
-Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often being partially but not completely incorrect.
-Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
+%There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
+%Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
+%Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often making predictions that are partially but not completely incorrect.
+%Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
 
-All tested models had difficulty discerning the vegetation classes (herbaceous, trees, crops, shrubs) between each other, and less so with the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
-This was exacerbated by the imbalance in class distribution: the urban class is simple to predict well due to the fact that it is rare and very rarely forms a majority, therefore a guess of 0\% is usually a good one.
-In contrast, the tree class had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
-Most models showed good improvement in predicting tree cover compared to the intercept model, indicating that the covariates used were useful in distinguishing tree cover from other classes and quantifying its proportion in the pixel.
-In contrast, there were only moderate improvements for predicting the herbaceous vegetation fraction, as it is more difficult for the classifiers to distinguish this class with the given input covariates.
+All tested models had lower accuracy when estimating the cover fractions of vegetation classes (herbaceous, trees, crops, shrubs), compared to the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
+This was exacerbated by the imbalance in class distribution: the built-up class is rare and very rarely forms a majority, therefore a prediction of 0\% leads to a perfect prediction most of the times, and even if it does not, the error is low.
+In contrast, tree cover had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
+The \gls{RRMSE} and \gls{RMAE} statistics show this (see supplementary material, figure S1): the tree cover prediction has low relative error given its high mean value (32\%), whereas the built-up class is the most challenging to predict according to \gls{RRMSE} given its low mean value (3\%).
+The shrub cover fraction is also very challenging to predict, since it had the highest \gls{RMAE} and none of the models showed large improvements in \gls{RMSE} compared to the equal proportions model.
+There were only moderate improvements for predicting the herbaceous vegetation fraction as well.
+In contrast, most models were a significant improvement in predicting tree, built-up, water and bare cover fractions compared to the equal proportion model, indicating that the features used to train the models were useful to distinguish these classes from the others and to quantify their proportions in each pixel.
 
-\subsection{Adjustment for zero inflation}
+\subsection{Zero inflation adjustment with multi-step models and median voting}
 
-Since \gls{RF} regression resulted in the lowest RMSE, it was used for testing the effect of the multi-step approach detailed in section \ref{sec-multistep}.
-The overall results can be seen at the bottom of table \ref{tab-accuracy}.
+We investigated two ways to adjust for the issue of zero inflation in the training data.
+First, we used a median vote instead of a mean vote for ensembling decision tree votes in \gls{RF} regression, so as to predict exactly 0\% and 100\% more often.
+Second, we used our proposed multi-step model approach as detailed in section \ref{sec-multistep}.
+The overall results can be seen at the bottom of table \ref{tab-accuracy}, whereas per-class results can be seen in figure \ref{fig-randomforest} and supplementary material, figure S2.
 
-For an \gls{RF} regression that uses the mean statistic for ensembling the tree votes, using a two-step approach improved the \gls{MAE} of the model, while increasing \gls{RMSE} (see figure \ref{fig-randomforest}).
-This is because the false positives in the first step of the two-step model (the binary zero/non-zero classification) result in some high fractions nevertheless being set to zero.
-But on average it does improve the accuracy, since without this adjustment, \gls{RF} regression would almost never assign the precise value of 0\% to a fraction, and thus result in some error for all pixels.
-With this adjustment, it is possible to predict the result perfectly in the cases where the fraction is 0\%, which is the most common case.
+The median voting approach resulted in significantly more predictions of 0\% and 100\% class fractions, therefore the output looked closer to a discrete classification map compared to mean voting.
+Since these two most common values were predicted precisely much more often, \gls{MAE} reduced.
+However, in cases of high uncertainty, the median vote was also much more likely to predict a 100\% fraction compared to the mean vote, and thus was also more likely to predict 100\% of the wrong class.
+This resulted in increased \gls{RMSE}.
+Therefore the overall effect of using the median vote is a change in the balance of \gls{RMSE} and \gls{MAE}.
 
-However, the two-step approach combined with binary relevance also introduced a drawback in that there were cases where all land cover class fractions were predicted to be 0\% in the first step.
-In such cases it was not possible to determine the land cover fractions, and to make them sum up to 100\%, they were all set to equal parts ($100\% / 7$), which introduced further error.
+We observed a similar effect when using the multi-step approach for an \gls{RF} regression that uses the mean statistic for ensembling the tree votes.
+Using a two-step approach decreased the \gls{MAE} of the model, while increasing \gls{RMSE}.
+This is because the false positives in the first step of the two-step model (the binary zero/non-zero classification) result in some high fractions nevertheless being predicted as zero.
+The two-step approach combined with binary relevance also introduced a drawback in that there were cases where all land cover class fractions were predicted to be 0\% in the first step.
+In such cases it was not possible to determine the land cover fractions.
+Therefore, to make them sum up to 100\%, we set these cases to equal proportions ($100\% / 7$), which introduced further error.
 
-The three-step approach solved this issue, since the first step would not predict zeroes but rather purity.
-In the case of pure pixels, they would be passed to \gls{RF} classification, which would always return the most likely pure class.
-For mean vote \gls{RF}, the result was not significantly different from the two-step approach.
-While this approach improved also the case of predicting the 100\% fractions precisely, the addition of another model caused extra error that would propagate through the process, negating the gain.
+The three-step approach solved this issue, since the first step predicts purity, rather than zeroes.
+In this approach, pure pixels are passed to \gls{RF} classification, which always predicts the most likely pure class.
+When the three-step approach was applied to mean vote \gls{RF}, the result was a slight decrease in \gls{RMSE} across most classes compared to the two-step approach.
+However, there was also an increase in \gls{MAE} of the predicted crop cover.
+Therefore the three-step model offsets the increase in \gls{RMSE} as seen in the two-step model case, and does not cause a high increase in \gls{MAE}, thus leading to a better balance between the two measures.
 
-Another option explored for dealing with the issue of zero inflation was to use the median statistic for ensembling tree votes in \gls{RF}, instead of the mean.
-This had a comparable effect to using the multi-step approach, as it resulted in a lot more precise predictions for fractions with 0\%.
-However, in the cases with high uncertainty, the median vote was much more likely to predict a 100\% fraction, and thus was also more likely to predict 100\% of the wrong class.
-Thus it resulted in much higher \gls{RMSE} and much lower \gls{MAE}.
-Using a multi-step approach together with median vote resulted in little difference, as the two approaches have a similar effect.
-The three-step approach was an overall improvement to the single-step approach, albeit a small one.
+The median vote had a stronger effect on decreasing \gls{MAE} (and increasing \gls{RMSE}) compared to the multi-step approach with mean voting.
+When both approaches were combined, no further decrease in \gls{MAE} could be achieved, however, the combination of the three-step model and median voting decreased \gls{RMSE} compared to the single-step median model, therefore reducing large errors.
+
+We also investigated the separate steps of the multi-step model in more detail to better understand the accuracy of each model.
+One parameter in the three-step model is the purity threshold: how high should the cover fraction be for a pixel to be considered ``pure'' and be subject to classification rather than regression.
+Decreasing the purity threshold led to a result closer to the one-step model, i.e. reduced \gls{RMSE} and increased \gls{MAE}, as the classification model was used less often.
+The purity binary classifier, when only 100\% cover is considered ``pure'', achieved 78\% overall accuracy, and the accuracy decreased when the purity threshold was decreased.
+The classification model achieved 87\% overall accuracy, but shrub and built-up classes had very low users' accuracy.
+These classes are highly heterogeneous, therefore there were too few observations to train the classifier to identify these classes.
+Decreasing the purity threshold resulted in a lower overall accuracy, but an increase in the users' accuracy in the classifier model for these particular classes.
+The regression step by itself had lower accuracy than the combined three-step model, with 21.38 \gls{RMSE} and 10.49 \gls{MAE} (using median voting), as the middle of the range is the most difficult to predict correctly.
+Decreasing the purity threshold led to a lower \gls{RMSE}, as the value range of the regressor training data gets decreased, but a higher \gls{MAE}, as the amount of training data for the regressor decreased.
+Overall for the whole multi-step model, lowering the purity threshold increased \gls{RMSE} and slightly increased \gls{MAE} as well.
+
+All in all, both median voting and the multi-step approach successfully result in more correctly predicted 0\% and 100\% fractions, thus lowering \gls{MAE} and increasing the \gls{SCM} \gls{OA}.
+While it also increases \gls{RMSE}, combining the two concepts together leads to a lower increase in \gls{RMSE}.
 
@@
     \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-04-rf-comparison-bar}
-    \caption{Comparison of \gls{RF} regression models (intercept model shown as a reference).
+    \caption{Comparison of \gls{RF} regression models (equal proportion model shown as a reference).
 1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
@@
 
-\subsection{Covariate importance}
-
-\Gls{RF} permutation importance was performed to test which variables were important for \gls{RF} regression predictions.
-The single-step \gls{RF} model was selected for this, as it achieved the lowest \gls{RMSE}.
-The results are shown in figure \ref{fig-varimp}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=14cm]{article-figures/heatmaps/2020-06-19-varimp-heatmap-top5}
-    \caption{Random Forest single-step approach variable importance, top 5 covariates per category.
-Categories are ordered by cumulative importance.
-The values represent increase in RMSE for the given class when the given covariate is permuted.}
-    \label{fig-varimp}
-\end{figure}
-
-The overall most important variables were the maximum and minimum \gls{NDVI} over the whole time series of PROBA-V imagery.
-They were followed by the median \gls{NDMI} and \gls{NIRv} over the whole time series.
-These vegetation indices were of lower importance only for the shrub and crop classes.
-These two classes are complicated to distinguish from other natural vegetation, and thus benefited from a wide variety of additional covariates, including location and climate information.
-
-To distinguish crops in particular, soil information such as pH, bulk density and sand fraction was useful for the model.
-Crops are usually grown in fertile soils that can sustain them, and due to the fact that cropland is further managed, the soil is also altered to be more fertile, which affects these soil properties.
-In addition, the crop prediction model benefited from harmonic metrics derived from time series the most.
-The second order amplitude and sine of the harmonic model of the time series allows the model to detect areas with a double harvest throughout the year, which is indicative of crops.
-
-Climate data was also most useful for predicting crops, especially the mean temperature and the closely (exponentially) related mean of water vapour pressure.
-All of the natural vegetation class predictions benefited from (log-transformed) annual precipitation, solar irradiance in January and isothermality information.
-
-Location data was most useful for crop prediction, and to a lesser extent for grassland, shrubland and tree prediction.
-These classes tend to be spatially clustered into large areas.
-Both absolute and regular latitude was useful for the prediction, and latitude was more useful than longitude for the predictions.
-In contrast, location information was not useful for inland water prediction, as it is a rare and spatially dispersed class that takes little area.
-
-Terrain information was the least useful covariate category.
-It was most useful for crops and water, as crops are inconvenient to grow on slopes or in high altitudes.
-Aspect and \gls{TPI} were only useful for the water class, as standing water has no aspect and a \gls{TPI} of zero, as it has a slope of zero too.
-
 \subsection{Spatial predictions and accuracy}
 
-We used the single-step \gls{RF} model to predict the land cover at a global scale to obtain wall-to-wall global land cover fraction maps.
-See figure \ref{fig-walltowall} for a visualisation (sampled at 0.2 degrees) of all of the fraction layers separately, and the supplementary materials for the full resolution raster itself.
+To visually demonstrate the predictions of the model with the highest accuracy by \gls{MAE}, we used the three-step median \gls{RF} model to predict land cover fractions at a global scale (100~m resolution, but sampled every 0.2~degrees).
+See figure \ref{fig-walltowall} for a visualisation of all of the fraction layers separately, and the supplementary materials for the output GeoTIFF file itself.
 The wall-to-wall fraction maps reveal how land cover fraction mapping is capable of expressing gradients and mixed land cover.
 
-In addition, to see whether there were any patterns in the accuracy of the predictions globally, we produced spatial accuracy maps based on the predictions in locations where we had independent validation data.
+In addition, to examine whether there were any patterns in the accuracy of the predictions globally, we produced spatial accuracy maps based on the predictions in locations where we had independent validation data.
 This is also presented in raster format in figure \ref{fig-walltowall}.
 
-The global patterns of the land cover fractions are shown to express biotic gradients.
+Biotic gradients can be seen in the global patterns of the land cover fractions.
 For instance, gradients are visible between communities dominated by shrubs and ones dominated by herbaceous vegetation, such as in south and east Africa.
 Likewise, the gradient of tree cover from 100\% in the African tropics to 0\% in the sub-Sahara region is evident.
+The tropical forest edge appears with a hard edge when using median voting and the three-step approach, as forest occurs in discrete patches due to human activity, rather than changing gradually over space.
+The tree cover in the transition zone towards savannah is much more mixed and gradual.
 Herbaceous cover in sub-Sahara shows an asymmetric gradient: the cover is highest at around 14-15\textdegree{} N and decreases quickly towards the north, becoming zero around 18\textdegree{} N; but decreases slowly towards the south, reaching all the way to 5\textdegree{} N.
-Inland water shows up as more discrete, as it naturally forms discrete patches and mixed pixels with water are more rare.
-Built-up area is relatively rare and rarely forms a 100\% fraction, as urban areas tend to include both built-up area and greenery within the footprint of the pixel.
+Inland water shows up as more discrete, as it naturally forms discrete patches.
+Mixed pixels that include water are uncommon.
+Built-up area is also relatively rare worldwide.
+It rarely forms a 100\% fraction, as urban areas tend to include both built-up area and greenery within the footprint of the 100~m pixel.
 
-The spatial pattern of map accuracy shows that the land cover of areas with pure land cover was the easiest to predict, such as tropical forests (100\% tree cover) and deserts (100\% bare land).
-Conversely, it was more difficult to predict fractions in areas with mixed land cover, as isolating individual fractions from pixel-level information is more challenging.
-In addition, land cover fractions in the extreme latitudes were more difficult to predict as well.
-In these areas less training data was available, owing to the lack of high-resolution imagery there.
+The spatial pattern of map accuracy shows that the land cover fractions in areas with pure land cover, such as tropical forests (100\% tree cover) and deserts (100\% bare land), were predicted with the highest accuracy.
+Conversely, fractions in areas with mixed land cover were predicted less accurately, as isolating individual fractions from pixel-level information is more challenging.
+In addition, land cover fractions in the extreme latitudes were predicted less accurately as well.
+In these areas, less training data was available, owing to the lack of high-resolution imagery there.
 
-The distribution of the predictions was relatively even across the whole range for herbaceous vegetation and trees, but uneven for bare land and especially built-up area.
-Built-up fraction was never predicted over 65\% and thus was more often underestimating in cases where the predicted fraction was above 15\%.
-The predictions of each class showed underestimation of high fractions and overestimation of low fractions, i.e. tending towards the mean.
-As we can see from the box widths of the boxplots in figure \ref{fig-walltowall}, the number of predictions were much more even across the entire range for herbaceous vegetation compared to shrubs.
-Shrub fractions were never predicted above 85\%, and the low fraction predictions dominated, showing that the model expected that shrubs would be mixed with other classes and form small fractions only.
+As we can see from the box widths of the boxplots in figure \ref{fig-walltowall}, the distribution of the predictions was relatively even across the whole range for herbaceous vegetation and trees, but uneven for bare land, inland water and built-up area.
+The number of predictions was much more even across the entire range for herbaceous vegetation compared to shrubs.
+The model overestimated the fractions of trees and herbaceous cover, as the medians of each box are below the 1:1 line almost throughout the entire range, but it underestimated built-up and crop fractions.
+The overall \gls{ME} is below 0.001, which means that overall the model is not biased.
+
+Lastly, we compared the \gls{RF} three-step median model predictions with existing global products that correspond to a particular land cover class in our classification.
+The results are given in table \ref{tab-products}.
+The \gls{RF} three-step median model performed worse than the \gls{GSW} water occurrence history product, but better than the \gls{GFCC} tree cover and \gls{GHSL} built-up products.
+There was a tie when comparing the \gls{RF} three-step median model with the \gls{FROM-GLC} impervious surface product when it comes to \gls{MAE}, but \gls{FROM-GLC} was slightly more accurate according to \gls{RMSE} and less biased according to \gls{ME}.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{article-figures/maps/2020-07-29-walltowall.png}
+    \includegraphics[width=\textwidth]{article-figures/maps/2020-10-12-walltowall.png}
     \caption{Random Forest single model predictions per class.
@@
 
+\begin{table}
+    \centering
+    \adjustbox{max width=\textwidth}{
+    \begin{tabular}{lllllllll}
+    \toprule
+    \textbf{Fraction and source} & \textbf{\glsentryshort{RMSE} (\%)} & \textbf{\glsentryshort{MAE} (\%)} & \textbf{\glsentryshort{ME} (\%)} & \textbf{\glsentryshort{RRMSE}} & \textbf{\glsentryshort{RMAE}} & \textbf{\glsentryshort{RME}} & $\mathbf{R^2_{NSE}}$ & $\mathbf{R^2_{OLS}}$ \\
+    \midrule
+    %Inland water (RF 1-step) &
+    %11.65 & 3.48 & -0.81 &
+    %2.07 & 0.62 & -0.14 &
+    %0.72 & 0.73 \\
+    Inland water (RF 3-step median) &
+    12.90 & 2.25 & -1.17 &
+    2.29 & 0.40 & -0.21 &
+    0.65 & 0.67 \\
+    \glsentryshort{GSW}, year 2015 \citep{pekel_high-resolution_2016} &
+    \textbf{10.22} & \textbf{1.94} & \textbf{-0.63} &
+    \textbf{1.81} & \textbf{0.34} & \textbf{-0.11} &
+    \textbf{0.78} & \textbf{0.78} \\
+    \midrule
+    %Trees (RF 1-step) &
+    %\textbf{19.29} & 12.90 & \textbf{-0.56} &
+    %\textbf{0.61} & 0.41 & \textbf{-0.02} &
+    %\textbf{0.77} & \textbf{0.77} \\
+    Trees (RF 3-step median) &
+    \textbf{21.18} & \textbf{10.68} & \textbf{1.81} &
+    \textbf{0.67} & \textbf{0.34} & \textbf{0.06} &
+    \textbf{0.72} & \textbf{0.74} \\
+    \glsentryshort{GFCC}, epoch 2015 \citep{townshend_global_2017} &
+    28.80 & 18.33 & -12.57 &
+    0.91 & 0.58 & -0.40 &
+    0.48 & 0.61 \\
+    \midrule
+    %Built-up (RF 1-step) &
+    %\textbf{9.80} & 3.01 & -1.11 &
+    %\textbf{3.33} & 1.02 & -0.38 &
+    %\textbf{0.38} & 0.51 \\
+    Built-up (RF 3-step median) &
+    11.28 & \textbf{2.57} & -2.33 &
+    3.83 & \textbf{0.87} & -0.79 &
+    0.19 & 0.22 \\
+    \glsentryshort{GHSL} built-up, 2014 \citep{corbane_ghs-built_2018} &
+    18.67 & 5.43 & 4.80 &
+    6.34 & 1.84 & 1.63 &
+    -1.23 & 0.52 \\
+    \glsentryshort{FROM-GLC} impervious surface, 2015 \citep{gong_annual_2020} &
+    \textbf{11.18} & \textbf{2.57} & \textbf{0.93} &
+    \textbf{3.79} & \textbf{0.87} & \textbf{0.32} &
+    \textbf{0.20} & \textbf{0.65} \\
+    \bottomrule
+    \end{tabular}
+    }
+    \caption{Accuracy comparison between our results and existing global land cover fraction maps. Highest accuracy and lowest bias results are highlighted.}
+    \label{tab-products}
+\end{table}
+
+\subsection{Feature importance}
+
+We performed \Gls{RF} permutation importance on the three-step median model to test which features were important for \gls{RF} predictions.
+The results are shown in figure \ref{fig-varimp}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=14cm]{article/article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
+    \caption{Random Forest single-step approach variable importance, top 5 features per category.
+Categories are ordered by cumulative importance.
+The values represent increase in MAE for the given class when the given feature is permuted.}
+    \label{fig-varimp}
+\end{figure}
+
+The overall most important variables were the maximum and minimum \gls{NDVI} over the whole time series of PROBA-V imagery.
+They were followed by the median \gls{NIRv} and \gls{NDMI} over the whole time series.
+The importance of remote sensing information can also be seen from the accuracy statistics of a single-step \gls{RF} regression model, when only vegetation index and temporal features were input into the model (see table \ref{tab-accuracy}, ``only RS features'').
+After excluding all the static features, \gls{RMSE} and \gls{MAE} increased only by around 1 percentage point.
+
+The remote sensing data was of lower importance (and sometimes even confounding) only for the shrub class.
+This class is complicated to distinguish from other natural vegetation, and thus when trying to predict shrub cover fractions, the model benefited from a wide variety of additional features, including location and climate information.
+
+To distinguish crops in particular, soil information, such as pH, bulk density and sand fraction, was useful for the model.
+Crops are usually grown in fertile soils that can sustain them, and due to the fact that cropland is further managed, the soil is also altered to be more fertile, which affects these soil properties.
+In addition, harmonic metrics derived from time series benefited crop cover fraction estimation the most of all the classes.
+The second order amplitude and sine of the harmonic model of the time series allows the model to detect areas with a double harvest throughout the year, which is indicative of crops.
+
+Climate data, especially the mean temperature and the closely (exponentially) related mean of water vapour pressure, was also most useful for predicting crop cover fractions.
+It was also beneficial for estimating built-up fractions.
+Location features were the most important for predicting the cover fractions of these two classes as well.
+These classes tend to be spatially clustered, when looking at a large scale.
+While the models for predicting the cover fraction of classes made the best use of the location features, all of the classes could benefit from them to some extent, thus it is beneficial to include these intrinsic parameters in the model.
+Both absolute and regular latitude were used by the model to improve prediction accuracy, and latitude was more beneficial than longitude for increasing the prediction accuracy.
+
+Terrain information was the least useful feature category.
+It was most useful for inland water fraction prediction, since it typically has little to no slope, and rarely occurs at high altitudes.
+The aspect was the one feature that does not appear to have contributed to model accuracy.
+
 \section{Discussion}
 
-\subsection{Methods and their accuracies}
+\subsection{Land cover fraction mapping methods}
 
-In this study, we have tested a number of machine learning methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale.
-Since global land cover fraction mapping is a novel field, there are few other studies or products that our results can be directly compared with.
-Most products in this field are made available as layers for a particular land cover class fraction, rather than being thematically comprehensive and guaranteeing a sum of all fractions to 100\%.
-In addition, the documentation for these products rarely report accuracy metrics, and none use an independent validation set for this purpose.
-Lastly, these products are generated for different time scales than our study, both in terms of input time series as well as the time for which the model training data has been collected.
+In this study, we have tested a number of methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale.
+We found that of the methods that we tested, without additional measures to take the data imbalance into account, \gls{RF} regression and Cubist regression produce output of the highest accuracy by \gls{RMSE} and \gls{MAE}, respectively.
+Our findings agree with those of \citet{li_monitoring_2018}, who compared Cubist and \gls{RF} regression for water fraction classification and found that Cubist performs slightly better than \gls{RF} regression for this particular class.
+Their Cubist regression result achieved 7.52\% \gls{MAE} and 10.39\% \gls{RMSE}.
+Our respective results using Cubist regression were 2.57\% \gls{MAE} and 11.51\% \gls{RMSE}.
+While Cubist works best for the inland water class, when considering all of the classes, \gls{RF} regression nevertheless results in higher accuracy (see table \ref{tab-accuracy}).
+The difference in the reported numbers may be due to the differences in scale (regional vs global) and training data balance.
+As we focused on predicting multiple classes, our training dataset intrinsically had higher zero inflation.
 
-Compared to a global land cover product that specialises in a specific land cover class, namely tree cover, the \gls{RF} one-step mean model had a higher \gls{RMSE} (19.2\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (16.8\%).
-However, the methodology for the validation differed significantly from the one used in our study, as it compared the predicted fractions to fractions as computed from lidar datasets within several local study areas, rather than using global expert image interpretation as we did.
-A previous study on the MODIS vegetation continuous fields product by \citet{montesano_modis_2009} used a more similar methodology as we did, i.e.
-expert interpreters estimating cover fractions from a 100 by 100 point subdivisions of pixels using high-resolution imagery, and reported an R² of 0.57, \gls{RMSE} of 13.4, RMSD of 21.3, slope from a linear regression of 0.5 and intercept of 18.4.
-In comparison, our results of a single-step \gls{RF} regression for the tree cover class was 0.77 for both \gls{NSE} and linear regression R², \gls{RMSE} of 19.3, slope of 0.7 and intercept of 8.8.
-Thus while our model appears to perform better, it is important to note that the scopes of the studies were different, as \citet{montesano_modis_2009} only evaluated boreal regions rather than the entire globe, and used MODIS 500 m data rather than PROBA-V 100 m data.
-
-Our results showed that \gls{RF} regression using binary relevance performed the best of all the tested algorithms.
-It was also adaptive to two different goals: minimising \gls{RMSE} (with the default, mean voting) and minimising \gls{MAE} (with median voting).
-Both of these goals may be valuable to different user communities, depending on whether areas with high uncertainty in the predicted land cover should be treated as a mix of covers, or a guess should be made at a single most likely cover.
-The former is more likely to be useful for the modelling community that deals with uncertainty with probabilistic frameworks, and the latter may be more useful for policymakers and land owners who are more concerned with what land cover is most likely to be on the ground.
-
-Our results also showed that the three-step approach introduced in this study, when combined with median voting, resulted in a decrease in \gls{RMSE} without affecting \gls{MAE}.
-Thus, it achieved the best combination that is optimised towards reducing \gls{MAE}.
-The three-step approach can also be used with any algorithm that provides both classification and regression modes, or with two separate unrelated classification and regression algorithms.
-
-Compared to other studies that compared different machine learning algorithms for land cover fraction classification, we have achieved a much lower \gls{MAE}.
-\citet{li_monitoring_2018} compared Cubist and \gls{RF} regression for water fraction classification and achieved \gls{MAE} as low as 7.52\% and \gls{RMSE} of 10.39\% with Cubist.
-Our respective results were 2.57\% \gls{MAE} and 11.51\% RMSE.
-Our findings confirm their conclusion that Cubist works slightly better than \gls{RF} regression for this class (\gls{RF} regression result is 11.65\% \gls{RMSE}, 3.48\% \gls{MAE}), but overall when considering all of the classes, \gls{RF} regression nevertheless edges out (see table \ref{tab-accuracy}).
-One important difference between the studies is that we focused on thematically comprehensive land cover, rather than a particular class, which has an effect on the training data balance.
-Also, our focus was on a global rather than regional scale, which has an effect on the resulting statistics due to higher sample site diversity.
-
-At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which was also observed as performing better than \gls{RF} regression and \glspl{SVM}.
-Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the urban class, and we also observed that Cubist performed slightly better than the other two algorithms for this class.
+At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which he also noted as performing better than \gls{RF} regression and \glspl{SVM}.
+Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the built-up class, and we also noted that Cubist performed slightly better than the other two algorithms for this class.
 Both \citet{okujeni_comparison_2014} and \citet{foody1997fuzzynnet} also performed land cover fraction mapping over an urban area.
-The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF}.
-This could be due to much finer resolution (3.6–9 m) used in their study, and the associated difference in class definitions.
+The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF} regression.
+This could be due to much finer spatial resolution (3.6–9 m) data used in their study, and the associated difference in class definitions.
 \citet{foody1997fuzzynnet} reported an \gls{RMSE} of 16.6\% using \gls{MLP} \glspl{NN}, lower than our results (22.7\%).
@@
 
-\subsection{Variable importance}
+\subsection{Multi-step approach for dealing with data imbalance}
 
-Remote sensing variables were the most important variables in our study, especially the maximum and minimum \gls{NDVI} over the entire time series.
+After determining the baseline methods with the highest, we tested our newly proposed multi-step approach on \gls{RF} regression, comparing the effect with that of median voting instead of mean voting.
+One challenge in land cover fraction prediction is the tendency of the models to favour the mean over the extremes, as that minimises \gls{RMSE} that may arise from incorrect predictions.
+However, that leads to increased fuzziness of the result, where pixels with high uncertainty are marked as a mix of many classes, and fractions of 0\% are rarely, if ever, predicted.
+Our proposed multi-step approach adjusts the balance the other way.
+As it is a combination of one or two classification models and one regression model, it predicts significantly more pure pixels compared to a single regression model.
+
+Therefore, the multi-step approach was successful in reducing \gls{MAE} (and improving the related \gls{SCM} metrics), as the particularly common case of 0\% fractions was captured better.
+On the other hand, it comes at a cost of higher \gls{RMSE} and lower $R^2$.
+This is because in highly uncertain, but pure cases, the model makes a best guess of a pure class, and due to the high uncertainty, the prediction is often incorrect.
+This leads to 100\% error in those cases, which is highly penalised by \gls{RMSE}.
+Due to this effect, the resulting map is closer to a discrete classification map, with less expressive transition gradients between land cover classes.
+
+A similar effect was seen when using techniques such as median voting for \gls{RF} regression.
+Median voting also resulted in more correct 0\% fraction predictions, but likewise increased the chances of predicting 100\% of the wrong class.
+Our results showed that median voting has a stronger effect for reducing \gls{MAE} than the multi-step approach.
+Therefore it may be a suitable choice in cases where it is computationally infeasible to train three models.
+However, when the median voting approach was combined with the three-step approach, the \gls{RMSE} decreased without affecting \gls{MAE}.
+Thus, three-step median \gls{RF} achieved the best combination that is optimised towards reducing \gls{MAE}, doing so without increasing \gls{RMSE} as much as the single-step median vote model does.
+
+These findings show that the use of a multi-step approach depends on what is more important for the user.
+If an occasional prediction of 100\% of the wrong class is acceptable, then a multi-step model provides an overall more accurate result, especially for zero fractions.
+On the other hand, a single-step approach emphasises the strength of land cover fraction mapping by expressing gradual changes over space better, and avoids large errors.
+The latter is more likely to be useful for the modelling community that deals with uncertainty with probabilistic frameworks, and the former may be more useful for policymakers and land owners who are more concerned with what land cover is most likely to be present on the ground.
+In addition, we expect a multi-step approach to be more suitable for fine resolution mapping, where more pure pixels can be expected, and a single-step approach to be more useful for coarse resolution mapping, where mixed pixels are the norm.
+
+It is also worth noting that the multi-step approach is flexible and can be used with any algorithm that provides both classification and regression modes, or with two separate unrelated classification and regression algorithms.
+Therefore there may be some combinations of models that have not been tested yet, but could achieve even higher accuracy.
+
+\subsection{Comparison with global land cover products}
+
+To gain insight into how well our proposed multi-step median vote \gls{RF} model performs, we compared it to existing global land cover fraction products.
+These products, that only focus on a single land cover class, had varying accuracy compared to our model (see table \ref{tab-products}).
+Some products, like \gls{GSW} water occurrence \citep{pekel_high-resolution_2016}, had a higher accuracy.
+Others, like \gls{GFCC} forest cover \citep{townshend_global_2017} and \gls{GHSL} built-up \citep{corbane_automated_2019}, had a lower accuracy.
+The accuracy of the impervious surface class fraction from \gls{FROM-GLC} mostly matches that of the built-up cover fraction of our proposed model.
+
+The accuracies of other global products reported in literature likewise varied compared to our results.
+However, these comparisons are much more difficult to make, as the validation methods and scope vary significantly between the studies.
+For example, our \gls{RF} three-step median model had a higher \gls{RMSE} for the tree class (21.1\%) than the one reported by \citet{sexton_global_2013} for their vegetation continuous fields product (16.8\%).
+However, \citet{sexton_global_2013} validated their data using lidar datasets within several local study areas, rather than using global image interpretation data as we did.
+The study by \citet{montesano_modis_2009} that used a validation approach closer to ours to validate the MODIS tree cover product, reported  an $R^2$ of 0.57, \gls{RMSE} of 13.4\%, RMSD of 21.3\%, slope from a linear regression of 0.5 and intercept of 18.4.
+In comparison, our results for the three-step median \gls{RF} model for the tree cover class had an $R^2_{OLS}$ of 0.74, \gls{RMSE} of 21.1\%, slope of 0.87 and intercept of 5.82.
+Thus while our model appears to perform better, \citet{montesano_modis_2009} only evaluated boreal regions rather than the entire globe, and used MODIS 500 m data rather than PROBA-V 100 m data.
+
+\subsection{Feature importance}
+
+Remote sensing features were the most important features for the multi-step median \gls{RF} model, especially the maximum and minimum \gls{NDVI} over the entire time series.
 Note that these values are taken from a time series that has undergone temporal outlier removal, as detailed in section \ref{sec-temporal-filter}, and thus roughly correspond to the 5th and 95th percentiles of the data without additional temporal filtering.
-Multiple vegetation indices were useful, with \gls{NDMI} and \gls{NIRv} median over the time series being still much more important than any other variable from other groups.
-Nevertheless, even though a lot of the other covariates were of much lower importance, they  contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of covariates that were correlated over 0.9 as explained in section \ref{sec-covariate-selection}).
-Methods based on decision trees help with effectively using covariates that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any covariates as non-informative.
-This is also due to the large variety of land cover classes in the study, since a covariate is useful if it helps predict any of the land cover classes better.
+Multiple vegetation indices were useful for increasing the model accuracy: both \gls{NDMI} and \gls{NIRv} median over the time series were much more important than any other feature from other groups.
+Nevertheless, even though a lot of the other features were of much lower importance, they contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of collinear features as explained in section \ref{sec-covariate-selection}).
+Methods based on decision trees help with effectively using features that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any features as non-informative.
+This is also due to the large variety of land cover classes in the study, since a feature is useful if it helps predict any of the land cover classes better.
 
-The result that including more covariates is beneficial, but remote sensing data is the most important, is in line with the conclusions of e.g.
+The result that including more training features is beneficial, but remote sensing data is the most important, is in line with the conclusions of e.g.
 \citet{li_monitoring_2018} and \citet{hengl_soilgrids250m_2017}.
-Remote sensing data is also unique in that it forms a time series, which both allows us to extract more information from it, such as the temporal metrics used in this study, but also to monitor land cover change over time.
+Remote sensing data is also unique in that it forms a time series, which enables us to both calculate additional temporal metrics and to monitor land cover change over time.
 In addition to remote sensing data, \citet{hengl_soilgrids250m_2017} also noted high importance of climate data, however, it is focused on soils, whereas climate has more effect on long-term processes such as soil formation than on land cover.
+Climate data was the second most important in our case, mostly for the crop fraction estimation, which is also closely linked with soils.
+Note that the feature importance may be different at different scales, e.g. if predictions are made at a finer spatial resolution, such as 30~m or 10~m.
 
@@
 
-One challenge in land cover fraction prediction is the tendency of the models to favour the mean over the extremes, as that minimises \gls{RMSE} that may arise from incorrect predictions.
-However, that leads to increased fuzziness of the result, where pixels with high uncertainty are marked as a mix of many classes, and zero fractions are rarely predicted.
-Our proposed multi-step model adjusts the balance the other way round.
-As it is a combination of one or two classification models and one regression model, the prediction has a lot more pure pixels compared to a single regression model.
-This works very well in reducing \gls{MAE} (and improving the related metrics from the \gls{SCM}), as the especially common case of zero is captured better.
-On the other hand, it comes at a cost of higher \gls{RMSE} and lower \gls{NSE}.
-This is because in highly uncertain cases, the model does a best guess at a pure class, and due to the high uncertainty often the guess is incorrect.
-This leads to 100\% error in those cases, which is highly penalised by \gls{RMSE}.
-Due to this effect, the resulting map is closer to a discrete classification map, with less expressive transition gradients between land cover classes.
-Thus we chose to present the single-model \gls{RF} regression result in figure \ref{fig-walltowall}.
-
-A similar effect is seen when using techniques such as median vote for \gls{RF} regression.
-This allows for predicting zero fractions correctly more often, but likewise more often causes a prediction of 100\% of the wrong class.
-Our results showed that median vote has a stronger effect for reducing \gls{MAE} than the multi-step approach.
-Combining both methods is beneficial to further reduce \gls{RMSE} for median vote predictions.
-
-These findings show that the use of a multi-step approach depends on what is more important for the user.
-If an occasional prediction of 100\% of the wrong class is acceptable, then a multi-step model provides an overall more accurate result, especially for zero fractions.
-On the other hand, a single-step approach emphasises the strength of land cover fraction mapping in being able to express gradual changes over space better, and avoids large errors.
-We expect a multi-step approach to be more suitable for fine resolution mapping, where more pure pixels can be expected, and single-step approach to be more useful for coarse resolution mapping where mixed pixels are the norm.
-
 Machine learning algorithms pose several challenges that are inherent to how the models are constructed.
 The trade-off between minimising \gls{RMSE} and minimising \gls{MAE} comes from the chosen loss function.
-In addition, the result always tends towards the mean in cases of high uncertainty.
-The models predict in areas with a high cover of a fraction, such as shrublands, a lower proportion than expected, whereas in cases with a low cover, such as for fraction of built-up in dark-soil deserts, higher fractions are predicted.
-This is due to a guess of the mean being less penalising than predicting the extremes incorrectly; e.g.
+Typically, in cases of high uncertainty, the loss function is minimised when the predictions tend towards the mean.
+In that case, the models predict in areas with a high cover of a fraction, such as mixed shrublands, a lower fraction of the class than expected, whereas in cases with a low cover, such as for fraction of built-up, higher fractions are predicted than expected.
+This is due to a prediction of the mean being less penalising than predicting the extremes incorrectly; e.g.
 for a case of 50\% shrub cover, predicting 100\% shrubs would be a larger mistake (and thus lead to higher \gls{RMSE}) than predicting 15\% shrubs.
-Likewise, predicting 0\% built-up in dark bare soils risks a case where it truly would be 100\% built-up, so on average predicting 15\% built-up in this case lowers the possible error.
-The three-step approach tilts the balance in the other direction, as the first step determines that the pixel is pure, and the classification step determines that it is more likely to be bare soil than built-up area, but it may also misclassify in other cases.
-If not pure, the regression step still tends to predict towards the mean due to the loss function, so the challenge of underestimating large fractions and overestimating small ones remains.
+Likewise, predicting 0\% built-up in areas covered by dark bare soils risks a case where it truly would be 100\% built-up, so on average predicting 15\% built-up in this case lowers the possible error.
+The three-step approach (or median vote) tilts the balance in the other direction, as e.g.
+the first step determines that the pixel is pure, and the classification step determines that it is more likely to be bare soil than built-up area.
+But there may also be cases when the classifier predicts the wrong class.
+If the pixel is not pure, the regression step still tends to predict towards the mean due to the loss function, so the challenge of underestimating large fractions and overestimating small ones remains.
+However, with the multi-step approach, it is now possible to influence the decision process of the model to tweak it towards the desired outcome.
+In addition, using a median vote in tree-based ensemble models makes the model tend towards the median, which may still be one of the extreme values.
+In that case, the model typically overestimates or underestimates values for most of the value range, but on average is still not biased, as the predicted values are more polarised but balance each other out.
+The multi-step approach could be improved further by exploring the option of using different machine learning models for each step, and by gathering more features that would increase the accuracy of the models at each of the steps.
 
-There is an inherent challenge in land cover mapping for discerning classes that are related, e.g. herbaceous vegetation and shrubs.
-Herbaceous vegetation was particularly difficult to map, in part due to confusion between herbaceous vegetation, cropland and shrubs.
-These classes are similar not just to regression algorithms, but also to expert interpreters, which may lead to higher uncertainties also in the training and validation data for these classes.
-The difference between the definition shrubs and herbaceous vegetation largely comes down to height, therefore dynamic information about vegetation height would allow mapping these classes more accurately.
+There is another inherent challenge in land cover mapping for discerning classes that are related, e.g. herbaceous vegetation and shrubs.
+These classes were particularly difficult to map, in part due to their heterogeneity, and in part due to confusion between herbaceous vegetation, cropland and shrubs.
+As shown in figure \ref{fig-varimp}, specific types of features can be used to discriminate between these classes better.
+Specifically, the identification of herbaceous vegetation is primarily based on vegetation indices, identification of shrubs relies mostly on climate data, whereas cropland identification is more data-intensive and makes use of vegetation indices, temporal metrics, climate, soil and location information.
+These three classes are challenging to discriminate between not just for regression algorithms, but also for expert interpreters, which may lead to higher uncertainties also in the training and validation data for these classes.
+It is even more challenging to discern between these classes if they mix within the area of a single pixel, which is common in grasslands and shrublands, as well as in smallholder agriculture.
+The difference between the definition of trees, shrubs and herbaceous vegetation largely comes down to plant height, therefore dynamic information about vegetation height would allow mapping these classes more accurately.
 However, this would require either photogrammetry techniques (that typically rely on much finer spatial resolution and more overlap between the scenes) to reconstruct vegetation height, or the use of non-optical sensors such as \gls{SAR} interferometry.
-Another way to differentiate between the vegetation classes could be to make use of hyperspectral data, which allows differentiation between different kinds of vegetation based on their photosynthesis processes, which affect light absorption.
-Emerging new high level hyperspectral products, such as \gls{SIF} or \gls{GPP}, such as ones based on the upcoming FLEX satellite, could allow for an easy way to incorporate this extra information into land cover models.
+Another way to differentiate between the vegetation classes could be to make use of hyperspectral data, which allows differentiation between different kinds of vegetation based on e.g.
+their photosynthesis processes or water content, which affect light absorption.
+Emerging new high level hyperspectral products, such as \gls{SIF} or \gls{GPP}, such as ones based on the upcoming FLEX satellite, could allow for a straightforward way to incorporate this extra information into land cover models.
+Hyperspectral data could be useful for better differentiation of non-vegetated classes as well, such as bare soil from urban, e.g.
+by separating the spectral signature of sand from concrete or asphalt.
+In addition, land cover time series information could help track land cover change over time, as the land cover at one time step depends on the land cover at the previous step.
+This information would allow the regressor to limit the predicted values to a smaller range, and thus reduce the noise in the predictions.
 
@@
 For example, the built-up class rarely forms a 100\% fraction.
-That makes it easy to classify, as a fraction of 0\% is in most cases a good guess, but that makes it not useful for user needs.
+That makes it simple to achieve a high prediction accuracy according to absolute statistics, as a fraction of 0\% is in most cases not far off from the true value.
+However, a prediction of 0\% in every pixel makes the fraction map not useful for user needs.
 This challenge is further exacerbated by the training dataset containing relatively few points in built-up areas to begin with.
-As such, having a more balanced training dataset may lead to a better representation of the accuracy of each class and help with class predictions.
+Therefore, having a more balanced training dataset may further increase the accuracy of the models.
+However, the issue of value imbalance within the class will always remain for land cover fraction mapping, therefore the multi-step model approach will be relevant, especially if the legend involves even more classes, or if the land cover is more homogeneous at the level of the mapping unit.
 
 In this study, model optimisation and dataset sizes were also a challenge.
-Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming covariates or replacing missing values).
+Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming features or replacing missing values).
 Since the large dataset size increases the training time for the models, parameters to tune have to be chosen carefully.
@@
 
-The scale of the study poses performance challenges, as covariates that the models use need to be available at a global scale.
-Our results showed that all of the covariate groups used in the study were useful, therefore leaving a covariate out means sacrificing some predictive power.
-The gain would be less time needed for processing, as that covariate would no longer need to be downloaded, preprocessed and processed.
+Another challenging aspect about land cover fraction mapping is the available algorithms that would suit this use case.
+In its core, it is a multivariate regression problem, but most regression algorithms are univariate.
+Our results showed that the binary relevance method works well to facilitate the use of univariate regression methods for this task, as the highest accuracy was achieved by algorithms that used the binary relevance method.
+This further expands the field of possible algorithms that could be used.
+In addition, both the binary relevance method and the multi-step method can accept different algorithms for its submodels, i.e.
+for different classes in the case of binary relevance and for different steps in the case of the multi-step approach.
+There may be a combination of models that would work even better than the models we tested.
+We could not cover the entire range of methods in the scope of this study, therefore there is still room for improvement.
+For example, classical spectral unmixing methods, such as spectral mixture analysis, has been often used in the past (e.g. in \citet{adams_classification_1995}).
+However, due to the limitations of the spectral mixture analysis methods \citep{somers_endmember_2011}, we could not use it with the features we selected (as listed in the supplementary material, table S1), as they do not form a linear mixture representing the land cover fractions.
+Modern machine learning methods are advantageous for their ability to model complex nonlinear relationships between variables and thus make use of all of the features as shown in figure \ref{fig-varimp}.
+
+The scale of the study poses performance challenges, as features that the models use need to be available at a global scale.
+Our results showed that all of the feature groups used in the study were useful, therefore leaving a feature out means sacrificing some predictive power.
+The gain would be less time needed for processing, as that feature would no longer need to be downloaded, preprocessed and processed.
 In addition, model training performance (time for training and memory usage) may be an important consideration for global land cover mapping, as it may limit the scope of what models may be used for this task.
+For example, ensemble learning techniques like the super learner are very resource-intensive and may take weeks to train.
+Therefore it is important to take into account not just the model accuracy, but also whether the improvement of the accuracy is worth the increase in processing time or computing resource usage.
 
 Several more challenges are yet to be tackled in this field, but doing so is becoming more and more feasible over time.
-Finer spatial resolution mapping, such as 10 m mapping using Sentinel-2 data, is a future research direction, where the pixel footprint will be even more likely to be covered by 100\% of a single class.
-Such future developments would be more likely to benefit from a multi-model approach or optimisation for \gls{MAE}.
-Alternatively, finer spatial resolution sensors would allow mapping fractions at coarser resolutions in a more precise way, as it would be possible for the model to directly make use of data from the sensor about subpixels of the mapped pixels.
+%Studies on methods such as ours contribute to increasing the accuracy of land cover products and making them more suitable for target user audiences, without the need of gathering and processing additional data.
+%The three-step method proposed and described in this study has already been adopted in production of the \gls{CGLS-LC100} product.
+%While this study presents the original idea of, and independent research into, the multi-step method, \citet{buchhorn_copernicus_2020} describes how the method was applied in the production process of the \gls{CGLS-LC100} product.
+%
+Finer spatial resolution mapping, such as 10 m mapping using Sentinel-2 data, is a future research direction, where the pixel footprint will more likely cover homogeneous land.
+Therefore, due to an increase in 0\%/100\% fractions, such future developments would be more likely to benefit from a multi-step approach or optimisation for \gls{MAE}.
+The multi-step approach is fully portable to finer scales, but more research is needed to determine the effect of the different scales on the purity of the pixels, and how much benefit does a combination of classification and regression bring compared to doing only classification or only regression.
+Finer spatial resolution sensors can also be used for mapping fractions at coarser resolutions in a more precise way, by performing aggregation of the finer resolution pixels to estimate the land cover fraction at a coarser resolution.
+This is likely to become the norm if even finer resolution data (e.g. 1~m) becomes available globally.
 
@@
 As PROBA-V only measures four spectral bands, the amount of information that can be retrieved from them is limited.
-Sensors such as Sentinel-2 \gls{MSI} would allow for computing a more diverse range of \glspl{VI}, such as ones based on the slope of the red edge, potentially allowing for a better distinction between different vegetation types.
+Sensors such as Sentinel-2 \gls{MSI} have a much wider range of spectral bands that could be used both directly as features, as well as enable computing a more diverse range of \glspl{VI}, such as ones based on the slope of the red edge.
+This could potentially improve the distinction between different land cover classes.
 
-With more availability of such additional remote sensing data, it becomes increasingly more feasible to perform land cover change mapping, to make it possible to monitor land cover changes over time.
-Land cover fraction information would allow better tracking of gradual changes, such as regrowth, over time.
+With more availability of such additional remote sensing data, it becomes increasingly more feasible to perform land cover monitoring and change mapping.
+Mapping land cover fractions is a great opportunity to track gradual changes, such as regrowth, better.
 The challenge here is that the higher uncertainty about fraction estimates may cause the time series of land cover fractions to fluctuate, making it difficult to determine robust trends.
@@
 
-In this study, we compared three groups of algorithms for land cover fraction mapping and created a demonstration map showing the global distribution of land cover in fraction layers, using the best performing model according to \gls{RMSE}.
-We also introduced a multi-step approach to tackle the issue of accurately predicting the extreme fraction values of 0\% and 100\%.
+In this study, we compared multiple groups of algorithms for land cover fraction mapping.
 The best performing models were based on \gls{CART} tree ensembles, with \gls{RF} regression performing the best overall.
-A single-step \gls{RF} regression model using majority voting resulted in the best results for RMSE (17.3\%) and \gls{NSE} (0.66).
-Using median voting combined with the three-step approach obtained the best results for \gls{MAE} (7.9\%) and \gls{SCM} \gls{OA} (72±2\%).
-Remote sensing covariates were the most important in the models, although all other types of covariates (climate, soil, terrain) also contributed to the model significantly.
+A single-step \gls{RF} regression model using majority voting resulted in the best results for RMSE (17.3\%) and $R^2_{NSE}$ (0.66).
+In addition, we investigated ways to tackle the issue of accurately predicting the extreme fraction values of 0\% and 100\% by proposing a hierarchical multi-step approach combining classification and regression models.
+We compared and combined this approach with \gls{RF} median voting.
+The combined \gls{RF} median three-step approach obtained the best results for \gls{MAE} (7.9\%) and \gls{SCM} \gls{OA} (72±2\%).
+Based on this model, we created a demonstration map showing the global distribution of land cover in separate land cover fraction layers.
+Remote sensing features were the most important for model accuracy, although all other types of features (climate, soil, terrain) also contributed significantly.
 
-These findings directly contribute towards the operationalisation of global land cover fraction mapping by analysing thematically exhaustive global land cover fraction mapping using currently available methods.
-Information on land cover fractions offers better precision than discrete land cover maps, and allows the users to define their own thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
+These findings directly contribute to the operationalisation of global land cover fraction mapping by analysing and advancing currently available methods for thematically exhaustive global land cover fraction mapping.
+Information on land cover fractions offers better precision than discrete land cover maps, and allows the users to manually define thresholds to generate discrete classifications of their own choosing, based on their classes on interest.
 Furthermore, given the recent advances in optical sensor spatial resolution and the resulting increase in pixel purity, the multi-step model approach may become more important in the future.
@@
 In addition, we thank Bruno Smets for ideas and support from the \gls{CGLS-LC100} project, Benjamin Kellenberger for help on tuning \gls{NN} parametrisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
+We also thank the three anonymous reviewers for their valuable and detailed comments.
 

### 2020-11-18 article/highlights.tex
Draft 2 of first review round

@@
 \begin{itemize}
-    \item 3 machine learning model groups tested for deriving 7 land cover fractions globally
+    \item 4 machine learning model groups tested for deriving 7 land cover fractions globally
+    \item Multi-step approach proposed for dealing with zero inflation in fraction data
     \item Random forest regression performed the best
     \item Optical remote sensing data was the most important input for model predictions
-    \item Independent validation showed highest accuracies for built-up and water fractions
     \item Multi-fraction land cover mapping shown as an alternative to discrete classification

### 2020-11-19 article/article.tex
Fix paths to graphics
Minor fixes

@@
     \centering
-    \includegraphics[width=14cm]{article/article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
-    \caption{Random Forest single-step approach variable importance, top 5 features per category.
+    \includegraphics[width=14cm]{article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
+    \caption{Random Forest three-step median approach variable importance, top 5 features per category.
 Categories are ordered by cumulative importance.
@@
 
-After determining the baseline methods with the highest, we tested our newly proposed multi-step approach on \gls{RF} regression, comparing the effect with that of median voting instead of mean voting.
+After determining the baseline methods with the highest accuracy, we tested our newly proposed multi-step approach on \gls{RF} regression, comparing the effect with that of median voting instead of mean voting.
 One challenge in land cover fraction prediction is the tendency of the models to favour the mean over the extremes, as that minimises \gls{RMSE} that may arise from incorrect predictions.

### 2021-02-04 article/article.tex
Move parts to appendix

@@
 \newacronym{ME}{ME}{mean error}
+\newacronym{OLS}{OLS}{ordinary least squares}
 \newacronym{RRMSE}{RRMSE}{relative root mean squared error}
@@
 
-We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping.
-For the model with the lowest MAE, we assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
+A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated due to how common the 0\% fraction is, but as regression favours the mean, 0\% and 100\% fractions are difficult to predict accurately.
+We proposed a new solution by combining three models: a binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.
+
+We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping, and for our proposed three-step model, assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
 The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points, provided by the \gls{CGLS-LC100} project.
@@
 
-A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated, but regression favours the mean, thus 0\% and 100\% fractions are difficult to predict accurately.
-We proposed a new solution by combining three models.
-A binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.
-
-Results showed that from the compared models, random forest regression reached the lowest RMSE of 17.3\%.
-Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest with our proposed three-model approach.
-This model was driven primarily by optical remote sensing data, followed by climate data.
-
-This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions.
+Results showed that the random forest regression model reached the lowest RMSE of 17.3\%.
+Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest regression with our proposed three-model approach and median vote.
+This model was driven primarily by optical remote sensing data.
+
+This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions, and tuned for user needs.
 Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which enables customising the result for a diverse range of uses.
@@
 
-In addition to classification methods, one aspect that is important for classification accuracy is the data (features) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
-In addition to optical features (such as spectral bands), temporal features (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
-\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are commonly used to improve land cover class separability, e.g.
-for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
-Elevation features (elevation, slope, aspect) are also known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
-%In this paper, we also used such features for increasing the accuracy of land cover fraction regression model predictions.
-%, and their importance for the predictions was analysed.
-
 Lastly, model accuracy assessment by itself is often challenging, especially at a global scale.
@@
 
-In this study we investigated and compared a variety of methods for performing global land cover fraction mapping.
-Furthermore, we proposed a method to enhance land cover fraction mapping further by dealing with data imbalance, which is inherent in land cover fraction training data, and tested it on the method that achieved the highest accuracy in the method comparison.
-Lastly, we investigated the effect of a wide variety of input features on each mapped class fraction, to gain insight into what data was key for the model with the highest accuracy to achieve the best result.
-Therefore the objectives of our study were to:
+In this study, after investigating and comparing a variety of methods for performing global land cover fraction mapping, we proposed a method to enhance land cover fraction mapping by dealing with the inherent data imbalance issue.
+We assessed the accuracy of the result both thematically and spatially, by calculating model accuracy spatially across the globe.
+Lastly, we investigated the effect of input features on each mapped class fraction.
+Therefore the primary objectives of our study were to:
 
 \begin{enumerate}
- \item Compare the performance of a variety of machine learning regression models for mapping fractions of land cover on a global scale.
  \item Investigate approaches for reducing bias in the model predictions with regards to zero inflation and predictions tending towards the mean.
- \item Analyse the importance of features input into the highest accuracy model for the prediction of each land cover fraction.
  \item Assess the accuracy of the models from a thematic and spatial point of view, comparing it to existing global land cover products.
@@
 
-To train the models and predict land cover fractions in unsampled locations, six groups of features were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see supplementary material, table S1).
+To train the models and predict land cover fractions in unsampled locations, six groups of features were used: vegetation indices, temporal metrics, terrain metrics, soil metrics, climate metrics and location data (see \ref{sec-inputdata}, table \ref{tab-inputdata}).
 These features had to be preprocessed before they could be input into the models.
@@
 We used these sine, cosine, trend and intercept components, as well as the phase and amplitude of the two harmonic orders, as temporal features in our study.
-For a list of features that were ultimately used as input to the models, see supplementary material, table S1.
+For a list of features that were ultimately used as input to the models, see \ref{sec-inputdata}, table \ref{tab-inputdata}.
 
@@
 These features include data from each of the feature categories.
-See supplementary material, table S1 for an overview of all of the features that remained and thus used in model training and prediction.
+See \ref{sec-inputdata}, table \ref{tab-inputdata} for an overview of all of the features that remained and thus used in model training and prediction.
 
@@
 In addition, as a baseline we also compared the results with a trivial equal proportion model (all fractions always predicted to be equal, namely $\frac{100\%}{7}\approx14.29\%$).
+Any useful model has to perform better than the equal proportion model, and by comparison to it, it is possible to tell how much better a model performs in estimating the fraction of each class.
 We tuned each algorithm to select optimal parameters, and postprocessed the output of each algorithm as necessary to ensure that all land cover fractions in each pixel add up to 100\%.
@@
 All of the model building and data analysis was performed using the free and open-source statistical software R \citep{r_2019}.
+The results showed that \gls{RF} regression performed the best, with the lowest \gls{RMSE} value of 17.3\%.
+For more technical details and in-depth results of the regression model accuracy comparison, see \ref{sec-model-comparison}.
 
@@
 \centering
-    \begin{tabular}{m{4cm}ll}
+  \begin{adjustbox}{max width=\textwidth}
+    \begin{tabular}{m{4cm}lll}
          \toprule
-         \textbf{Category} & \textbf{Name} & \textbf{Reference} \tabularnewline
+         \textbf{Category} & \textbf{Name} & \textbf{Reference} & \textbf{R package and authors} \tabularnewline
          \midrule
@@
          \end{tabular}
+         &
+         \begin{tabular}{@{}l@{}}
+         \cran{GSIF} \citep{hengl2004fuzzycmeans} \tabularnewline
+         stats \citep{r_2019} \tabularnewline
+         \cran{pls} \citep{pls} \tabularnewline
+         \cran{glmnet} \citep{glmnet} \tabularnewline
+         \cran{nnet} \citep{nnet}
+         \end{tabular}
          \tabularnewline \midrule
@@
          \end{tabular}
+         &
+         \begin{tabular}{@{}l@{}}
+         \cran{ranger} \citep{ranger} \tabularnewline
+         \cran{Cubist} \citep{cubistpackage}
+         \end{tabular}
          \tabularnewline \midrule
@@
          \citealt{suykens_least_1999}
+         \end{tabular}
+         & 
+         \begin{tabular}{@{}l@{}}
+         \cran{keras} \citep{keras} \tabularnewline
+         \cran{liquidSVM} \citep{liquidSVM}
          \end{tabular}
@@
          \citealt{laan_super_2007}
+         &
+         \cran{sl3} \citep{sl3}
          \tabularnewline \bottomrule
     \end{tabular}
+  \end{adjustbox}
     \caption{List of regression methods for land cover fraction estimation tested in this study.}
@@
 \end{table}
-
-\subsubsection{Linear models}
-
-We chose to compare five types of linear models, to have a baseline for a comparison with the nonlinear machine learning models.
-The most simple model we selected was the \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression.
-It is an extension to the standard linear regression that allows for multiple outcomes, and is implemented in base R \citep{r_2019}.
-Next, we tested two linear models that include input data regularisation in the model itself: lasso regression \citep{tibshirani_regression_1996}, implemented in the \cran{glmnet} package \citep{glmnet}, and \gls{PLS} regression \citep{wold_pls-regression_2001}, implemented in the \cran{pls} package \citep{pls}.
-We also tested \gls{MLR} \citep{theil_multinomial_1969}, provided by the \cran{nnet} package \citep{nnet}.
-\gls{MLR} is usually used for classification and is fit using land cover class labels rather than fractions, but the output includes probabilities for each class that add up to 100\%.
-We fit \gls{MLR} using the dominant land cover class as a label for the pixel and used the class probabilities as a proxy for land cover fractions.
-Lastly, we tested \gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, which is provided in the \cran{GSIF} package \citep{hengl2004fuzzycmeans}.
-It is a simple regression method, where the land cover fractions in a pixel are determined by the distance of the pixel from the centroids of each class in feature space.
-
-%The \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to the standard linear regression that allows for multiple outcomes.
-%In this study, we used a \gls{GLM} as a baseline model.
-%We also tested whether dropping any additional features from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction \citep{greenhouse_methods_1959} to compare models with different features.
-%However, only three features could be dropped and it did not improve the predictions, therefore we decided to keep all the features in the model.
-%\Gls{GLM} is implemented in base R \citep{r_2019}.
-
-%Ridge regression is a linear model regularisation method which reduces large coefficients of model features in order to prevent overfitting.
-%Lasso regression is based on the same principle, except it reduces the coefficients of each model feature to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
-%Elastic net regression is a method combining the two, where model coefficients may be reduced or set to zero.
-%This has the effect of the coefficients of correlated features being set to a similar value.
-%We chose lasso regression \citep{tibshirani_regression_1996}, so that it would be possible to identify if there are any features that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
-%The $\lambda$ parameter was chosen using cross-validation.
-%The lasso regression was performed using the \cran{glmnet} package \citep{glmnet}.
-
-%\Gls{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input features into latent variables that explain the most variance of the response variable in multidimensional space.
-%As such, it is well suited for multicollinear data.
-%\ac{PLS} regression was performed using the package \cran{pls} \citep{pls}.
-
-%All of the aforementioned linear models do not normalise the output to add up to 100\%, and also allow negative predictions.
-%Therefore for these models, all predictions < 0 were set to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
-
-%\Gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
-%This method always produces output that sums to 100\%.
-%In this study, we used an implementation adapted from the package \cran{GSIF} \citep{hengl2004fuzzycmeans}.
-%We tested both using a logistic regression to estimate the class centroids, as well as using a weighted average of the training data to do so, and ended up using the latter method as the result was more accurate.
-
-%Lastly, we tested \gls{MLR} \citep{theil_multinomial_1969}.
-%While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
-%This probability can be used as a proxy for the fraction of the pixel that the class covers.
-%These probabilities always add up to 100\%.
-%However, an \gls{MLR} can only be trained on labelled data, rather than fractions.
-%Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
-%This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
-%The logistic regression implementation was provided by the \cran{nnet} package \citep{nnet}.
-
-\subsubsection{Non-tree machine learning methods}
-
-\Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
-In this study, after performing tuning, we ended up using a \glsentryfull{MLP} \citep{dreyfus_artificial_1990} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
-We used the Nadam optimiser \citep{dozat_incorporating_2016} with \gls{MAE} as the loss function to optimise the \gls{NN}, and softmax activation for the output.
-The models were trained using the \cran{keras} package \citep{keras}, built upon TensorFlow that enables the use of a graphics processing unit to accelerate the \gls{NN} training process.
-
-\Glsentryfullpl{SVM} are machine learning models that attempt to find the optimal boundary between the class clusters in feature space by constructing a dividing hyperplane.
-For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM} \citep{suykens_least_1999}.
-As \gls{SVM} models are univariate, we used the binary relevance method \citep{karalas2016br}: training separate models per class that predict a single class, and then combining the results.
-We found the package \cran{liquidSVM} \citep{liquidSVM} suitable for this purpose, as it is optimised for processing large datasets such as ours.
-
-\subsubsection{Tree-based machine learning methods}
-
-\Glsentryfull{RF} regression is a popular method for land cover classification that works by building a number of \gls{CART} decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual decision trees \citep{breiman2001random}.
-\ac{RF} is univariate, therefore we again used the binary relevance approach.
-The processing was done using the \cran{ranger} package \citep{ranger}.
-
-Lastly, we tested Cubist regression \citep{cubist}.
-It is based on \gls{RF} regression, but instead of using a threshold of a feature to split the decision tree, Cubist uses a linear regression based on a subset of the data relevant for the split in question.
-In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previously generated ones.
-Model tuning led to us using 10 committees.
-Cubist predictions were also made using the binary relevance method.
-The algorithm is provided by the \cran{Cubist} package \citep{cubistpackage}.
-
-\subsubsection{Ensemble learning}
-
-Lastly, we made an ensemble from the two machine learning models that produced the lowest \gls{RMSE} and \gls{MAE} respectively.
-We used the \cran{sl3} package \citep{sl3} to create a hierarchical ensemble, where the two models were cross-validated using 10-fold cross-validation to obtain relative weights of each model for each land cover class, following the super learner algorithm \citep{laan_super_2007}.
-The predictions of the two models, along with the weights of the models, were then used as input features for another \gls{RF} regression metalearner.
-The output from this metalearner is the final prediction of the whole ensemble.
-As the ensembled methods are univariate, we used the binary relevance method in this case as well.
-
-%\subsubsection{Other methods}
-
-%Additional methods were considered, but in the end turned out to be unworkable with our dataset.
-%Multivariate \gls{RF} is available in R provided by the \cran{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
-%However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
-
-%Another model that was considered was \cran{SuperLearner} \citep{SuperLearner}, an ensembling method that can combine multiple of the tested methods into one model, giving models different weights according to how well they are known to perform for a given task.
-%Here we ran into similar issues, as cross-validation is required for each model to train, which increases processing time dramatically.
-%Attempts to reduce training time by lowering the number of folds resulted in worse results than when using a single classifier.
-
-%Lastly, multiple-endmember spectral unmixing using non-negative least squares \citep{franc2005sequential} was also attempted, but the results were significantly worse than even the equal proportions model, possibly because the class fractions are not a linear combination of the features.
 
@@
 To assess the performance of the models, we used a number of statistical measures.
-We started with the statistics of assessing land cover fraction model accuracy that are the most commonly used in this field: \glsentryfull{RMSE}, \glsentryfull{MAE} and \glsentryfull{ME}.
+We started with the statistics of assessing land cover fraction model accuracy that are the most commonly used in this field: \glsentryfull{RMSE}, \gls{MAE} and \gls{ME}.
 \gls{MAE} represents the average difference between the predicted and the reference land cover fractions.
@@
 The value range of \gls{NSE} is $(-\infty, 1]$, where a value of 0 means that the predicted values are no better than predicting the mean value.
-The second way is to calculate the $R^2$ of an Ordinary Least Squares regression that estimates the intercept and slope, rather than presetting it.
+The second way is to calculate the $R^2$ of an \gls{OLS} regression that estimates the intercept and slope, rather than presetting it.
 This method always gives higher $R^2$ values than \gls{NSE}, as it allows the regression line to have more flexibility.
@@
 The 1:1 line indicates how well do the distributions of the predictions and the reference data match.
+We also created additional plots showing how \gls{RMSE}, \gls{MAE} and \gls{ME} change over these bins.
 
@@
 
-\subsection{Feature importance}
-
-To calculate feature importance, we performed permutation importance on the model that had the lowest \gls{MAE}.
-We shuffled the values of each feature in turn, and the models made predictions based on all of the features, including the shuffled one.
-This results in a decrease in the accuracy of predictions, as the feature in question no longer contains meaningful information.
-We then recorded the resulting increase in \gls{MAE} compared to the validation set as a measure of feature importance, both per class and overall for all classes combined.
+Lastly, we investigated how features affect model accuracy. See \ref{sec-varimp} for more information on feature importance.
 
@@
 
-\subsection{Method accuracy comparison}
-
-The overall accuracy statistics of the compared models are reported in table \ref{tab-accuracy}, and per-class statistics in figure \ref{fig-models}.
+\subsection{Regression model accuracy comparison}
+
 Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote, and by \gls{MAE} when using a median vote.
-The combination of the three-step model and the median vote resulted in a decrease in \gls{RMSE} without impacting \gls{MAE}.
-
-The results show that the $R^2$ statistics, especially $R^2_{NSE}$, are in agreement with the \gls{RMSE} statistics, and the \glsentryfull{OA} from the \glsentryfull{SCM} is in agreement with the \gls{MAE} statistics.
-$R^2$ is a representation of how closely the predictions correlate with the validation data, therefore big outliers have a large effect on the value.
-The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
-
-The baseline models that performed the best were the two tree-based machine learning models: \gls{RF} regression and Cubist.
-Both of these models are univariate, therefore predictions per class had to be made using the binary relevance method (one model per class).
-This shows that despite the limitations of the binary relevance method, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and the need for a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models and the flexibility of fitting the models for each class outweigh the disadvantages of the binary relevance method.
-
-The super learner method that combines both \gls{RF} regression and Cubist regression resulted in a model that is in between the two ensembled models in terms of accuracy.
-Its \gls{RMSE} was below that of Cubist regression, but above that of \gls{RF} regression.
-Conversely, the \gls{MAE} was below that of \gls{RF} regression (single model), but above that of Cubist regression.
-This shows that the metalearner (another \gls{RF} model) could not differentiate between the two ensembled models well enough to select the model that is the most accurate for a given class.
-Rather, it weighted in both of the models' predictions, losing the advantages of the models when taken separately.
-
-The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed the worst, as they are limited to a linear approach.
-The results of all of the linear models were very similar.
-There was no added value to \gls{PLS} and Lasso regressions over the basic \gls{GLM}.
-This is likely due to the additional preprocessing step of feature selection, as detailed in section \ref{sec-covariate-selection}.
-Lasso and \gls{PLS} regressions are extensions to \gls{GLM} that include a regularisation step, which proved to be unnecessary if all of the highly correlated features are already manually removed.
-
-\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
-This indicates that the shape of the feature data in feature space is too complex to capture merely with a centroid approach.
-This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
-
-\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods.
-Nevertheless, it performed better than all of the multivariate linear models.
-
-\begin{table}
-\centering
-%\adjustbox{max width=0.5\textwidth}{
-\begin{tabular}{lllllll}
-\toprule
-\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & $\mathbf{R^2_{NSE}}$ & $\mathbf{R^2_{OLS}}$ & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
-\midrule
-Equal proportions
-& 29.9  & 21.4  & 0     & — & 26±5  & 0.13±0.07 \\
-\Gls{FNC}
-& 24.4  & 13.5  & 0.33  & 0.35 & 53±4  & 0.42±0.06 \\
-\Gls{GLM}, \Gls{PLS}, Lasso
-& 21.6  & 12.7  & 0.48  & 0.49 & 56±4  & 0.43±0.05 \\
-\Gls{MLR}
-& 21.6  & 12.1  & 0.48  & 0.48 & 58±4  & 0.46±0.06 \\
-\Gls{MLP} \glspl{NN}
-& 22.7  & 9.2   & 0.43  & 0.52 & 68±1  & 0.57±0.02 \\
-\Gls{SVM} regression
-& 20.7  & 8.9   & 0.52  & 0.56 & 69±2  & 0.58±0.03 \\
-Cubist regression
-& 18.1  & 8.1   & 0.63  & 0.65 & \textbf{72±2}  & 0.63±0.03 \\
-\Gls{RF} regression
-& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} & 67±4  & 0.57±0.05 \\
-\ensuremath{''}
-only RS features
-& 18.4  & 10.3  & 0.62  & 0.64 & 64±4  & 0.54±0.05 \\
-\ensuremath{''} two-step
-& 19.9  & 8.2   & 0.56  & 0.60 & 71±2  & 0.62±0.02 \\
-\ensuremath{''} three-step
-& 19.4  & 8.4   & 0.58  & 0.60 & 71±3  & 0.62±0.04 \\
-\ensuremath{''} median vote
-& 20.7  & \textbf{7.9}   & 0.52  & 0.60 & 71±1  & 0.63±0.02 \\
-\ensuremath{''} \ensuremath{''} two-step
-& 20.0  & 8.1   & 0.54  & 0.60 & \textbf{72±1}  & 0.63±0.02 \\
-\ensuremath{''} \ensuremath{''} three-step
-& 20.2  & \textbf{7.9}   & 0.54  & 0.60 & \textbf{72±2}  & \textbf{0.64±0.02} \\
-Cubist + RF ensemble
-& 17.7  & 8.6   & 0.65  & 0.65 & 70±3 & 0.61±0.04 \\
-\bottomrule
-\end{tabular}
-%}
-\caption{Accuracy statistics of the tested models. Best performing statistics are highlighted. ``Only RS features'' stands for a model trained only with \glspl{VI} and temporal metrics.}
-\label{tab-accuracy}
-\end{table}
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
-    \caption{Comparison of absolute \gls{RMSE} (top) and absolute \gls{MAE} (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
-    \label{fig-models}
-\end{figure}
-
-%There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
-%Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
-%Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often making predictions that are partially but not completely incorrect.
-%Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
-
-All tested models had lower accuracy when estimating the cover fractions of vegetation classes (herbaceous, trees, crops, shrubs), compared to the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
-This was exacerbated by the imbalance in class distribution: the built-up class is rare and very rarely forms a majority, therefore a prediction of 0\% leads to a perfect prediction most of the times, and even if it does not, the error is low.
-In contrast, tree cover had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
-The \gls{RRMSE} and \gls{RMAE} statistics show this (see supplementary material, figure S1): the tree cover prediction has low relative error given its high mean value (32\%), whereas the built-up class is the most challenging to predict according to \gls{RRMSE} given its low mean value (3\%).
-The shrub cover fraction is also very challenging to predict, since it had the highest \gls{RMAE} and none of the models showed large improvements in \gls{RMSE} compared to the equal proportions model.
-There were only moderate improvements for predicting the herbaceous vegetation fraction as well.
-In contrast, most models were a significant improvement in predicting tree, built-up, water and bare cover fractions compared to the equal proportion model, indicating that the features used to train the models were useful to distinguish these classes from the others and to quantify their proportions in each pixel.
+For more details about the results of the accuracy comparison between the tested regression models, see \ref{sec-model-comparison}.
 
@@
 Second, we used our proposed multi-step model approach as detailed in section \ref{sec-multistep}.
-The overall results can be seen at the bottom of table \ref{tab-accuracy}, whereas per-class results can be seen in figure \ref{fig-randomforest} and supplementary material, figure S2.
+The overall results can be seen in table \ref{tab-accuracy}, whereas per-class results can be seen in figure \ref{fig-randomforest} and the relative statistics in \ref{sec-randomforest-relative}, figure \ref{fig-randomforest-relative}.
 
@@
 
+\begin{table}
+\centering
+%\adjustbox{max width=0.5\textwidth}{
+\begin{tabular}{lllllll}
+\toprule
+\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & $\mathbf{R^2_{NSE}}$ & $\mathbf{R^2_{OLS}}$ (slope/int) & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
+\midrule
+\Gls{RF} regression
+& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} (1.09/-1.22) & 67±4  & 0.57±0.05 \\
+\ensuremath{''} two-step
+& 19.9  & 8.2   & 0.56  & 0.60 (0.78/3.48) & 71±2  & 0.62±0.02 \\
+\ensuremath{''} three-step
+& 19.4  & 8.4   & 0.58  & 0.60 (0.84/2.34) & 71±3  & 0.62±0.04 \\
+\ensuremath{''} median vote
+& 20.7  & \textbf{7.9}   & 0.52  & 0.60 (0.74/3.80) & 71±1  & 0.63±0.02 \\
+\ensuremath{''} \ensuremath{''} two-step
+& 20.0  & 8.1   & 0.54  & 0.60 (0.77/3.67) & \textbf{72±1}  & 0.63±0.02 \\
+\ensuremath{''} \ensuremath{''} three-step
+& 20.2  & \textbf{7.9}   & 0.54  & 0.60 (0.77/3.34) & \textbf{72±2}  & \textbf{0.64±0.02} \\
+\bottomrule
+\end{tabular}
+%}
+\caption{Accuracy statistics of multi-step models. Best performing statistics are highlighted. ``slope'' refers to the OLS-estimated slope, ``int'' refers to the OLS-estimated intercept.}
+\label{tab-accuracy}
+\end{table}
+
 \begin{figure}
@@
 To visually demonstrate the predictions of the model with the highest accuracy by \gls{MAE}, we used the three-step median \gls{RF} model to predict land cover fractions at a global scale (100~m resolution, but sampled every 0.2~degrees).
-See figure \ref{fig-walltowall} for a visualisation of all of the fraction layers separately, and the supplementary materials for the output GeoTIFF file itself.
+See figure \ref{fig-walltowall} for a visualisation of all of the fraction layers separately, and the supplementary material for the output GeoTIFF file itself.
 The wall-to-wall fraction maps reveal how land cover fraction mapping is capable of expressing gradients and mixed land cover.
@@
 The overall \gls{ME} is below 0.001, which means that overall the model is not biased.
+For an alternative visualisation showing the change of \gls{RMSE}, \gls{MAE} and \gls{ME} overall and per class, see \ref{sec-apu}, figure \ref{fig-apu}.
+
+We also analysed the effect that the training features had on the model prediction, and found that features obtained from remote sensing were the most important, but all of the feature categories contributed to improving the model predictions.
+For more detailed information, see \ref{sec-varimp}.
 
@@
 
-\subsection{Feature importance}
-
-We performed \Gls{RF} permutation importance on the three-step median model to test which features were important for \gls{RF} predictions.
-The results are shown in figure \ref{fig-varimp}.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=14cm]{article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
-    \caption{Random Forest three-step median approach variable importance, top 5 features per category.
-Categories are ordered by cumulative importance.
-The values represent increase in MAE for the given class when the given feature is permuted.}
-    \label{fig-varimp}
-\end{figure}
-
-The overall most important variables were the maximum and minimum \gls{NDVI} over the whole time series of PROBA-V imagery.
-They were followed by the median \gls{NIRv} and \gls{NDMI} over the whole time series.
-The importance of remote sensing information can also be seen from the accuracy statistics of a single-step \gls{RF} regression model, when only vegetation index and temporal features were input into the model (see table \ref{tab-accuracy}, ``only RS features'').
-After excluding all the static features, \gls{RMSE} and \gls{MAE} increased only by around 1 percentage point.
-
-The remote sensing data was of lower importance (and sometimes even confounding) only for the shrub class.
-This class is complicated to distinguish from other natural vegetation, and thus when trying to predict shrub cover fractions, the model benefited from a wide variety of additional features, including location and climate information.
-
-To distinguish crops in particular, soil information, such as pH, bulk density and sand fraction, was useful for the model.
-Crops are usually grown in fertile soils that can sustain them, and due to the fact that cropland is further managed, the soil is also altered to be more fertile, which affects these soil properties.
-In addition, harmonic metrics derived from time series benefited crop cover fraction estimation the most of all the classes.
-The second order amplitude and sine of the harmonic model of the time series allows the model to detect areas with a double harvest throughout the year, which is indicative of crops.
-
-Climate data, especially the mean temperature and the closely (exponentially) related mean of water vapour pressure, was also most useful for predicting crop cover fractions.
-It was also beneficial for estimating built-up fractions.
-Location features were the most important for predicting the cover fractions of these two classes as well.
-These classes tend to be spatially clustered, when looking at a large scale.
-While the models for predicting the cover fraction of classes made the best use of the location features, all of the classes could benefit from them to some extent, thus it is beneficial to include these intrinsic parameters in the model.
-Both absolute and regular latitude were used by the model to improve prediction accuracy, and latitude was more beneficial than longitude for increasing the prediction accuracy.
-
-Terrain information was the least useful feature category.
-It was most useful for inland water fraction prediction, since it typically has little to no slope, and rarely occurs at high altitudes.
-The aspect was the one feature that does not appear to have contributed to model accuracy.
-
 \section{Discussion}
 
-\subsection{Land cover fraction mapping methods}
-
-In this study, we have tested a number of methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale.
-We found that of the methods that we tested, without additional measures to take the data imbalance into account, \gls{RF} regression and Cubist regression produce output of the highest accuracy by \gls{RMSE} and \gls{MAE}, respectively.
-Our findings agree with those of \citet{li_monitoring_2018}, who compared Cubist and \gls{RF} regression for water fraction classification and found that Cubist performs slightly better than \gls{RF} regression for this particular class.
-Their Cubist regression result achieved 7.52\% \gls{MAE} and 10.39\% \gls{RMSE}.
-Our respective results using Cubist regression were 2.57\% \gls{MAE} and 11.51\% \gls{RMSE}.
-While Cubist works best for the inland water class, when considering all of the classes, \gls{RF} regression nevertheless results in higher accuracy (see table \ref{tab-accuracy}).
-The difference in the reported numbers may be due to the differences in scale (regional vs global) and training data balance.
-As we focused on predicting multiple classes, our training dataset intrinsically had higher zero inflation.
-
-At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which he also noted as performing better than \gls{RF} regression and \glspl{SVM}.
-Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the built-up class, and we also noted that Cubist performed slightly better than the other two algorithms for this class.
-Both \citet{okujeni_comparison_2014} and \citet{foody1997fuzzynnet} also performed land cover fraction mapping over an urban area.
-The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF} regression.
-This could be due to much finer spatial resolution (3.6–9 m) data used in their study, and the associated difference in class definitions.
-\citet{foody1997fuzzynnet} reported an \gls{RMSE} of 16.6\% using \gls{MLP} \glspl{NN}, lower than our results (22.7\%).
-The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, namely their focus on fine scale predictions over a local area for the urban class in particular, in contrast to our global focus with comprehensive classes, leading to a different balance of the datasets between the studies.
-
 \subsection{Multi-step approach for dealing with data imbalance}
 
-After determining the baseline methods with the highest accuracy, we tested our newly proposed multi-step approach on \gls{RF} regression, comparing the effect with that of median voting instead of mean voting.
 One challenge in land cover fraction prediction is the tendency of the models to favour the mean over the extremes, as that minimises \gls{RMSE} that may arise from incorrect predictions.
@@
 Thus while our model appears to perform better, \citet{montesano_modis_2009} only evaluated boreal regions rather than the entire globe, and used MODIS 500 m data rather than PROBA-V 100 m data.
-
-\subsection{Feature importance}
-
-Remote sensing features were the most important features for the multi-step median \gls{RF} model, especially the maximum and minimum \gls{NDVI} over the entire time series.
-Note that these values are taken from a time series that has undergone temporal outlier removal, as detailed in section \ref{sec-temporal-filter}, and thus roughly correspond to the 5th and 95th percentiles of the data without additional temporal filtering.
-Multiple vegetation indices were useful for increasing the model accuracy: both \gls{NDMI} and \gls{NIRv} median over the time series were much more important than any other feature from other groups.
-Nevertheless, even though a lot of the other features were of much lower importance, they contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of collinear features as explained in section \ref{sec-covariate-selection}).
-Methods based on decision trees help with effectively using features that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any features as non-informative.
-This is also due to the large variety of land cover classes in the study, since a feature is useful if it helps predict any of the land cover classes better.
-
-The result that including more training features is beneficial, but remote sensing data is the most important, is in line with the conclusions of e.g.
-\citet{li_monitoring_2018} and \citet{hengl_soilgrids250m_2017}.
-Remote sensing data is also unique in that it forms a time series, which enables us to both calculate additional temporal metrics and to monitor land cover change over time.
-In addition to remote sensing data, \citet{hengl_soilgrids250m_2017} also noted high importance of climate data, however, it is focused on soils, whereas climate has more effect on long-term processes such as soil formation than on land cover.
-Climate data was the second most important in our case, mostly for the crop fraction estimation, which is also closely linked with soils.
-Note that the feature importance may be different at different scales, e.g. if predictions are made at a finer spatial resolution, such as 30~m or 10~m.
 
@@
 These classes were particularly difficult to map, in part due to their heterogeneity, and in part due to confusion between herbaceous vegetation, cropland and shrubs.
-As shown in figure \ref{fig-varimp}, specific types of features can be used to discriminate between these classes better.
+As shown in \ref{sec-varimp}, figure \ref{fig-varimp}, specific types of features can be used to discriminate between these classes better.
 Specifically, the identification of herbaceous vegetation is primarily based on vegetation indices, identification of shrubs relies mostly on climate data, whereas cropland identification is more data-intensive and makes use of vegetation indices, temporal metrics, climate, soil and location information.
@@
 
-In this study, model optimisation and dataset sizes were also a challenge.
-Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming features or replacing missing values).
-Since the large dataset size increases the training time for the models, parameters to tune have to be chosen carefully.
-As such, it is possible to miss a particular parameter combination that may further improve the results.
-On the other hand, our results showed that the differences between the best performing methods are not very high.
-The differences in model parametrisations are expected to be even lower.
-
-Another challenging aspect about land cover fraction mapping is the available algorithms that would suit this use case.
-In its core, it is a multivariate regression problem, but most regression algorithms are univariate.
-Our results showed that the binary relevance method works well to facilitate the use of univariate regression methods for this task, as the highest accuracy was achieved by algorithms that used the binary relevance method.
-This further expands the field of possible algorithms that could be used.
-In addition, both the binary relevance method and the multi-step method can accept different algorithms for its submodels, i.e.
-for different classes in the case of binary relevance and for different steps in the case of the multi-step approach.
-There may be a combination of models that would work even better than the models we tested.
-We could not cover the entire range of methods in the scope of this study, therefore there is still room for improvement.
-For example, classical spectral unmixing methods, such as spectral mixture analysis, has been often used in the past (e.g. in \citet{adams_classification_1995}).
-However, due to the limitations of the spectral mixture analysis methods \citep{somers_endmember_2011}, we could not use it with the features we selected (as listed in the supplementary material, table S1), as they do not form a linear mixture representing the land cover fractions.
-Modern machine learning methods are advantageous for their ability to model complex nonlinear relationships between variables and thus make use of all of the features as shown in figure \ref{fig-varimp}.
-
-The scale of the study poses performance challenges, as features that the models use need to be available at a global scale.
-Our results showed that all of the feature groups used in the study were useful, therefore leaving a feature out means sacrificing some predictive power.
-The gain would be less time needed for processing, as that feature would no longer need to be downloaded, preprocessed and processed.
-In addition, model training performance (time for training and memory usage) may be an important consideration for global land cover mapping, as it may limit the scope of what models may be used for this task.
-For example, ensemble learning techniques like the super learner are very resource-intensive and may take weeks to train.
-Therefore it is important to take into account not just the model accuracy, but also whether the improvement of the accuracy is worth the increase in processing time or computing resource usage.
-
 Several more challenges are yet to be tackled in this field, but doing so is becoming more and more feasible over time.
@@
 
-% NOTE: Appendix is in supplement.tex
-
 %\section*{Abbreviations}
@@
 
-%\listoffigures
+\appendix
+
+\section{Regression model comparison} \label{sec-model-comparison}
+
+\subsection{Regression models tested in this study}
+
+To test the performance of various regression models for land cover fraction mapping, we tested four types of models: linear models, models based on decision trees, machine learning models not based on decision trees, and ensemble learning (for an overview, see table \ref{tab-methods}).
+
+%\subsubsection{Linear models}
+
+First, we chose to compare five types of linear models, to have a baseline for a comparison with the nonlinear machine learning models.
+The most simple model we selected was the \glsentryfull{GLM}, also known as multivariate linear regression.
+It is an extension to the standard linear regression that allows for multiple outcomes.
+Next, we tested two linear models that include input data regularisation in the model itself: lasso regression and \gls{PLS} regression.
+We also tested \glsentryfull{MLR}, which is usually used for classification and is fit using land cover class labels rather than fractions, but the output includes probabilities for each class that add up to 100\%.
+We fit \gls{MLR} using the dominant land cover class as a label for the pixel and used the class probabilities as a proxy for land cover fractions.
+Lastly, we tested \glsentryfull{FNC} regression, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means.
+It is a simple regression method, where the land cover fractions in a pixel are determined by the distance of the pixel from the centroids of each class in feature space.
+
+%The \gls{GLM} \citep{neter_applied_1996}, also known as multivariate linear regression, is an extension to the standard linear regression that allows for multiple outcomes.
+%In this study, we used a \gls{GLM} as a baseline model.
+%We also tested whether dropping any additional features from the model would result in better predictions, using a sphericity test with Greenhouse-Geisser correction \citep{greenhouse_methods_1959} to compare models with different features.
+%However, only three features could be dropped and it did not improve the predictions, therefore we decided to keep all the features in the model.
+%\Gls{GLM} is implemented in base R \citep{r_2019}.
+
+%Ridge regression is a linear model regularisation method which reduces large coefficients of model features in order to prevent overfitting.
+%Lasso regression is based on the same principle, except it reduces the coefficients of each model feature to zero if the absolute sum of coefficients is above a threshold, thus acting as a variable selection method.
+%Elastic net regression is a method combining the two, where model coefficients may be reduced or set to zero.
+%This has the effect of the coefficients of correlated features being set to a similar value.
+%We chose lasso regression \citep{tibshirani_regression_1996}, so that it would be possible to identify if there are any features that could be further omitted in addition to ones dropped in section \ref{sec-covariate-selection}.
+%The $\lambda$ parameter was chosen using cross-validation.
+%The lasso regression was performed using the \cran{glmnet} package \citep{glmnet}.
+
+%\Gls{PLS} regression \citep{wold_pls-regression_2001} is a concept related to principal component regression and also aims at reducing the input features into latent variables that explain the most variance of the response variable in multidimensional space.
+%As such, it is well suited for multicollinear data.
+%\ac{PLS} regression was performed using the package \cran{pls} \citep{pls}.
+
+%All of the aforementioned linear models do not normalise the output to add up to 100\%, and also allow negative predictions.
+%Therefore for these models, all predictions < 0 were set to 0, and then all values were linearly rescaled per each prediction to add up to 100\%.
+
+%\Gls{FNC} regression \citep{keller_fuzzy_1985}, also called fuzzy nearest prototype, fuzzy $c$-means or fuzzy $k$-means, is a simple regression method where the pixel's membership of a class is determined by the distance of the pixel from each class centroid in feature space.
+%This method always produces output that sums to 100\%.
+%In this study, we used an implementation adapted from the package \cran{GSIF} \citep{hengl2004fuzzycmeans}.
+%We tested both using a logistic regression to estimate the class centroids, as well as using a weighted average of the training data to do so, and ended up using the latter method as the result was more accurate.
+
+%Lastly, we tested \gls{MLR} \citep{theil_multinomial_1969}.
+%While logistic regressions are most commonly used for classification tasks, the output of a multinomial regression itself is the probability of a pixel to belong to each of the defined classes.
+%This probability can be used as a proxy for the fraction of the pixel that the class covers.
+%These probabilities always add up to 100\%.
+%However, an \gls{MLR} can only be trained on labelled data, rather than fractions.
+%Therefore the target (label) input for the \ac{MLR} was the name of the dominant land cover fraction of each pixel.
+%This does not guarantee that the pixels that the model was trained on were pure endmembers, but it allows the model to learn from a much larger dataset compared to using just the pure pixels.
+%The logistic regression implementation was provided by the \cran{nnet} package \citep{nnet}.
+
+%\subsubsection{Non-tree machine learning methods}
+
+The second group tested in this study was machine learning methods not based on decision trees.
+\Glsentryfullpl{NN} are a promising technique for land cover fraction mapping, as they allow both multiple inputs and multiple outputs, and, using the softmax activation function, also ensures that the result sums up to 100\% with no need for additional postprocessing.
+In this study, after performing tuning, we ended up using a \glsentryfull{MLP} with three hidden layers with 128, 64 and 32 neurons per layer respectively.
+We used the Nadam optimiser \citep{dozat_incorporating_2016} with \gls{MAE} as the loss function to optimise the \gls{NN}, and softmax activation for the output.
+The models were trained using the \cran{keras} package, built upon TensorFlow that enables the use of a graphics processing unit to accelerate the \gls{NN} training process.
+The other method in this category that we tested was \Glsentryfullpl{SVM}, which are machine learning models that attempt to find the optimal boundary between the class clusters in feature space by constructing a dividing hyperplane.
+For land cover fraction classification, we used \gls{SVM} regression based on Least Squares \gls{SVM}.
+As \gls{SVM} models are univariate, we used the binary relevance method \citep{karalas2016br}: training separate models per class that predict a single class, and then combining the results.
+
+%\subsubsection{Tree-based machine learning methods}
+
+The third group we tested were tree-based machine learning models.
+\Glsentryfull{RF} regression is a popular method for land cover classification that works by building a number of \gls{CART} decision trees based on random subsets of the input training data, and taking the mean or median of the ``votes'' of these individual decision trees.
+\ac{RF} is univariate, therefore we again used the binary relevance approach.
+Next, we tested Cubist regression.
+It is based on \gls{RF} regression, but instead of using a threshold of a feature to split the decision tree, Cubist uses a linear regression based on a subset of the data relevant for the split in question.
+In addition, it features committees, a boosting technique that iteratively trains trees so as to learn from the previously generated ones.
+Model tuning led to us using 10 committees.
+Cubist predictions were also made using the binary relevance method.
+
+%\subsubsection{Ensemble learning}
+
+Lastly, we made an ensemble from the two machine learning models that produced the lowest \gls{RMSE} and \gls{MAE} respectively: \gls{RF} and Cubist regressions.
+We used the super learner algorithm \citep{laan_super_2007} to create a hierarchical ensemble, where the two models were cross-validated using 10-fold cross-validation to obtain relative weights of each model for each land cover class.
+The predictions of the two models, along with the weights of the models, were then used as input features for another \gls{RF} regression metalearner.
+The output from this metalearner is the final prediction of the whole ensemble.
+As the ensembled methods are univariate, we used the binary relevance method in this case as well.
+
+%\subsubsection{Other methods}
+
+%Additional methods were considered, but in the end turned out to be unworkable with our dataset.
+%Multivariate \gls{RF} is available in R provided by the \cran{MutivariateRandomForest} package \citep{MultivariateRandomForest}, which would potentially allow better results than using regular \gls{RF} with binary relevance.
+%However, it is not optimised for large datasets, which did not allow to train it on more than 2000 samples at a time, making it incomparable to the rest of the tested methods.
+
+%Lastly, multiple-endmember spectral unmixing using non-negative least squares \citep{franc2005sequential} was also attempted, but the results were significantly worse than even the equal proportions model, possibly because the class fractions are not a linear combination of the features.
+
+\subsection{Comparison results and discussion}
+
+The overall accuracy statistics of the compared models are reported in table \ref{tab-accuracy-linear}, and per-class statistics in figure \ref{fig-models}.
+
+The results show that the $R^2$ statistics, especially $R^2_{NSE}$, are in agreement with the \gls{RMSE} statistics, and the \glsentryfull{OA} from the \glsentryfull{SCM} is in agreement with the \gls{MAE} statistics.
+$R^2$ is a representation of how closely the predictions correlate with the validation data, therefore big outliers have a large effect on the value.
+The \gls{SCM} does not take this into account, as it does a comparison on the overlap of fractions in each pixel, which doesn't apply an extra penalty for large errors.
+
+The baseline models that performed the best were the two tree-based machine learning models: \gls{RF} regression and Cubist.
+Both of these models are univariate, therefore predictions per class had to be made using the binary relevance method (one model per class).
+This shows that despite the limitations of the binary relevance method, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and the need for a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models and the flexibility of fitting the models for each class outweigh the disadvantages of the binary relevance method.
+
+The super learner method that combines both \gls{RF} regression and Cubist regression resulted in a model that is in between the two ensembled models in terms of accuracy.
+Its \gls{RMSE} was below that of Cubist regression, but above that of \gls{RF} regression.
+Conversely, the \gls{MAE} was below that of \gls{RF} regression (single model), but above that of Cubist regression.
+This shows that the metalearner (another \gls{RF} model) could not differentiate between the two ensembled models well enough to select the model that is the most accurate for a given class.
+Rather, it weighted in both of the models' predictions, losing the advantages of the models when taken separately.
+
+The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed the worst, as they are limited to a linear approach.
+The results of all of the linear models were very similar.
+There was no added value to \gls{PLS} and Lasso regressions over the basic \gls{GLM}.
+This is likely due to the additional preprocessing step of feature selection, as detailed in section \ref{sec-covariate-selection}.
+Lasso and \gls{PLS} regressions are extensions to \gls{GLM} that include a regularisation step, which proved to be unnecessary if all of the highly correlated features are already manually removed.
+
+\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
+This indicates that the shape of the feature data in feature space is too complex to capture merely with a centroid approach.
+This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
+
+\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods.
+Nevertheless, it performed better than all of the multivariate linear models.
+
+%There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
+%Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
+%Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often making predictions that are partially but not completely incorrect.
+%Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
+
+All tested models had lower accuracy when estimating the cover fractions of vegetation classes (herbaceous, trees, crops, shrubs), compared to the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
+This was exacerbated by the imbalance in class distribution: the built-up class is rare and very rarely forms a majority, therefore a prediction of 0\% leads to a perfect prediction most of the times, and even if it does not, the error is low.
+In contrast, tree cover had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
+The \gls{RRMSE} and \gls{RMAE} statistics show this (see figure \ref{fig-models-relative}): the tree cover prediction has low relative error given its high mean value (32\%), whereas the built-up class is the most challenging to predict according to \gls{RRMSE} given its low mean value (3\%).
+The shrub cover fraction is also very challenging to predict, since it had the highest \gls{RMAE} and none of the models showed large improvements in \gls{RMSE} compared to the equal proportions model.
+There were only moderate improvements for predicting the herbaceous vegetation fraction as well.
+In contrast, most models were a significant improvement in predicting tree, built-up, water and bare cover fractions compared to the equal proportion model, indicating that the features used to train the models were useful to distinguish these classes from the others and to quantify their proportions in each pixel.
+
+\begin{table}
+\centering
+%\adjustbox{max width=0.5\textwidth}{
+\begin{tabular}{lllllll}
+\toprule
+\textbf{Model} & \textbf{\ac{RMSE} (\%)} & \textbf{\ac{MAE} (\%)} & $\mathbf{R^2_{NSE}}$ & $\mathbf{R^2_{OLS}}$ (slope/int) & \textbf{\ac{OA} (\%)} & \textbf{Kappa} \\
+\midrule
+Equal proportions
+& 29.9  & 21.4  & 0     & — & 26±5  & 0.13±0.07 \\
+\Glsentryshort{FNC}
+& 24.4  & 13.5  & 0.33  & 0.35 (0.80/2.89) & 53±4  & 0.42±0.06 \\
+\Glsentryshort{GLM}, \Glsentryshort{PLS}, Lasso
+& 21.6  & 12.7  & 0.48  & 0.49 (1.10/-1.42) & 56±4  & 0.43±0.05 \\
+\Glsentryshort{MLR}
+& 21.6  & 12.1  & 0.48  & 0.48 (0.96/0.62) & 58±4  & 0.46±0.06 \\
+\Glsentryshort{MLP} \glsentryshortpl{NN}
+& 22.7  & 9.2   & 0.43  & 0.52 (0.70/4.25) & 68±1  & 0.57±0.02 \\
+\Glsentryshort{SVM} regression
+& 20.7  & 8.9   & 0.52  & 0.56 (0.79/3.02) & 69±2  & 0.58±0.03 \\
+Cubist regression
+& 18.1  & \textbf{8.1}   & 0.63  & 0.65 (0.88/1.77) & \textbf{72±2}  & \textbf{0.63±0.03} \\
+\Glsentryshort{RF} regression
+& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} (1.09/-1.22) & 67±4  & 0.57±0.05 \\
+\ensuremath{''}
+only RS features
+& 18.4  & 10.3  & 0.62  & 0.64 (1.09/-1.25) & 64±4  & 0.54±0.05 \\
+Cubist + RF ensemble
+& 17.7  & 8.6   & 0.65  & 0.65 (0.95/0.73) & 70±3 & 0.61±0.04 \\
+\bottomrule
+\end{tabular}
+%}
+\caption{Accuracy statistics of the tested regression models.
+Best performing statistics are highlighted.
+``slope'' refers to the OLS-estimated slope, ``int'' refers to the OLS-estimated intercept.
+``Only RS features'' stands for a model trained only with \glspl{VI} and temporal metrics.}
+\label{tab-accuracy-linear}
+\end{table}
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
+    \caption{Comparison of absolute \gls{RMSE} (top) and absolute \gls{MAE} (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
+    \label{fig-models}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-11-02-model-comparison-bar}
+    \caption{Comparison of relative RMSE (top) and relative MAE (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
+    \label{fig-models-relative}
+\end{figure}
+
+In this study, we have tested a number of methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale.
+We found that of the methods that we tested, without additional measures to take the data imbalance into account, \gls{RF} regression and Cubist regression produce output of the highest accuracy by \gls{RMSE} and \gls{MAE}, respectively.
+Our findings agree with those of \citet{li_monitoring_2018}, who compared Cubist and \gls{RF} regression for water fraction classification and found that Cubist performs slightly better than \gls{RF} regression for this particular class.
+Their Cubist regression result achieved 7.52\% \gls{MAE} and 10.39\% \gls{RMSE}.
+Our respective results using Cubist regression were 2.57\% \gls{MAE} and 11.51\% \gls{RMSE}.
+While Cubist works best for the inland water class, when considering all of the classes, \gls{RF} regression nevertheless results in higher accuracy (see table \ref{tab-accuracy-linear}).
+The difference in the reported numbers may be due to the differences in scale (regional vs global) and training data balance.
+As we focused on predicting multiple classes, our training dataset intrinsically had higher zero inflation.
+
+At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which he also noted as performing better than \gls{RF} regression and \glspl{SVM}.
+Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the built-up class, and we also noted that Cubist performed slightly better than the other two algorithms for this class.
+Both \citet{okujeni_comparison_2014} and \citet{foody1997fuzzynnet} also performed land cover fraction mapping over an urban area.
+The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF} regression.
+This could be due to much finer spatial resolution (3.6–9 m) data used in their study, and the associated difference in class definitions.
+\citet{foody1997fuzzynnet} reported an \gls{RMSE} of 16.6\% using \gls{MLP} \glspl{NN}, lower than our results (22.7\%).
+The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, namely their focus on fine scale predictions over a local area for the urban class in particular, in contrast to our global focus with comprehensive classes, leading to a different balance of the datasets between the studies.
+
+In this study, model optimisation and dataset sizes were also a challenge.
+Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming features or replacing missing values).
+Since the large dataset size increases the training time for the models, parameters to tune have to be chosen carefully.
+As such, it is possible to miss a particular parameter combination that may further improve the results.
+On the other hand, our results showed that the differences between the best performing methods are not very high.
+The differences in model parametrisations are expected to be even lower.
+
+Another challenging aspect about land cover fraction mapping is the available algorithms that would suit this use case.
+In its core, it is a multivariate regression problem, but most regression algorithms are univariate.
+Our results showed that the binary relevance method works well to facilitate the use of univariate regression methods for this task, as the highest accuracy was achieved by algorithms that used the binary relevance method.
+This further expands the field of possible algorithms that could be used.
+In addition, both the binary relevance method and the multi-step method can accept different algorithms for its submodels, i.e.
+for different classes in the case of binary relevance and for different steps in the case of the multi-step approach.
+There may be a combination of models that would work even better than the models we tested.
+We could not cover the entire range of methods in the scope of this study, therefore there is still room for improvement.
+For example, classical spectral unmixing methods, such as spectral mixture analysis, has been often used in the past (e.g. in \citet{adams_classification_1995}).
+However, due to the limitations of the spectral mixture analysis methods \citep{somers_endmember_2011}, we could not use it with the features we selected (as listed in \ref{sec-inputdata}, table \ref{tab-inputdata}), as they do not form a linear mixture representing the land cover fractions.
+Modern machine learning methods are advantageous for their ability to model complex nonlinear relationships between variables and thus make use of all of the features as shown in \ref{sec-varimp}, figure \ref{fig-varimp}.
+
+\section{Multi-step model relative error comparison} \label{sec-randomforest-relative}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-11-02-rf-comparison-bar}
+    \caption{Comparison of relative accuracy metrics of RF regression models (equal proportion model shown as a reference).
+1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
+Mean and median are the tree vote summary statistics.}
+    \label{fig-randomforest-relative}
+\end{figure}
+
+\section{Model accuracy changes per predicted fraction} \label{sec-apu}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=\textwidth]{article-figures/apuplots/2021-01-13-apu-rf3step.pdf}
+    \caption{Change in accuracy statistics of the \gls{RF} three-step median model per predicted land cover fraction, overlaid over the histogram of predictions.
+Bins containing less than 10 predictions were removed due to too low sample size leading to unreliable error statistics.}
+    \label{fig-apu}
+\end{figure}
+
+\section{Feature importance} \label{sec-varimp}
+
+In addition to classification methods, one aspect that is important for classification accuracy is the data (features) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
+In addition to optical features (such as spectral bands), temporal features (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
+\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are commonly used to improve land cover class separability, e.g.
+for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
+Elevation features (elevation, slope, aspect) are also known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
+%In this paper, we also used such features for increasing the accuracy of land cover fraction regression model predictions.
+%, and their importance for the predictions was analysed.
+
+To calculate feature importance, we performed permutation importance on the model that had the lowest \gls{MAE}.
+We shuffled the values of each feature in turn, and the models made predictions based on all of the features, including the shuffled one.
+This results in a decrease in the accuracy of predictions, as the feature in question no longer contains meaningful information.
+We then recorded the resulting increase in \gls{MAE} compared to the validation set as a measure of feature importance, both per class and overall for all classes combined.
+
+We performed \Gls{RF} permutation importance on the three-step median model to test which features were important for \gls{RF} predictions.
+The results are shown in figure \ref{fig-varimp}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=14cm]{article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
+    \caption{Random Forest three-step median approach variable importance, top 5 features per category.
+Categories are ordered by cumulative importance.
+The values represent increase in MAE for the given class when the given feature is permuted.}
+    \label{fig-varimp}
+\end{figure}
+
+The overall most important variables were the maximum and minimum \gls{NDVI} over the whole time series of PROBA-V imagery.
+They were followed by the median \gls{NIRv} and \gls{NDMI} over the whole time series.
+The importance of remote sensing information can also be seen from the accuracy statistics of a single-step \gls{RF} regression model, when only vegetation index and temporal features were input into the model (see table \ref{tab-accuracy}, ``only RS features'').
+After excluding all the static features, \gls{RMSE} and \gls{MAE} increased only by around 1 percentage point.
+
+The remote sensing data was of lower importance (and sometimes even confounding) only for the shrub class.
+This class is complicated to distinguish from other natural vegetation, and thus when trying to predict shrub cover fractions, the model benefited from a wide variety of additional features, including location and climate information.
+
+To distinguish crops in particular, soil information, such as pH, bulk density and sand fraction, was useful for the model.
+Crops are usually grown in fertile soils that can sustain them, and due to the fact that cropland is further managed, the soil is also altered to be more fertile, which affects these soil properties.
+In addition, harmonic metrics derived from time series benefited crop cover fraction estimation the most of all the classes.
+The second order amplitude and sine of the harmonic model of the time series allows the model to detect areas with a double harvest throughout the year, which is indicative of crops.
+
+Climate data, especially the mean temperature and the closely (exponentially) related mean of water vapour pressure, was also most useful for predicting crop cover fractions.
+It was also beneficial for estimating built-up fractions.
+Location features were the most important for predicting the cover fractions of these two classes as well.
+These classes tend to be spatially clustered, when looking at a large scale.
+While the models for predicting the cover fraction of classes made the best use of the location features, all of the classes could benefit from them to some extent, thus it is beneficial to include these intrinsic parameters in the model.
+Both absolute and regular latitude were used by the model to improve prediction accuracy, and latitude was more beneficial than longitude for increasing the prediction accuracy.
+
+Terrain information was the least useful feature category.
+It was most useful for inland water fraction prediction, since it typically has little to no slope, and rarely occurs at high altitudes.
+The aspect was the one feature that does not appear to have contributed to model accuracy.
+
+Remote sensing features were the most important features for the multi-step median \gls{RF} model, especially the maximum and minimum \gls{NDVI} over the entire time series.
+Note that these values are taken from a time series that has undergone temporal outlier removal, as detailed in section \ref{sec-temporal-filter}, and thus roughly correspond to the 5th and 95th percentiles of the data without additional temporal filtering.
+Multiple vegetation indices were useful for increasing the model accuracy: both \gls{NDMI} and \gls{NIRv} median over the time series were much more important than any other feature from other groups.
+Nevertheless, even though a lot of the other features were of much lower importance, they contributed to prediction accuracy enough so that they could not be easily excluded from the models as redundant (after the removal of collinear features as explained in section \ref{sec-covariate-selection}).
+Methods based on decision trees help with effectively using features that may also have an overlap in the information that they provide, although linear models likewise tended to not exclude any features as non-informative.
+This is also due to the large variety of land cover classes in the study, since a feature is useful if it helps predict any of the land cover classes better.
+
+The result that including more training features is beneficial, but remote sensing data is the most important, is in line with the conclusions of e.g.
+\citet{li_monitoring_2018} and \citet{hengl_soilgrids250m_2017}.
+Remote sensing data is also unique in that it forms a time series, which enables us to both calculate additional temporal metrics and to monitor land cover change over time.
+In addition to remote sensing data, \citet{hengl_soilgrids250m_2017} also noted high importance of climate data, however, it is focused on soils, whereas climate has more effect on long-term processes such as soil formation than on land cover.
+Climate data was the second most important in our case, mostly for the crop fraction estimation, which is also closely linked with soils.
+Note that the feature importance may be different at different scales, e.g. if predictions are made at a finer spatial resolution, such as 30~m or 10~m.
+
+The scale of the study poses performance challenges, as features that the models use need to be available at a global scale.
+Our results showed that all of the feature groups used in the study were useful, therefore leaving a feature out means sacrificing some predictive power.
+The gain would be less time needed for processing, as that feature would no longer need to be downloaded, preprocessed and processed.
+In addition, model training performance (time for training and memory usage) may be an important consideration for global land cover mapping, as it may limit the scope of what models may be used for this task.
+For example, ensemble learning techniques like the super learner are very resource-intensive and may take weeks to train.
+Therefore it is important to take into account not just the model accuracy, but also whether the improvement of the accuracy is worth the increase in processing time or computing resource usage.
+
+\newpage
+
+\section{List of features used in regression models} \label{sec-inputdata}
+
+\begin{table}[h!]
+\begin{center}
+\begin{adjustbox}{max width=8cm}
+\begin{tabular}{m{12cm}l}
+\toprule
+\multicolumn{1}{l}{Category \& data source}&\multicolumn{1}{c}{Covariate}\tabularnewline
+\midrule
+Location (3, intrinsic)&
+\begin{tabular}{@{}l@{}}longitude\tabularnewline
+latitude\tabularnewline
+absolute latitude
+\end{tabular}\tabularnewline
+\midrule
+Vegetation indices (22, derived from PROBA-V 100m top-of-canopy reflectance v1.02 \citep{dierckx2014probav})&
+\begin{tabular}{@{}l@{}}
+minimum NDVI\tabularnewline
+maximum NDVI\tabularnewline
+median NDMI\tabularnewline
+NDMI yearly IQR\tabularnewline
+NDMI March-May IQR\tabularnewline
+NDMI June-August IQR\tabularnewline
+NDMI September-November IQR\tabularnewline
+NDMI December-February IQR\tabularnewline
+OSAVI March-May IQR\tabularnewline
+OSAVI June-August IQR\tabularnewline
+OSAVI September-November IQR\tabularnewline
+OSAVI December-February IQR\tabularnewline
+EVI March-May IQR\tabularnewline
+EVI June-August IQR\tabularnewline
+EVI September-November IQR\tabularnewline
+EVI December-February IQR\tabularnewline
+median NIRv\tabularnewline
+NIRv yearly IQR\tabularnewline
+NIRv March-May IQR\tabularnewline
+NIRv June-August IQR\tabularnewline
+NIRv September-November IQR\tabularnewline
+NIRv December-February IQR
+\end{tabular}\tabularnewline
+\midrule
+Temporal metrics (9, derived from a harmonic model over time series of PROBA-V 100m top-of-canopy reflectance v1.02 \citep{dierckx2014probav})&
+\begin{tabular}{@{}l@{}}
+NDVI order 1 cosine\tabularnewline
+NDVI order 1 sine\tabularnewline
+NDVI order 2 cosine\tabularnewline
+NDVI order 2 sine\tabularnewline
+NDVI trend coefficient\tabularnewline
+NDVI order 1 phase\tabularnewline
+NDVI order 1 amplitude\tabularnewline
+NDVI order 2 phase\tabularnewline
+NDVI order 2 amplitude
+\end{tabular}\tabularnewline
+\midrule
+Terrain (4, ASTER GDEM V003 \citep{ASTGTM003})&
+\begin{tabular}{@{}l@{}}
+elevation\tabularnewline
+slope (log-transformed)\tabularnewline
+aspect\tabularnewline
+terrain position index
+\end{tabular}\tabularnewline
+\midrule
+Climate (21, WorldClim 2.0 \citep{worldclim2})&
+\begin{tabular}{@{}l@{}}
+January precipitation (log)\tabularnewline
+April precipitation (log)\tabularnewline
+July precipitation (log)\tabularnewline
+October precipitation (log)\tabularnewline
+January solar irradiance\tabularnewline
+July solar irradiance\tabularnewline
+mean temperature\tabularnewline
+temperature monthly range\tabularnewline
+isothermality\tabularnewline
+temperature annual range\tabularnewline
+annual precipitation (log)\tabularnewline
+temperature seasonality\tabularnewline
+minimum solar irradiance\tabularnewline
+maximum solar irradiance\tabularnewline
+mean solar irradiance\tabularnewline
+mean windspeed\tabularnewline
+mean water vapour pressure\tabularnewline
+coldest month precipitation (log)\tabularnewline
+warmest month precipitation (log)\tabularnewline
+wettest month solar irradiance\tabularnewline
+driest month solar irradiance
+\end{tabular}\tabularnewline
+\midrule
+Soil (8, SoilGrids \citep{hengl_soilgrids250m_2017})&
+\begin{tabular}{@{}l@{}}
+soil available water\tabularnewline
+soil bulk density\tabularnewline
+soil cation exchange capacity (log)\tabularnewline
+soil clay fraction\tabularnewline
+soil coarse fragments (log)\tabularnewline
+soil pH\tabularnewline
+soil sand fraction\tabularnewline
+soil water wilting point
+\end{tabular}\tabularnewline
+\bottomrule
+\end{tabular}
+\end{adjustbox}
+\caption{List of features used as input for the models tested in this study, and their data sources.}
+\label{tab-inputdata}
+\end{center}
+\end{table}
+
+\listoffigures
 

### 2021-02-04 article/highlights.tex
Move parts to appendix

@@
 \begin{itemize}
-    \item 4 machine learning model groups tested for deriving 7 land cover fractions globally
+    %\item 4 machine learning model groups tested for deriving 7 land cover fractions globally
+    \item 7 land cover fractions derived globally using machine learning algorithms
     \item Multi-step approach proposed for dealing with zero inflation in fraction data
-    \item Random forest regression performed the best
-    \item Optical remote sensing data was the most important input for model predictions
+    %\item 1234567890123456789012345678901234567890123456789012345678901234567890123456789012345
+    \item The approach provides a middle ground between discrete and fractional mapping
+    \item The approach gives the best guess and improves estimation of 0\% and 100\% fractions
+    \item Comprehensive spatial accuracy assessment based on over 20\,000 points
     \item Multi-fraction land cover mapping shown as an alternative to discrete classification

### 2021-02-08 article/article.tex
Revision round 3 draft 1

@@
 
-After determining the method with the lowest \gls{RMSE}, we attempted to improve it further in an attempt to solve the dataset balance issue, namely the high frequency of 0\% and 100\% fractions.
+As \gls{RF} regression performed the best in the regression model comparison, we attempted to improve it further in an attempt to solve the dataset balance issue, namely the high frequency of 0\% and 100\% fractions.
 As the dataset describes fractions of each land cover class at each point, most of the locations consist of a mix of only a few classes, and the fraction of the rest of the classes is zero at that location.
@@
 
-Lastly, we produced spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point.
-We rasterised the result into a map for ease of view.
+To show how the model accuracy varies in space, we produced spatial residuals, i.e. the overestimation and underestimation of each class fraction for each validation point.
+We rasterised the result into global maps for ease of view.
 If multiple points fell into the same raster cell, we reported the mean value.
@@
 
-Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote, and by \gls{MAE} when using a median vote.
+Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote (\gls{RMSE}: 17.3\%, \gls{MAE}: 9.4\%), and by \gls{MAE} when using a median vote (\gls{RMSE}: 20.7\%, MAE: 7.9\%).
 For more details about the results of the accuracy comparison between the tested regression models, see \ref{sec-model-comparison}.
@@
 Second, we used our proposed multi-step model approach as detailed in section \ref{sec-multistep}.
-The overall results can be seen in table \ref{tab-accuracy}, whereas per-class results can be seen in figure \ref{fig-randomforest} and the relative statistics in \ref{sec-randomforest-relative}, figure \ref{fig-randomforest-relative}.
+The overall results can be seen in table \ref{tab-accuracy}, whereas per-class results can be seen in figure \ref{fig-randomforest} and the relative statistics in figure \ref{fig-randomforest-relative}.
 
@@
 Decreasing the purity threshold resulted in a lower overall accuracy, but an increase in the users' accuracy in the classifier model for these particular classes.
-The regression step by itself had lower accuracy than the combined three-step model, with 21.38 \gls{RMSE} and 10.49 \gls{MAE} (using median voting), as the middle of the range is the most difficult to predict correctly.
+The regression step by itself had lower accuracy than the combined three-step model, with 21.38\%~\gls{RMSE} and 10.49\%~\gls{MAE} (using median voting), as the middle of the range is the most difficult to predict correctly.
 Decreasing the purity threshold led to a lower \gls{RMSE}, as the value range of the regressor training data gets decreased, but a higher \gls{MAE}, as the amount of training data for the regressor decreased.
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-11-02-rf-comparison-bar}
+    \caption{Comparison of relative accuracy metrics of RF regression models (equal proportion model shown as a reference).
+1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
+Mean and median are the tree vote summary statistics.}
+    \label{fig-randomforest-relative}
+\end{figure}
+
 \subsection{Spatial predictions and accuracy}
@@
 The overall \gls{ME} is below 0.001, which means that overall the model is not biased.
-For an alternative visualisation showing the change of \gls{RMSE}, \gls{MAE} and \gls{ME} overall and per class, see \ref{sec-apu}, figure \ref{fig-apu}.
+For an alternative visualisation showing the change of \gls{RMSE}, \gls{MAE} and \gls{ME} overall and per class over each prediction bin, see \ref{sec-apu}, figure \ref{fig-apu}.
 
@@
 Median voting also resulted in more correct 0\% fraction predictions, but likewise increased the chances of predicting 100\% of the wrong class.
-Our results showed that median voting has a stronger effect for reducing \gls{MAE} than the multi-step approach.
+Our results showed that median voting has a stronger effect in reducing \gls{MAE} than the multi-step approach.
 Therefore it may be a suitable choice in cases where it is computationally infeasible to train three models.
@@
 The accuracy of the impervious surface class fraction from \gls{FROM-GLC} mostly matches that of the built-up cover fraction of our proposed model.
+This shows that our proposed method, coupled with the large training dataset that we used, achieves similar and sometimes even better performance compared to specialised land cover fraction products, but brings the advantage of producing fractions for a variety of land cover classes that sum up to 100\%.
 
@@
 However, with the multi-step approach, it is now possible to influence the decision process of the model to tweak it towards the desired outcome.
-In addition, using a median vote in tree-based ensemble models makes the model tend towards the median, which may still be one of the extreme values.
-In that case, the model typically overestimates or underestimates values for most of the value range, but on average is still not biased, as the predicted values are more polarised but balance each other out.
+In addition, using a median vote in tree-based ensemble models makes the model tend towards the median, which is often one of the extreme values.
+In that case, we see the inverse pattern compared to mean vote: small fractions are underestimated (i.e. predicted as 0\%), and large fractions are overestimated (i.e. predicted as 100\%).
+On average, the model is still not biased, as the predicted values are more polarised but balance each other out.
 The multi-step approach could be improved further by exploring the option of using different machine learning models for each step, and by gathering more features that would increase the accuracy of the models at each of the steps.
 
-There is another inherent challenge in land cover mapping for discerning classes that are related, e.g. herbaceous vegetation and shrubs.
+Another challenge inherent in land cover mapping is discerning classes that are related, e.g. herbaceous vegetation and shrubs.
 These classes were particularly difficult to map, in part due to their heterogeneity, and in part due to confusion between herbaceous vegetation, cropland and shrubs.
@@
 The difference between the definition of trees, shrubs and herbaceous vegetation largely comes down to plant height, therefore dynamic information about vegetation height would allow mapping these classes more accurately.
-However, this would require either photogrammetry techniques (that typically rely on much finer spatial resolution and more overlap between the scenes) to reconstruct vegetation height, or the use of non-optical sensors such as \gls{SAR} interferometry.
+However, this would require either photogrammetry techniques (that typically rely on much finer spatial resolution and more overlap between the scenes) to reconstruct vegetation height, or the use of non-optical sensors such as \gls{SAR} interferometry or lidar data.
 Another way to differentiate between the vegetation classes could be to make use of hyperspectral data, which allows differentiation between different kinds of vegetation based on e.g.
@@
 
-In this study, we compared multiple groups of algorithms for land cover fraction mapping.
-The best performing models were based on \gls{CART} tree ensembles, with \gls{RF} regression performing the best overall.
-A single-step \gls{RF} regression model using majority voting resulted in the best results for RMSE (17.3\%) and $R^2_{NSE}$ (0.66).
-In addition, we investigated ways to tackle the issue of accurately predicting the extreme fraction values of 0\% and 100\% by proposing a hierarchical multi-step approach combining classification and regression models.
-We compared and combined this approach with \gls{RF} median voting.
+We investigated ways to tackle the issue of accurately predicting the extreme fraction values of 0\% and 100\% by proposing a hierarchical multi-step approach combining classification and regression models.
+This approach was applied to an \gls{RF} regression model, which our tests showed to have the highest accuracy (RMSE: 17.3\%) for land cover fraction mapping compared to other regression algorithms.
+We also combined this approach with \gls{RF} median voting.
 The combined \gls{RF} median three-step approach obtained the best results for \gls{MAE} (7.9\%) and \gls{SCM} \gls{OA} (72±2\%).
+The proposed approach results in predictions of the most likely pure land cover class, when the class is uncertain.
+This is in contrast to predicting a mix of classes when a standard one-step model is used, and therefore is useful for users who are more interested in the most likely class, rather than class probabilities.
 Based on this model, we created a demonstration map showing the global distribution of land cover in separate land cover fraction layers.
@@
 First, we chose to compare five types of linear models, to have a baseline for a comparison with the nonlinear machine learning models.
-The most simple model we selected was the \glsentryfull{GLM}, also known as multivariate linear regression.
+The most simple model we selected was the \gls{GLM}, also known as multivariate linear regression.
 It is an extension to the standard linear regression that allows for multiple outcomes.
@@
 The overall accuracy statistics of the compared models are reported in table \ref{tab-accuracy-linear}, and per-class statistics in figure \ref{fig-models}.
-
 The results show that the $R^2$ statistics, especially $R^2_{NSE}$, are in agreement with the \gls{RMSE} statistics, and the \glsentryfull{OA} from the \glsentryfull{SCM} is in agreement with the \gls{MAE} statistics.
@@
 Both of these models are univariate, therefore predictions per class had to be made using the binary relevance method (one model per class).
-This shows that despite the limitations of the binary relevance method, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and the need for a rescale step to make sure the fractions sum up to 100\%, the advantages of these machine learning models and the flexibility of fitting the models for each class outweigh the disadvantages of the binary relevance method.
+This shows that the disadvantages of the binary relevance method, namely that each model is only trained on the fractions of its own class and has no knowledge of the fractions of the other classes, and the need for a rescale step to make sure the fractions sum up to 100\%, are outweighed by the advantages of these machine learning models and the flexibility of training separate models for each class.
 
@@
 Its \gls{RMSE} was below that of Cubist regression, but above that of \gls{RF} regression.
-Conversely, the \gls{MAE} was below that of \gls{RF} regression (single model), but above that of Cubist regression.
+Conversely, the \gls{MAE} was below that of \gls{RF} regression, but above that of Cubist regression.
 This shows that the metalearner (another \gls{RF} model) could not differentiate between the two ensembled models well enough to select the model that is the most accurate for a given class.
 Rather, it weighted in both of the models' predictions, losing the advantages of the models when taken separately.
-
-The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed the worst, as they are limited to a linear approach.
-The results of all of the linear models were very similar.
-There was no added value to \gls{PLS} and Lasso regressions over the basic \gls{GLM}.
-This is likely due to the additional preprocessing step of feature selection, as detailed in section \ref{sec-covariate-selection}.
-Lasso and \gls{PLS} regressions are extensions to \gls{GLM} that include a regularisation step, which proved to be unnecessary if all of the highly correlated features are already manually removed.
-
-\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
-This indicates that the shape of the feature data in feature space is too complex to capture merely with a centroid approach.
-This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
-
-\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods.
-Nevertheless, it performed better than all of the multivariate linear models.
-
-%There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
-%Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
-%Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often making predictions that are partially but not completely incorrect.
-%Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
-
-All tested models had lower accuracy when estimating the cover fractions of vegetation classes (herbaceous, trees, crops, shrubs), compared to the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
-This was exacerbated by the imbalance in class distribution: the built-up class is rare and very rarely forms a majority, therefore a prediction of 0\% leads to a perfect prediction most of the times, and even if it does not, the error is low.
-In contrast, tree cover had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
-The \gls{RRMSE} and \gls{RMAE} statistics show this (see figure \ref{fig-models-relative}): the tree cover prediction has low relative error given its high mean value (32\%), whereas the built-up class is the most challenging to predict according to \gls{RRMSE} given its low mean value (3\%).
-The shrub cover fraction is also very challenging to predict, since it had the highest \gls{RMAE} and none of the models showed large improvements in \gls{RMSE} compared to the equal proportions model.
-There were only moderate improvements for predicting the herbaceous vegetation fraction as well.
-In contrast, most models were a significant improvement in predicting tree, built-up, water and bare cover fractions compared to the equal proportion model, indicating that the features used to train the models were useful to distinguish these classes from the others and to quantify their proportions in each pixel.
 
@@
 \begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
+    \includegraphics[width=0.85\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
     \caption{Comparison of absolute \gls{RMSE} (top) and absolute \gls{MAE} (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
@@
 
-\begin{figure}
-    \includegraphics[width=\textwidth]{article-figures/barplots/2020-11-02-model-comparison-bar}
-    \caption{Comparison of relative RMSE (top) and relative MAE (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
-    \label{fig-models-relative}
-\end{figure}
-
-In this study, we have tested a number of methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale.
-We found that of the methods that we tested, without additional measures to take the data imbalance into account, \gls{RF} regression and Cubist regression produce output of the highest accuracy by \gls{RMSE} and \gls{MAE}, respectively.
+The linear statistical models (\gls{GLM}, \gls{PLS}, Lasso regression and \gls{MLR}) performed the worst, as they are limited to a linear approach.
+The results of all of the linear models were very similar.
+There was no added value to \gls{PLS} and Lasso regressions over the basic \gls{GLM}.
+This is likely due to the additional preprocessing step of feature selection, as detailed in section \ref{sec-covariate-selection}.
+Lasso and \gls{PLS} regressions are extensions to \gls{GLM} that include a regularisation step, which proved to be unnecessary if all of the highly correlated features are already manually removed.
+
+\Gls{FNC} was the model that performed the worst out of the non-trivial models tested.
+This indicates that the shape of the feature data in feature space is too complex to capture merely with a centroid approach.
+This is further evidenced by a much better result obtained by the binary relevance \gls{SVM} model, which works on a similar principle but can capture more complex shapes.
+
+\Gls{MLP} \glspl{NN}, despite being a multivariate machine learning model, performed worse than all of the tested binary relevance methods.
+Nevertheless, it performed better than all of the multivariate linear models.
+
+%There was a trade-off between optimisation for \gls{MAE} and \gls{RMSE}, i.e. between penalising for large errors or not.
+%Models that had lower \gls{MAE} were more crisp, whereas models with lower \gls{RMSE} were more fuzzy.
+%Models that achieved lowest \gls{MAE}, in cases of high uncertainty tended to predict 100\% fractions more often, but risk predicting 100\% of the wrong class, whereas models with lowest \gls{RMSE} would predict a mix of all classes that could be within the pixel, thus more often making predictions that are partially but not completely incorrect.
+%Therefore models optimised for \gls{RMSE} are closer to a probabilistic view of land cover, whereas models optimised for \gls{MAE} are more representative of the best guess of the model for the most likely pure class of the pixel.
+
+All tested models had lower accuracy when estimating the cover fractions of vegetation classes (herbaceous, trees, crops, shrubs), compared to the non-vegetated land cover (inland water, built-up, bare land), as shown in figure \ref{fig-models}.
+This was exacerbated by the imbalance in class distribution: the built-up class is rare and very rarely forms a majority, therefore a prediction of 0\% leads to a perfect prediction most of the times, and even if it does not, the error is low.
+In contrast, tree cover had a lot more balanced distribution of fractions, including a large number of pure pixels, which leaves no such trivial solution.
+The \gls{RRMSE} and \gls{RMAE} statistics show this (see figure \ref{fig-models-relative}): the tree cover prediction has low relative error given its high mean value (32\%), whereas the built-up class is the most challenging to predict according to \gls{RRMSE} given its low mean value (3\%).
+The shrub cover fraction is also very challenging to predict, since it had the highest \gls{RMAE} and none of the models showed large improvements in \gls{RMSE} compared to the equal proportions model.
+There were only moderate improvements in the prediction of the herbaceous vegetation fraction as well.
+In contrast, most models were a significant improvement in predicting tree, built-up, water and bare cover fractions compared to the equal proportion model, indicating that the features used to train the models were useful to distinguish these classes from the others and to quantify their proportions in each pixel.
+
+To sum up, we tested a number of regression methods for characterising land cover by means of predicting thematically comprehensive land cover fractions at the global scale, and found that \gls{RF} regression and Cubist regression produce output of the highest accuracy by \gls{RMSE} and \gls{MAE}, respectively.
 Our findings agree with those of \citet{li_monitoring_2018}, who compared Cubist and \gls{RF} regression for water fraction classification and found that Cubist performs slightly better than \gls{RF} regression for this particular class.
@@
 
-At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which he also noted as performing better than \gls{RF} regression and \glspl{SVM}.
-Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the built-up class, and we also noted that Cubist performed slightly better than the other two algorithms for this class.
-Both \citet{okujeni_comparison_2014} and \citet{foody1997fuzzynnet} also performed land cover fraction mapping over an urban area.
-The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF} regression.
-This could be due to much finer spatial resolution (3.6–9 m) data used in their study, and the associated difference in class definitions.
-\citet{foody1997fuzzynnet} reported an \gls{RMSE} of 16.6\% using \gls{MLP} \glspl{NN}, lower than our results (22.7\%).
-The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, namely their focus on fine scale predictions over a local area for the urban class in particular, in contrast to our global focus with comprehensive classes, leading to a different balance of the datasets between the studies.
-
-In this study, model optimisation and dataset sizes were also a challenge.
-Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming features or replacing missing values).
-Since the large dataset size increases the training time for the models, parameters to tune have to be chosen carefully.
-As such, it is possible to miss a particular parameter combination that may further improve the results.
-On the other hand, our results showed that the differences between the best performing methods are not very high.
-The differences in model parametrisations are expected to be even lower.
-
-Another challenging aspect about land cover fraction mapping is the available algorithms that would suit this use case.
-In its core, it is a multivariate regression problem, but most regression algorithms are univariate.
-Our results showed that the binary relevance method works well to facilitate the use of univariate regression methods for this task, as the highest accuracy was achieved by algorithms that used the binary relevance method.
-This further expands the field of possible algorithms that could be used.
+\begin{figure}
+    \includegraphics[width=0.85\textwidth]{article-figures/barplots/2020-11-02-model-comparison-bar}
+    \caption{Comparison of relative RMSE (top) and relative MAE (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
+    \label{fig-models-relative}
+\end{figure}
+
+%At a local scale, \citet{walton2008subpixelrf} assessed the accuracy of machine learning algorithms for urban land cover fraction mapping and indicated 14.5\% \gls{MAE} and 20.0\% \gls{RMSE} for Cubist, which he also noted as performing better than \gls{RF} regression and \glspl{SVM}.
+%Our results were 9.70\% \gls{RMSE} and 2.43\% \gls{MAE} for Cubist over the built-up class, and we also noted that Cubist performed slightly better than the other two algorithms for this class.
+%Both \citet{okujeni_comparison_2014} and \citet{foody1997fuzzynnet} also performed land cover fraction mapping over an urban area.
+%The results of \citet{okujeni_comparison_2014} indicated that \gls{SVM} regression performed better than \gls{RF} regression.
+%This could be due to much finer spatial resolution (3.6–9 m) data used in their study, and the associated difference in class definitions.
+%\citet{foody1997fuzzynnet} reported an \gls{RMSE} of 16.6\% using \gls{MLP} \glspl{NN}, lower than our results (22.7\%).
+%The differences in the absolute numbers of \gls{RMSE} and \gls{MAE} between these studies and ours are most likely due to the different scope, namely their focus on fine scale predictions over a local area for the urban class in particular, in contrast to our global focus with comprehensive classes, leading to a different balance of the datasets between the studies.
+
+%In this study, model optimisation and dataset sizes were also a challenge.
+%Each model may have multiple parameters that can be tuned, and the input to each method can also be modified (e.g. by log-transforming features or replacing missing values).
+%Since the large dataset size increases the training time for the models, parameters to tune have to be chosen carefully.
+%As such, it is possible to miss a particular parameter combination that may further improve the results.
+%On the other hand, our results showed that the differences between the best performing methods are not very high.
+%The differences in model parametrisations are expected to be even lower.
+
+%Another challenging aspect about land cover fraction mapping is the available algorithms that would suit this use case.
+%In its core, it is a multivariate regression problem, but most regression algorithms are univariate.
+Our results showed that the binary relevance method works well to facilitate the use of univariate regression methods for the global land cover fraction mapping task, as the highest accuracy was achieved by algorithms that used the binary relevance method.
+This further expands the field of possible algorithms that could be used for the task.
 In addition, both the binary relevance method and the multi-step method can accept different algorithms for its submodels, i.e.
@@
 For example, classical spectral unmixing methods, such as spectral mixture analysis, has been often used in the past (e.g. in \citet{adams_classification_1995}).
-However, due to the limitations of the spectral mixture analysis methods \citep{somers_endmember_2011}, we could not use it with the features we selected (as listed in \ref{sec-inputdata}, table \ref{tab-inputdata}), as they do not form a linear mixture representing the land cover fractions.
-Modern machine learning methods are advantageous for their ability to model complex nonlinear relationships between variables and thus make use of all of the features as shown in \ref{sec-varimp}, figure \ref{fig-varimp}.
-
-\section{Multi-step model relative error comparison} \label{sec-randomforest-relative}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=\textwidth]{article-figures/barplots/2020-11-02-rf-comparison-bar}
-    \caption{Comparison of relative accuracy metrics of RF regression models (equal proportion model shown as a reference).
-1-step models use no adjustment for zero inflation, 2-step models perform classification on zeroes and regression for non-zeroes, 3-step models perform a classification into pure and non-pure pixels, and then a regression or classification based on that.
-Mean and median are the tree vote summary statistics.}
-    \label{fig-randomforest-relative}
-\end{figure}
+However, due to the limitations of the spectral mixture analysis methods \citep{somers_endmember_2011}, we could not use it with the features we selected, as they do not form a linear mixture representing the land cover fractions.
+%(as listed in \ref{sec-inputdata}, table \ref{tab-inputdata})
+Modern machine learning methods are advantageous for their ability to model complex nonlinear relationships between variables and thus make use of all of the features.
+%as shown in \ref{sec-varimp}, figure \ref{fig-varimp}.
 
@@
 
-In addition to classification methods, one aspect that is important for classification accuracy is the data (features) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
-In addition to optical features (such as spectral bands), temporal features (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
-\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are commonly used to improve land cover class separability, e.g.
+%In addition to classification methods, one aspect that is important for classification accuracy is the data (features) that the algorithm is trained on and predict land cover from \citep{yu2014metadiscoveries}.
+%In addition to optical features (such as spectral bands), temporal features (growing season start, duration, growing intensity) are commonly used to improve the separability of vegetation, such as crop types \citep{jakubauskas2001harmonic}.
+%\Glspl{VI} condense data and reduce the noise inherent in single bands, and thus are commonly used to improve land cover class separability, e.g.
 for wetlands \citep{sader1995wetlands}, as well as for defining classification rules in rule-based fuzzy classification \citep{baraldi2006rulebased}.
-Elevation features (elevation, slope, aspect) are also known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
+%Elevation features (elevation, slope, aspect) are also known to improve the separability between classes, such as trees \citep{burrough2001fuzzy}.
 %In this paper, we also used such features for increasing the accuracy of land cover fraction regression model predictions.
@@
 
-To calculate feature importance, we performed permutation importance on the model that had the lowest \gls{MAE}.
-We shuffled the values of each feature in turn, and the models made predictions based on all of the features, including the shuffled one.
+To gain insight on what features drive the \gls{RF} three-step median model, we performed permutation importance on it:
+we shuffled the values of each feature in turn, and the model made predictions based on all of the features, including the shuffled one.
 This results in a decrease in the accuracy of predictions, as the feature in question no longer contains meaningful information.
 We then recorded the resulting increase in \gls{MAE} compared to the validation set as a measure of feature importance, both per class and overall for all classes combined.
-
-We performed \Gls{RF} permutation importance on the three-step median model to test which features were important for \gls{RF} predictions.
+%
+%We performed \Gls{RF} permutation importance on the three-step median model to test which features were important for \gls{RF} predictions.
 The results are shown in figure \ref{fig-varimp}.
@@
     \centering
-    \includegraphics[width=14cm]{article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
+    \includegraphics[width=11cm]{article-figures/heatmaps/2020-11-06-varimp-heatmap-top5}
     \caption{Random Forest three-step median approach variable importance, top 5 features per category.
@@
 They were followed by the median \gls{NIRv} and \gls{NDMI} over the whole time series.
-The importance of remote sensing information can also be seen from the accuracy statistics of a single-step \gls{RF} regression model, when only vegetation index and temporal features were input into the model (see table \ref{tab-accuracy}, ``only RS features'').
-After excluding all the static features, \gls{RMSE} and \gls{MAE} increased only by around 1 percentage point.
+The importance of remote sensing information can also be seen from the accuracy statistics of a single-step \gls{RF} regression model, when only vegetation index and temporal features were input into the model (see table \ref{tab-accuracy}, ``only RS features''):
+\gls{RMSE} and \gls{MAE} increased only by around 1 percentage point.
 
@@
 Climate data was the second most important in our case, mostly for the crop fraction estimation, which is also closely linked with soils.
-Note that the feature importance may be different at different scales, e.g. if predictions are made at a finer spatial resolution, such as 30~m or 10~m.
-
-The scale of the study poses performance challenges, as features that the models use need to be available at a global scale.
-Our results showed that all of the feature groups used in the study were useful, therefore leaving a feature out means sacrificing some predictive power.
-The gain would be less time needed for processing, as that feature would no longer need to be downloaded, preprocessed and processed.
+%Note that the feature importance may be different at different scales, e.g. if predictions are made at a finer spatial resolution, such as 30~m or 10~m.
+
+%The scale of the study poses performance challenges, as features that the models use need to be available at a global scale.
+The feature importance results showed that all of the feature groups used in the study were useful, therefore leaving a feature out means sacrificing some predictive power.
+On the other hand, leaving out some features would be beneficial in that less time would be needed for processing, as that feature would no longer need to be downloaded, preprocessed and processed.
+This is an important consideration, given that for global land cover mapping, all features need to be available for the whole globe as well.
 In addition, model training performance (time for training and memory usage) may be an important consideration for global land cover mapping, as it may limit the scope of what models may be used for this task.
 For example, ensemble learning techniques like the super learner are very resource-intensive and may take weeks to train.
-Therefore it is important to take into account not just the model accuracy, but also whether the improvement of the accuracy is worth the increase in processing time or computing resource usage.
+Therefore for operational land cover product creation purposes, it is important to take into account not just the model accuracy, but also whether the improvement of the accuracy is worth the increase in processing time or computing resource usage.
 

### 2021-02-18 article/article.tex
Final updates for second revision round

@@
 
-A major challenge for land cover fraction mapping models is data sparsity: the dataset is zero-inflated due to how common the 0\% fraction is, but as regression favours the mean, 0\% and 100\% fractions are difficult to predict accurately.
+A major challenge for land cover fraction mapping models is data sparsity.
+Land cover fraction data is by its nature zero-inflated due to how common the 0\% fraction is.
+As regression favours the mean, 0\% and 100\% fractions are difficult for regression models to predict accurately.
 We proposed a new solution by combining three models: a binary model determines whether a pixel is pure; if so, it is processed using a classification model; otherwise with a regression model.
 
-We compared the performance of multiple types of linear and machine learning regression algorithms for land cover fraction mapping, and for our proposed three-step model, assessed the importance of six categories of features, evaluated the spatial and per-class accuracy, and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe.
+We compared multiple regression algorithms and implemented our proposed three-step model on the algorithm with the lowest RMSE.
+We further evaluated the spatial and per-class accuracy of the model and demonstrated a wall-to-wall prediction of seven land cover fractions over the globe. 
 The models were trained on over 138\,000 points and validated on a separate dataset of over 20\,000 points, provided by the \gls{CGLS-LC100} project.
@@
 Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest regression with our proposed three-model approach and median vote.
-This model was driven primarily by optical remote sensing data.
-
-This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions, and tuned for user needs.
+
+This research proves that machine learning algorithms can be applied globally to map a wide variety of land cover fractions.
 Fraction mapping expresses land cover more precisely, and empowers users to create their own discrete maps using user-defined thresholds and rules, which enables customising the result for a diverse range of uses.
-The three-step approach enables mapping 0\% and 100\% fractions more accurately, and thus has already been taken up in the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project.
-This study also includes an accuracy assessment of land cover fraction maps both thematically and spatially, which could be taken up by future land cover fraction mapping efforts.
+The three-step approach is useful for addressing the zero-inflation issue and mapping 0\% and 100\% fractions more accurately, and thus has already been taken up in the operational production of global land cover fraction layers within the \gls{CGLS-LC100} project.
+Furthermore, this study contributes to the accuracy assessment of land cover fraction maps both thematically and spatially, and these methods could be taken up by future land cover fraction mapping efforts.
 \end{abstract}
@@
 
-Except for the cover fraction layers of the \ac{CGLS-LC100} product, all other global land cover products that include all major land cover classes, such as the ones described by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single land cover class.
+Except for the cover fraction layers of the \ac{CGLS-LC100} product, all other global land cover products that include major land cover classes, such as the ones described by \citet{bartholome2005glc2000, friedl2010modis, arino2007globcover, see2015hybrid, chen2015globeland30}, are provided with discrete classes (also known as ``hard'' or ``crisp'' classification), where each pixel of the map can only represent a single land cover class.
 Such discrete classification oversimplifies reality, as mixed pixels that are covered by multiple land cover classes are a common occurrence. 
@@
 \citealp{walton2008subpixelrf}), \gls{SVM} regression \citep{walton2008subpixelrf}, Cubist regression \citep{walton2008subpixelrf}, \gls{MLP} \glspl{NN} \citep{zhang2001fullyfuzzy}, genetic algorithms \citep{stavrakoudis_boosted_2011} and wavelet transformation \citep{uma_shankar_wavelet-fuzzy_2011}.
-The previous studies have only compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.
+The previous studies have only used or compared a few of these methods at once, and never with a thematically complete set of land cover classes nor at global scale.
 
 A common issue with the use of land cover fraction data as input into regression models is data imbalance.
-The more classes are mapped, the more more likely it is that one or more classes are not present in a given pixel (have a 0\% fraction), leading to zero inflation.
+The more classes are mapped, the more likely it is that one or more classes are not present in a given pixel (have a 0\% fraction), leading to zero inflation.
 This is especially the case in homogeneous areas, where we can find not only an inflation of 0\% fractions, but also an inflation in 100\% fractions.
-This leaves little data in the middle for the regression models to learn from.
+This data imbalance leaves little training data in the middle for the regression models to learn from.
 Conversely, regression models tend to favour predictions closer to the mean and rarely predict extreme values.
@@
 
-Lastly, model accuracy assessment by itself is often challenging, especially at a global scale.
+Model accuracy assessment by itself is often challenging, especially at a global scale.
 It requires a comprehensive dataset across the globe that would be comparable with the training data, and yet independent of it.
 Accuracy assessment of land cover fractions is even more challenging, as it requires a dataset that either provides fraction information, or fine spatial resolution data from which it can be calculated.
-Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited.
-This makes it difficult for the users to decide whether a given product suits their needs.
+Because of these challenges, the information about the accuracy of the existing global land cover fraction products is often limited, which makes it difficult for the users to decide whether a given product suits their needs.
 Users would also benefit from knowing the spatial variation of accuracy, as models may be more accurate at certain locations of the world and less so at others.
@@
 
-In this study, after investigating and comparing a variety of methods for performing global land cover fraction mapping, we proposed a method to enhance land cover fraction mapping by dealing with the inherent data imbalance issue.
+In this study, after investigating and comparing a variety of methods for global land cover fraction mapping, we proposed an approach to enhance land cover fraction mapping by dealing with the inherent data imbalance issue.
 We assessed the accuracy of the result both thematically and spatially, by calculating model accuracy spatially across the globe.
@@
 
-\subsection{Multi-step approach to account for value imbalance}
+\subsection{Multi-step approach to account for zero inflation}
 \label{sec-multistep}
 
-As \gls{RF} regression performed the best in the regression model comparison, we attempted to improve it further in an attempt to solve the dataset balance issue, namely the high frequency of 0\% and 100\% fractions.
+As \gls{RF} regression performed the best by \gls{RMSE} in the regression model comparison, we tested whether it could be improved further by attempting to solve the training dataset balance issue, namely the high frequency of 0\% and 100\% fractions.
 As the dataset describes fractions of each land cover class at each point, most of the locations consist of a mix of only a few classes, and the fraction of the rest of the classes is zero at that location.
@@
 
-Taking the algorithm with the lowest \gls{RMSE}, we compared three approaches.
+We compared three approaches using \gls{RF} models:
 (a): a single regression model trained on all data.
@@
 
-\subsection{Regression model accuracy comparison}
-
-Overall, \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote (\gls{RMSE}: 17.3\%, \gls{MAE}: 9.4\%), and by \gls{MAE} when using a median vote (\gls{RMSE}: 20.7\%, MAE: 7.9\%).
+\subsection{Zero inflation adjustment with multi-step models and median voting}
+
+Regression model comparison showed that \gls{RF} regression achieved the highest accuracy: by \gls{RMSE} when using a single model and a mean vote (\gls{RMSE}: 17.3\%, \gls{MAE}: 9.4\%), and by \gls{MAE} when using a median vote (\gls{RMSE}: 20.7\%, MAE: 7.9\%).
+Therefore we chose \gls{RF} regression to test our proposed multi-step approach.
 For more details about the results of the accuracy comparison between the tested regression models, see \ref{sec-model-comparison}.
 
-\subsection{Zero inflation adjustment with multi-step models and median voting}
-
-We investigated two ways to adjust for the issue of zero inflation in the training data.
-First, we used a median vote instead of a mean vote for ensembling decision tree votes in \gls{RF} regression, so as to predict exactly 0\% and 100\% more often.
-Second, we used our proposed multi-step model approach as detailed in section \ref{sec-multistep}.
-The overall results can be seen in table \ref{tab-accuracy}, whereas per-class results can be seen in figure \ref{fig-randomforest} and the relative statistics in figure \ref{fig-randomforest-relative}.
+The overall accuracy statistics of all of the \gls{RF} regression models that we tested, including two-step, three-step, as well as median voting approaches and their combinations, can be seen in table \ref{tab-accuracy}.
+Per-class results can be seen in figure \ref{fig-randomforest} and the relative statistics in figure \ref{fig-randomforest-relative}.
 
@@
 This is because the false positives in the first step of the two-step model (the binary zero/non-zero classification) result in some high fractions nevertheless being predicted as zero.
-The two-step approach combined with binary relevance also introduced a drawback in that there were cases where all land cover class fractions were predicted to be 0\% in the first step.
-In such cases it was not possible to determine the land cover fractions.
+The two-step approach combined with binary relevance also had a drawback: in some cases all land cover class fractions were predicted to be 0\% in the first step, making it impossible to determine the land cover fractions.
 Therefore, to make them sum up to 100\%, we set these cases to equal proportions ($100\% / 7$), which introduced further error.
@@
 
-To visually demonstrate the predictions of the model with the highest accuracy by \gls{MAE}, we used the three-step median \gls{RF} model to predict land cover fractions at a global scale (100~m resolution, but sampled every 0.2~degrees).
+We selected the three-step median vote \gls{RF} model for further analysis, as it represents a good balance between the \gls{RMSE} and \gls{MAE} statistics, and its analysis helps further understand the three-step model.
+To visually demonstrate the model, we used it to predict land cover fractions at a global scale (100~m resolution, but sampled every 0.2~degrees).
 See figure \ref{fig-walltowall} for a visualisation of all of the fraction layers separately, and the supplementary material for the output GeoTIFF file itself.
 The wall-to-wall fraction maps reveal how land cover fraction mapping is capable of expressing gradients and mixed land cover.
-
-In addition, to examine whether there were any patterns in the accuracy of the predictions globally, we produced spatial accuracy maps based on the predictions in locations where we had independent validation data.
-This is also presented in raster format in figure \ref{fig-walltowall}.
+In addition, the spatial accuracy maps that we produced based on the model predictions show the variation in the model accuracy globally, and are also presented in raster format in figure \ref{fig-walltowall}.
+
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[width=\textwidth]{article-figures/maps/2020-10-12-walltowall.png}
+    \caption{Random Forest single model predictions per class.
+Top row: predicted fractions, each given its own colour (``Overall'' shows a hardened map of dominant land cover using the same colour scheme).
+Middle row: absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.
+Bottom row: distribution of predicted versus true values, shown as box plots with bins each 10\%, widths representing sample size.}
+    \label{fig-walltowall}
+\end{sidewaysfigure}
 
@@
 
-We also analysed the effect that the training features had on the model prediction, and found that features obtained from remote sensing were the most important, but all of the feature categories contributed to improving the model predictions.
+Our analysis of the effect of training features on model prediction accuracy showed that features obtained from remote sensing were the most important, but all of the feature categories contributed to improving the model predictions.
 For more detailed information, see \ref{sec-varimp}.
 
-Lastly, we compared the \gls{RF} three-step median model predictions with existing global products that correspond to a particular land cover class in our classification.
-The results are given in table \ref{tab-products}.
+Lastly, the results of the comparison of the \gls{RF} three-step median model predictions with existing global products that correspond to a particular land cover class in our classification are given in table \ref{tab-products}.
 The \gls{RF} three-step median model performed worse than the \gls{GSW} water occurrence history product, but better than the \gls{GFCC} tree cover and \gls{GHSL} built-up products.
 There was a tie when comparing the \gls{RF} three-step median model with the \gls{FROM-GLC} impervious surface product when it comes to \gls{MAE}, but \gls{FROM-GLC} was slightly more accurate according to \gls{RMSE} and less biased according to \gls{ME}.
-
-\begin{sidewaysfigure}
-    \centering
-    \includegraphics[width=\textwidth]{article-figures/maps/2020-10-12-walltowall.png}
-    \caption{Random Forest single model predictions per class.
-Top row: predicted fractions, each given its own colour (``Overall'' shows a hardened map of dominant land cover using the same colour scheme).
-Middle row: absolute errors per class, based on predictions at point locations of validation data, displayed aggregated over 1 by 1 degree cells using the mean function.
-Bottom row: distribution of predicted versus true values, shown as box plots with bins each 10\%, widths representing sample size.}
-    \label{fig-walltowall}
-\end{sidewaysfigure}
 
@@
 A similar effect was seen when using techniques such as median voting for \gls{RF} regression.
-Median voting also resulted in more correct 0\% fraction predictions, but likewise increased the chances of predicting 100\% of the wrong class.
+Like the three-step model, median voting resulted in more correct 0\% fraction predictions, but likewise increased the chances of predicting 100\% of the wrong class.
 Our results showed that median voting has a stronger effect in reducing \gls{MAE} than the multi-step approach.
 Therefore it may be a suitable choice in cases where it is computationally infeasible to train three models.
-However, when the median voting approach was combined with the three-step approach, the \gls{RMSE} decreased without affecting \gls{MAE}.
+However, this approach is only applicable to \gls{RF} models, whereas the three-step approach is generic and can be used with any combination of models.
+In addition, when the median voting approach was combined with the three-step approach, the \gls{RMSE} decreased without affecting \gls{MAE}.
 Thus, three-step median \gls{RF} achieved the best combination that is optimised towards reducing \gls{MAE}, doing so without increasing \gls{RMSE} as much as the single-step median vote model does.
@@
 Based on this model, we created a demonstration map showing the global distribution of land cover in separate land cover fraction layers.
-Remote sensing features were the most important for model accuracy, although all other types of features (climate, soil, terrain) also contributed significantly.
+Remote sensing features were the most important for model accuracy, although all other types of features (climate, soil, terrain) also contributed significantly for some classes and thus could not be omitted without negatively affecting the overall model accuracy.
 
@@
 
-\minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {PROBA-V} data, and IIASA for providing the model training dataset.
-In addition, we thank Bruno Smets for ideas and support from the \gls{CGLS-LC100} project, Benjamin Kellenberger for help on tuning \gls{NN} parametrisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
+\minisection{Acknowledgements} We would like to thank VITO and Terrascope for providing us with computational facilities to access and process {PROBA-V} data, and the team of Steffen Fritz (IIASA) for providing the model training dataset.
+In addition, we thank Bruno Smets for ideas and support from the \gls{CGLS-LC100} project, Devis Tuia for the initial idea for the multi-step approach, Benjamin Kellenberger for help on tuning \gls{NN} parametrisation, as well as Linlin Li and Tomislav Hengl for suggestions on additional machine learning methods to test.
 We also thank the three anonymous reviewers for their valuable and detailed comments.
@@
 
+\newpage
+
 \appendix
 
+\section{List of features used in regression models} \label{sec-inputdata}
+
+\begin{table}[h!]
+\begin{center}
+\begin{adjustbox}{max width=8cm}
+\begin{tabular}{m{12cm}l}
+\toprule
+\multicolumn{1}{l}{Category \& data source}&\multicolumn{1}{c}{Covariate}\tabularnewline
+\midrule
+Location (3, intrinsic)&
+\begin{tabular}{@{}l@{}}longitude\tabularnewline
+latitude\tabularnewline
+absolute latitude
+\end{tabular}\tabularnewline
+\midrule
+Vegetation indices (22, derived from PROBA-V 100m top-of-canopy reflectance v1.02 \citep{dierckx2014probav})&
+\begin{tabular}{@{}l@{}}
+minimum NDVI\tabularnewline
+maximum NDVI\tabularnewline
+median NDMI\tabularnewline
+NDMI yearly IQR\tabularnewline
+NDMI March-May IQR\tabularnewline
+NDMI June-August IQR\tabularnewline
+NDMI September-November IQR\tabularnewline
+NDMI December-February IQR\tabularnewline
+OSAVI March-May IQR\tabularnewline
+OSAVI June-August IQR\tabularnewline
+OSAVI September-November IQR\tabularnewline
+OSAVI December-February IQR\tabularnewline
+EVI March-May IQR\tabularnewline
+EVI June-August IQR\tabularnewline
+EVI September-November IQR\tabularnewline
+EVI December-February IQR\tabularnewline
+median NIRv\tabularnewline
+NIRv yearly IQR\tabularnewline
+NIRv March-May IQR\tabularnewline
+NIRv June-August IQR\tabularnewline
+NIRv September-November IQR\tabularnewline
+NIRv December-February IQR
+\end{tabular}\tabularnewline
+\midrule
+Temporal metrics (9, derived from a harmonic model over time series of PROBA-V 100m top-of-canopy reflectance v1.02 \citep{dierckx2014probav})&
+\begin{tabular}{@{}l@{}}
+NDVI order 1 cosine\tabularnewline
+NDVI order 1 sine\tabularnewline
+NDVI order 2 cosine\tabularnewline
+NDVI order 2 sine\tabularnewline
+NDVI trend coefficient\tabularnewline
+NDVI order 1 phase\tabularnewline
+NDVI order 1 amplitude\tabularnewline
+NDVI order 2 phase\tabularnewline
+NDVI order 2 amplitude
+\end{tabular}\tabularnewline
+\midrule
+Terrain (4, ASTER GDEM V003 \citep{ASTGTM003})&
+\begin{tabular}{@{}l@{}}
+elevation\tabularnewline
+slope (log-transformed)\tabularnewline
+aspect\tabularnewline
+terrain position index
+\end{tabular}\tabularnewline
+\midrule
+Climate (21, WorldClim 2.0 \citep{worldclim2})&
+\begin{tabular}{@{}l@{}}
+January precipitation (log)\tabularnewline
+April precipitation (log)\tabularnewline
+July precipitation (log)\tabularnewline
+October precipitation (log)\tabularnewline
+January solar irradiance\tabularnewline
+July solar irradiance\tabularnewline
+mean temperature\tabularnewline
+temperature monthly range\tabularnewline
+isothermality\tabularnewline
+temperature annual range\tabularnewline
+annual precipitation (log)\tabularnewline
+temperature seasonality\tabularnewline
+minimum solar irradiance\tabularnewline
+maximum solar irradiance\tabularnewline
+mean solar irradiance\tabularnewline
+mean windspeed\tabularnewline
+mean water vapour pressure\tabularnewline
+coldest month precipitation (log)\tabularnewline
+warmest month precipitation (log)\tabularnewline
+wettest month solar irradiance\tabularnewline
+driest month solar irradiance
+\end{tabular}\tabularnewline
+\midrule
+Soil (8, SoilGrids \citep{hengl_soilgrids250m_2017})&
+\begin{tabular}{@{}l@{}}
+soil available water\tabularnewline
+soil bulk density\tabularnewline
+soil cation exchange capacity (log)\tabularnewline
+soil clay fraction\tabularnewline
+soil coarse fragments (log)\tabularnewline
+soil pH\tabularnewline
+soil sand fraction\tabularnewline
+soil water wilting point
+\end{tabular}\tabularnewline
+\bottomrule
+\end{tabular}
+\end{adjustbox}
+\caption{List of features used as input for the models tested in this study, and their data sources.}
+\label{tab-inputdata}
+\end{center}
+\end{table}
+
 \section{Regression model comparison} \label{sec-model-comparison}
@@
 
-To test the performance of various regression models for land cover fraction mapping, we tested four types of models: linear models, models based on decision trees, machine learning models not based on decision trees, and ensemble learning (for an overview, see table \ref{tab-methods}).
+Before applying the multi-step approach, we tested the performance of various regression models for land cover fraction mapping.
+We tested four types of models: linear models, models based on decision trees, machine learning models not based on decision trees, and ensemble learning (for an overview, see table \ref{tab-methods}).
 
@@
 \begin{figure}
-    \includegraphics[width=0.85\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-06-03-model-comparison-bar}
     \caption{Comparison of absolute \gls{RMSE} (top) and absolute \gls{MAE} (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
@@
 \begin{figure}
-    \includegraphics[width=0.85\textwidth]{article-figures/barplots/2020-11-02-model-comparison-bar}
+    \includegraphics[width=\textwidth]{article-figures/barplots/2020-11-02-model-comparison-bar}
     \caption{Comparison of relative RMSE (top) and relative MAE (bottom) per class of the best performing models in their category, and the equal proportions solution as a reference.}
@@
 
-\section{Model accuracy changes per predicted fraction} \label{sec-apu}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=\textwidth]{article-figures/apuplots/2021-01-13-apu-rf3step.pdf}
-    \caption{Change in accuracy statistics of the \gls{RF} three-step median model per predicted land cover fraction, overlaid over the histogram of predictions.
-Bins containing less than 10 predictions were removed due to too low sample size leading to unreliable error statistics.}
-    \label{fig-apu}
-\end{figure}
-
 \section{Feature importance} \label{sec-varimp}
@@
 
+\section{Model accuracy changes per predicted fraction} \label{sec-apu}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=\textwidth]{article-figures/apuplots/2021-01-13-apu-rf3step.pdf}
+    \caption{Change in accuracy statistics of the \gls{RF} three-step median model per predicted land cover fraction, overlaid over the histogram of predictions.
+Bins containing less than 10 samples were omitted due to too low sample size leading to unreliable error statistics.}
+    \label{fig-apu}
+\end{figure}
+
 \newpage
-
-\section{List of features used in regression models} \label{sec-inputdata}
-
-\begin{table}[h!]
-\begin{center}
-\begin{adjustbox}{max width=8cm}
-\begin{tabular}{m{12cm}l}
-\toprule
-\multicolumn{1}{l}{Category \& data source}&\multicolumn{1}{c}{Covariate}\tabularnewline
-\midrule
-Location (3, intrinsic)&
-\begin{tabular}{@{}l@{}}longitude\tabularnewline
-latitude\tabularnewline
-absolute latitude
-\end{tabular}\tabularnewline
-\midrule
-Vegetation indices (22, derived from PROBA-V 100m top-of-canopy reflectance v1.02 \citep{dierckx2014probav})&
-\begin{tabular}{@{}l@{}}
-minimum NDVI\tabularnewline
-maximum NDVI\tabularnewline
-median NDMI\tabularnewline
-NDMI yearly IQR\tabularnewline
-NDMI March-May IQR\tabularnewline
-NDMI June-August IQR\tabularnewline
-NDMI September-November IQR\tabularnewline
-NDMI December-February IQR\tabularnewline
-OSAVI March-May IQR\tabularnewline
-OSAVI June-August IQR\tabularnewline
-OSAVI September-November IQR\tabularnewline
-OSAVI December-February IQR\tabularnewline
-EVI March-May IQR\tabularnewline
-EVI June-August IQR\tabularnewline
-EVI September-November IQR\tabularnewline
-EVI December-February IQR\tabularnewline
-median NIRv\tabularnewline
-NIRv yearly IQR\tabularnewline
-NIRv March-May IQR\tabularnewline
-NIRv June-August IQR\tabularnewline
-NIRv September-November IQR\tabularnewline
-NIRv December-February IQR
-\end{tabular}\tabularnewline
-\midrule
-Temporal metrics (9, derived from a harmonic model over time series of PROBA-V 100m top-of-canopy reflectance v1.02 \citep{dierckx2014probav})&
-\begin{tabular}{@{}l@{}}
-NDVI order 1 cosine\tabularnewline
-NDVI order 1 sine\tabularnewline
-NDVI order 2 cosine\tabularnewline
-NDVI order 2 sine\tabularnewline
-NDVI trend coefficient\tabularnewline
-NDVI order 1 phase\tabularnewline
-NDVI order 1 amplitude\tabularnewline
-NDVI order 2 phase\tabularnewline
-NDVI order 2 amplitude
-\end{tabular}\tabularnewline
-\midrule
-Terrain (4, ASTER GDEM V003 \citep{ASTGTM003})&
-\begin{tabular}{@{}l@{}}
-elevation\tabularnewline
-slope (log-transformed)\tabularnewline
-aspect\tabularnewline
-terrain position index
-\end{tabular}\tabularnewline
-\midrule
-Climate (21, WorldClim 2.0 \citep{worldclim2})&
-\begin{tabular}{@{}l@{}}
-January precipitation (log)\tabularnewline
-April precipitation (log)\tabularnewline
-July precipitation (log)\tabularnewline
-October precipitation (log)\tabularnewline
-January solar irradiance\tabularnewline
-July solar irradiance\tabularnewline
-mean temperature\tabularnewline
-temperature monthly range\tabularnewline
-isothermality\tabularnewline
-temperature annual range\tabularnewline
-annual precipitation (log)\tabularnewline
-temperature seasonality\tabularnewline
-minimum solar irradiance\tabularnewline
-maximum solar irradiance\tabularnewline
-mean solar irradiance\tabularnewline
-mean windspeed\tabularnewline
-mean water vapour pressure\tabularnewline
-coldest month precipitation (log)\tabularnewline
-warmest month precipitation (log)\tabularnewline
-wettest month solar irradiance\tabularnewline
-driest month solar irradiance
-\end{tabular}\tabularnewline
-\midrule
-Soil (8, SoilGrids \citep{hengl_soilgrids250m_2017})&
-\begin{tabular}{@{}l@{}}
-soil available water\tabularnewline
-soil bulk density\tabularnewline
-soil cation exchange capacity (log)\tabularnewline
-soil clay fraction\tabularnewline
-soil coarse fragments (log)\tabularnewline
-soil pH\tabularnewline
-soil sand fraction\tabularnewline
-soil water wilting point
-\end{tabular}\tabularnewline
-\bottomrule
-\end{tabular}
-\end{adjustbox}
-\caption{List of features used as input for the models tested in this study, and their data sources.}
-\label{tab-inputdata}
-\end{center}
-\end{table}
-
 \listoffigures

### 2021-02-18 article/highlights.tex
Final updates for second revision round

--- 2020-11-18 article/highlights.tex
@@
 \begin{itemize}
-    \item 4 machine learning model groups tested for deriving 7 land cover fractions globally
+    \item 7 land cover fractions derived using machine learning models and validated globally
     \item Multi-step approach proposed for dealing with zero inflation in fraction data
-    \item Random forest regression performed the best
-    \item Optical remote sensing data was the most important input for model predictions
+    \item Independent spatial accuracy assessment based on over 20\,000 reference points
+   %\item 1234567890123456789012345678901234567890123456789012345678901234567890123456789012345
+    \item Multi-step approach gives a best guess and improves 0\% and 100\% fraction estimation
     \item Multi-fraction land cover mapping shown as an alternative to discrete classification

### 2021-02-18 article/article.tex
UTF-8 to LaTeX

@@
 
-\author[WURGRS]{Dainius Masiliūnas\corref{ca}}
+\author[WURGRS]{Dainius Masili\={u}nas\corref{ca}}
 \ead{dainius.masiliunas@wur.nl}
@@
 Results showed that the random forest regression model reached the lowest RMSE of 17.3\%.
-Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest regression with our proposed three-model approach and median vote.
+Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest with our proposed three-model approach and median vote.
 

### 2021-02-18 article/article.tex
Replace UTF-8 characters

@@
 
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc}
+%\usepackage[utf8]{inputenc}
 \usepackage{lineno}
@@
 \usepackage{textcomp} % for degree sign
+\usepackage[separate-uncertainty=true, multi-part-units=repeat, detect-weight=true]{siunitx}
 %\usepackage[sort&compress,sectionbib]{natbib} % already in elsarticle
@@
 Results showed that the random forest regression model reached the lowest RMSE of 17.3\%.
-Lowest MAE (7.9\%) and highest overall accuracy (72±2\%) was achieved using random forest with our proposed three-model approach and median vote.
+Lowest MAE (7.9\%) and highest overall accuracy (\SI{72 \pm 2}{\percent}) was achieved using random forest with our proposed three-model approach and median vote.
 
@@
 \Gls{RF} regression
-& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} (1.09/-1.22) & 67±4  & 0.57±0.05 \\
+& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} (1.09/-1.22) & \num{67\pm4}  & \num{0.57\pm0.05} \\
 \ensuremath{''} two-step
-& 19.9  & 8.2   & 0.56  & 0.60 (0.78/3.48) & 71±2  & 0.62±0.02 \\
+& 19.9  & 8.2   & 0.56  & 0.60 (0.78/3.48) & \num{71\pm2}  & \num{0.62\pm0.02} \\
 \ensuremath{''} three-step
-& 19.4  & 8.4   & 0.58  & 0.60 (0.84/2.34) & 71±3  & 0.62±0.04 \\
+& 19.4  & 8.4   & 0.58  & 0.60 (0.84/2.34) & \num{71\pm3}  & \num{0.62\pm0.04} \\
 \ensuremath{''} median vote
-& 20.7  & \textbf{7.9}   & 0.52  & 0.60 (0.74/3.80) & 71±1  & 0.63±0.02 \\
+& 20.7  & \textbf{7.9}   & 0.52  & 0.60 (0.74/3.80) & \num{71\pm1}  & \num{0.63\pm0.02} \\
 \ensuremath{''} \ensuremath{''} two-step
-& 20.0  & 8.1   & 0.54  & 0.60 (0.77/3.67) & \textbf{72±1}  & 0.63±0.02 \\
+& 20.0  & 8.1   & 0.54  & 0.60 (0.77/3.67) & \textbf{\num{72\pm1}}  & \num{0.63\pm0.02} \\
 \ensuremath{''} \ensuremath{''} three-step
-& 20.2  & \textbf{7.9}   & 0.54  & 0.60 (0.77/3.34) & \textbf{72±2}  & \textbf{0.64±0.02} \\
+& 20.2  & \textbf{7.9}   & 0.54  & 0.60 (0.77/3.34) & \textbf{\num{72\pm2}}  & \textbf{\num{0.64\pm0.02}} \\
 \bottomrule
@@
 We also combined this approach with \gls{RF} median voting.
-The combined \gls{RF} median three-step approach obtained the best results for \gls{MAE} (7.9\%) and \gls{SCM} \gls{OA} (72±2\%).
+The combined \gls{RF} median three-step approach obtained the best results for \gls{MAE} (7.9\%) and \gls{SCM} \gls{OA} (\SI{72\pm2}{\percent}).
 The proposed approach results in predictions of the most likely pure land cover class, when the class is uncertain.
@@
 Equal proportions
-& 29.9  & 21.4  & 0     & — & 26±5  & 0.13±0.07 \\
+& 29.9  & 21.4  & 0     & — & \num{26\pm5}  & \num{0.13\pm0.07} \\
 \Glsentryshort{FNC}
-& 24.4  & 13.5  & 0.33  & 0.35 (0.80/2.89) & 53±4  & 0.42±0.06 \\
+& 24.4  & 13.5  & 0.33  & 0.35 (0.80/2.89) & \num{53\pm4}  & \num{0.42\pm0.06} \\
 \Glsentryshort{GLM}, \Glsentryshort{PLS}, Lasso
-& 21.6  & 12.7  & 0.48  & 0.49 (1.10/-1.42) & 56±4  & 0.43±0.05 \\
+& 21.6  & 12.7  & 0.48  & 0.49 (1.10/-1.42) & \num{56\pm4}  & \num{0.43\pm0.05} \\
 \Glsentryshort{MLR}
-& 21.6  & 12.1  & 0.48  & 0.48 (0.96/0.62) & 58±4  & 0.46±0.06 \\
+& 21.6  & 12.1  & 0.48  & 0.48 (0.96/0.62) & \num{58\pm4}  & \num{0.46\pm0.06} \\
 \Glsentryshort{MLP} \glsentryshortpl{NN}
-& 22.7  & 9.2   & 0.43  & 0.52 (0.70/4.25) & 68±1  & 0.57±0.02 \\
+& 22.7  & 9.2   & 0.43  & 0.52 (0.70/4.25) & \num{68\pm1}  & \num{0.57\pm0.02} \\
 \Glsentryshort{SVM} regression
-& 20.7  & 8.9   & 0.52  & 0.56 (0.79/3.02) & 69±2  & 0.58±0.03 \\
+& 20.7  & 8.9   & 0.52  & 0.56 (0.79/3.02) & \num{69\pm2}  & \num{0.58\pm0.03} \\
 Cubist regression
-& 18.1  & \textbf{8.1}   & 0.63  & 0.65 (0.88/1.77) & \textbf{72±2}  & \textbf{0.63±0.03} \\
+& 18.1  & \textbf{8.1}   & 0.63  & 0.65 (0.88/1.77) & \textbf{\num{72\pm2}}  & \textbf{\num{0.63\pm0.03}} \\
 \Glsentryshort{RF} regression
-& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} (1.09/-1.22) & 67±4  & 0.57±0.05 \\
+& \textbf{17.3}  & 9.4   & \textbf{0.66}  & \textbf{0.67} (1.09/-1.22) & \num{67\pm4}  & \num{0.57\pm0.05} \\
 \ensuremath{''}
 only RS features
-& 18.4  & 10.3  & 0.62  & 0.64 (1.09/-1.25) & 64±4  & 0.54±0.05 \\
+& 18.4  & 10.3  & 0.62  & 0.64 (1.09/-1.25) & \num{64\pm4}  & \num{0.54\pm0.05} \\
 Cubist + RF ensemble
-& 17.7  & 8.6   & 0.65  & 0.65 (0.95/0.73) & 70±3 & 0.61±0.04 \\
+& 17.7  & 8.6   & 0.65  & 0.65 (0.95/0.73) & \num{70\pm3} & \num{0.61\pm0.04} \\
 \bottomrule
